### 2022-10-31 manuscript/manuscript.qmd
Add Manuscript

---
title: Open-Source machine learning BANTER acoustic classification of a cryptic echolocating species
author:
  - name: Shannon Rankin
    email: shannon.rankin@noaa.gov
    affiliations: 
        - id: SWFSC
          name: Southwest Fisheries Science Center
          department: Marine Mammal & Turtle Division
          address: 8901 La Jolla Shores Dr.
          city: La Jolla
          state: CA
          postal-code: 92037
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Taiki Sakai
    email: taiki.sakai@noaa.gov
    affiliations:
        - id: SWFSC
          name: Southwest Fisheries Science Center
          department: Marine Mammal & Turtle Division
          address: 8901 La Jolla Shores Dr.
          city: La Jolla
          state: CA
          postal-code: 92037
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Frederick Archer
    email: eric.archer@noaa.gov
    affiliations:
        - id: SWFSC
          name: Southwest Fisheries Science Center
          department: Marine Mammal & Turtle Division
          address: 8901 La Jolla Shores Dr.
          city: La Jolla
          state: CA
          postal-code: 92037
    note: Yet another author footnote.
  - name: Jay Barlow
    email: jaybarlow33@yahoo.com
    affilations:
        - id: SWFSC
          name: BarlowAGAD
          department: 
          address: 
          city: 
          state: 
          postal-code: 
        - ref: Barlow AGAD
  - name: Danielle Cholewiak
    email: danielle.cholewiak@noaa.gov
    affilations:
        - id: NEFSC
          name: Northeast Fisheries Science Center
          department: 
          address: 166 Water Street
          city: Woods Hole
          state: MA
          postal-code: 02543
  - name: Annamaria DeAngelis
    email: annamaria.deangelis@noaa.gov
    affilations:
        - id: NEFSC
          name: Northeast Fisheries Science Center
          department: 
          address: 166 Water Street
          city: Woods Hole
          state: MA
          postal-code: 02543
  - name: Jennifer L. K. McCullough
    email: jennifer.mccullough@noaa.gov
    affilations:
        - id: PIFSC
          name: Pacific Islands Fisheries Science Center
          department: 
          address: 1845 Wasp Blvd
          city: Honolulu
          state: HI
          postal-code: 96818
  - name: Erin Oleson
    email: erin.oleson@noaa.gov
    affilations:
        - id: PIFSC
          name: Pacific Islands Fisheries Science Center
          department: 
          address: 1845 Wasp Blvd
          city: Honolulu
          state: HI
          postal-code: 96818
  - name: Melissa Soldevilla
    email: melissa.soldevilla@noaa.gov
    affilations:
        - id: SEFSC
          name: Southeast Fisheries Science Center
          department: 
          address: 75 Virginia Beach Drive
          city: Miami
          state: FL
          postal-code: 33143
abstract: |
  Passive acoustic monitoring is increasingly used for assessing populations of marine mammals; however, analysis of large datasets is limited by our ability to easily classify sounds detected. Classification of beaked whale acoustic events, in particular, require evaluation of multiple lines of evidence by expert analysts. Here we present a highly automated approach to acoustic detection and classification using supervised machine learning and open source software methods. Data from four large scale surveys of beaked whales (North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and BANTER (hierarchical random forest classifier). Overall correct classification results ranged from 86% for the South Atlantic data to 99% for the North Atlantic. Results for many species could likely be improved with increased sample sizes and consideration of alternative automated detectors. These methods provide a highly automated approach to acoustic detection and classification using open source methods that can be readily adopted for species and geographic regions.
keywords: 
  - bioacoustics
  - machine learning
  - random forest
  - species classification
  - marine mammals
  - passive acoustic monitoring
  - beaked whale
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

# Introduction

Passive acoustic monitoring (PAM) from has proven to be a valuable tool for studying populations of marine mammals (Van Parijs et al., 2009); however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring. Historically, experienced acousticians manually identify stereotyped sounds that could be reliably attributed to a given species, based on their spectral or temporal characteristics (Baumann-Pickering et al., 2013; Bittle and Duncan, 2013; Rankin and Barlow, 2005; Soldevilla et al., 2008). The dramatic increase in recordings make it impossible for experienced acousticians to manually annotate all data, and there has been an increasing need for the development of automated classification routines that can provide accurate determinations of the species responsible for the sounds.

Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods (MacLeod, 2018). However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species (Baumann-Pickering et al. 2013). Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species (Barlow et al., 2021a; Baumann-Pickering et al., 2014; Simonis et al., 2020). Unfortunately, manual classification of beaked whale echolocation clicks requires analysis by trained observers using a number of visual aids to examine the call characteristics. Development of automated classification routines, if accurate, serve to improve the efficiency and decrease the cost of analyzing large datasets.

BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings (Rankin et al., 2017). This hierarchical random forest event classifier relies on an initial call classifier (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier. The event classifier considers the probability of assignment to species for each call in the call classifier, along with any event-level characteristics (such as call rate). BANTER is very flexible and can accommodate any number of measures from any number of detectors. While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results (Rankin et al., 2017). In this case, settings for a whistle & moan detector were modified to improve performance at detecting burst pulses. This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the overall number of detections was different. This approach was successfully applied to classified echolocation clicks for narwhal and belugas (Zahn et al., 2021). For species who primarily (or exclusively) produce echolocation clicks, BANTER may serve as an option for automated machine learning classification.

Here we applied BANTER acoustic classification to beaked whale detections from four large acoustic datasets off the US East Coast, Hawaii, and the US West Coast. Acoustic data were analyzed by experienced acousticians to determine species identity, and these 'ground truth' classifications served as training data for the supervised BANTER trials. Our goals are to identify an efficient and accurate automated approach to acoustic classification of beaked whales, and to provide a framework for analysis that may serve for other PAM studies.

# Materials and Methods

## Field Data Collection

Passive acoustic data was collected during three surveys conducted by National Oceanographic and Atmospheric Administrations' research operations off the US East Coast, the Hawaiian Islands (Hawaii) and the US West Coast.

Recordings from the western North Atlantic Ocean (U.S. east coast) were collected using towed hydrophone arrays during the 2016 AMAPPS II (Atlantic Marine Assessment Program for Protected Species) survey (cite). AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey area. The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, cite) and recordings used in this analysis consisted of a single hydrophone (HTI-96-min, High Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see DeAngelis et al., 2018 for more information). The SAtlantic area ranged from Delaware south to central Florida (GU1605, cite) and recordings used in this analysis consisted of a single hydrophone (Reson etc) recorded 500 kHz sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass filter (see XX for more information).

Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, Yano et al., 2018).  Drifting recording buoys were deployed at randomly selected locations within the main Hawaiian Islands. Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA). Soundtrap recorders were sampled at 288 kHz with a duty cycle of 2 min on for every 10 minutes. SM3M data were sampled at 256 kHz with continuous recordings. Individual buoys were deployed for 10 to 50 days (see Yano et al., 2018 for more detailed information).

Recordings from the Eastern North Pacific Ocean off the U.S. west coast (EPacific) were collected using drifting acoustic recorders during the 2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels) survey (Keating et al., 2018). Buoys were deployed at a number of stations in the California Current, including pre-determined locations (for abundance estimation studies) and in close proximity to seamounts (Ad Hoc Seamount Experiment). Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA). Recordings from pre-determined locations were duty cycled and recorded 2 minutes out of a variable 'off' time; recordings from the Seamount Experiment were continuous. Soundtrap recorders were sampled at 288 kHz and the SM3M recorders were recorded at 256 sample rate (with some sampling at 96 kHz for ocean noise measurements). Individual buoys were deployed for 11-23 days (see Keating et al. 2018 for more detailed information).

## Acoustic Data Analysis

All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c) with a suite of generic click classifiers within the Click Detector module (see Keating and Barlow, 2013).  Classifier sets were saved such that any click may be classified as more than one click type. The general click classifiers were effectively click detectors, as each implemented a simple spectral band settings (e.g., 2- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz). An additional detector/classifier within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see Keating and Barlow, 2013).

Acoustic detection of beaked whale events were then classified to species by experienced acoustic researchers using multiple lines of evidence. If beaked whales were sighted in close proximity to the recording device, the visual confirmation of species identify was used to determine species identity. Otherwise, experienced acousticians followed strict protocol to define species identify based on acoustic characteristics. When the available characteristics were inconclusive, the species was considered an unidentified beaked whale; only acoustic events classified to species were considered for this study. Acoustic researchers with experience in classifying beaked whale echolocation clicks examined the bearing-time display for the click detector module in PAMGuard to assess the species identity of the acoustic event. For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see DeAngelis et al., 2018). Data were saved within the PAMGuard database and binaries for downstream processing using PAMpal package (Sakai, 2021) in R programming language (R Core Team 2014).

Standard click calculations were measured from echolocation clicks using default values in PAMpal.  For EPacific data, any events \> 2 minutes were subdivided into 2-minute events to ensure that all events were of equal value (because most of our data was duty cycled into 2-minute blocks). Inter-click interval (ICI) was calculated using the 'calculateICI' function in PAMpal; the mode of the ICI was identified for each beaked whale acoustic event. In addition, for EPacific data, used the 'browseEdinfo' and 'matchEnvData' function in PAMpal to identify appropriate ERDAPP environmental data (cite), including: sea surface temperature (from jplMURSST41), productivity (from   erdMH1pp8day dataset), chlorophyll (from erdMBchla8day dataset), and seafloor gradient (x,y, and gradient magnitude from rdSrtm30plusSeafloorGradient).

Multiple trials were conducted to better understand the value (and potential limitations) of the different suite of measurements. Data subsets for each trial were exported for BANTER classification using the export_banter function in PAMpal. At a minimum, each trial included standard click calculations for all clicks in a beaked whale acoustic event (EC); and some trials included the mode of the ICI for each event (EC_ICI). For the EPACIFIC data, additional trials that also included water depth (EC_ICI_D) and environmental data (EC_ICI_ENV) were run for each event.

## BANTER Acoustic Classification

BANTER models was created for each trial using the BANTER package (cite) in R programming language (R Core Team 2014). Standard click calculations were used for the BANTER call classifier; event level measures of ICI and environmental features (for EPacific data) were applied to the BANTER event classifier. A minimum of two events per species and complete suite of variables (absence of NAs) were required for training and testing of the classification model. The BANTER random forest models contain two parameters, ntree and sampsize, which must be selected to ensure stability of the model. The sample size (sampsize) is the number of events considered for each trial of the model, and the ntree is the number of trees (iterations of the model) run in a given model. These values were modified to obtain the best results, with an emphasis on lower sample size and higher number of trees for computational performance (Rankin and Archer, 2021).

The model output include the confusion matrix, and a number of visualizations provided by the rfpermute package (cite) in R programming language (R Core Team 2014).

# Results

Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study; species included: Baird's beaked whales (*Berardius bairdii*), Blainsville's beaked whales (*Mesoplodon densirostris*), Cuvier's beaked whales (*Ziphius cavirostris*), Cross Seamount beaked whale (McDonald et al., 2009), Gervais' beaked whales (*Mesoplodon europaeus*), Longman's beaked whale (*Indopacetus pacificus*), Sowerby's beaked whales (*Mesoplodon bidens*), True's beaked whales (*Mesoplodon mirus*), Stejneger's beaked whale (*Mesoplodon stejnegeri*), and unidentified beaked whale 'BW43' (Baumann-Pickering et al., 2013). Analyses were conducted separately for the different study areas due to differences in the data collection methods that precluded combining of datasets. Specific species combinations and sample sizes varied by study area.

## NAtlantic

\[*Note: Last reprocess had both models with same classification results (though ICI was 'important', it did not really change the results). I need to reword a few things to accommodate this (and waiting on confirmation that re: ptoential changes to one function in PAMpal*)\]

Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n=70), Sowerby's beaked whales (n=5), True's beaked whales (n=65), and Gervais' beaked whales (n=3). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

For the NAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (EC_ICI, Detector Model: sampsize = 4, ntrees = 30,000; BANTER Model: sampsize = 4, trees=300,000) and provided an overall correct classification rate of 100% for all four species (Fig. 1). There were insufficient samples of Gervais' beaked whales to include in the NAtlantic model (n=2). All classification results were greater than the expected error rate (Prior in Fig. 1a). Results from the EC model are presented here; results from the EC_ICI model can be found in supplementary materials (Supplement Fig. 1).

\[Insert Figure 1\]\[Figure 1. BANTER classification results from the NAtlantic dataset (EC). Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.

The proximity plot (Fig.1b) provides a view of the distribution of events within the feature space of the two 'most important' dimensions, or predictors. For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result. The importance heat map (Fig. 1c) shows that the most important features for predicting each species.

The strength of the classifications by species can be seen in the plot votes (Fig. 1d), which show the istribution of the votes for each of the 300,000 trials. For each plot, the events are represented as vertical slices (with plot subdivided along the x-axes according to the event sample size), and the percentage of votes for each species is represented by their color along that vertical slice. In a perfect scenario, all 300,000 trees would have voted for the correct species and therefore the classification would be correct AND the strength of these classifications would be maximized. The plot vote for the NAtlantic dataset show that all species

## SAtlantic

Species encountered in the SAtlantic study dataset include Blainsville's beaked whales (n=35), Cuvier's beaked whales (n=66), Gervais' beaked whales (n=45), Sowerby's beaked whales (n=2), and True's beaked whales (n-12). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

For the SAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (Detector Model: sampsize = 2, ntrees = 30,000; BANTER Model: sampsize = 1, ntrees = 300,000) and provided an overall correct classification rate of 86.9% for all four species. The sample sizes for the BANTER model was kept low (sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2. Analysis with a larger sample size resulted in improved classification results for Cuvier's beaked whale (see Supplement Fig. 3 for EC_ICI_ALT), but resulted in the loss of Sowerby's in the final model. Classification scores ranged from a low of 74.2% (Cuvier's beaked whale) to 100% (Sowerby's and True's beaked whale, Fig. 2); all classification results were greater than the expected error rate (Prior in Fig. 2a). Results from the EC_ICI model are presented here; results from the EC model can be found in supplementary materials (Supplement Fig. 2).

\[Insert Figure 2\]\[Figure 2. BANTER classification results from the SAtlantic dataset. Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color. \]

For SAtlantic, the proximity plot (Fig. 2b) shows that True's beaked whales are differentiated from the other species based solely on the two most important features. The importance heat map (Fig. 2c) shows that Cuvier's, Gervais' and Sowerby's beaked whales rely on a number of predictors that are lower in importance and there is likely greater distinction between these species than is represented by the proximity plot (Fig. 2b). The plot votes (Fig. 2d) shows a moderate number of trees in the forest were misclassified and these misclassifications included all species in the model.

## Hawaii

The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (Yano et al. 2018); species included Longman's beaked whale (n=121), Blainsville's beaked whale (n=518), Cross Seamount beaked whale (n=84) and Cuvier's beaked whale (n=201). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

The most accurate model considered the echolocation clicks only (Detector Model: sampsize = 10, ntrees = 30,000; BANTER model: sampsize = 4, ntree = 200,000) and provided an overall correct classification rate of 89.7% for all four species. Classification scores ranged from a low of 82.1% (Cross Seamount Beaked Whale) to 96.7% (Longman's beaked whale, Fig. 3).  All classification results were greater than the expected error rate (Prior in Fig. 3a). Results from the EC model are presented here; results from EC_ICI model can be found in supplementary materials (Supplement Fig. 4).

\[Insert Figure 3\]\[Figure 3. BANTER classification results from the Hawaii dataset. Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color. Species codes include Cross Seamount beaked whale (BWC), Longman's beaked whale (IP), Blainsville's beaked whale (MD) and Cuvier's beaked whale (ZC).

There are clear distinctions in the primary features used to classify three of the four species as shown in the Proximity Plot (Fig. 3b) which shows the classification results based on the two most 'important' predictors. Distinctive clusters based on the two most important features can be found for Longman's (IP), Blainsville's (MD), and Cuvier's beaked whales (ZC), and there is overlap in the feature space of the clusters for Cuvier's and Cross Seamount beaked whales (BWC). This overlap is resolved with different features for ZC (3^rd^, 7^th^ in Fig. 3c heat map) and BWC (5^th^, 8^th^ feature in heat map), as evidence in the low misclassification rates for these species (Fig. 3a confusion matrix). The plot votes graph (Fig. 3d) show complicated results with variation in the proportion of trees that correctly classified the species for each event.

## Eastern Pacific

The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States (Keating et al. 2018); species included Baird's beaked whale (n=91), Unidentified Beaked Whale 'BW43' (n=11), Stejneger's beaked whale (n=41) and Cuvier's beaked whale (n=1,332). BANTER classification trials included (1) echolocation clicks (EC), (2) echolocation clicks and ICI (EC_ICI), (3) echolocation clicks, ICI, and water depth (EC_ICI_D), and (4) echolocation clicks, ICI, water depth, sea surface salinity, productivity, and chlorophyll (EC_ICI_ENV).

The best model considered the echolocation clicks and the ICI values (Detector Model: sampsize = 5, ntrees = 30,000; BANTER model: sampsize = 4, ntree = 200,000) and provided an overall correct classification rate of 93.7% for all four species. Classification scores ranged from a low of 93.2% (Cuvier's beaked whale) to 100% (BW43, Fig. 4a). All classification results were greater than the expected error rate (Prior in Fig. 4a). Results from the EC_ICI model are presented here; complete results can be found in supplementary materials (Supplement Figs. 5-7).

\[Insert Figure 4\]\[Figure 4. BANTER classification results from the EPacific dataset. Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).  Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color. Species codes include Baird's beaked whale (BB), unidentified beaked whale (BW43), Stejneger's beaked whale (MS) and Cuvier's beaked whale (ZC).\]

There are clear distinctions in the primary features used to classify the species as shown in the Proximity Plot (Fig. 4b) which shows the classification results based on the two most 'important' predictors. The species each form distinctive clusters based on these two features; however, many events fall outside these distinct bounds (detections shown in area between species classification clusters). Both BB and BW43 relied on a wide range of variables for classification (Fig. 4c heat map). The plot votes (Fig. 4d) show strong classification results for most events of most species.

# Discussion

The results for the four study areas varied from a low of 86.9% correct classification in the SAtlantic dataset to a high of 100% correct classification in the NAtlantic dataset. All species in all datasets had classifications rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), BWC (Hawaii), and Blainsville's beaked whales (Hawaii) had classification rates below 90%. There were insufficient sample sizes of Gervais to include in the NAtlantic dataset.

Several of our datasets included very low sample sizes for some species; however, low sample sizes did not always result in low classification scores. In the Atlantic dataset, Sowerby's beaked whale was consistently correctly classified in both models (NAtlantic and SAtlantic); however, while the plot votes showed very strong classification strength in the NAtlantic, the classification strength (represented by the number of trees that voted for the correct classification) was poor in the SAtlantic study area.This is likely due to both a small sample size of Sowerby's detected in the S. study area (n=2) combined with the small sampsize used in the BANTER model. In general, larger sample sizes should lead to improved overall classification results, and improved strength of these classification results (as identified by plot vote graph).

While some species have distinctive differences in their click characteristics that lead to strong classification despite small sample sizes (e.g. Sowerby's beaked whales in NAtlantic); other species, such as Cuvier's beaked whales, have significant variation in their click measurements.  For species with high variability in the predictor variables and an overlap in the range of these variables with other species, a large sample size is required to describe the true variability of these call measurements. Even with reasonably large sample sizes, the classification may suffer (e.g, SAtlantic Cuvier's beaked whale = 74%). We found by increasing the sampsize in BANTER, we could increase the classification rate for this species in this area to (97%, see Supplement Fig.3); however, this resulted in the loss of Sowerby's beaked whale in the final model. So, an increased sample size in Sowerby's would likely lead to increased classification results for other species in the model.

Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (long-beaked and short-beaked common dolphins, Rankin et al., 2017). This suggests that large increases in sample size may improve classification results for Cuvier's beaked whales.

Researchers have noted several 'types' of click characteristics for Cuvier's beaked whales (ADA, pers. comm) which may be related to variability in the propagation environment, different animal behaviors, gender or age classes, or acoustic populations. Future work should examine if these Cuvier's beaked whale events can be clustered in an objective and stable manner; if so, then these can be input into BANTER models as separate types of Cuvier's beaked whales. We recommend caution in this approach- this should only be considered if there is reasonable evidence that these different groups represent authentic cluster. We do not recommend segregation of misclassified subgroups simply to improve classification scores.

Gervais' and True's beaked whales have similar click characteristics that make them difficult to differentiate, and require considerable expertise to classify manually, using a variety of available visuals and metrics (DeAngelis et al., 2018). Our results suggest that BANTER may serve as an efficient and effective means of classifying and differentiating these two species. Despite modest sample sizes, Gervais' and True's beaked whales showed high classification scores (95.6% and 100%, respectively). Unfortunately, we were unable to combine the N and S Atlantic datasets (which would increase sample sizes of True's beaked whales) due to differences in hydrophone characteristics that result in differences in call measures. Calibration of signals (to make them comparable) may allow for combining datasets to improve sample sizes.

For the SAtlantic models, ICI was found to be an important predictor for many species \[*talk about the 1 change in sowerby's when the ICI was added- do you think it was because of an error in ICI?*\] . For Hawaii and the EPacific, however, we found that adding ICI into the BANTER classifier did not improve final classification results. The differences here are likely due to different analysis methods for identifying events. Beaked whale events were defined by encounter time for both the Hawaii and EPacific data, which were collected using drifting acoustic recorders where animals could not be localized. The Atlantic data were collected using a towed hydrophone array with localization capabilities. Each event was defined as an individual click train, so ICI for these events more likely represented true ICI, and was found to be more informative. Baumann-Pickering et al. (2013) showed an overlap in the ICI between Cuvier's and Blainsville's beaked whales, but a strong difference in ICI between these species and BWC. Likewise, Baumann-Pickering (2013) found a difference in ICI between Baird's beaked whales and Cuvier's beaked whales. Subdividing click trains in PAMGuard is time consuming, but alternative options include consideration of the PAMGuard click train detector module in BANTER, or by developing a similar function in R to apply to events. Integration of ICI from individual click trains may serve to improve classification results for some species in Hawaii and EPacific.

Gervais were eliminated from the NAtlantic BANTER model due to low sample size (n=2); results were extremely poor when they were included in the model. Analysis notes indicate uncertainty in manual species classification on one of the two events in this dataset. While small sample size can at times provide good classification scores, as we found in Sowerby's beaked whale in this SAtlantic dataset, they can be devastating if there are inaccuracies in the training data. These two examples of small sample sizes highlights a conflict in preferred protocol. Ideally, (1) all [species]{.underline} would be included in a classification model, so as to better represent the local species diversity, (2) all [events]{.underline} would ideally be included in the classification model, so as to better represent the variability found in the area, and (3) only confident 'ground truth' classifications would be considered in the training model. Unfortunately, in species where identity must be determined based on call characteristics (rather than visual confirmation of species identity in the field), it can sometimes be difficult to confidently determine species identity when calls are highly variable and do not include at least one call that provides high confidence of species identity. In the case of beaked whales, we recommend that training data include a reasonable level of confidence for inclusion. When sample sizes are low, such as in these cases, it may be wise to consider the strength of the classifications when including them for model training (as we have done here). An alternative is to require agreement from multiple analysts, as provided by the EPacific datasets. This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified. For example, Barlow et al. (2021b) recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (including the EPacific data presented here).

While BANTER provides an efficient and consistent approach to classification, there are significant limitations that must be considered. BANTER is a supervised machine learning tool and requires reliable training data for success. Training data should consist of labels with strong confidence in species identify (ideally determined by agreement from more than one analyst) and sample sizes should be large enough to explain the natural variability in the data. In fact, consideration of the strength of classification scores may help identify cases where the training data were of poor quality, or where there may be evidence to identify different acoustic populations or behavior states.

The workflow presented here provides a highly automated approach to detection of acoustic events (PAMGuard), integration of environmental data (PAMpal) and acoustic event classification (BANTER). These methods significantly reduce manual analysis, provide more consistent classification results with fewer biases, and provide an estimate of classification error.  The greatest improvement to classification results for beaked whales would likely result from improved sample sizes and examination of individual click trains in measuring ICI. Consideration of additional detectors (e.g. matched filter detector, click train detector) or additional environmental variables may further improve classification results. These highly automated methods may allow for analysis of data across large spatial and temporal scales to address large ecological and population level questions.

# Acknowledgements

The authors would like to acknowledge the large number of scientists and shipboard crew who were responsible for data collection during multiple large scale studies. Funding for data collection and analysis were provided by U.S. Navy, Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration. The manuscript was improved thanks to reviews from \[XXXX\].

Here are two sample references: @Feynman1963118 @Dirac1953888.

By default, natbib will be used with the `authoryear` style, set in `classoption` variable in YAML. You can sets extra options with `natbiboptions` variable in YAML header. Example

    natbiboptions: longnamesfirst,angle,semicolon

There are various more specific bibliography styles available at <https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files>. To use one of these, add it in the header using, for example, `biblio-style: model1-num-names`.

## Using CSL

If `cite-method` is set to `citeproc` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`. In this case, the `csl` option is used to format the references. By default, this template will provide an appropriate style, but alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>. These can be downloaded and stored locally, or the url can be used as in the example header.

# Figures and tables

@fig-meaningless is generated using an R chunk.

```{r}
#| label: fig-meaningless
#| fig-cap: A meaningless scatterplot
#| fig-width: 5
#| fig-height: 5
#| fig-align: center
#| out-width: 50%
#| echo: false
plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in @tbl-simple example.

```{r}
#| label: tbl-simple
#| tbl-cap: Caption centered above table
#| echo: true
knitr::kable(head(mtcars)[,1:4])
```

# References {.unnumbered}

Barlow, J., Cárdenas‐Hinojosa, G., Henderson, E.E., Breese, D., López‐Arzate, D., Hidalgo Pla, E. and Taylor, B.L., 2021. Unique morphological and acoustic characteristics of beaked whales (*Mesoplodon* sp.) off the west coast of Baja California, Mexico. Marine Mammal Science. https://doi.org/10.1111/mms.12853

Barlow, J., Moore, J.E., McCullough, J.L. and Griffiths, E.T., 2021. Acoustic‐based estimates of Cuvier's beaked whale (*Ziphius cavirostris*) density and abundance along the US West Coast from drifting hydrophone recorders. Mar. Mamm. Sci. <https://doi.org/10.1111/mms.12872>

Baumann-Pickering, S., McDonald, M.A., Simonis, A.E., Solsona Berga, A., Merkens, K.P., Oleson, E.M., Roch, M.A., Wiggins, S.M., Rankin, S., Yack, T.M. and Hildebrand, J.A., 2013. Species-specific beaked whale echolocation signals. J. Acoust. Soc. Am. 134(3), 2293-2301. https://doi.org/10.1121/1.4823843

Baumann-Pickering, S., Roch, M.A., Brownell Jr, R.L., Simonis, A.E., McDonald, M.A., Solsona-Berga, A., Oleson, E.M., Wiggins, S.M. and Hildebrand, J.A., 2014. Spatio-temporal patterns of beaked whale
echolocation signals in the North Pacific. PLoS One.  9(1), e86072. https://doi.org/10.1371/journal.pone.0086072

Baumann-Pickering, S., Wiggins, S.M., Roth, E.H., Roch, M.A., Schnitzler, H.U. and Hildebrand, J.A., 2010. Echolocation signals of a beaked whale at Palmyra Atoll. J. Acoust. Soc. Am. 127(6),
3790-3799. https://doi.org/10.1121/1.3409478

Bittle, M. and Duncan, A., 2013. A review of current marine mammal detection and classification algorithms for use in automated passive acoustic monitoring. In Proceedings of Acoustics: Science, Technology and Amenity. 2013, 1-8.

DeAngelis, A.I., Stanistreet, J.E., Baumann-Pickering, S. and Cholewiak, D.M., 2018. A description of echolocation clicks recorded in the presence of True's beaked whale (*Mesoplodon mirus*). J. Acoust. Soc. Am. 144(5), 2691-2700. https://doi.org/10.1121/1.5067379

Keating JL, Barlow J., 2013. Click detectors and classifiers used during the 2012 southern California behavioral response study. NOAA Tech Memo NOAA-TM-NMFS-SWFSC-517:1-14.

MacLeod, C.D., 2018. Beaked whales, overview, in: Encyclopedia of marine mammals. Academic Press, pp 80-83.

McDonald, M.A., Hildebrand, J.A., Wiggins, S.M., Johnston, D.W. and Polovina, J.J., 2009. An acoustic survey of beaked whales at Cross Seamount near Hawaii. J. Acoust. Soc. Am.  125(2), 624-627. https://doi.org/10.1121/1.3050317

R Core Team, 2014. R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL http://www.R-project.org/

Rankin, S., and Archer, F., 2021. BANTER: A User\'s Guide to Acoustic Classification. https://taikisan21.github.io/PAMpal/banterGuide.html

Rankin, S., Archer, F., Keating, J.L., Oswald, J.N., Oswald, M., Curtis, A. and Barlow, J., 2017. Acoustic classification of dolphins in the California Current using whistles, echolocation clicks, and burst pulses. Mar. Mamm. Sci. 33(2), 520-540. https://doi.org/10.1111/mms.12381

Rankin, S. and Barlow, J., 2005. Source of the North Pacific \"boing\" sound attributed to minke whales. J. Acoust. Soc. Am. 118(5), 3346-3351. https://doi.org/10.1121/1.2046747

Soldevilla, M.S., Henderson, E.E., Campbell, G.S., Wiggins, S.M., Hildebrand, J.A. and Roch, M.A., 2008. Classification of Risso\'s and Pacific white-sided dolphins using spectral properties of echolocation clicks. J. Acoust. Soc. Am. 124(1), 609-624. https://doi.org/10.1121/1.2932059

Sakai, T. 2021. PAMpal: Load and Process Passive Acoustic Data. R package version 0.13.0.

Simonis, A.E., Brownell Jr, R.L., Thayre, B.J., Trickey, J.S., Oleson, E.M., Huntington, R. and Baumann-Pickering, S., 2020. Co-occurrence of beaked whale strandings and naval sonar in the Mariana Islands, Western Pacific. Proc. Roy. Soc. B. 287(1921), p.20200070. https://doi.org/10.1098/rspb.2020.0070

Van Parijs, S.M., Clark, C.W., Sousa-Lima, R.S., Parks, S.E., Rankin, S., Risch, D. and Van Opzeeland, I.C., 2009. Management and research applications of real-time and archival passive acoustic sensors over varying temporal and spatial scales. Mar. Ecol. Progr. Ser. 395, pp.21-36. https://doi.org/10.3354/meps08123

Zahn, M.J., Rankin, S., McCullough, J.L., Koblitz, J.C., Archer, F., Rasmussen, M.H. and Laidre, K.L., 2021. Acoustic differentiation and classification of wild belugas and narwhals using echolocation clicks. Sci.
Rep. 11(1), pp.1-16. https://doi.org/10.1038/s41598-021-01441-w

### 2022-10-31 manuscript/manuscript.tex
Add Manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Open-Source machine learning BANTER acoustic classification of a cryptic echolocating species},
  pdfauthor={Shannon Rankin; Taiki Sakai; Frederick Archer; Jay Barlow; Danielle Cholewiak; Annamaria DeAngelis; Jennifer L. K. McCullough; Erin Oleson; Melissa Soldevilla},
  pdfkeywords={bioacoustics, machine learning, random forest, species
classification, marine mammals, passive acoustic monitoring, beaked
whale},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Open-Source machine learning BANTER acoustic classification of a
cryptic echolocating species}
\author[1]{Shannon Rankin%
\corref{cor1}%
\fnref{fn1}}
 \ead{shannon.rankin@noaa.gov} 
\author[1]{Taiki Sakai%
%
\fnref{fn2}}
 \ead{taiki.sakai@noaa.gov} 
\author[1]{Frederick Archer%
%
\fnref{fn3}}
 \ead{eric.archer@noaa.gov} 
\author[]{Jay Barlow%
%
}
 \ead{jaybarlow33@yahoo.com} 
\author[]{Danielle Cholewiak%
%
}
 \ead{danielle.cholewiak@noaa.gov} 
\author[]{Annamaria DeAngelis%
%
}
 \ead{annamaria.deangelis@noaa.gov} 
\author[]{Jennifer L. K. McCullough%
%
}
 \ead{jennifer.mccullough@noaa.gov} 
\author[]{Erin Oleson%
%
}
 \ead{erin.oleson@noaa.gov} 
\author[]{Melissa Soldevilla%
%
}
 \ead{melissa.soldevilla@noaa.gov}

\affiliation[1]{organization={Southwest Fisheries Science Center, Marine
Mammal \& Turtle Division},addressline={8901 La Jolla Shores
Dr.},city={La Jolla},postcode={92037},postcodesep={}}

\cortext[cor1]{Corresponding author}
\fntext[fn1]{This is the first author footnote.}
\fntext[fn2]{Another author footnote, this is a very long footnote and
it should be a really long footnote. But this footnote is not yet
sufficiently long enough to make two lines of footnote text.}
\fntext[fn3]{Yet another author footnote.}

\begin{abstract}
Passive acoustic monitoring is increasingly used for assessing
populations of marine mammals; however, analysis of large datasets is
limited by our ability to easily classify sounds detected.
Classification of beaked whale acoustic events, in particular, require
evaluation of multiple lines of evidence by expert analysts. Here we
present a highly automated approach to acoustic detection and
classification using supervised machine learning and open source
software methods. Data from four large scale surveys of beaked whales
(North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were
analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and
BANTER (hierarchical random forest classifier). Overall correct
classification results ranged from 86\% for the South Atlantic data to
99\% for the North Atlantic. Results for many species could likely be
improved with increased sample sizes and consideration of alternative
automated detectors. These methods provide a highly automated approach
to acoustic detection and classification using open source methods that
can be readily adopted for species and geographic regions.
\end{abstract}

\begin{keyword}
    bioacoustics \sep machine learning \sep random forest \sep species
classification \sep marine mammals \sep passive acoustic
monitoring \sep 
    beaked whale
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

Please make sure that your manuscript follows the guidelines in the
Guide for Authors of the relevant journal. It is not necessary to
typeset your manuscript in exactly the same way as an article, unless
you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see
\url{https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions}

\hypertarget{bibliography-styles}{%
\section{Bibliography styles}\label{bibliography-styles}}

Here are two sample references: \citet{Feynman1963118}
\citet{Dirac1953888}.

By default, natbib will be used with the \texttt{authoryear} style, set
in \texttt{classoption} variable in YAML. You can sets extra options
with \texttt{natbiboptions} variable in YAML header. Example

\begin{verbatim}
natbiboptions: longnamesfirst,angle,semicolon
\end{verbatim}

There are various more specific bibliography styles available at
\url{https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files}.
To use one of these, add it in the header using, for example,
\texttt{biblio-style:\ model1-num-names}.

\hypertarget{using-csl}{%
\subsection{Using CSL}\label{using-csl}}

If \texttt{cite-method} is set to \texttt{citeproc} in
\texttt{elsevier\_article()}, then pandoc is used for citations instead
of \texttt{natbib}. In this case, the \texttt{csl} option is used to
format the references. By default, this template will provide an
appropriate style, but alternative \texttt{csl} files are available from
\url{https://www.zotero.org/styles?q=elsevier}. These can be downloaded
and stored locally, or the url can be used as in the example header.

\hypertarget{equations}{%
\section{Equations}\label{equations}}

Here is an equation: \[ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
\]

Inline equations work as well: \(\sum_{i = 2}^\infty\{\alpha_i^\beta\}\)

\hypertarget{figures-and-tables}{%
\section{Figures and tables}\label{figures-and-tables}}

Figure~\ref{fig-meaningless} is generated using an R chunk.

\begin{figure}

{\centering \includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/figure-pdf/fig-meaningless-1.pdf}

}

\caption{\label{fig-meaningless}A meaningless scatterplot}

\end{figure}

\hypertarget{tables-coming-from-r}{%
\section{Tables coming from R}\label{tables-coming-from-r}}

Tables can also be generated using R chunks, as shown in
Table~\ref{tbl-simple} example.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(}\FunctionTok{head}\NormalTok{(mtcars)[,}\DecValTok{1}\SpecialCharTok{:}\DecValTok{4}\NormalTok{])}
\end{Highlighting}
\end{Shaded}

\hypertarget{tbl-simple}{}
\begin{longtable}[]{@{}lrrrr@{}}
\caption{\label{tbl-simple}Caption centered above table}\tabularnewline
\toprule()
& mpg & cyl & disp & hp \\
\midrule()
\endfirsthead
\toprule()
& mpg & cyl & disp & hp \\
\midrule()
\endhead
Mazda RX4 & 21.0 & 6 & 160 & 110 \\
Mazda RX4 Wag & 21.0 & 6 & 160 & 110 \\
Datsun 710 & 22.8 & 4 & 108 & 93 \\
Hornet 4 Drive & 21.4 & 6 & 258 & 110 \\
Hornet Sportabout & 18.7 & 8 & 360 & 175 \\
Valiant & 18.1 & 6 & 225 & 105 \\
\bottomrule()
\end{longtable}

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

\end{document}

### 2022-11-29 beakerBanter_hawaii_tns.rmd
adding my (taiki) version of hawaii data processing RMD and sample html output

---
title: "beakerbanter_hawaii_tns"
author: "Taiki Sakai"
date: "2022-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# FALSE means it will used saved .rds files instead of re-processing
# TRUE will run everything from scratch - this will take a long time
freshRun <- FALSE
```

## Data Processing for Beaked Whales: Hawaii

Start by loading the required packages

```{r}
library(PAMpal)
library(banter)
library(rfPermute)
```

### PAMpal Processing

Set up our PPS for the Hawaii dataset

```{r, eval=freshRun}
pps <- PAMpalSettings(db='hawaii/Database_JMc/', 
                      binaries = 'hawaii/Binaries_new/',
                      sr_hz='auto', 
                      winLen_sec=.0025, 
                      filterfrom_khz=10, 
                      filterto_khz=NULL)
```

Process data and save to file so we don't have to process again, comment out
processing and saving lines and uncomment reading line if you've already done this.
We'll also check to see if there were any warning messages during processing.

```{r, include=FALSE, eval=!freshRun}
data <- readRDS('hawaii/hiceas_study.rds')
```

```{r, eval=freshRun}
data <- processPgDetections(pps, mode='db', id='Hawaii_BBW')
saveRDS(data, 'hawaii/hiceas_study.rds')
# Double check warning messages
print(getWarnings(data)$message)
```

Assign species IDs according to original Pamguard labels, then relabel them 
for consistency across projects

```{r}
data <- setSpecies(data, 'pamguard')
reSpecies <- readRDS('hawaii/species.rds')
print(reSpecies)
data <- setSpecies(data, 'reassign', value=reSpecies)
```

Filtering out unwanted data before sending off to BANTER. First we only want
a subset of species (and we also rename these for clarity later). Second the
Hawaii data has issues with the second hydrophone, so we filter down to only
the first channel

```{r}
goodSpecies<- c("ZC", "MD", "BW", "BWC", "IP", "BW", "possBW")
data <- filter(data, species %in% goodSpecies)
data <- setSpecies(data, method='reassign',
                   value=data.frame(old=c('ZC', 'MD', 'BWC', 'IP'), new=c("Cuviers", "Blainsvilles", "CrossSeamount", "Indopacetus")))
ch1only <- filter(data, Channel == '1')
```

We also want to use ICI in one of our models, so we'll calculate that.

```{r}
ch1only <- calculateICI(ch1only, time='peakTime')
```

Now we can export for BANTER, dropping species codes that we won't use for
training. We want to compare ICI vs non-ICI models, so we'll export one without
ICI data. We'll save both of these so we have an easy starting point for BANTER.

```{r, eval=freshRun}
bntDataIci <- export_banter(ch1only, dropSpecies = c("BW", "possBW"), training=TRUE)
bntData <- export_banter(ch1only, dropSpecies = c('BW', 'possBW'), 
                         dropVars = c('All_ici'), training=TRUE)
saveRDS(bntData, file='hawaii/hiceas_bntData.rds')
saveRDS(bntData, file='hawaii/hiceas_bntDataIci.rds')
```

### BANTER Training

Train our non-ICI model first, check stability of detector models before running
full banter.

```{r, include=FALSE, eval=!freshRun}
bntMdl <- readRDS('hawaii/hiceas_bntMdl_5e31_10e34.rds')
```

```{r, eval=freshRun}
bntMdl <- initBanterModel(bntData$events)
# was 30e3 / 1 in SRs initial, too much memory. 5e3/1 is 210mb and takes ~10min
bntMdl <- addBanterDetector(bntMdl, bntData$detectors, ntree=5e3, sampsize=1, importance = TRUE)
```

Check for stability to see if we need more trees

```{r}
plotDetectorTrace(bntMdl, detector = paste0('Click_Detector_', 0:2))
plotDetectorTrace(bntMdl, detector = paste0('Click_Detector_', 3:5))
plotDetectorTrace(bntMdl, detector = paste0('Click_Detector_', 6))
summary(bntMdl)
```

Looks fine, run full event banter model

```{r, eval=freshRun}
bntMdl <- runBanterModel(bntMdl, ntree=10e3, sampsize=4)
summary(bntMdl)
```

Looks fine so lets save it. We're storing the tree/sampsize info in the filename

```{r, eval=freshRun}
saveRDS(bntMdl, 'hawaii/hiceas_bntMdl_5e31_10e34.rds')
```

Repeat for ICI model

```{r, include=FALSE, eval=!freshRun}
bntMdlIci <- readRDS('hawaii/hiceas_bntMdlIci_5e31_10e34.rds')
```

```{r, eval=freshRun}
bntMdlIci <- initBanterModel(bntDataIci$events)
# was 30e3 / 1 in SRs initial, too much memory. 5e3/1 is 210mb and takes ~10min
bntMdlIci <- addBanterDetector(bntMdlIci, bntDataIci$detectors, ntree=5e3, sampsize=1, importance = TRUE)
```

Check for stability to see if we need more trees

```{r}
plotDetectorTrace(bntMdlIci, detector = paste0('Click_Detector_', 0:2))
plotDetectorTrace(bntMdlIci, detector = paste0('Click_Detector_', 3:5))
plotDetectorTrace(bntMdlIci, detector = paste0('Click_Detector_', 6))
summary(bntMdlIci)
```

```{r, eval=freshRun}
bntMdlIci <- runBanterModel(bntMdlIci, ntree=10e3, sampsize=4)
summary(bntMdlIci)
```

Looks fine so lets save it. We're storing the tree/sampsize info in the filename

```{r, eval=freshRun}
saveRDS(bntMdlIci, 'hawaii/hiceas_bntMdlIci_5e31_10e34.rds')
```

### BANTER Analytics

First extract just the Random Forest part of our BANTER model for analysis. We'll
use the model with ICI included since that performed better.

```{r}
bntMdlRf <- getBanterModel(bntMdlIci)
```

Confusion Matrix

```{r, include=TRUE}
confusionMatrix(bntMdlRf)
```

Class Priors (Expected Error Rate)

```{r, include=TRUE}
classPriors(bntMdlRf, NULL)
```

Plot Confusion Matrix Heat Map

```{r, include=TRUE}
plotConfMat(bntMdlRf, title="ConfusionMatrix", plot=TRUE)
```

Plot Error Trace for Banter Model

```{r, include=TRUE}
plot(bntMdlRf)
```

PlotVotes

```{r, include=TRUE}
plotVotes(bntMdlRf)
```

Importance Heatmap

```{r, include=TRUE}
plotImportance(bntMdlRf, plot.type="heatmap")
```

Proximity Plot

```{r, include=TRUE}
plotProximity(bntMdlRf)
```

Plot Predicted Probabilities

```{r, include=TRUE}
plotPredictedProbs(bntMdlRf, bins=30, plot=TRUE)
```

### 2022-11-30 manuscript/manuscript.tex

@@
 \usepackage{xcolor}
+\usepackage[normalem]{ulem}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
-Please make sure that your manuscript follows the guidelines in the
-Guide for Authors of the relevant journal. It is not necessary to
-typeset your manuscript in exactly the same way as an article, unless
-you are submitting to a camera-ready copy (CRC) journal.
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-For detailed instructions regarding the elsevier article class, see
-\url{https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions}
+Passive acoustic monitoring (PAM) from has proven to be a valuable tool
+for studying populations of marine mammals (Van Parijs et al., 2009);
+however, the value of these studies depends on our ability to identify
+the sources of the sounds we are monitoring. Historically, experienced
+acousticians manually identify stereotyped sounds that could be reliably
+attributed to a given species, based on their spectral or temporal
+characteristics (Baumann-Pickering et al., 2013; Bittle and Duncan,
+2013; Rankin and Barlow, 2005; Soldevilla et al., 2008). The dramatic
+increase in recordings make it impossible for experienced acousticians
+to manually annotate all data, and there has been an increasing need for
+the development of automated classification routines that can provide
+accurate determinations of the species responsible for the sounds.
 
-\hypertarget{bibliography-styles}{%
-\section{Bibliography styles}\label{bibliography-styles}}
+Beaked whales are deep diving marine mammals found in offshore waters;
+their long dive intervals and cryptic surfacing behavior make them
+difficult to study using typical shipboard visual observation methods
+(MacLeod, 2018). However, beaked whales make stereotyped echolocation
+clicks that have temporal and spatial characteristics that can be used
+to differentiate species (Baumann-Pickering et al.~2013). Detection of
+beaked whale signals in large datasets is greatly expanding our
+understanding of their population structure and the potential impact of
+human activities on these species (Barlow et al., 2021a;
+Baumann-Pickering et al., 2014; Simonis et al., 2020). Unfortunately,
+manual classification of beaked whale echolocation clicks requires
+analysis by trained observers using a number of visual aids to examine
+the call characteristics. Development of automated classification
+routines, if accurate, serve to improve the efficiency and decrease the
+cost of analyzing large datasets.
+
+BANTER is a supervised machine learning acoustic event classifier
+originally developed for classifying dolphin schools from acoustic
+recordings (Rankin et al., 2017). This hierarchical random forest event
+classifier relies on an initial call classifier (in the dolphin case
+study, this included one each for whistles, echolocation clicks, and
+burst pulses) followed by an event classifier. The event classifier
+considers the probability of assignment to species for each call in the
+call classifier, along with any event-level characteristics (such as
+call rate). BANTER is very flexible and can accommodate any number of
+measures from any number of detectors. While BANTER relies on multiple
+call types, relatively minor changes to a call detector could yield
+different detector results (Rankin et al., 2017). In this case, settings
+for a whistle \& moan detector were modified to improve performance at
+detecting burst pulses. This approach of applying multiple call
+detectors (of the same type but with different settings) suggests that
+perhaps BANTER could be used on species that only produce one call type
+(or where only one call type was analyzed), as long as the settings of
+the automated call detector were modified such that the overall number
+of detections was different. This approach was successfully applied to
+classified echolocation clicks for narwhal and belugas (Zahn et al.,
+2021). For species who primarily (or exclusively) produce echolocation
+clicks, BANTER may serve as an option for automated machine learning
+classification.
+
+Here we applied BANTER acoustic classification to beaked whale
+detections from four large acoustic datasets off the US East Coast,
+Hawaii, and the US West Coast. Acoustic data were analyzed by
+experienced acousticians to determine species identity, and these
+`ground truth' classifications served as training data for the
+supervised BANTER trials. Our goals are to identify an efficient and
+accurate automated approach to acoustic classification of beaked whales,
+and to provide a framework for analysis that may serve for other PAM
+studies.
+
+\hypertarget{materials-and-methods}{%
+\section{Materials and Methods}\label{materials-and-methods}}
+
+\hypertarget{field-data-collection}{%
+\subsection{Field Data Collection}\label{field-data-collection}}
+
+Passive acoustic data was collected during three surveys conducted by
+National Oceanographic and Atmospheric Administrations' research
+operations off the US East Coast, the Hawaiian Islands (Hawaii) and the
+US West Coast.
+
+Recordings from the western North Atlantic Ocean (U.S. east coast) were
+collected using towed hydrophone arrays during the 2016 AMAPPS II
+(Atlantic Marine Assessment Program for Protected Species) survey
+(cite). AMAPPS II was subdivided into the northern (NAtlantic) and
+southern (SAtlantic) survey area. The NAtlantic study area ranged from
+Massachusetts south to New Jersey (HB1603, cite) and recordings used in
+this analysis consisted of a single hydrophone (HTI-96-min, High Tech
+Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high
+pass filter (National Instruments USB-6356 A/D card, see DeAngelis et
+al., 2018 for more information). The SAtlantic area ranged from Delaware
+south to central Florida (GU1605, cite) and recordings used in this
+analysis consisted of a single hydrophone (Reson etc) recorded 500 kHz
+sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass
+filter (see XX for more information).
+
+Recordings from Hawaii were collected using drifting acoustic recorders
+during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment
+Survey, Yano et al., 2018).~ Drifting recording buoys were deployed at
+randomly selected locations within the main Hawaiian Islands. Two types
+of drifting recorders were included: (1) Soundtrap ST4300 recorders
+(Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M
+recorders (Wildlife Acoustics, Maynard, MA). Soundtrap recorders were
+sampled at 288 kHz with a duty cycle of 2 min on for every 10 minutes.
+SM3M data were sampled at 256 kHz with continuous recordings. Individual
+buoys were deployed for 10 to 50 days (see Yano et al., 2018 for more
+detailed information).
+
+Recordings from the Eastern North Pacific Ocean off the U.S. west coast
+(EPacific) were collected using drifting acoustic recorders during the
+2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels)
+survey (Keating et al., 2018). Buoys were deployed at a number of
+stations in the California Current, including pre-determined locations
+(for abundance estimation studies) and in close proximity to seamounts
+(Ad Hoc Seamount Experiment). Two types of drifting recorders were
+included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New
+Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics,
+Maynard, MA). Recordings from pre-determined locations were duty cycled
+and recorded 2 minutes out of a variable `off' time; recordings from the
+Seamount Experiment were continuous. Soundtrap recorders were sampled at
+288 kHz and the SM3M recorders were recorded at 256 sample rate (with
+some sampling at 96 kHz for ocean noise measurements). Individual buoys
+were deployed for 11-23 days (see Keating et al.~2018 for more detailed
+information).
+
+\hypertarget{acoustic-data-analysis}{%
+\subsection{Acoustic Data Analysis}\label{acoustic-data-analysis}}
+
+All acoustic recordings were analyzed using PAMGuard software (version
+2.00.15c) with a suite of generic click classifiers within the Click
+Detector module (see Keating and Barlow, 2013).~ Classifier sets were
+saved such that any click may be classified as more than one click type.
+The general click classifiers were effectively click detectors, as each
+implemented a simple spectral band settings (e.g., 2- 15 kHz, 15 -- 30
+kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \textgreater{} 80 kHz). An
+additional detector/classifier within the 30 -- 50 kHz peak frequency
+range considered the presence of a frequency sweep that is
+characteristic of beaked whale pulses (see Keating and Barlow, 2013).
+
+Acoustic detection of beaked whale events were then classified to
+species by experienced acoustic researchers using multiple lines of
+evidence. If beaked whales were sighted in close proximity to the
+recording device, the visual confirmation of species identify was used
+to determine species identity. Otherwise, experienced acousticians
+followed strict protocol to define species identify based on acoustic
+characteristics. When the available characteristics were inconclusive,
+the species was considered an unidentified beaked whale; only acoustic
+events classified to species were considered for this study. Acoustic
+researchers with experience in classifying beaked whale echolocation
+clicks examined the bearing-time display for the click detector module
+in PAMGuard to assess the species identity of the acoustic event. For
+NEFSC data, acoustic detections of beaked whales using towed hydrophone
+arrays were further subdivided into click trains based on identification
+of consecutive clicks along the same bearing angle in the bearing time
+plot (see DeAngelis et al., 2018). Data were saved within the PAMGuard
+database and binaries for downstream processing using PAMpal package
+(Sakai, 2021) in R programming language (R Core Team 2014).
+
+Standard click calculations were measured from echolocation clicks using
+default values in PAMpal. ~For EPacific data, any events \textgreater{}
+2 minutes were subdivided into 2-minute events to ensure that all events
+were of equal value (because most of our data was duty cycled into
+2-minute blocks). Inter-click interval (ICI) was calculated using the
+`calculateICI' function in PAMpal; the mode of the ICI was identified
+for each beaked whale acoustic event. In addition, for EPacific data,
+used the `browseEdinfo' and `matchEnvData' function in PAMpal to
+identify appropriate ERDAPP environmental data (cite), including: sea
+surface temperature (from jplMURSST41), productivity (from~
+~erdMH1pp8day dataset), chlorophyll (from erdMBchla8day dataset), and
+seafloor gradient (x,y, and gradient magnitude from
+rdSrtm30plusSeafloorGradient).
+
+Multiple trials were conducted to better understand the value (and
+potential limitations) of the different suite of measurements. Data
+subsets for each trial were exported for BANTER classification using the
+export\_banter function in PAMpal. At a minimum, each trial included
+standard click calculations for all clicks in a beaked whale acoustic
+event (EC); and some trials included the mode of the ICI for each event
+(EC\_ICI). For the EPACIFIC data, additional trials that also included
+water depth (EC\_ICI\_D) and environmental data (EC\_ICI\_ENV) were run
+for each event.
+
+\hypertarget{banter-acoustic-classification}{%
+\subsection{BANTER Acoustic
+Classification}\label{banter-acoustic-classification}}
+
+BANTER models was created for each trial using the BANTER package (cite)
+in R programming language (R Core Team 2014). Standard click
+calculations were used for the BANTER call classifier; event level
+measures of ICI and environmental features (for EPacific data) were
+applied to the BANTER event classifier. A minimum of two events per
+species and complete suite of variables (absence of NAs) were required
+for training and testing of the classification model. The BANTER random
+forest models contain two parameters, ntree and sampsize, which must be
+selected to ensure stability of the model. The sample size (sampsize) is
+the number of events considered for each trial of the model, and the
+ntree is the number of trees (iterations of the model) run in a given
+model. These values were modified to obtain the best results, with an
+emphasis on lower sample size and higher number of trees for
+computational performance (Rankin and Archer, 2021).
+
+The model output include the confusion matrix, and a number of
+visualizations provided by the rfpermute package (cite) in R programming
+language (R Core Team 2014).
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed
+for this study; species included: Baird's beaked whales (\emph{Berardius
+bairdii}), Blainsville's beaked whales (\emph{Mesoplodon densirostris}),
+Cuvier's beaked whales (\emph{Ziphius cavirostris}), Cross Seamount
+beaked whale (McDonald et al., 2009), Gervais' beaked whales
+(\emph{Mesoplodon europaeus}), Longman's beaked whale (\emph{Indopacetus
+pacificus}), Sowerby's beaked whales (\emph{Mesoplodon bidens}), True's
+beaked whales (\emph{Mesoplodon mirus}), Stejneger's beaked whale
+(\emph{Mesoplodon stejnegeri}), and unidentified beaked whale `BW43'
+(Baumann-Pickering et al., 2013). Analyses were conducted separately for
+the different study areas due to differences in the data collection
+methods that precluded combining of datasets. Specific species
+combinations and sample sizes varied by study area.
+
+\hypertarget{natlantic}{%
+\subsection{NAtlantic}\label{natlantic}}
+
+{[}\emph{Note: Last reprocess had both models with same classification
+results (though ICI was `important', it did not really change the
+results). I need to reword a few things to accommodate this (and waiting
+on confirmation that re: ptoential changes to one function in
+PAMpal}){]}
+
+Species encountered in the NAtlantic dataset include Cuvier's beaked
+whales (n=70), Sowerby's beaked whales (n=5), True's beaked whales
+(n=65), and Gervais' beaked whales (n=3). BANTER classification trials
+included (1) echolocation clicks (EC), and (2) echolocation clicks and
+ICI (EC\_ICI).
+
+For the NAtlantic dataset, the most accurate model considered the
+echolocation clicks and the ICI for an event (EC\_ICI, Detector Model:
+sampsize = 4, ntrees = 30,000; BANTER Model: sampsize = 4,
+trees=300,000) and provided an overall correct classification rate of
+100\% for all four species (Fig. 1). There were insufficient samples of
+Gervais' beaked whales to include in the NAtlantic model (n=2). All
+classification results were greater than the expected error rate (Prior
+in Fig. 1a). Results from the EC model are presented here; results from
+the EC\_ICI model can be found in supplementary materials (Supplement
+Fig. 1).
+
+{[}Insert Figure 1{]}{[}Figure 1. BANTER classification results from the
+NAtlantic dataset (EC). Confusion matrix (a) provides the percent
+correct classification or each species (pct.correct), lower confidence
+intervals (LCI\_0.95), upper confidence intervals (UCI\_0.95), and
+priors (expected error rate). Proximity plot (b) for species events from
+BANTER model (central dot color represent true species identity; color
+of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of variable importance for each species; colors
+scale from most important predictors (dark red) to least important
+predictors (dark blue). Vote Plot (d) shows the vote distribution for
+each event (vertical slice) for each species; distribution of votes by
+species is shown by their representative color.
+
+The proximity plot (Fig.1b) provides a view of the distribution of
+events within the feature space of the two `most important' dimensions,
+or predictors. For each event in the plot, the color of the central dot
+represents the true species identity, and the circle represents the
+BANTER classification result. The importance heat map (Fig. 1c) shows
+that the most important features for predicting each species.
+
+The strength of the classifications by species can be seen in the plot
+votes (Fig. 1d), which show the istribution of the votes for each of the
+300,000 trials. For each plot, the events are represented as vertical
+slices (with plot subdivided along the x-axes according to the event
+sample size), and the percentage of votes for each species is
+represented by their color along that vertical slice. In a perfect
+scenario, all 300,000 trees would have voted for the correct species and
+therefore the classification would be correct AND the strength of these
+classifications would be maximized. The plot vote for the NAtlantic
+dataset show that all species
+
+\hypertarget{satlantic}{%
+\subsection{SAtlantic}\label{satlantic}}
+
+Species encountered in the SAtlantic study dataset include Blainsville's
+beaked whales (n=35), Cuvier's beaked whales (n=66), Gervais' beaked
+whales (n=45), Sowerby's beaked whales (n=2), and True's beaked whales
+(n-12). BANTER classification trials included (1) echolocation clicks
+(EC), and (2) echolocation clicks and ICI (EC\_ICI).
+
+For the SAtlantic dataset, the most accurate model considered the
+echolocation clicks and the ICI for an event (Detector Model: sampsize =
+2, ntrees = 30,000; BANTER Model: sampsize = 1, ntrees = 300,000) and
+provided an overall correct classification rate of 86.9\% for all four
+species. The sample sizes for the BANTER model was kept low (sampsize =
+1) to retain Sowerby's beaked whale, which had a sample size of 2.
+Analysis with a larger sample size resulted in improved classification
+results for Cuvier's beaked whale (see Supplement Fig. 3 for
+EC\_ICI\_ALT), but resulted in the loss of Sowerby's in the final model.
+Classification scores ranged from a low of 74.2\% (Cuvier's beaked
+whale) to 100\% (Sowerby's and True's beaked whale, Fig. 2); all
+classification results were greater than the expected error rate (Prior
+in Fig. 2a). Results from the EC\_ICI model are presented here; results
+from the EC model can be found in supplementary materials (Supplement
+Fig. 2).
+
+\begin{figure}
+
+\begin{minipage}[b]{0.50\linewidth}
+
+{\centering
+
+\begin{longtable}[]{@{}lllll@{}}
+\toprule()
+& Chinstrap & Gentoo & Adelie & \textbf{\% Correct} \\
+\midrule()
+\endhead
+Chinstrap & 70 & 0 & 0 & 100.0 \\
+Gentoo & 0 & 5 & 0 & 100.0 \\
+Adelie & 0 & 0 & 65 & 100.0 \\
+\textbf{Overall} & - & - & - & 100.0 \\
+\bottomrule()
+\end{longtable}
+
+}
+
+\end{minipage}%
+%
+\begin{minipage}[b]{0.50\linewidth}
+
+{\centering
+
+\begin{longtable}[]{@{}lll@{}}
+\toprule()
+LCI 0.95 & UCI 0.95 & Prior \\
+\midrule()
+\endhead
+94.9 & 100.0 & 50.0 \\
+47.8 & 100.0 & 3.6 \\
+94.5 & 100.0 & 46.4 \\
+97.4 & 100.0 & 46.7 \\
+\bottomrule()
+\end{longtable}
+
+}
+
+\end{minipage}%
+\newline
+\begin{minipage}[b]{0.33\linewidth}
+
+{\centering
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi})}
+\tightlist
+\item
+  \includegraphics{Figures/myplot1.jpg}
+\end{enumerate}
+
+}
+
+\end{minipage}%
+%
+\begin{minipage}[b]{0.33\linewidth}
+
+{\centering
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi})}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  \includegraphics{Figures/myplot2.jpg}
+\end{enumerate}
+
+}
+
+\end{minipage}%
+%
+\begin{minipage}[b]{0.33\linewidth}
+
+{\centering
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi})}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  \includegraphics{Figures/myplot3.jpg}
+\end{enumerate}
+
+}
+
+\end{minipage}%
+\newline
+\begin{minipage}[b]{0.33\linewidth}
+
+{\centering
+
+Figure 2. BANTER classification results from the SAtlantic dataset.
+Confusion matrix provides the percent correct classification for each
+species (pct.correct), lower confidence intervals (LCI\_0.95), upper
+confidence intervals (UCI\_0.95), priors (expected error rate).
+Proximity plot (a) for species events from BANTER model (central dot
+color represents true species identity; color of circle surrounding dot
+represents BANTER species classification). Heat map (b) for ranks of
+variable importance for each species; colors scale from most important
+predictors (dark red) to least important predictors (dark blue). Vote
+Plot (c) shows the vote distribution for each event (vertical slice) for
+each species; distribution of votes by species is shown by their
+representative color.
+
+}
+
+\end{minipage}%
+
+\end{figure}
+
+For SAtlantic, the proximity plot (Fig. 2b) shows that True's beaked
+whales are differentiated from the other species based solely on the two
+most important features. The importance heat map (Fig. 2c) shows that
+Cuvier's, Gervais' and Sowerby's beaked whales rely on a number of
+predictors that are lower in importance and there is likely greater
+distinction between these species than is represented by the proximity
+plot (Fig. 2b). The plot votes (Fig. 2d) shows a moderate number of
+trees in the forest were misclassified and these misclassifications
+included all species in the model.
+
+\hypertarget{hawaii}{%
+\subsection{Hawaii}\label{hawaii}}
+
+The Hawaii dataset consisted of 13 drifting buoys deployed within the
+main Hawaiian Islands (Yano et al.~2018); species included Longman's
+beaked whale (n=121), Blainsville's beaked whale (n=518), Cross Seamount
+beaked whale (n=84) and Cuvier's beaked whale (n=201). BANTER
+classification trials included (1) echolocation clicks (EC), and (2)
+echolocation clicks and ICI (EC\_ICI).
+
+The most accurate model considered the echolocation clicks only
+(Detector Model: sampsize = 10, ntrees = 30,000; BANTER model: sampsize
+= 4, ntree = 200,000) and provided an overall correct classification
+rate of 89.7\% for all four species. Classification scores ranged from a
+low of 82.1\% (Cross Seamount Beaked Whale) to 96.7\% (Longman's beaked
+whale, Fig. 3). ~All classification results were greater than the
+expected error rate (Prior in Fig. 3a). Results from the EC model are
+presented here; results from EC\_ICI model can be found in supplementary
+materials (Supplement Fig. 4).
+
+{[}Insert Figure 3{]}{[}Figure 3. BANTER classification results from the
+Hawaii dataset. Confusion matrix (a) provides the percent correct
+classification for each species (pct.correct), lower confidence
+intervals (LCI\_0.95), upper confidence intervals (UCI\_0.95), priors
+(expected error rate). Proximity plot (b) for species events from BANTER
+model (central dot color represents true species identity; color of
+circle surrounding dot represents BANTER species classification). Heat
+map (c) for ranks of variable importance for each species; colors scale
+from most important predictors (dark red) to least important predictors
+(dark blue). Vote Plot (d) shows the vote distribution for each event
+(vertical slice) for each species; distribution of votes by species is
+shown by their representative color. Species codes include Cross
+Seamount beaked whale (BWC), Longman's beaked whale (IP), Blainsville's
+beaked whale (MD) and Cuvier's beaked whale (ZC).
+
+There are clear distinctions in the primary features used to classify
+three of the four species as shown in the Proximity Plot (Fig. 3b) which
+shows the classification results based on the two most `important'
+predictors. Distinctive clusters based on the two most important
+features can be found for Longman's (IP), Blainsville's (MD), and
+Cuvier's beaked whales (ZC), and there is overlap in the feature space
+of the clusters for Cuvier's and Cross Seamount beaked whales (BWC).
+This overlap is resolved with different features for ZC
+(3\textsuperscript{rd}, 7\textsuperscript{th} in Fig. 3c heat map) and
+BWC (5\textsuperscript{th}, 8\textsuperscript{th} feature in heat map),
+as evidence in the low misclassification rates for these species (Fig.
+3a confusion matrix). The plot votes graph (Fig. 3d) show complicated
+results with variation in the proportion of trees that correctly
+classified the species for each event.
+
+\hypertarget{eastern-pacific}{%
+\subsection{Eastern Pacific}\label{eastern-pacific}}
+
+The EPacific dataset consisted of 25 drifting recording buoys (20
+stations plus 5 seamount experiment drifts) deployed off the west coast
+of the United States (Keating et al.~2018); species included Baird's
+beaked whale (n=91), Unidentified Beaked Whale `BW43' (n=11),
+Stejneger's beaked whale (n=41) and Cuvier's beaked whale (n=1,332).
+BANTER classification trials included (1) echolocation clicks (EC), (2)
+echolocation clicks and ICI (EC\_ICI), (3) echolocation clicks, ICI, and
+water depth (EC\_ICI\_D), and (4) echolocation clicks, ICI, water depth,
+sea surface salinity, productivity, and chlorophyll (EC\_ICI\_ENV).
+
+The best model considered the echolocation clicks and the ICI values
+(Detector Model: sampsize = 5, ntrees = 30,000; BANTER model: sampsize =
+4, ntree = 200,000) and provided an overall correct classification rate
+of 93.7\% for all four species. Classification scores ranged from a low
+of 93.2\% (Cuvier's beaked whale) to 100\% (BW43, Fig. 4a). All
+classification results were greater than the expected error rate (Prior
+in Fig. 4a). Results from the EC\_ICI model are presented here; complete
+results can be found in supplementary materials (Supplement Figs. 5-7).
+
+{[}Insert Figure 4{]}{[}Figure 4. BANTER classification results from the
+EPacific dataset. Confusion matrix (a) provides the percent correct
+classification for each species (pct.correct), lower confidence
+intervals (LCI\_0.95), upper confidence intervals (UCI\_0.95), priors
+(expected error rate).~ Proximity plot (b) for species events from
+BANTER model (central dot color represents true species identity; color
+of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of variable importance for each species; colors
+scale from most important predictors (dark red) to least important
+predictors (dark blue). Vote Plot (d) shows the vote distribution for
+each event (vertical slice) for each species; distribution of votes by
+species is shown by their representative color. Species codes include
+Baird's beaked whale (BB), unidentified beaked whale (BW43), Stejneger's
+beaked whale (MS) and Cuvier's beaked whale (ZC).{]}
+
+There are clear distinctions in the primary features used to classify
+the species as shown in the Proximity Plot (Fig. 4b) which shows the
+classification results based on the two most `important' predictors. The
+species each form distinctive clusters based on these two features;
+however, many events fall outside these distinct bounds (detections
+shown in area between species classification clusters). Both BB and BW43
+relied on a wide range of variables for classification (Fig. 4c heat
+map). The plot votes (Fig. 4d) show strong classification results for
+most events of most species.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+The results for the four study areas varied from a low of 86.9\% correct
+classification in the SAtlantic dataset to a high of 100\% correct
+classification in the NAtlantic dataset. All species in all datasets had
+classifications rates well above those expected by chance; however
+Cuvier's (SAtlantic, Hawaii), BWC (Hawaii), and Blainsville's beaked
+whales (Hawaii) had classification rates below 90\%. There were
+insufficient sample sizes of Gervais to include in the NAtlantic
+dataset.
+
+Several of our datasets included very low sample sizes for some species;
+however, low sample sizes did not always result in low classification
+scores. In the Atlantic dataset, Sowerby's beaked whale was consistently
+correctly classified in both models (NAtlantic and SAtlantic); however,
+while the plot votes showed very strong classification strength in the
+NAtlantic, the classification strength (represented by the number of
+trees that voted for the correct classification) was poor in the
+SAtlantic study area.This is likely due to both a small sample size of
+Sowerby's detected in the S. study area (n=2) combined with the small
+sampsize used in the BANTER model. In general, larger sample sizes
+should lead to improved overall classification results, and improved
+strength of these classification results (as identified by plot vote
+graph).
+
+While some species have distinctive differences in their click
+characteristics that lead to strong classification despite small sample
+sizes (e.g.~Sowerby's beaked whales in NAtlantic); other species, such
+as Cuvier's beaked whales, have significant variation in their click
+measurements. ~For species with high variability in the predictor
+variables and an overlap in the range of these variables with other
+species, a large sample size is required to describe the true
+variability of these call measurements. Even with reasonably large
+sample sizes, the classification may suffer (e.g, SAtlantic Cuvier's
+beaked whale = 74\%). We found by increasing the sampsize in BANTER, we
+could increase the classification rate for this species in this area to
+(97\%, see Supplement Fig.3); however, this resulted in the loss of
+Sowerby's beaked whale in the final model. So, an increased sample size
+in Sowerby's would likely lead to increased classification results for
+other species in the model.
+
+Previous application of BANTER to dolphin species in the California
+Current found that large sample sizes could result in strong
+classification of species where experienced acousticians are unable to
+differentiate species (long-beaked and short-beaked common dolphins,
+Rankin et al., 2017). This suggests that large increases in sample size
+may improve classification results for Cuvier's beaked whales.
+
+Researchers have noted several `types' of click characteristics for
+Cuvier's beaked whales (ADA, pers. comm) which may be related to
+variability in the propagation environment, different animal behaviors,
+gender or age classes, or acoustic populations. Future work should
+examine if these Cuvier's beaked whale events can be clustered in an
+objective and stable manner; if so, then these can be input into BANTER
+models as separate types of Cuvier's beaked whales. We recommend caution
+in this approach- this should only be considered if there is reasonable
+evidence that these different groups represent authentic cluster. We do
+not recommend segregation of misclassified subgroups simply to improve
+classification scores.
+
+Gervais' and True's beaked whales have similar click characteristics
+that make them difficult to differentiate, and require considerable
+expertise to classify manually, using a variety of available visuals and
+metrics (DeAngelis et al., 2018). Our results suggest that BANTER may
+serve as an efficient and effective means of classifying and
+differentiating these two species. Despite modest sample sizes, Gervais'
+and True's beaked whales showed high classification scores (95.6\% and
+100\%, respectively). Unfortunately, we were unable to combine the N and
+S Atlantic datasets (which would increase sample sizes of True's beaked
+whales) due to differences in hydrophone characteristics that result in
+differences in call measures. Calibration of signals (to make them
+comparable) may allow for combining datasets to improve sample sizes.
+
+For the SAtlantic models, ICI was found to be an important predictor for
+many species {[}\emph{talk about the 1 change in sowerby's when the ICI
+was added- do you think it was because of an error in ICI?}{]} . For
+Hawaii and the EPacific, however, we found that adding ICI into the
+BANTER classifier did not improve final classification results. The
+differences here are likely due to different analysis methods for
+identifying events. Beaked whale events were defined by encounter time
+for both the Hawaii and EPacific data, which were collected using
+drifting acoustic recorders where animals could not be localized. The
+Atlantic data were collected using a towed hydrophone array with
+localization capabilities. Each event was defined as an individual click
+train, so ICI for these events more likely represented true ICI, and was
+found to be more informative. Baumann-Pickering et al.~(2013) showed an
+overlap in the ICI between Cuvier's and Blainsville's beaked whales, but
+a strong difference in ICI between these species and BWC. Likewise,
+Baumann-Pickering (2013) found a difference in ICI between Baird's
+beaked whales and Cuvier's beaked whales. Subdividing click trains in
+PAMGuard is time consuming, but alternative options include
+consideration of the PAMGuard click train detector module in BANTER, or
+by developing a similar function in R to apply to events. Integration of
+ICI from individual click trains may serve to improve classification
+results for some species in Hawaii and EPacific.
+
+Gervais were eliminated from the NAtlantic BANTER model due to low
+sample size (n=2); results were extremely poor when they were included
+in the model. Analysis notes indicate uncertainty in manual species
+classification on one of the two events in this dataset. While small
+sample size can at times provide good classification scores, as we found
+in Sowerby's beaked whale in this SAtlantic dataset, they can be
+devastating if there are inaccuracies in the training data. These two
+examples of small sample sizes highlights a conflict in preferred
+protocol. Ideally, (1) all \uline{species} would be included in a
+classification model, so as to better represent the local species
+diversity, (2) all \uline{events} would ideally be included in the
+classification model, so as to better represent the variability found in
+the area, and (3) only confident `ground truth' classifications would be
+considered in the training model. Unfortunately, in species where
+identity must be determined based on call characteristics (rather than
+visual confirmation of species identity in the field), it can sometimes
+be difficult to confidently determine species identity when calls are
+highly variable and do not include at least one call that provides high
+confidence of species identity. In the case of beaked whales, we
+recommend that training data include a reasonable level of confidence
+for inclusion. When sample sizes are low, such as in these cases, it may
+be wise to consider the strength of the classifications when including
+them for model training (as we have done here). An alternative is to
+require agreement from multiple analysts, as provided by the EPacific
+datasets. This concern about accurate labeling of training data is
+further complicated by the potential existence of species that have not
+yet been identified. For example, Barlow et al.~(2021b) recently
+discovered what appears to be a new species of beaked whale off Baja
+California, Mexico; this putative new species may have been detected but
+misclassified in other datasets (including the EPacific data presented
+here).
+
+While BANTER provides an efficient and consistent approach to
+classification, there are significant limitations that must be
+considered. BANTER is a supervised machine learning tool and requires
+reliable training data for success. Training data should consist of
+labels with strong confidence in species identify (ideally determined by
+agreement from more than one analyst) and sample sizes should be large
+enough to explain the natural variability in the data. In fact,
+consideration of the strength of classification scores may help identify
+cases where the training data were of poor quality, or where there may
+be evidence to identify different acoustic populations or behavior
+states.
+
+The workflow presented here provides a highly automated approach to
+detection of acoustic events (PAMGuard), integration of environmental
+data (PAMpal) and acoustic event classification (BANTER). These methods
+significantly reduce manual analysis, provide more consistent
+classification results with fewer biases, and provide an estimate of
+classification error.~ The greatest improvement to classification
+results for beaked whales would likely result from improved sample sizes
+and examination of individual click trains in measuring ICI.
+Consideration of additional detectors (e.g.~matched filter detector,
+click train detector) or additional environmental variables may further
+improve classification results. These highly automated methods may allow
+for analysis of data across large spatial and temporal scales to address
+large ecological and population level questions.
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+The authors would like to acknowledge the large number of scientists and
+shipboard crew who were responsible for data collection during multiple
+large scale studies. Funding for data collection and analysis were
+provided by U.S. Navy, Bureau of Ocean Energy Management, and the
+National Oceanographic and Atmospheric Administration. The manuscript
+was improved thanks to reviews from {[}XXXX{]}.
 
@@
 and stored locally, or the url can be used as in the example header.
-
-\hypertarget{equations}{%
-\section{Equations}\label{equations}}
-
-Here is an equation: \[ 
-  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
-  \left(\frac{x}{\beta}\right)^{\alpha-1}
-  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
-  \alpha,\beta,x > 0 .
-\]
-
-Inline equations work as well: \(\sum_{i = 2}^\infty\{\alpha_i^\beta\}\)
 
@@
 
-\renewcommand\refname{References}
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+Barlow, J., Cárdenas‐Hinojosa, G., Henderson, E.E., Breese, D.,
+López‐Arzate, D., Hidalgo Pla, E. and Taylor, B.L., 2021. Unique
+morphological and acoustic characteristics of beaked whales
+(\emph{Mesoplodon} sp.) off the west coast of Baja California, Mexico.
+Marine Mammal Science. https://doi.org/10.1111/mms.12853
+
+Barlow, J., Moore, J.E., McCullough, J.L. and Griffiths, E.T., 2021.
+Acoustic‐based estimates of Cuvier's beaked whale (\emph{Ziphius
+cavirostris}) density and abundance along the US West Coast from
+drifting hydrophone recorders. Mar.~Mamm. Sci.
+\url{https://doi.org/10.1111/mms.12872}
+
+Baumann-Pickering, S., McDonald, M.A., Simonis, A.E., Solsona Berga, A.,
+Merkens, K.P., Oleson, E.M., Roch, M.A., Wiggins, S.M., Rankin, S.,
+Yack, T.M. and Hildebrand, J.A., 2013. Species-specific beaked whale
+echolocation signals.~J. Acoust. Soc. Am.~134(3), 2293-2301.
+https://doi.org/10.1121/1.4823843
+
+Baumann-Pickering, S., Roch, M.A., Brownell Jr, R.L., Simonis, A.E.,
+McDonald, M.A., Solsona-Berga, A., Oleson, E.M., Wiggins, S.M. and
+Hildebrand, J.A., 2014. Spatio-temporal patterns of beaked whale
+echolocation signals in the North Pacific.~PLoS One. ~9(1), e86072.
+https://doi.org/10.1371/journal.pone.0086072
+
+Baumann-Pickering, S., Wiggins, S.M., Roth, E.H., Roch, M.A.,
+Schnitzler, H.U. and Hildebrand, J.A., 2010. Echolocation signals of a
+beaked whale at Palmyra Atoll.~J. Acoust. Soc. Am.~127(6), 3790-3799.
+https://doi.org/10.1121/1.3409478
+
+Bittle, M. and Duncan, A., 2013. A review of current marine mammal
+detection and classification algorithms for use in automated passive
+acoustic monitoring. In Proceedings of Acoustics: Science, Technology
+and Amenity. 2013, 1-8.
+
+DeAngelis, A.I., Stanistreet, J.E., Baumann-Pickering, S. and Cholewiak,
+D.M., 2018. A description of echolocation clicks recorded in the
+presence of True's beaked whale (\emph{Mesoplodon mirus}). J. Acoust.
+Soc. Am.~144(5), 2691-2700. https://doi.org/10.1121/1.5067379
+
+Keating JL, Barlow J., 2013. Click detectors and classifiers used during
+the 2012 southern California behavioral response study. NOAA Tech Memo
+NOAA-TM-NMFS-SWFSC-517:1-14.
+
+MacLeod, C.D., 2018. Beaked whales, overview, in: Encyclopedia of marine
+mammals. Academic Press, pp 80-83.
+
+McDonald, M.A., Hildebrand, J.A., Wiggins, S.M., Johnston, D.W. and
+Polovina, J.J., 2009. An acoustic survey of beaked whales at Cross
+Seamount near Hawaii. J. Acoust. Soc. Am.~ 125(2), 624-627.
+https://doi.org/10.1121/1.3050317
+
+R Core Team, 2014. R: A language and environment for statistical
+computing. R Foundation for Statistical Computing, Vienna, Austria. URL
+http://www.R-project.org/
+
+Rankin, S., and Archer, F., 2021. BANTER: A User's Guide to Acoustic
+Classification. https://taikisan21.github.io/PAMpal/banterGuide.html
+
+Rankin, S., Archer, F., Keating, J.L., Oswald, J.N., Oswald, M., Curtis,
+A. and Barlow, J., 2017. Acoustic classification of dolphins in the
+California Current using whistles, echolocation clicks, and burst
+pulses.~Mar.~Mamm. Sci. 33(2), 520-540.
+https://doi.org/10.1111/mms.12381
+
+Rankin, S. and Barlow, J., 2005. Source of the North Pacific ``boing''
+sound attributed to minke whales. J. Acoust. Soc. Am.~118(5), 3346-3351.
+https://doi.org/10.1121/1.2046747
+
+Soldevilla, M.S., Henderson, E.E., Campbell, G.S., Wiggins, S.M.,
+Hildebrand, J.A. and Roch, M.A., 2008. Classification of Risso's and
+Pacific white-sided dolphins using spectral properties of echolocation
+clicks. J. Acoust. Soc. Am.~124(1), 609-624.
+https://doi.org/10.1121/1.2932059
+
+Sakai, T. 2021. PAMpal: Load and Process Passive Acoustic Data. R
+package version 0.13.0.
+
+Simonis, A.E., Brownell Jr, R.L., Thayre, B.J., Trickey, J.S., Oleson,
+E.M., Huntington, R. and Baumann-Pickering, S., 2020. Co-occurrence of
+beaked whale strandings and naval sonar in the Mariana Islands, Western
+Pacific. Proc. Roy. Soc. B. 287(1921), p.20200070.
+https://doi.org/10.1098/rspb.2020.0070
+
+Van Parijs, S.M., Clark, C.W., Sousa-Lima, R.S., Parks, S.E., Rankin,
+S., Risch, D. and Van Opzeeland, I.C., 2009. Management and research
+applications of real-time and archival passive acoustic sensors over
+varying temporal and spatial scales. Mar.~Ecol. Progr. Ser. 395,
+pp.21-36. https://doi.org/10.3354/meps08123
+
+Zahn, M.J., Rankin, S., McCullough, J.L., Koblitz, J.C., Archer, F.,
+Rasmussen, M.H. and Laidre, K.L., 2021. Acoustic differentiation and
+classification of wild belugas and narwhals using echolocation clicks.
+Sci. Rep.~11(1), pp.1-16. https://doi.org/10.1038/s41598-021-01441-w
+
   \bibliography{bibliography.bib}

### 2022-12-14 hawaii/beakerBanter_hawaii.Rmd
Add Hawaii Dataset
Run Hawaii Dataset. Gitignore should include the binaries/database folders in the Hawaii folder dataset

---
title: "beakerbanter_hawaii"
author: "Shannon Rankin"
date: "2022-12-07"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = TRUE)
# When this rmarkdown file is knit:
# FALSE means it will used saved .rds files instead of re-processing
# TRUE will run everything from scratch - this will take a long time
freshRun <- FALSE
```

## Data Processing for Beaked Whales: Hawaii

Start by loading the required packages

```{r, results=FALSE}
library(PAMpal)
library(banter)
library(rfPermute)
library(here)
here()
```

### PAMpal Processing

1.  Set up our PPS (PAMPal Settings Object) for the Hawaii dataset

```{r, eval=freshRun}
pps <- PAMpalSettings(db='Database/', 
                      binaries = 'Binaries/',
                      sr_hz='auto', 
                      winLen_sec=.0025, 
                      filterfrom_khz=10, 
                      filterto_khz=NULL)
```

2.  Process data and save to file to eliminate repeated processing.

***If*** this is the initial processing, ensure you have set 'freshRun =
TRUE' in code chunk at the top of this document so that you will run the
processPGDetections code and save the data as an rds file. In this case,
it will take some time to run, and we will check for warning messages
during processing.

```{r, eval=freshRun}
data <- processPgDetections(pps, mode='db', id='hawaii_bw')
saveRDS(data, 'hawaii_study.rds')
# Double check warning messages
print(getWarnings(data)$message)
```

***If*** you have already run the processing code, ensure you have set
'freshRun = FALSE' in code chunk at the top of this document so that you
will read in the existing .rds file and check for warning messages.

```{r, include=FALSE, eval=!freshRun}
data <- readRDS('hawaii_study.rds')
# Double check warning messages
print(getWarnings(data)$message)
```

3.  Assign species identity according to original PAMguard labels, then
    relabel for consistency across projects.

```{r}
data <- setSpecies(data, 'pamguard')
reSpecies <- readRDS('species.rds')
print(reSpecies)
data <- setSpecies(data, 'reassign', value=reSpecies)
```

4.  Filter out unwanted data: (1) subset and rename species for BANTER
    model and (2) filter to retain only Channel 1.

```{r}
goodSpecies<- c("ZC", "MD", "BW", "BWC", "IP", "BW", "possBW")
data <- filter(data, species %in% goodSpecies)
data <- setSpecies(data, method='reassign',
                   value=data.frame(old=c('ZC', 'MD', 'BWC', 'IP'), new=c("Cuviers", "Blainsvilles", "CrossSeamount", "Indopacetus")))
data_ch1only <- filter(data, Channel == '1')
```

5.  Calculate Inter-Click Interval (ICI).

```{r}
data_ch1only <- calculateICI(data_ch1only, time='peakTime')
```

6.  Export data for BANTER (and drop species codes that will not be used
    for training). We will create two datasets: one with ICI and one
    without ICI, and save these for import into BANTER.

```{r, eval=freshRun}
banter_data <- export_banter(data_ch1only, dropSpecies = c('BW', 'possBW'), 
                         dropVars = c('All_ici'), training=TRUE)
saveRDS(banter_data, file='hawaii_banter_data.rds')

banter_data_ici <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), training=TRUE)
saveRDS(banter_data_ici, file='hawaii_banter_data_ici.rds')
```

### Build a BANTER Classification Model

#### *Model without ICI*

Load data, initialize a BANTER model, and create an initial BANTER
detector model (stage 1).

```{r, include=FALSE, eval=!freshRun}
banter_model <- readRDS('hawaii_banter_model_t5e3s10_t10e3s4.rds')
```

```{r, eval=freshRun}
banter_model <- initBanterModel(banter_data$events)
banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=5e3, sampsize=10, importance = TRUE)
```

Check for stability (are more trees needed?) before moving on

```{r}
plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 0:2))
plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 3:5))
plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 6))
summary(banter_model)
```

Run BANTER Event Model (stage 2)

```{r, eval=freshRun}
banter_model <- runBanterModel(banter_model, ntree=10e3, sampsize=4)
summary(banter_model)
```

Check for stability (are more trees needed?). If acceptable, save model
with tree/sampsize info in the filename.

```{r, eval=freshRun}
saveRDS(banter_model, 'hawaii_banter_model_t5e3s10_t10e3s4.rds')
```

#### *ICI Model*

Load data, initialize a BANTER model, and create an initial BANTER
detector model (stage 1).

```{r, include=FALSE, eval=!freshRun}
banter_model_ici <- readRDS('hawaii_banter_model_ici_t5e3s5_t10e3s4.rds')
```

```{r, eval=freshRun}
banter_model_ici <- initBanterModel(banter_data_ici$events)
# was 30e3 / 1 in SRs initial, too much memory. 5e3/1 is 210mb and takes ~10min
banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=5e3, sampsize=5, importance = TRUE)
```

Check for stability to see if we need more trees

```{r}
plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 0:2))
plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 3:5))
plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 6))
summary(banter_model_ici)
```

Run BANTER Event Model (stage 2)

```{r, eval=freshRun}
banter_model_ici <- runBanterModel(banter_model_ici, ntree=10e3, sampsize=4)
summary(banter_model_ici)
```

Check for stability (are more trees needed?). If acceptable, save model
with tree/sampsize info in the filename.

```{r, eval=freshRun}
saveRDS(banter_model_ici, 'hawaii_banter_model_ici_t5e3s5_t10e3s4.rds')
```

### BANTER Analytics

There are a number of visualizations/data products that allow us to
visualize our BANTER classifier; most use the rfPermute package (see
[BANTER
Guidelines](https://taikisan21.github.io/PAMpal/banterGuide.html) for
more information)

First, identify the model you would like to examine (comment out the
model you [do not]{.underline} want to examine).

```{r}
#model <- banter_model_ici
model <- banter_model
```

Extract the Random Forest model object from our BANTER model for
analysis.

```{r}
banter_model_RF <- getBanterModel(model)
```

Confusion Matrix

```{r, include=TRUE}
hawaii_confuseMatrix <- confusionMatrix(banter_model_RF)
```

Class Priors (Expected Error Rate)

```{r, include=TRUE}
hawaii_priors <- classPriors(banter_model_RF, NULL)
```

Plot Confusion Matrix Heat Map

```{r, include=TRUE}
plotConfMat(banter_model_RF, title="ConfusionMatrix", plot=TRUE)
```

Plot Error Trace for Banter Model

```{r, include=TRUE}
plot(banter_model_RF)
```

PlotVotes

```{r, include=TRUE}
png("hawaii_votes.png", width = 20, height = 20, units = 'cm',  res = 300)
plotVotes(banter_model_RF)
dev.off()
```

Importance Heatmap

```{r, include=TRUE}
png("hawaii_importance.png", width = 30, height = 25, units = 'cm',  res = 300)
plotImportance(banter_model_RF, plot.type="heatmap")
dev.off()
```

Proximity Plot

```{r, include=TRUE}
png("hawaii_proximity.png", width = 20, height = 20, units = 'cm',  res = 300)
hawaii_proximityPlot <- plotProximity(banter_model_RF)
dev.off()
```

Plot Predicted Probabilities

```{r, include=TRUE}
plotPredictedProbs(banter_model_RF, bins=30, plot=TRUE)
```

### 2022-12-14 manuscript/manuscript.qmd
Add Hawaii Dataset
Run Hawaii Dataset. Gitignore should include the binaries/database folders in the Hawaii folder dataset

@@
 format:
+  pdf: default
   elsevier-pdf:
@@
     journal:
-      name: Journal Name
+      name: Ecological Informatics
       formatting: preprint
-      model: 3p
+      model: 1p
       cite-style: authoryear
@@
 
-\[Insert Figure 2\]\[Figure 2.
+::: {layout="[[100,100], [100,100,100]]" layout-valign="bottom"}
+|             | Chinstrap | Gentoo | Adelie | **% Correct** |
+|-------------|-----------|--------|--------|---------------|
+| Chinstrap   | 70        | 0      | 0      | 100.0         |
+| Gentoo      | 0         | 5      | 0      | 100.0         |
+| Adelie      | 0         | 0      | 65     | 100.0         |
+| **Overall** | \-        | \-     | \-     | 100.0         |
+
+| LCI 0.95 | UCI 0.95 | Prior |
+|----------|----------|-------|
+| 94.9     | 100.0    | 50.0  |
+| 47.8     | 100.0    | 3.6   |
+| 94.5     | 100.0    | 46.4  |
+| 97.4     | 100.0    | 46.7  |
+
+::: columns
+a)  ![Fig.1](Figures/myplot1.jpg){#fig-1111}
+:::
+
+::: columns
+b)  ![Fig.2](Figures/myplot2.jpg){#fig-2222}
+:::
+
+::: columns
+c)  ![Fig.3](Figures/myplot3.jpg){#fig-3333}
+:::
+
+Figure 2.
 BANTER classification results from the SAtlantic dataset.
-Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).
-Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
-\]
+Confusion matrix provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).
+Proximity plot (a) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (b) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (c) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+:::
 
@@
 
-The most accurate model considered the echolocation clicks only (Detector Model: sampsize = 10, ntrees = 30,000; BANTER model: sampsize = 4, ntree = 200,000) and provided an overall correct classification rate of 89.7% for all four species.
-Classification scores ranged from a low of 82.1% (Cross Seamount Beaked Whale) to 96.7% (Longman's beaked whale, Fig.
+The most accurate model considered the echolocation clicks only (Detector Model: sampsize = 10, ntrees = 5,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 91.8% for all four species.
+Classification scores ranged from a low of 84.5% (Cross Seamount Beaked Whale) to 97.5% (Longman's beaked whale, Fig.
 3).
@@
 4).
+
+::: kable(hawaii/hawaii_confuseMatrix) :::
+
+::: columns
+a)  [Fig.1](hawaii/hawaii_proximity.png)
+:::
+
+::: columns
+b)  [Fig.2](hawaii/hawaii_importance.png)
+:::
+
+::: columns
+c)  [Fig.3](hawaii/hawaii_votes.png)
+:::
 
@@
 
-Baumann-Pickering, S., Roch, M.A., Brownell Jr, R.L., Simonis, A.E., McDonald, M.A., Solsona-Berga, A., Oleson, E.M., Wiggins, S.M. and Hildebrand, J.A., 2014. Spatio-temporal patterns of beaked whale
-echolocation signals in the North Pacific. PLoS One.  9(1), e86072. https://doi.org/10.1371/journal.pone.0086072
+Baumann-Pickering, S., Roch, M.A., Brownell Jr, R.L., Simonis, A.E., McDonald, M.A., Solsona-Berga, A., Oleson, E.M., Wiggins, S.M.
+and Hildebrand, J.A., 2014.
+Spatio-temporal patterns of beaked whale echolocation signals in the North Pacific. PLoS One.
+ 9(1), e86072.
+https://doi.org/10.1371/journal.pone.0086072
 
-Baumann-Pickering, S., Wiggins, S.M., Roth, E.H., Roch, M.A., Schnitzler, H.U. and Hildebrand, J.A., 2010. Echolocation signals of a beaked whale at Palmyra Atoll. J. Acoust. Soc. Am. 127(6),
-3790-3799. https://doi.org/10.1121/1.3409478
+Baumann-Pickering, S., Wiggins, S.M., Roth, E.H., Roch, M.A., Schnitzler, H.U.
+and Hildebrand, J.A., 2010.
+Echolocation signals of a beaked whale at Palmyra Atoll. J.
+Acoust.
+Soc.
+Am. 127(6), 3790-3799.
+https://doi.org/10.1121/1.3409478
 
@@
 
-Rankin, S., and Archer, F., 2021. BANTER: A User\'s Guide to Acoustic Classification. https://taikisan21.github.io/PAMpal/banterGuide.html
+Rankin, S., and Archer, F., 2021. BANTER: A User's Guide to Acoustic Classification. https://taikisan21.github.io/PAMpal/banterGuide.html
 
@@
 
-Rankin, S. and Barlow, J., 2005. Source of the North Pacific \"boing\" sound attributed to minke whales. J. Acoust. Soc. Am. 118(5), 3346-3351. https://doi.org/10.1121/1.2046747
+Rankin, S. and Barlow, J., 2005. Source of the North Pacific "boing" sound attributed to minke whales. J. Acoust. Soc. Am. 118(5), 3346-3351. https://doi.org/10.1121/1.2046747
 
@@
 and Roch, M.A., 2008.
-Classification of Risso\'s and Pacific white-sided dolphins using spectral properties of echolocation clicks.
+Classification of Risso's and Pacific white-sided dolphins using spectral properties of echolocation clicks.
 J.
@@
 Sci.
-Rep. 11(1), pp.1-16. https://doi.org/10.1038/s41598-021-01441-w
+Rep.
+11(1), pp.1-16.
+https://doi.org/10.1038/s41598-021-01441-w

### 2022-12-14 manuscript/manuscript.tex
Add Hawaii Dataset
Run Hawaii Dataset. Gitignore should include the binaries/database folders in the Hawaii folder dataset

@@
 \documentclass[
-  authoryear,
-  preprint,
-  3p]{elsarticle}
+  letterpaper,
+  DIV=11,
+  numbers=noendperiod]{scrartcl}
 
@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
@@
 \makeatother
-
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 \makeatother
-\journal{Journal Name}
 \ifLuaTeX
@@
 \fi
-\usepackage[]{natbib}
-\bibliographystyle{elsarticle-harv}
 \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
@@
 
-\setlength{\parindent}{6pt}
-\begin{document}
-
-\begin{frontmatter}
 \title{Open-Source machine learning BANTER acoustic classification of a
 cryptic echolocating species}
-\author[1]{Shannon Rankin%
-\corref{cor1}%
-\fnref{fn1}}
- \ead{shannon.rankin@noaa.gov} 
-\author[1]{Taiki Sakai%
-%
-\fnref{fn2}}
- \ead{taiki.sakai@noaa.gov} 
-\author[1]{Frederick Archer%
-%
-\fnref{fn3}}
- \ead{eric.archer@noaa.gov} 
-\author[]{Jay Barlow%
-%
-}
- \ead{jaybarlow33@yahoo.com} 
-\author[]{Danielle Cholewiak%
-%
-}
- \ead{danielle.cholewiak@noaa.gov} 
-\author[]{Annamaria DeAngelis%
-%
-}
- \ead{annamaria.deangelis@noaa.gov} 
-\author[]{Jennifer L. K. McCullough%
-%
-}
- \ead{jennifer.mccullough@noaa.gov} 
-\author[]{Erin Oleson%
-%
-}
- \ead{erin.oleson@noaa.gov} 
-\author[]{Melissa Soldevilla%
-%
-}
- \ead{melissa.soldevilla@noaa.gov}
-
-\affiliation[1]{organization={Southwest Fisheries Science Center, Marine
-Mammal \& Turtle Division},addressline={8901 La Jolla Shores
-Dr.},city={La Jolla},postcode={92037},postcodesep={}}
-
-\cortext[cor1]{Corresponding author}
-\fntext[fn1]{This is the first author footnote.}
-\fntext[fn2]{Another author footnote, this is a very long footnote and
-it should be a really long footnote. But this footnote is not yet
-sufficiently long enough to make two lines of footnote text.}
-\fntext[fn3]{Yet another author footnote.}
-
+\author{Shannon Rankin \and Taiki Sakai \and Frederick Archer \and Jay
+Barlow \and Danielle Cholewiak \and Annamaria DeAngelis \and Jennifer L.
+K. McCullough \and Erin Oleson \and Melissa Soldevilla}
+\date{12/12/22}
+
+\begin{document}
+\maketitle
 \begin{abstract}
@@
 \end{abstract}
-
-\begin{keyword}
-    bioacoustics \sep machine learning \sep random forest \sep species
-classification \sep marine mammals \sep passive acoustic
-monitoring \sep 
-    beaked whale
-\end{keyword}
-\end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 The most accurate model considered the echolocation clicks only
-(Detector Model: sampsize = 10, ntrees = 30,000; BANTER model: sampsize
-= 4, ntree = 200,000) and provided an overall correct classification
-rate of 89.7\% for all four species. Classification scores ranged from a
-low of 82.1\% (Cross Seamount Beaked Whale) to 96.7\% (Longman's beaked
+(Detector Model: sampsize = 10, ntrees = 5,000; BANTER model: sampsize =
+4, ntree = 10,000) and provided an overall correct classification rate
+of 91.8\% for all four species. Classification scores ranged from a low
+of 84.5\% (Cross Seamount Beaked Whale) to 97.5\% (Longman's beaked
 whale, Fig. 3). ~All classification results were greater than the
@@
 materials (Supplement Fig. 4).
+
+::: kable(hawaii/hawaii\_confuseMatrix) :::
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi})}
+\tightlist
+\item
+  \href{hawaii/hawaii_proximity.png}{Fig.1}
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi})}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  \href{hawaii/hawaii_importance.png}{Fig.2}
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi})}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  \href{hawaii/hawaii_votes.png}{Fig.3}
+\end{enumerate}
 
@@
 
-Here are two sample references: \citet{Feynman1963118}
-\citet{Dirac1953888}.
+Here are two sample references: Feynman and Vernon Jr. (1963) Dirac
+(1953).
 
@@
 
-  \bibliography{bibliography.bib}
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-Dirac1953888}{}}%
+Dirac, P. A. M. 1953. {``The {Lorentz} Transformation and Absolute
+Time.''} \emph{Physica} 19 (1-\/--12): 888--96.
+\url{https://doi.org/10.1016/S0031-8914(53)80099-6}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Feynman1963118}{}}%
+Feynman, R. P, and F. L Vernon Jr. 1963. {``The Theory of a General
+Quantum System Interacting with a Linear Dissipative System.''}
+\emph{Annals of Physics} 24: 118--73.
+\url{https://doi.org/10.1016/0003-4916(63)90068-X}.
+
+\end{CSLReferences}
 

### 2022-12-14 test.qmd
Add Hawaii Dataset
Run Hawaii Dataset. Gitignore should include the binaries/database folders in the Hawaii folder dataset

---
title: Open-Source machine learning BANTER acoustic classification of a cryptic echolocating species
author:
  - name: Shannon Rankin
    email: shannon.rankin@noaa.gov
    affiliations: 
        - id: SWFSC
          name: Southwest Fisheries Science Center
          department: Marine Mammal & Turtle Division
          address: 8901 La Jolla Shores Dr.
          city: La Jolla
          state: CA
          postal-code: 92037
    attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
  Passive acoustic monitoring drives me batty.
keywords: 
  - bioacoustics
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2022-12-14 hawaii/beakerBanter_hawaii.Rmd
Modify Hawaii RMD file
Modify to clean up

@@
 ---
-title: "beakerbanter_hawaii"
+title: "BANTER for Beaked Whales: Hawaii"
 author: "Shannon Rankin"
@@
 
-```{r setup, include=FALSE, cache=TRUE}
+```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-## Data Processing for Beaked Whales: Hawaii
+## PAMpal Data Processing
 
@@
 
-```{r, results=FALSE}
+```{r, load packages, message=FALSE}
 library(PAMpal)
@@
 
-### PAMpal Processing
+Set up our PPS (PAMPal Settings Object) for the Hawaii dataset
 
-1.  Set up our PPS (PAMPal Settings Object) for the Hawaii dataset
-
-```{r, eval=freshRun}
+```{r, pampal settings, eval=freshRun}
 pps <- PAMpalSettings(db='Database/', 
@@
 
-***If*** this is the initial processing, ensure you have set 'freshRun =
-TRUE' in code chunk at the top of this document so that you will run the
-processPGDetections code and save the data as an rds file. In this case,
-it will take some time to run, and we will check for warning messages
-during processing.
+[If]{.underline} you have already run the processing code, ensure you
+have set 'freshRun = FALSE' at top of this document to read in the
+existing .rds file for downstream processing.
 
-```{r, eval=freshRun}
-data <- processPgDetections(pps, mode='db', id='hawaii_bw')
-saveRDS(data, 'hawaii_study.rds')
+```{r, read acoustic study, include=FALSE, eval=!freshRun}
+data <- readRDS('hawaii_study.rds')
 # Double check warning messages
@@
 
-***If*** you have already run the processing code, ensure you have set
-'freshRun = FALSE' in code chunk at the top of this document so that you
-will read in the existing .rds file and check for warning messages.
+[If]{.underline} this is the [initial processing]{.underline}, ensure
+you have set 'freshRun = TRUE' at top of this document to process and
+save data. This will take some time to run.
 
-```{r, include=FALSE, eval=!freshRun}
-data <- readRDS('hawaii_study.rds')
+```{r, process detections, eval=freshRun}
+data <- processPgDetections(pps, mode='db', id='hawaii_bw')
+saveRDS(data, 'hawaii_study.rds')
 # Double check warning messages
@@
 
-```{r}
+```{r, assign species ID}
 data <- setSpecies(data, 'pamguard')
@@
 
-```{r}
+```{r, filter unwanted data}
 goodSpecies<- c("ZC", "MD", "BW", "BWC", "IP", "BW", "possBW")
@@
 data <- setSpecies(data, method='reassign',
-                   value=data.frame(old=c('ZC', 'MD', 'BWC', 'IP'), new=c("Cuviers", "Blainsvilles", "CrossSeamount", "Indopacetus")))
+                   value=data.frame(old=c('ZC', 'MD', 'BWC', 'IP'), new=c("Cuviers", "Blainsvilles", "CrossSeamount", "Longmans")))
 data_ch1only <- filter(data, Channel == '1')
@@
 
-```{r}
+```{r, calculate ICI}
 data_ch1only <- calculateICI(data_ch1only, time='peakTime')
@@
 
-```{r, eval=freshRun}
+```{r, banter export, eval=freshRun}
 banter_data <- export_banter(data_ch1only, dropSpecies = c('BW', 'possBW'), 
@@
 
-### Build a BANTER Classification Model
+## Build a BANTER Classification Model
 
-#### *Model without ICI*
+### *Model without ICI*
 
@@
 
-```{r, include=FALSE, eval=!freshRun}
+```{r, read banter model, include=FALSE, eval=!freshRun}
 banter_model <- readRDS('hawaii_banter_model_t5e3s10_t10e3s4.rds')
@@
 
-```{r, eval=freshRun}
+```{r, banter detector model, eval=freshRun}
 banter_model <- initBanterModel(banter_data$events)
@@
 
-```{r}
+```{r, banter detector model stability}
 plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 0:2))
@@
 
-```{r, eval=freshRun}
+```{r, banter event model, eval=freshRun}
 banter_model <- runBanterModel(banter_model, ntree=10e3, sampsize=4)
@@
 
-```{r, eval=freshRun}
+```{r,save banter model, eval=freshRun}
 saveRDS(banter_model, 'hawaii_banter_model_t5e3s10_t10e3s4.rds')
@@
 
-#### *ICI Model*
+### *ICI Model*
 
@@
 
-```{r, include=FALSE, eval=!freshRun}
+```{r, read banter ici model, include=FALSE, eval=!freshRun}
 banter_model_ici <- readRDS('hawaii_banter_model_ici_t5e3s5_t10e3s4.rds')
@@
 
-```{r, eval=freshRun}
+```{r, banter ici detector model, eval=freshRun}
 banter_model_ici <- initBanterModel(banter_data_ici$events)
-# was 30e3 / 1 in SRs initial, too much memory. 5e3/1 is 210mb and takes ~10min
 banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=5e3, sampsize=5, importance = TRUE)
@@
 
-```{r}
+```{r, banter ici detector model stability}
 plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 0:2))
@@
 
-```{r, eval=freshRun}
+```{r, banter ici event model, eval=freshRun}
 banter_model_ici <- runBanterModel(banter_model_ici, ntree=10e3, sampsize=4)
@@
 
-```{r, eval=freshRun}
+```{r, save banter ici model, eval=freshRun}
 saveRDS(banter_model_ici, 'hawaii_banter_model_ici_t5e3s5_t10e3s4.rds')
@@
 
-### BANTER Analytics
+## BANTER Analytics
 
@@
 
-```{r}
+```{r, identify model}
 #model <- banter_model_ici
@@
 
-```{r}
+```{r, extract RF model}
 banter_model_RF <- getBanterModel(model)
@@
 
-```{r, include=TRUE}
+```{r, confusion matrix, include=TRUE}
 hawaii_confuseMatrix <- confusionMatrix(banter_model_RF)
@@
 
-```{r, include=TRUE}
+```{r, class priors, include=TRUE}
 hawaii_priors <- classPriors(banter_model_RF, NULL)
@@
 
-```{r, include=TRUE}
+```{r, confusion matrix heat map, include=TRUE}
 plotConfMat(banter_model_RF, title="ConfusionMatrix", plot=TRUE)
@@
 
-```{r, include=TRUE}
+```{r, model error trace, include=TRUE}
 plot(banter_model_RF)
@@
 
-```{r, include=TRUE}
+```{r, plot votes, include=TRUE}
 png("hawaii_votes.png", width = 20, height = 20, units = 'cm',  res = 300)
@@
 dev.off()
+plotVotes(banter_model_RF)
 ```
@@
 
-```{r, include=TRUE}
+```{r, importance heat map, include=TRUE}
 png("hawaii_importance.png", width = 30, height = 25, units = 'cm',  res = 300)
@@
 dev.off()
+plotImportance(banter_model_RF, plot.type="heatmap")
 ```
@@
 
-```{r, include=TRUE}
+```{r, proximity plot, include=TRUE}
 png("hawaii_proximity.png", width = 20, height = 20, units = 'cm',  res = 300)
@@
 dev.off()
+hawaii_proximityPlot <- plotProximity(banter_model_RF)
 ```
@@
 
-```{r, include=TRUE}
+```{r, predicted probabilities, include=TRUE}
 plotPredictedProbs(banter_model_RF, bins=30, plot=TRUE)

### 2023-02-07 hawaii/beakerBanter_hawaii.Rmd
Update manuscript and hawaii data
Extensive work on Manuscript and preparation of Hawaii data (in folder). Hawaii data *should* be good to knit. Manuscript has references, needs updated results, figures, inline data, and updated discussion.

@@
 author: "Shannon Rankin"
-date: "2022-12-07"
+date: "2022-02-06"
 output: html_document
@@
 library(rfPermute)
+library(kableExtra)
 library(here)
@@
 
-Set up our PPS (PAMPal Settings Object) for the Hawaii dataset
+1.  Set up our PPS (PAMPal Settings Object) for the Hawaii dataset
 
@@
 
-4.  Filter out unwanted data: (1) subset and rename species for BANTER
+3.  Filter out unwanted data: (1) subset and rename species for BANTER
     model and (2) filter to retain only Channel 1.
@@
 ```{r, identify model}
-#model <- banter_model_ici
 model <- banter_model
+modelname <- "banter_model"
+
+model_ici <- banter_model_ici
+modelname_ici <- "banter_model_ici"
 ```
@@
 banter_model_RF <- getBanterModel(model)
+banter_model_ici_RF <- getBanterModel(model_ici)
 ```
@@
 ```{r, confusion matrix, include=TRUE}
-hawaii_confuseMatrix <- confusionMatrix(banter_model_RF)
+hawaii_confuseMatrix <- rfPermute::confusionMatrix(banter_model_RF)
+hawaii_confuseMatrix <- kable(hawaii_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(5, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(5,hline_after = TRUE)%>%
+  row_spec(5, bold = TRUE)%>%
+  save_kable(paste(modelname, "hawaii_confuseMatrix.png", sep="_"), zoom = 2)
+
+hawaii_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
+hawaii_ici_confuseMatrix <- kable(hawaii_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(5, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(5,hline_after = TRUE)%>%
+  row_spec(5, bold = TRUE)%>%
+  save_kable(paste(modelname_ici, "hawaii_ici_confuseMatrix.png", sep="_"), zoom = 2)
+
+```
+
+Proximity Plot
+
+```{r, proximity plot, include=TRUE}
+png(paste(modelname, "hawaii_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
+hawaii_proximityPlot <- plotProximity(banter_model_RF)
+dev.off()
+hawaii_proximityPlot <- plotProximity(banter_model_RF)
+
+png(paste(modelname_ici, "hawaii_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
+dev.off()
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
+```
+
+Importance Heatmap
+
+```{r, importance heat map, include=TRUE}
+png(paste(modelname, "hawaii_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
+plotImportance(banter_model_RF, plot.type="heatmap")
+dev.off()
+hawaii_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
+
+png(paste(modelname_ici, "hawaii_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
+plotImportance(banter_model_ici_RF, plot.type="heatmap")
+dev.off()
+ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+```
+
+PlotVotes
+
+```{r, plot votes, include=TRUE}
+png(paste(modelname, "hawaii_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_RF)
+dev.off()
+hawaii_votes <- plotVotes(banter_model_RF)
+
+png(paste(modelname_ici, "hawaii_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_ici_RF)
+dev.off()
+ici_hawaii_votes <- plotVotes(banter_model_ici_RF)
 ```
@@
 
-PlotVotes
-
-```{r, plot votes, include=TRUE}
-png("hawaii_votes.png", width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_RF)
-dev.off()
-plotVotes(banter_model_RF)
-```
-
-Importance Heatmap
-
-```{r, importance heat map, include=TRUE}
-png("hawaii_importance.png", width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_RF, plot.type="heatmap")
-dev.off()
-plotImportance(banter_model_RF, plot.type="heatmap")
-```
-
-Proximity Plot
-
-```{r, proximity plot, include=TRUE}
-png("hawaii_proximity.png", width = 20, height = 20, units = 'cm',  res = 300)
-hawaii_proximityPlot <- plotProximity(banter_model_RF)
-dev.off()
-hawaii_proximityPlot <- plotProximity(banter_model_RF)
-```
-
 Plot Predicted Probabilities

### 2023-02-07 manuscript/manuscript.qmd
Update manuscript and hawaii data
Extensive work on Manuscript and preparation of Hawaii data (in folder). Hawaii data *should* be good to knit. Manuscript has references, needs updated results, figures, inline data, and updated discussion.

@@
 ---
-
+```{r, hide=TRUE}
+library(tidyverse)
+library(knitr)
+library(here)
+```
 # Introduction
 
-Passive acoustic monitoring (PAM) from has proven to be a valuable tool for studying populations of marine mammals (Van Parijs et al., 2009); however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring.
-Historically, experienced acousticians manually identify stereotyped sounds that could be reliably attributed to a given species, based on their spectral or temporal characteristics (Baumann-Pickering et al., 2013; Bittle and Duncan, 2013; Rankin and Barlow, 2005; Soldevilla et al., 2008).
+Passive acoustic monitoring (PAM) from has proven to be a valuable tool for studying populations of marine mammals [@parijs2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring.
+Historically, experienced acousticians manually identify stereotyped sounds that could be reliably attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering2013]; [@bittle2013]; [@rankin2005]; [@soldevilla2008].
 The dramatic increase in recordings make it impossible for experienced acousticians to manually annotate all data, and there has been an increasing need for the development of automated classification routines that can provide accurate determinations of the species responsible for the sounds.
 
-Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods (MacLeod, 2018).
-However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species (Baumann-Pickering et al.
-2013).
-Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species (Barlow et al., 2021a; Baumann-Pickering et al., 2014; Simonis et al., 2020).
+Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods ([@macleod2018]).
+However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering2013a].
+Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow2022]; [@baumann-pickering2014]; [@simonis2020].
 Unfortunately, manual classification of beaked whale echolocation clicks requires analysis by trained observers using a number of visual aids to examine the call characteristics.
@@
 
-BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings (Rankin et al., 2017).
+BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings [@rankin2017].
 This hierarchical random forest event classifier relies on an initial call classifier (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier.
@@
 BANTER is very flexible and can accommodate any number of measures from any number of detectors.
-While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results (Rankin et al., 2017).
+While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results [@rankin2017a].
 In this case, settings for a whistle & moan detector were modified to improve performance at detecting burst pulses.
 This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the overall number of detections was different.
-This approach was successfully applied to classified echolocation clicks for narwhal and belugas (Zahn et al., 2021).
+This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn2021].
 For species who primarily (or exclusively) produce echolocation clicks, BANTER may serve as an option for automated machine learning classification.
@@
 AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey area.
-The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, cite) and recordings used in this analysis consisted of a single hydrophone (HTI-96-min, High Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see DeAngelis et al., 2018 for more information).
-The SAtlantic area ranged from Delaware south to central Florida (GU1605, cite) and recordings used in this analysis consisted of a single hydrophone (Reson etc) recorded 500 kHz sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass filter (see XX for more information).
+The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (HTI-96-min, High Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see [@deangelis2018b] for more information).
+The SAtlantic area ranged from Delaware south to central Florida (GU1605, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (Reson \[MODEL\], Teledyne Marine, Slangerup, Denmark) recorded 500 kHz sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass filter (see \[CITE\] for more information).
 
-Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, Yano et al., 2018).  Drifting recording buoys were deployed at randomly selected locations within the main Hawaiian Islands.
+Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, [@yano2018]).  Drifting recording buoys were deployed at randomly selected locations within the main Hawaiian Islands.
 Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA).
@@
 SM3M data were sampled at 256 kHz with continuous recordings.
-Individual buoys were deployed for 10 to 50 days (see Yano et al., 2018 for more detailed information).
+Individual buoys were deployed for 10 to 50 days (see [@yano2018]for more detailed information).
 
 Recordings from the Eastern North Pacific Ocean off the U.S.
-west coast (EPacific) were collected using drifting acoustic recorders during the 2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels) survey (Keating et al., 2018).
+west coast (EPacific) were collected using drifting acoustic recorders during the 2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels) survey [@keating2018].
 Buoys were deployed at a number of stations in the California Current, including pre-determined locations (for abundance estimation studies) and in close proximity to seamounts (Ad Hoc Seamount Experiment).
@@
 Soundtrap recorders were sampled at 288 kHz and the SM3M recorders were recorded at 256 sample rate (with some sampling at 96 kHz for ocean noise measurements).
-Individual buoys were deployed for 11-23 days (see Keating et al.
-2018 for more detailed information).
+Individual buoys were deployed for 11-23 days (see [@keating2018a] for more detailed information).
 
@@
 
-All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c) with a suite of generic click classifiers within the Click Detector module (see Keating and Barlow, 2013).  Classifier sets were saved such that any click may be classified as more than one click type.
+All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c) with a suite of generic click classifiers within the Click Detector module (see [@keating2013]).  Classifier sets were saved such that any click may be classified as more than one click type.
 The general click classifiers were effectively click detectors, as each implemented a simple spectral band settings (e.g., 2- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz).
-An additional detector/classifier within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see Keating and Barlow, 2013).
+An additional detector/classifier within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see [@keating2013a]).
 
@@
 Acoustic researchers with experience in classifying beaked whale echolocation clicks examined the bearing-time display for the click detector module in PAMGuard to assess the species identity of the acoustic event.
-For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see DeAngelis et al., 2018).
-Data were saved within the PAMGuard database and binaries for downstream processing using PAMpal package (Sakai, 2021) in R programming language (R Core Team 2014).
+For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see [@deangelis2018]).
+Data were saved within the PAMGuard database and binaries for downstream processing using PAMpal package [@sakai2021] in R programming language [@rcoreteam2022].
 
@@
 Inter-click interval (ICI) was calculated using the 'calculateICI' function in PAMpal; the mode of the ICI was identified for each beaked whale acoustic event.
-In addition, for EPacific data, used the 'browseEdinfo' and 'matchEnvData' function in PAMpal to identify appropriate ERDAPP environmental data (cite), including: sea surface temperature (from jplMURSST41), productivity (from   erdMH1pp8day dataset), chlorophyll (from erdMBchla8day dataset), and seafloor gradient (x,y, and gradient magnitude from rdSrtm30plusSeafloorGradient).
 
@@
 At a minimum, each trial included standard click calculations for all clicks in a beaked whale acoustic event (EC); and some trials included the mode of the ICI for each event (EC_ICI).
-For the EPACIFIC data, additional trials that also included water depth (EC_ICI_D) and environmental data (EC_ICI_ENV) were run for each event.
 
@@
 
-BANTER models was created for each trial using the BANTER package (cite) in R programming language (R Core Team 2014).
-Standard click calculations were used for the BANTER call classifier; event level measures of ICI and environmental features (for EPacific data) were applied to the BANTER event classifier.
+BANTER models was created for each trial using the BANTER package in R programming language [@archer2022].
+Standard click calculations were used for the BANTER call classifier; event level measures of ICI were applied to the BANTER event classifier.
 A minimum of two events per species and complete suite of variables (absence of NAs) were required for training and testing of the classification model.
@@
 The sample size (sampsize) is the number of events considered for each trial of the model, and the ntree is the number of trees (iterations of the model) run in a given model.
-These values were modified to obtain the best results, with an emphasis on lower sample size and higher number of trees for computational performance (Rankin and Archer, 2021).
+These values were modified to obtain the best results, with an emphasis on lower sample size and higher number of trees for computational performance [@rankin2021].
 
-The model output include the confusion matrix, and a number of visualizations provided by the rfpermute package (cite) in R programming language (R Core Team 2014).
+The model output include the confusion matrix, and a number of visualizations provided by the rfpermute package in R programming language [@archer2022a].
 
@@
 
-Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study; species included: Baird's beaked whales (*Berardius bairdii*), Blainsville's beaked whales (*Mesoplodon densirostris*), Cuvier's beaked whales (*Ziphius cavirostris*), Cross Seamount beaked whale (McDonald et al., 2009), Gervais' beaked whales (*Mesoplodon europaeus*), Longman's beaked whale (*Indopacetus pacificus*), Sowerby's beaked whales (*Mesoplodon bidens*), True's beaked whales (*Mesoplodon mirus*), Stejneger's beaked whale (*Mesoplodon stejnegeri*), and unidentified beaked whale 'BW43' (Baumann-Pickering et al., 2013).
+Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study; species included: Baird's beaked whales (*Berardius bairdii*), Blainsville's beaked whales (*Mesoplodon densirostris*), Cuvier's beaked whales (*Ziphius cavirostris*), Cross Seamount beaked whale (McDonald et al., 2009), Gervais' beaked whales (*Mesoplodon europaeus*), Longman's beaked whale (*Indopacetus pacificus*), Sowerby's beaked whales (*Mesoplodon bidens*), True's beaked whales (*Mesoplodon mirus*), Stejneger's beaked whale (*Mesoplodon stejnegeri*), and unidentified beaked whale 'BW43' [@baumann-pickering2013].
 Analyses were conducted separately for the different study areas due to differences in the data collection methods that precluded combining of datasets.
@@
 ## NAtlantic
-
-\[*Note: Last reprocess had both models with same classification results (though ICI was 'important', it did not really change the results).
-I need to reword a few things to accommodate this (and waiting on confirmation that re: ptoential changes to one function in PAMpal*)\]
 
@@
 
-::: {layout="[[100,100], [100,100,100]]" layout-valign="bottom"}
-|             | Chinstrap | Gentoo | Adelie | **% Correct** |
-|-------------|-----------|--------|--------|---------------|
-| Chinstrap   | 70        | 0      | 0      | 100.0         |
-| Gentoo      | 0         | 5      | 0      | 100.0         |
-| Adelie      | 0         | 0      | 65     | 100.0         |
-| **Overall** | \-        | \-     | \-     | 100.0         |
-
-| LCI 0.95 | UCI 0.95 | Prior |
-|----------|----------|-------|
-| 94.9     | 100.0    | 50.0  |
-| 47.8     | 100.0    | 3.6   |
-| 94.5     | 100.0    | 46.4  |
-| 97.4     | 100.0    | 46.7  |
-
-::: columns
-a)  ![Fig.1](Figures/myplot1.jpg){#fig-1111}
-:::
-
-::: columns
-b)  ![Fig.2](Figures/myplot2.jpg){#fig-2222}
-:::
-
-::: columns
-c)  ![Fig.3](Figures/myplot3.jpg){#fig-3333}
-:::
-
 Figure 2.
@@
 Vote Plot (c) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
-:::
 
@@
 ## Hawaii
-
 The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (Yano et al.
@@
 
-::: kable(hawaii/hawaii_confuseMatrix) :::
+::: {layout="[[100], [30,30,30]]"}
+a)  ![Fig.1](../hawaii/banter_model_hawaii_confuseMatrix.png){#fig-a}
 
-::: columns
-a)  [Fig.1](hawaii/hawaii_proximity.png)
-:::
+b)  ![Fig.1](../hawaii/banter_model_hawaii_proximity.png){#fig-b} :::
 
-::: columns
-b)  [Fig.2](hawaii/hawaii_importance.png)
-:::
+c)  ![Fig.1](../hawaii/banter_model_hawaii_importance.png)
 
-::: columns
-c)  [Fig.3](hawaii/hawaii_votes.png)
+d)  ![Fig.1](../hawaii/banter_model_hawaii_votes.png)
 :::
@@
 
-The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States (Keating et al.
-2018); species included Baird's beaked whale (n=91), Unidentified Beaked Whale 'BW43' (n=11), Stejneger's beaked whale (n=41) and Cuvier's beaked whale (n=1,332).
-BANTER classification trials included (1) echolocation clicks (EC), (2) echolocation clicks and ICI (EC_ICI), (3) echolocation clicks, ICI, and water depth (EC_ICI_D), and (4) echolocation clicks, ICI, water depth, sea surface salinity, productivity, and chlorophyll (EC_ICI_ENV).
+The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States ([@keating2018]); species included Baird's beaked whale (n=91), Unidentified Beaked Whale 'BW43' (n=11), Stejneger's beaked whale (n=41) and Cuvier's beaked whale (n=1,332).
+BANTER classification trials included (1) echolocation clicks (EC) and (2) echolocation clicks and ICI (EC_ICI).
 
@@
 
-Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (long-beaked and short-beaked common dolphins, Rankin et al., 2017).
+Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (long-beaked and short-beaked common dolphins, [@rankin2017b]).
 This suggests that large increases in sample size may improve classification results for Cuvier's beaked whales.
@@
 
-Gervais' and True's beaked whales have similar click characteristics that make them difficult to differentiate, and require considerable expertise to classify manually, using a variety of available visuals and metrics (DeAngelis et al., 2018).
+Gervais' and True's beaked whales have similar click characteristics that make them difficult to differentiate, and require considerable expertise to classify manually, using a variety of available visuals and metrics [@deangelis2018a].
 Our results suggest that BANTER may serve as an efficient and effective means of classifying and differentiating these two species.
@@
 Each event was defined as an individual click train, so ICI for these events more likely represented true ICI, and was found to be more informative.
-Baumann-Pickering et al.
-(2013) showed an overlap in the ICI between Cuvier's and Blainsville's beaked whales, but a strong difference in ICI between these species and BWC.
-Likewise, Baumann-Pickering (2013) found a difference in ICI between Baird's beaked whales and Cuvier's beaked whales.
+[@baumann-pickering2013] showed an overlap in the ICI between Cuvier's and Blainsville's beaked whales, but a strong difference in ICI between these species and BWC.
+Likewise, [@baumann-pickering2013] found a difference in ICI between Baird's beaked whales and Cuvier's beaked whales.
 Subdividing click trains in PAMGuard is time consuming, but alternative options include consideration of the PAMGuard click train detector module in BANTER, or by developing a similar function in R to apply to events.
@@
 This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified.
-For example, Barlow et al.
-(2021b) recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (including the EPacific data presented here).
+For example, (cite jays new paper) recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (including the EPacific data presented here).
 
@@
 # References {.unnumbered}
-
-Barlow, J., Cárdenas‐Hinojosa, G., Henderson, E.E., Breese, D., López‐Arzate, D., Hidalgo Pla, E.
-and Taylor, B.L., 2021.
-Unique morphological and acoustic characteristics of beaked whales (*Mesoplodon* sp.) off the west coast of Baja California, Mexico.
-Marine Mammal Science.
-https://doi.org/10.1111/mms.12853
-
-Barlow, J., Moore, J.E., McCullough, J.L.
-and Griffiths, E.T., 2021.
-Acoustic‐based estimates of Cuvier's beaked whale (*Ziphius cavirostris*) density and abundance along the US West Coast from drifting hydrophone recorders.
-Mar.
-Mamm.
-Sci.
-<https://doi.org/10.1111/mms.12872>
-
-Baumann-Pickering, S., McDonald, M.A., Simonis, A.E., Solsona Berga, A., Merkens, K.P., Oleson, E.M., Roch, M.A., Wiggins, S.M., Rankin, S., Yack, T.M.
-and Hildebrand, J.A., 2013.
-Species-specific beaked whale echolocation signals. J.
-Acoust.
-Soc.
-Am. 134(3), 2293-2301.
-https://doi.org/10.1121/1.4823843
-
-Baumann-Pickering, S., Roch, M.A., Brownell Jr, R.L., Simonis, A.E., McDonald, M.A., Solsona-Berga, A., Oleson, E.M., Wiggins, S.M.
-and Hildebrand, J.A., 2014.
-Spatio-temporal patterns of beaked whale echolocation signals in the North Pacific. PLoS One.
- 9(1), e86072.
-https://doi.org/10.1371/journal.pone.0086072
-
-Baumann-Pickering, S., Wiggins, S.M., Roth, E.H., Roch, M.A., Schnitzler, H.U.
-and Hildebrand, J.A., 2010.
-Echolocation signals of a beaked whale at Palmyra Atoll. J.
-Acoust.
-Soc.
-Am. 127(6), 3790-3799.
-https://doi.org/10.1121/1.3409478
-
-Bittle, M.
-and Duncan, A., 2013.
-A review of current marine mammal detection and classification algorithms for use in automated passive acoustic monitoring.
-In Proceedings of Acoustics: Science, Technology and Amenity.
-2013, 1-8.
-
-DeAngelis, A.I., Stanistreet, J.E., Baumann-Pickering, S.
-and Cholewiak, D.M., 2018.
-A description of echolocation clicks recorded in the presence of True's beaked whale (*Mesoplodon mirus*).
-J.
-Acoust.
-Soc.
-Am. 144(5), 2691-2700.
-https://doi.org/10.1121/1.5067379
-
-Keating JL, Barlow J., 2013. Click detectors and classifiers used during the 2012 southern California behavioral response study. NOAA Tech Memo NOAA-TM-NMFS-SWFSC-517:1-14.
-
-MacLeod, C.D., 2018. Beaked whales, overview, in: Encyclopedia of marine mammals. Academic Press, pp 80-83.
-
-McDonald, M.A., Hildebrand, J.A., Wiggins, S.M., Johnston, D.W.
-and Polovina, J.J., 2009.
-An acoustic survey of beaked whales at Cross Seamount near Hawaii.
-J.
-Acoust.
-Soc.
-Am.  125(2), 624-627.
-https://doi.org/10.1121/1.3050317
-
-R Core Team, 2014. R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL http://www.R-project.org/
-
-Rankin, S., and Archer, F., 2021. BANTER: A User's Guide to Acoustic Classification. https://taikisan21.github.io/PAMpal/banterGuide.html
-
-Rankin, S., Archer, F., Keating, J.L., Oswald, J.N., Oswald, M., Curtis, A.
-and Barlow, J., 2017.
-Acoustic classification of dolphins in the California Current using whistles, echolocation clicks, and burst pulses. Mar.
-Mamm.
-Sci.
-33(2), 520-540.
-https://doi.org/10.1111/mms.12381
-
-Rankin, S. and Barlow, J., 2005. Source of the North Pacific "boing" sound attributed to minke whales. J. Acoust. Soc. Am. 118(5), 3346-3351. https://doi.org/10.1121/1.2046747
-
-Soldevilla, M.S., Henderson, E.E., Campbell, G.S., Wiggins, S.M., Hildebrand, J.A.
-and Roch, M.A., 2008.
-Classification of Risso's and Pacific white-sided dolphins using spectral properties of echolocation clicks.
-J.
-Acoust.
-Soc.
-Am. 124(1), 609-624.
-https://doi.org/10.1121/1.2932059
-
-Sakai, T. 2021. PAMpal: Load and Process Passive Acoustic Data. R package version 0.13.0.
-
-Simonis, A.E., Brownell Jr, R.L., Thayre, B.J., Trickey, J.S., Oleson, E.M., Huntington, R.
-and Baumann-Pickering, S., 2020.
-Co-occurrence of beaked whale strandings and naval sonar in the Mariana Islands, Western Pacific.
-Proc.
-Roy.
-Soc.
-B.
-287(1921), p.20200070.
-https://doi.org/10.1098/rspb.2020.0070
-
-Van Parijs, S.M., Clark, C.W., Sousa-Lima, R.S., Parks, S.E., Rankin, S., Risch, D.
-and Van Opzeeland, I.C., 2009.
-Management and research applications of real-time and archival passive acoustic sensors over varying temporal and spatial scales.
-Mar.
-Ecol.
-Progr.
-Ser.
-395, pp.21-36.
-https://doi.org/10.3354/meps08123
-
-Zahn, M.J., Rankin, S., McCullough, J.L., Koblitz, J.C., Archer, F., Rasmussen, M.H.
-and Laidre, K.L., 2021.
-Acoustic differentiation and classification of wild belugas and narwhals using echolocation clicks.
-Sci.
-Rep.
-11(1), pp.1-16.
-https://doi.org/10.1038/s41598-021-01441-w

### 2023-02-07 manuscript/manuscript.tex
Update manuscript and hawaii data
Extensive work on Manuscript and preparation of Hawaii data (in folder). Hawaii data *should* be good to knit. Manuscript has references, needs updated results, figures, inline data, and updated discussion.

@@
 K. McCullough \and Erin Oleson \and Melissa Soldevilla}
-\date{12/12/22}
+\date{2/6/23}
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+-- Attaching packages --------------------------------------- tidyverse 1.3.2 --
+v ggplot2 3.4.0      v purrr   1.0.1 
+v tibble  3.1.8      v dplyr   1.0.10
+v tidyr   1.2.1      v stringr 1.5.0 
+v readr   2.1.3      v forcats 1.0.0 
+-- Conflicts ------------------------------------------ tidyverse_conflicts() --
+x dplyr::filter() masks stats::filter()
+x dplyr::lag()    masks stats::lag()
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(knitr)}
+\FunctionTok{library}\NormalTok{(here)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+here() starts at C:/Users/shannon.rankin/Documents/GitHub/BANTER_BeakedWhales
+\end{verbatim}
 
@@
 Passive acoustic monitoring (PAM) from has proven to be a valuable tool
-for studying populations of marine mammals (Van Parijs et al., 2009);
+for studying populations of marine mammals ((Parijs et al. 2009));
 however, the value of these studies depends on our ability to identify
@@
 attributed to a given species, based on their spectral or temporal
-characteristics (Baumann-Pickering et al., 2013; Bittle and Duncan,
-2013; Rankin and Barlow, 2005; Soldevilla et al., 2008). The dramatic
+characteristics ((Baumann-Pickering et al. 2013a); (Bittle and Duncan
+2013); (Rankin and Barlow 2005); (Soldevilla et al. 2008)). The dramatic
 increase in recordings make it impossible for experienced acousticians
@@
 difficult to study using typical shipboard visual observation methods
-(MacLeod, 2018). However, beaked whales make stereotyped echolocation
+((MacLeod 2018)). However, beaked whales make stereotyped echolocation
 clicks that have temporal and spatial characteristics that can be used
-to differentiate species (Baumann-Pickering et al.~2013). Detection of
-beaked whale signals in large datasets is greatly expanding our
+to differentiate species ((Baumann-Pickering et al. 2013b)). Detection
+of beaked whale signals in large datasets is greatly expanding our
 understanding of their population structure and the potential impact of
-human activities on these species (Barlow et al., 2021a;
-Baumann-Pickering et al., 2014; Simonis et al., 2020). Unfortunately,
+human activities on these species ((Barlow et al. 2022);
+(Baumann-Pickering et al. 2014); (Simonis et al. 2020)). Unfortunately,
 manual classification of beaked whale echolocation clicks requires
@@
 originally developed for classifying dolphin schools from acoustic
-recordings (Rankin et al., 2017). This hierarchical random forest event
-classifier relies on an initial call classifier (in the dolphin case
-study, this included one each for whistles, echolocation clicks, and
-burst pulses) followed by an event classifier. The event classifier
+recordings ((Rankin et al. 2017a)). This hierarchical random forest
+event classifier relies on an initial call classifier (in the dolphin
+case study, this included one each for whistles, echolocation clicks,
+and burst pulses) followed by an event classifier. The event classifier
 considers the probability of assignment to species for each call in the
@@
 call types, relatively minor changes to a call detector could yield
-different detector results (Rankin et al., 2017). In this case, settings
-for a whistle \& moan detector were modified to improve performance at
-detecting burst pulses. This approach of applying multiple call
-detectors (of the same type but with different settings) suggests that
-perhaps BANTER could be used on species that only produce one call type
-(or where only one call type was analyzed), as long as the settings of
-the automated call detector were modified such that the overall number
-of detections was different. This approach was successfully applied to
-classified echolocation clicks for narwhal and belugas (Zahn et al.,
-2021). For species who primarily (or exclusively) produce echolocation
-clicks, BANTER may serve as an option for automated machine learning
-classification.
+different detector results ((Rankin et al. 2017b)). In this case,
+settings for a whistle \& moan detector were modified to improve
+performance at detecting burst pulses. This approach of applying
+multiple call detectors (of the same type but with different settings)
+suggests that perhaps BANTER could be used on species that only produce
+one call type (or where only one call type was analyzed), as long as the
+settings of the automated call detector were modified such that the
+overall number of detections was different. This approach was
+successfully applied to classified echolocation clicks for narwhal and
+belugas ((Zahn et al. 2021)). For species who primarily (or exclusively)
+produce echolocation clicks, BANTER may serve as an option for automated
+machine learning classification.
 
@@
 southern (SAtlantic) survey area. The NAtlantic study area ranged from
-Massachusetts south to New Jersey (HB1603, cite) and recordings used in
-this analysis consisted of a single hydrophone (HTI-96-min, High Tech
-Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high
-pass filter (National Instruments USB-6356 A/D card, see DeAngelis et
-al., 2018 for more information). The SAtlantic area ranged from Delaware
-south to central Florida (GU1605, cite) and recordings used in this
-analysis consisted of a single hydrophone (Reson etc) recorded 500 kHz
+Massachusetts south to New Jersey (HB1603, {[}CITE{]}) and recordings
+used in this analysis consisted of a single hydrophone (HTI-96-min, High
+Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz
+high pass filter (National Instruments USB-6356 A/D card, see (DeAngelis
+et al. 2018a) for more information). The SAtlantic area ranged from
+Delaware south to central Florida (GU1605, {[}CITE{]}) and recordings
+used in this analysis consisted of a single hydrophone (Reson
+{[}MODEL{]}, Teledyne Marine, Slangerup, Denmark) recorded 500 kHz
 sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass
-filter (see XX for more information).
+filter (see {[}CITE{]} for more information).
 
@@
 during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment
-Survey, Yano et al., 2018).~ Drifting recording buoys were deployed at
+Survey, (Yano et al. 2018)).~ Drifting recording buoys were deployed at
 randomly selected locations within the main Hawaiian Islands. Two types
@@
 SM3M data were sampled at 256 kHz with continuous recordings. Individual
-buoys were deployed for 10 to 50 days (see Yano et al., 2018 for more
+buoys were deployed for 10 to 50 days (see (Yano et al. 2018)for more
 detailed information).
@@
 2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels)
-survey (Keating et al., 2018). Buoys were deployed at a number of
+survey ((J. Keating et al. 2018a)). Buoys were deployed at a number of
 stations in the California Current, including pre-determined locations
@@
 some sampling at 96 kHz for ocean noise measurements). Individual buoys
-were deployed for 11-23 days (see Keating et al.~2018 for more detailed
-information).
+were deployed for 11-23 days (see (J. Keating et al. 2018b) for more
+detailed information).
 
@@
 2.00.15c) with a suite of generic click classifiers within the Click
-Detector module (see Keating and Barlow, 2013).~ Classifier sets were
-saved such that any click may be classified as more than one click type.
-The general click classifiers were effectively click detectors, as each
-implemented a simple spectral band settings (e.g., 2- 15 kHz, 15 -- 30
-kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \textgreater{} 80 kHz). An
+Detector module (see (J. L. Keating and Barlow 2013a)).~ Classifier sets
+were saved such that any click may be classified as more than one click
+type. The general click classifiers were effectively click detectors, as
+each implemented a simple spectral band settings (e.g., 2- 15 kHz, 15 --
+30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \textgreater{} 80 kHz). An
 additional detector/classifier within the 30 -- 50 kHz peak frequency
 range considered the presence of a frequency sweep that is
-characteristic of beaked whale pulses (see Keating and Barlow, 2013).
+characteristic of beaked whale pulses (see (J. L. Keating and Barlow
+2013b)).
 
@@
 of consecutive clicks along the same bearing angle in the bearing time
-plot (see DeAngelis et al., 2018). Data were saved within the PAMGuard
+plot (see (DeAngelis et al. 2018b)). Data were saved within the PAMGuard
 database and binaries for downstream processing using PAMpal package
-(Sakai, 2021) in R programming language (R Core Team 2014).
+((Sakai 2021)) in R programming language ((R Core Team 2022)).
 
 Standard click calculations were measured from echolocation clicks using
-default values in PAMpal. ~For EPacific data, any events \textgreater{}
-2 minutes were subdivided into 2-minute events to ensure that all events
+default values in PAMpal. For EPacific data, any events \textgreater{} 2
+minutes were subdivided into 2-minute events to ensure that all events
 were of equal value (because most of our data was duty cycled into
@@
 `calculateICI' function in PAMpal; the mode of the ICI was identified
-for each beaked whale acoustic event. In addition, for EPacific data,
-used the `browseEdinfo' and `matchEnvData' function in PAMpal to
-identify appropriate ERDAPP environmental data (cite), including: sea
-surface temperature (from jplMURSST41), productivity (from~
-~erdMH1pp8day dataset), chlorophyll (from erdMBchla8day dataset), and
-seafloor gradient (x,y, and gradient magnitude from
-rdSrtm30plusSeafloorGradient).
+for each beaked whale acoustic event.
 
@@
 event (EC); and some trials included the mode of the ICI for each event
-(EC\_ICI). For the EPACIFIC data, additional trials that also included
-water depth (EC\_ICI\_D) and environmental data (EC\_ICI\_ENV) were run
-for each event.
+(EC\_ICI).
 
@@
 
-BANTER models was created for each trial using the BANTER package (cite)
-in R programming language (R Core Team 2014). Standard click
-calculations were used for the BANTER call classifier; event level
-measures of ICI and environmental features (for EPacific data) were
+BANTER models was created for each trial using the BANTER package in R
+programming language ((Archer 2022a)). Standard click calculations were
+used for the BANTER call classifier; event level measures of ICI were
 applied to the BANTER event classifier. A minimum of two events per
@@
 emphasis on lower sample size and higher number of trees for
-computational performance (Rankin and Archer, 2021).
+computational performance ((Rankin and Archer 2021)).
 
 The model output include the confusion matrix, and a number of
-visualizations provided by the rfpermute package (cite) in R programming
-language (R Core Team 2014).
+visualizations provided by the rfpermute package in R programming
+language ((Archer 2022b)).
 
@@
 (\emph{Mesoplodon stejnegeri}), and unidentified beaked whale `BW43'
-(Baumann-Pickering et al., 2013). Analyses were conducted separately for
-the different study areas due to differences in the data collection
+((Baumann-Pickering et al. 2013a)). Analyses were conducted separately
+for the different study areas due to differences in the data collection
 methods that precluded combining of datasets. Specific species
@@
 \subsection{NAtlantic}\label{natlantic}}
-
-{[}\emph{Note: Last reprocess had both models with same classification
-results (though ICI was `important', it did not really change the
-results). I need to reword a few things to accommodate this (and waiting
-on confirmation that re: ptoential changes to one function in
-PAMpal}){]}
 
@@
 
-\begin{figure}
-
-\begin{minipage}[b]{0.50\linewidth}
-
-{\centering
-
-\begin{longtable}[]{@{}lllll@{}}
-\toprule()
-& Chinstrap & Gentoo & Adelie & \textbf{\% Correct} \\
-\midrule()
-\endhead
-Chinstrap & 70 & 0 & 0 & 100.0 \\
-Gentoo & 0 & 5 & 0 & 100.0 \\
-Adelie & 0 & 0 & 65 & 100.0 \\
-\textbf{Overall} & - & - & - & 100.0 \\
-\bottomrule()
-\end{longtable}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[b]{0.50\linewidth}
-
-{\centering
-
-\begin{longtable}[]{@{}lll@{}}
-\toprule()
-LCI 0.95 & UCI 0.95 & Prior \\
-\midrule()
-\endhead
-94.9 & 100.0 & 50.0 \\
-47.8 & 100.0 & 3.6 \\
-94.5 & 100.0 & 46.4 \\
-97.4 & 100.0 & 46.7 \\
-\bottomrule()
-\end{longtable}
-
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[b]{0.33\linewidth}
-
-{\centering
-
-\begin{enumerate}
-\def\labelenumi{\alph{enumi})}
-\tightlist
-\item
-  \includegraphics{Figures/myplot1.jpg}
-\end{enumerate}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[b]{0.33\linewidth}
-
-{\centering
-
-\begin{enumerate}
-\def\labelenumi{\alph{enumi})}
-\setcounter{enumi}{1}
-\tightlist
-\item
-  \includegraphics{Figures/myplot2.jpg}
-\end{enumerate}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[b]{0.33\linewidth}
-
-{\centering
-
-\begin{enumerate}
-\def\labelenumi{\alph{enumi})}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  \includegraphics{Figures/myplot3.jpg}
-\end{enumerate}
-
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[b]{0.33\linewidth}
-
-{\centering
-
 Figure 2. BANTER classification results from the SAtlantic dataset.
@@
 
-}
-
-\end{minipage}%
-
-\end{figure}
-
 For SAtlantic, the proximity plot (Fig. 2b) shows that True's beaked
@@
 
-::: kable(hawaii/hawaii\_confuseMatrix) :::
+\begin{figure}
+
+\begin{minipage}[t]{\linewidth}
+
+{\centering
 
@@
 \def\labelenumi{\alph{enumi})}
-\tightlist
 \item
-  \href{hawaii/hawaii_proximity.png}{Fig.1}
+  \begin{figure}
+
+  {\centering \includegraphics{../hawaii/banter_model_hawaii_confuseMatrix.png}
+
+  }
+
+  \caption{\label{fig-a}Fig.1}
+
+  \end{figure}
+\item
+  \includegraphics{../hawaii/banter_model_hawaii_proximity.png} :::
+\item
+  \begin{figure}
+
+  {\centering \includegraphics{../hawaii/banter_model_hawaii_importance.png}
+
+  }
+
+  \caption{Fig.1}
+
+  \end{figure}
+\item
+  \begin{figure}
+
+  {\centering \includegraphics{../hawaii/banter_model_hawaii_votes.png}
+
+  }
+
+  \caption{Fig.1}
+
+  \end{figure}
 \end{enumerate}
 
-\begin{enumerate}
-\def\labelenumi{\alph{enumi})}
-\setcounter{enumi}{1}
-\tightlist
-\item
-  \href{hawaii/hawaii_importance.png}{Fig.2}
-\end{enumerate}
-
-\begin{enumerate}
-\def\labelenumi{\alph{enumi})}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  \href{hawaii/hawaii_votes.png}{Fig.3}
-\end{enumerate}
+}
+
+\end{minipage}%
+
+\end{figure}
 
@@
 stations plus 5 seamount experiment drifts) deployed off the west coast
-of the United States (Keating et al.~2018); species included Baird's
-beaked whale (n=91), Unidentified Beaked Whale `BW43' (n=11),
+of the United States ((J. Keating et al. 2018a)); species included
+Baird's beaked whale (n=91), Unidentified Beaked Whale `BW43' (n=11),
 Stejneger's beaked whale (n=41) and Cuvier's beaked whale (n=1,332).
-BANTER classification trials included (1) echolocation clicks (EC), (2)
-echolocation clicks and ICI (EC\_ICI), (3) echolocation clicks, ICI, and
-water depth (EC\_ICI\_D), and (4) echolocation clicks, ICI, water depth,
-sea surface salinity, productivity, and chlorophyll (EC\_ICI\_ENV).
+BANTER classification trials included (1) echolocation clicks (EC) and
+(2) echolocation clicks and ICI (EC\_ICI).
 
@@
 differentiate species (long-beaked and short-beaked common dolphins,
-Rankin et al., 2017). This suggests that large increases in sample size
-may improve classification results for Cuvier's beaked whales.
+(Rankin et al. 2017c)). This suggests that large increases in sample
+size may improve classification results for Cuvier's beaked whales.
 
@@
 expertise to classify manually, using a variety of available visuals and
-metrics (DeAngelis et al., 2018). Our results suggest that BANTER may
+metrics ((DeAngelis et al. 2018c)). Our results suggest that BANTER may
 serve as an efficient and effective means of classifying and
@@
 train, so ICI for these events more likely represented true ICI, and was
-found to be more informative. Baumann-Pickering et al.~(2013) showed an
+found to be more informative. (Baumann-Pickering et al. 2013a) showed an
 overlap in the ICI between Cuvier's and Blainsville's beaked whales, but
 a strong difference in ICI between these species and BWC. Likewise,
-Baumann-Pickering (2013) found a difference in ICI between Baird's
-beaked whales and Cuvier's beaked whales. Subdividing click trains in
-PAMGuard is time consuming, but alternative options include
+(Baumann-Pickering et al. 2013a) found a difference in ICI between
+Baird's beaked whales and Cuvier's beaked whales. Subdividing click
+trains in PAMGuard is time consuming, but alternative options include
 consideration of the PAMGuard click train detector module in BANTER, or
@@
 further complicated by the potential existence of species that have not
-yet been identified. For example, Barlow et al.~(2021b) recently
+yet been identified. For example, (cite jays new paper) recently
 discovered what appears to be a new species of beaked whale off Baja
@@
 
-Here are two sample references: Feynman and Vernon Jr. (1963) Dirac
-(1953).
+Here are two sample references: (\textbf{Feynman1963118?})
+(\textbf{Dirac1953888?}).
 
@@
 
-Barlow, J., Cárdenas‐Hinojosa, G., Henderson, E.E., Breese, D.,
-López‐Arzate, D., Hidalgo Pla, E. and Taylor, B.L., 2021. Unique
-morphological and acoustic characteristics of beaked whales
-(\emph{Mesoplodon} sp.) off the west coast of Baja California, Mexico.
-Marine Mammal Science. https://doi.org/10.1111/mms.12853
-
-Barlow, J., Moore, J.E., McCullough, J.L. and Griffiths, E.T., 2021.
-Acoustic‐based estimates of Cuvier's beaked whale (\emph{Ziphius
-cavirostris}) density and abundance along the US West Coast from
-drifting hydrophone recorders. Mar.~Mamm. Sci.
-\url{https://doi.org/10.1111/mms.12872}
-
-Baumann-Pickering, S., McDonald, M.A., Simonis, A.E., Solsona Berga, A.,
-Merkens, K.P., Oleson, E.M., Roch, M.A., Wiggins, S.M., Rankin, S.,
-Yack, T.M. and Hildebrand, J.A., 2013. Species-specific beaked whale
-echolocation signals.~J. Acoust. Soc. Am.~134(3), 2293-2301.
-https://doi.org/10.1121/1.4823843
-
-Baumann-Pickering, S., Roch, M.A., Brownell Jr, R.L., Simonis, A.E.,
-McDonald, M.A., Solsona-Berga, A., Oleson, E.M., Wiggins, S.M. and
-Hildebrand, J.A., 2014. Spatio-temporal patterns of beaked whale
-echolocation signals in the North Pacific.~PLoS One. ~9(1), e86072.
-https://doi.org/10.1371/journal.pone.0086072
-
-Baumann-Pickering, S., Wiggins, S.M., Roth, E.H., Roch, M.A.,
-Schnitzler, H.U. and Hildebrand, J.A., 2010. Echolocation signals of a
-beaked whale at Palmyra Atoll.~J. Acoust. Soc. Am.~127(6), 3790-3799.
-https://doi.org/10.1121/1.3409478
-
-Bittle, M. and Duncan, A., 2013. A review of current marine mammal
-detection and classification algorithms for use in automated passive
-acoustic monitoring. In Proceedings of Acoustics: Science, Technology
-and Amenity. 2013, 1-8.
-
-DeAngelis, A.I., Stanistreet, J.E., Baumann-Pickering, S. and Cholewiak,
-D.M., 2018. A description of echolocation clicks recorded in the
-presence of True's beaked whale (\emph{Mesoplodon mirus}). J. Acoust.
-Soc. Am.~144(5), 2691-2700. https://doi.org/10.1121/1.5067379
-
-Keating JL, Barlow J., 2013. Click detectors and classifiers used during
-the 2012 southern California behavioral response study. NOAA Tech Memo
-NOAA-TM-NMFS-SWFSC-517:1-14.
-
-MacLeod, C.D., 2018. Beaked whales, overview, in: Encyclopedia of marine
-mammals. Academic Press, pp 80-83.
-
-McDonald, M.A., Hildebrand, J.A., Wiggins, S.M., Johnston, D.W. and
-Polovina, J.J., 2009. An acoustic survey of beaked whales at Cross
-Seamount near Hawaii. J. Acoust. Soc. Am.~ 125(2), 624-627.
-https://doi.org/10.1121/1.3050317
-
-R Core Team, 2014. R: A language and environment for statistical
-computing. R Foundation for Statistical Computing, Vienna, Austria. URL
-http://www.R-project.org/
-
-Rankin, S., and Archer, F., 2021. BANTER: A User's Guide to Acoustic
-Classification. https://taikisan21.github.io/PAMpal/banterGuide.html
-
-Rankin, S., Archer, F., Keating, J.L., Oswald, J.N., Oswald, M., Curtis,
-A. and Barlow, J., 2017. Acoustic classification of dolphins in the
-California Current using whistles, echolocation clicks, and burst
-pulses.~Mar.~Mamm. Sci. 33(2), 520-540.
-https://doi.org/10.1111/mms.12381
-
-Rankin, S. and Barlow, J., 2005. Source of the North Pacific ``boing''
-sound attributed to minke whales. J. Acoust. Soc. Am.~118(5), 3346-3351.
-https://doi.org/10.1121/1.2046747
-
-Soldevilla, M.S., Henderson, E.E., Campbell, G.S., Wiggins, S.M.,
-Hildebrand, J.A. and Roch, M.A., 2008. Classification of Risso's and
-Pacific white-sided dolphins using spectral properties of echolocation
-clicks. J. Acoust. Soc. Am.~124(1), 609-624.
-https://doi.org/10.1121/1.2932059
-
-Sakai, T. 2021. PAMpal: Load and Process Passive Acoustic Data. R
-package version 0.13.0.
-
-Simonis, A.E., Brownell Jr, R.L., Thayre, B.J., Trickey, J.S., Oleson,
-E.M., Huntington, R. and Baumann-Pickering, S., 2020. Co-occurrence of
-beaked whale strandings and naval sonar in the Mariana Islands, Western
-Pacific. Proc. Roy. Soc. B. 287(1921), p.20200070.
-https://doi.org/10.1098/rspb.2020.0070
-
-Van Parijs, S.M., Clark, C.W., Sousa-Lima, R.S., Parks, S.E., Rankin,
-S., Risch, D. and Van Opzeeland, I.C., 2009. Management and research
-applications of real-time and archival passive acoustic sensors over
-varying temporal and spatial scales. Mar.~Ecol. Progr. Ser. 395,
-pp.21-36. https://doi.org/10.3354/meps08123
-
-Zahn, M.J., Rankin, S., McCullough, J.L., Koblitz, J.C., Archer, F.,
-Rasmussen, M.H. and Laidre, K.L., 2021. Acoustic differentiation and
-classification of wild belugas and narwhals using echolocation clicks.
-Sci. Rep.~11(1), pp.1-16. https://doi.org/10.1038/s41598-021-01441-w
-
 \hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-Dirac1953888}{}}%
-Dirac, P. A. M. 1953. {``The {Lorentz} Transformation and Absolute
-Time.''} \emph{Physica} 19 (1-\/--12): 888--96.
-\url{https://doi.org/10.1016/S0031-8914(53)80099-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Feynman1963118}{}}%
-Feynman, R. P, and F. L Vernon Jr. 1963. {``The Theory of a General
-Quantum System Interacting with a Linear Dissipative System.''}
-\emph{Annals of Physics} 24: 118--73.
-\url{https://doi.org/10.1016/0003-4916(63)90068-X}.
+\leavevmode\vadjust pre{\hypertarget{ref-archer2022}{}}%
+Archer, Frederick. 2022a. \emph{Banter: BioAcoustic eveNT classifiER}.
+\url{https://cran.r-project.org/web/packages/banter/index.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-archer2022a}{}}%
+---------. 2022b. \emph{rfPermute: Estimate Permutation p-Values for
+Random Forest Importance Metrics}.
+\url{https://cran.r-project.org/web/packages/rfPermute/index.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-barlow2022}{}}%
+Barlow, Jay, Jeffrey E. Moore, Jennifer L. K. McCullough, and Emily T.
+Griffiths. 2022. {``Acoustic-Based Estimates of Cuvier's Beaked Whale
+(Ziphius Cavirostris) Density and Abundance Along the U.S. West Coast
+from Drifting Hydrophone Recorders.''} \emph{Marine Mammal Science} 38
+(2): 517--38. \url{https://doi.org/10.1111/mms.12872}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-baumann-pickering2013}{}}%
+Baumann-Pickering, Simone, Mark A. McDonald, Anne E. Simonis, Alba
+Solsona Berga, Karlina P. B. Merkens, Erin M. Oleson, Marie A. Roch, et
+al. 2013a. {``Species-Specific Beaked Whale Echolocation Signals.''}
+\emph{The Journal of the Acoustical Society of America} 134 (3):
+2293--2301. \url{https://doi.org/10.1121/1.4817832}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-baumann-pickering2013a}{}}%
+---------, et al. 2013b. {``Species-Specific Beaked Whale Echolocation
+Signals.''} \emph{The Journal of the Acoustical Society of America} 134
+(3): 2293--2301. \url{https://doi.org/10.1121/1.4817832}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-baumann-pickering2014}{}}%
+Baumann-Pickering, Simone, Marie A. Roch, Robert L. Brownell Jr, Anne E.
+Simonis, Mark A. McDonald, Alba Solsona-Berga, Erin M. Oleson, Sean M.
+Wiggins, and John A. Hildebrand. 2014. {``Spatio-Temporal Patterns of
+Beaked Whale Echolocation Signals in the North Pacific.''} \emph{PLOS
+ONE} 9 (1): e86072. \url{https://doi.org/10.1371/journal.pone.0086072}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bittle2013}{}}%
+Bittle, Michael, and Alec Duncan. 2013. {``A Review of Current Marine
+Mammal Detection and Classification Algorithms for Use in Automated
+Passive Acoustic Monitoring.''} 8.
+
+\leavevmode\vadjust pre{\hypertarget{ref-deangelis2018}{}}%
+DeAngelis, Annamaria Izzi, Joy E. Stanistreet, Simone Baumann-Pickering,
+and Danielle M. Cholewiak. 2018b. {``A Description of Echolocation
+Clicks Recorded in the Presence of True's Beaked Whale (Mesoplodon
+Mirus).''} \emph{The Journal of the Acoustical Society of America} 144
+(5): 2691--2700. \url{https://doi.org/10.1121/1.5067379}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-deangelis2018a}{}}%
+---------. 2018c. {``A Description of Echolocation Clicks Recorded in
+the Presence of True's Beaked Whale (Mesoplodon Mirus).''} \emph{The
+Journal of the Acoustical Society of America} 144 (5): 2691--2700.
+\url{https://doi.org/10.1121/1.5067379}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-deangelis2018b}{}}%
+---------. 2018a. {``A Description of Echolocation Clicks Recorded in
+the Presence of True's Beaked Whale (Mesoplodon Mirus).''} \emph{The
+Journal of the Acoustical Society of America} 144 (5): 2691--2700.
+\url{https://doi.org/10.1121/1.5067379}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-keating2013}{}}%
+Keating, Jennifer L., and Jay Barlow. 2013a. {``Summary of PAMGuard
+Beaked Whale Click Detectors and Classifiers Used During the 2012
+Southern California Behavioral Response Study.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-keating2013a}{}}%
+---------. 2013b. {``Summary of PAMGuard Beaked Whale Click Detectors
+and Classifiers Used During the 2012 Southern California Behavioral
+Response Study.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-keating2018}{}}%
+Keating, Jennifer, Jay Barlow, Emily T. Griffiths, and Jeffrey E. Moore.
+2018a. {``Passive Acoustics Survey of Cetacean Abundance Levels
+(PASCAL-2016) Final Report.''}
+\url{https://www.boem.gov/sites/default/files/environmental-stewardship/Environmental-Studies/Pacific-Region/Studies/BOEM-2018-025.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-keating2018a}{}}%
+---------. 2018b. {``Passive Acoustics Survey of Cetacean Abundance
+Levels (PASCAL-2016) Final Report.''}
+\url{https://www.boem.gov/sites/default/files/environmental-stewardship/Environmental-Studies/Pacific-Region/Studies/BOEM-2018-025.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-macleod2018}{}}%
+MacLeod, Colin D. 2018. {``Beaked Whales, Overview.''} In, edited by
+Bernd Würsig, J. G. M. Thewissen, and Kit M. Kovacs, 80--83. Academic
+Press. \url{https://doi.org/10.1016/B978-0-12-804327-1.00062-5}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-parijs2009}{}}%
+Parijs, Sofie M. Van, Chris W. Clark, Renata S. Sousa-Lima, Susan E.
+Parks, Shannon Rankin, Denise Risch, and Ilse C. Van Opzeeland. 2009.
+{``Management and Research Applications of Real-Time and Archival
+Passive Acoustic Sensors over Varying Temporal and Spatial Scales.''}
+\emph{Marine Ecology Progress Series} 395 (December): 21--36.
+\url{https://doi.org/10.3354/meps08123}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2022}{}}%
+R Core Team. 2022. \emph{R: A Language and Environment for Statistical
+Computing}. Vienna, Austria: R Foundation for Statistical Computing.
+\url{http://www.R-project.org/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rankin2021}{}}%
+Rankin, Shannon, and Frederick Archer. 2021. {``BANTER: A User's Guide
+to Acoustic Classification.''}
+\url{https://taikisan21.github.io/PAMpal/banterGuide.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rankin2017}{}}%
+Rankin, Shannon, Frederick Archer, Jennifer L. Keating, Julie N. Oswald,
+Michael Oswald, Alex Curtis, and Jay Barlow. 2017a. {``Acoustic
+Classification of Dolphins in the California Current Using Whistles,
+Echolocation Clicks, and Burst Pulses.''} \emph{Marine Mammal Science} 2
+(33): 520--40. \url{https://doi.org/10.1111/mms.12381}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rankin2017a}{}}%
+---------. 2017b. {``Acoustic Classification of Dolphins in the
+California Current Using Whistles, Echolocation Clicks, and Burst
+Pulses.''} \emph{Marine Mammal Science} 2 (33): 520--40.
+\url{https://doi.org/10.1111/mms.12381}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rankin2017b}{}}%
+---------. 2017c. {``Acoustic Classification of Dolphins in the
+California Current Using Whistles, Echolocation Clicks, and Burst
+Pulses.''} \emph{Marine Mammal Science} 2 (33): 520--40.
+\url{https://doi.org/10.1111/mms.12381}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rankin2005}{}}%
+Rankin, Shannon, and Jay Barlow. 2005. {``Source of the North Pacific
+{``}Boing{''} Sound Attributed to Minke Whales.''} \emph{The Journal of
+the Acoustical Society of America} 118 (5): 3346--51.
+\url{https://doi.org/10.1121/1.2046747}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sakai2021}{}}%
+Sakai, Taiki. 2021. \emph{PAMpal: Load and Process Passive Acoustic
+Data}. \url{https://cran.r-project.org/web/packages/PAMpal/index.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-simonis2020}{}}%
+Simonis, Anne E., Robert L. Brownell, Bruce J. Thayre, Jennifer S.
+Trickey, Erin M. Oleson, Roderick Huntington, and Simone
+Baumann-Pickering. 2020. {``Co-Occurrence of Beaked Whale Strandings and
+Naval Sonar in the Mariana Islands, Western Pacific.''}
+\emph{Proceedings of the Royal Society B: Biological Sciences} 287
+(1921): 20200070. \url{https://doi.org/10.1098/rspb.2020.0070}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-soldevilla2008}{}}%
+Soldevilla, Melissa S., E. Elizabeth Henderson, Gregory S. Campbell,
+Sean M. Wiggins, John A. Hildebrand, and Marie A. Roch. 2008.
+{``Classification of Risso{'}s and Pacific White-Sided Dolphins Using
+Spectral Properties of Echolocation Clicks.''} \emph{The Journal of the
+Acoustical Society of America} 124 (1): 609--24.
+\url{https://doi.org/10.1121/1.2932059}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-yano2018}{}}%
+Yano, Kymberly M., Erin M. Oleson, Jennifer L. Keating, Lisa Ballance,
+Marie C. Hill, Amanda Bradford, Ann N. Allen, Trevor W. Joyce, Jeffrey
+E. Moore, and Annette E. Henry. 2018. {``Cetacean and Seabird Data
+Collected During the Hawaiian Islands Cetacean and Ecosystem Assessment
+Survey (HICEAS), July{\textendash}december 2017.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-zahn2021}{}}%
+Zahn, Marie J., Shannon Rankin, Jennifer L. K. McCullough, Jens C.
+Koblitz, Frederick Archer, Marianne H. Rasmussen, and Kristin L. Laidre.
+2021. {``Acoustic Differentiation and Classification of Wild Belugas and
+Narwhals Using Echolocation Clicks.''} \emph{Scientific Reports} 11 (1):
+22141. \url{https://doi.org/10.1038/s41598-021-01441-w}.
 

### 2023-02-07 natlantic/beakerBanter_natlantic.Rmd
Update manuscript and hawaii data
Extensive work on Manuscript and preparation of Hawaii data (in folder). Hawaii data *should* be good to knit. Manuscript has references, needs updated results, figures, inline data, and updated discussion.

---
title: 'BANTER for Beaked Whales: North Atlantic'
author: "Shannon Rankin"
date: "2023-01-26"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# When this rmarkdown file is knit:
# FALSE means it will used saved .rds files instead of re-processing
# TRUE will run everything from scratch - this will take a long time
freshRun <- FALSE
```

## PAMpal Data Processing

Start by loading the required packages

```{r, load packages, message=FALSE}
library (PAMpal)
library(banter)
library(rfPermute)
library(kableExtra)
library(readxl)
library(stringr)
library(here)
here()
```

1.  **Set up our PPS** (PAMPal Settings Object) for the NAtlantic
    dataset

```{r, pampal settings, eval=freshRun}
pps <- PAMpalSettings(db='Database/', 
                      binaries = 'Binaries/',
                      sr_hz='auto', 
                      winLen_sec=.0025, 
                      filterfrom_khz=10, 
                      filterto_khz=NULL)
```

2.  **Process data** and save to file to eliminate repeated processing.

[If]{.underline} you have already run the processing code, ensure you
have set 'freshRun = FALSE' at top of this document to read in the
existing .rds file for downstream processing.

```{r, read acoustic study, include=FALSE, eval=!freshRun}
data <- readRDS('natlantic_study.rds')
# Double check warning messages
print(getWarnings(data)$message)
```

[If]{.underline} this is the [initial processing]{.underline}, ensure
you have set 'freshRun = TRUE' at top of this document to process and
save data. This will take some time to run.

```{r, process detections, eval=freshRun}
data <- processPgDetections(pps, mode='db', id='natlantic_bw')
saveRDS(data, 'natlantic_study.rds')
# Double check warning messages
print(getWarnings(data)$message)
```

3.  **Assign species identity** according to original PAMguard labels,
    then relabel for consistency across projects.

    *Final species ID for beaked whale detections was provided by A.
    DeAngelis as an excel spreadsheet. The following code chunk pulls
    out the required information (event, species) in the format required
    for PAMpal.*

```{r, assign species ID}
HB1603_BW <- read_excel("HB1603_BWs.xlsx", sheet = "Alldata")#read in data
#create formatted event/species dataframe 
speciesID <- HB1603_BW[c("PGId", "Final Species Classification", "Database")] %>% 
  mutate("event" = paste(Database, PGId, sep=".OE")) %>% 
  rename("species" = "Final Species Classification") %>% 
  mutate("species" = str_replace_all(species, "'", ""),
         "species" = str_replace(species, "Gervais/Trues", "BW")) %>% 
  select(all_of(c("event", "species")))

#Assign species ID to data
data <- setSpecies(data, method='manual', value = speciesID)
#Note: 130 detections are not given species IDs and will be filtered out at a later step.
```

4.  **Filter out unwanted data**: Subset and rename species for BANTER
    model

```{r, filter unwanted data}
goodSpecies<- c(c("Trues", "Cuviers", "Sowerbys", "Gervais"))
data <- filter(data, species %in% goodSpecies)
```

5.  **Calculate Inter-Click Interval** (ICI).

```{r, calculate ICI}
data <- calculateICI(data, time='peakTime')
```

6.  **Export data for BANTER** (and drop species codes that will not be
    used for training). We will create two datasets: one with ICI and
    one without ICI, and save these for import into BANTER.

```{r, banter export, eval=freshRun}
banter_data <- export_banter(data, dropSpecies = 'unid', 
                         dropVars = c('All_ici'), training=TRUE)
saveRDS(banter_data, file='natlantic_banter_data.rds')

banter_data_ici <- export_banter(data, dropSpecies = 'unid', training=TRUE)
saveRDS(banter_data_ici, file='natlantic_banter_data_ici.rds')
```

## Build a BANTER Classification Model

### *Model without ICI*

1.  Load data, initialize a BANTER model, and create an initial BANTER
    detector model (stage 1).

```{r, read banter model, include=FALSE, eval=!freshRun}
banter_model <- readRDS('natlantic_banter_model_t5e3s3_t5e3s4.rds')
```

```{r, banter detector model, eval=freshRun}
banter_model <- initBanterModel(banter_data$events)
banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=5e3, sampsize=3, importance = TRUE)
```

Check for stability (are more trees needed?) before moving on

```{r, banter detector model stability}
plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 3:4))
plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 5:6))
summary(banter_model)
```

2.  Run BANTER Event Model (stage 2)

```{r, banter event model}
banter_model <- runBanterModel(banter_model, ntree=5e3, sampsize=4)
summary(banter_model)
```

Check for stability (are more trees needed?). If acceptable, save model
with tree/sampsize info in the filename.

```{r,save banter model, eval=freshRun}
saveRDS(banter_model, 'natlantic_banter_model_t5e3s3_t5e3s4.rds')
```

### *ICI Model*

1.  Load data, initialize a BANTER model, and create an initial BANTER
    detector model (stage 1).

```{r, read banter ici model, include=FALSE, eval=!freshRun}
banter_model_ici <- readRDS('natlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
```

```{r, banter ici detector model, eval=freshRun}
banter_model_ici <- initBanterModel(banter_data_ici$events)
banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=10e3, sampsize=2, importance = TRUE)
```

Check for stability to see if we need more trees

```{r, banter ici detector model stability}
plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 3:4))
plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 5:6))
summary(banter_model_ici)
```

2.  Run BANTER Event Model (stage 2)

```{r, banter ici event model}
banter_model_ici <- runBanterModel(banter_model_ici, ntree=10e3, sampsize=4)
summary(banter_model_ici)
```

Check for stability (are more trees needed?). If acceptable, save model
with tree/sampsize info in the filename.

```{r, save banter ici model, eval=freshRun}
saveRDS(banter_model_ici, 'natlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
```

## BANTER Analytics

There are a number of visualizations/data products that allow us to
visualize our BANTER classifier; most use the rfPermute package (see
[BANTER
Guidelines](https://taikisan21.github.io/PAMpal/banterGuide.html) for
more information)

First, identify the model you would like to examine (comment out the
model you [do not]{.underline} want to examine).

```{r, identify model}
model <- banter_model
modelname <- "banter_model"

model_ici <- banter_model_ici
modelname_ici <- "banter_model_ici"
```

Extract the Random Forest model object from our BANTER model for
analysis.

```{r, extract RF model, include=TRUE}
banter_model_RF <- getBanterModel(model)
banter_model_ici_RF <- getBanterModel(model_ici)
```

Confusion Matrix (simple model)

```{r, confusion matrix, include=TRUE}
natlantic_confuseMatrix <- rfPermute::confusionMatrix(banter_model_RF)
natlantic_confuseMatrix <- kable(natlantic_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
  kable_classic()%>%
  column_spec(4, border_right = TRUE)%>%
  row_spec(0, bold = TRUE)%>%
  row_spec(3,hline_after = TRUE)%>%
  row_spec(4, bold = TRUE)%>%
  save_kable(paste(modelname, "natlantic_confuseMatrix.png", sep="_"), zoom = 2)

natlantic_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
natlantic_ici_confuseMatrix <- kable(natlantic_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
  kable_classic()%>%
  column_spec(4, border_right = TRUE)%>%
  row_spec(0, bold = TRUE)%>%
  row_spec(3,hline_after = TRUE)%>%
  row_spec(4, bold = TRUE)%>%
  save_kable(paste(modelname_ici, "natlantic_ici_confuseMatrix.png", sep="_"), zoom = 2)

```

![](banter_model_natlantic_confuseMatrix.png)

Proximity Plot

```{r, proximity plot, include=TRUE}
png(paste(modelname, "natlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
natlantic_proximityPlot <- plotProximity(banter_model_RF)
dev.off()
natlantic_proximityPlot <- plotProximity(banter_model_RF)

png(paste(modelname_ici, "natlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
dev.off()
ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
```

Importance Heatmap

```{r, importance heat map, include=TRUE}
png(paste(modelname, "natlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
plotImportance(banter_model_RF, plot.type="heatmap")
dev.off()
natlantic_importance <- plotImportance(banter_model_RF, plot.type="heatmap")

png(paste(modelname_ici, "natlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
plotImportance(banter_model_ici_RF, plot.type="heatmap")
dev.off()
ici_natlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
```

PlotVotes

```{r, plot votes, include=TRUE}
png(paste(modelname, "natlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
plotVotes(banter_model_RF)
dev.off()
natlantic_votes <- plotVotes(banter_model_RF)

png(paste(modelname_ici, "natlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
plotVotes(banter_model_ici_RF)
dev.off()
ici_natlantic_votes <- plotVotes(banter_model_ici_RF)
```

Class Priors (Expected Error Rate)

```{r, class priors, include=TRUE}
natlantic_priors <- classPriors(banter_model_RF, NULL)
```

Plot Confusion Matrix Heat Map

```{r, confusion matrix heat map, include=TRUE}
plotConfMat(banter_model_RF, title="ConfusionMatrix", plot=TRUE)
```

Plot Error Trace for Banter Model

```{r, model error trace, include=TRUE}
plot(banter_model_RF)
```

Plot Predicted Probabilities

```{r, predicted probabilities, include=TRUE}
plotPredictedProbs(banter_model_RF, bins=30, plot=TRUE)
```

### 2023-02-07 satlantic/beakerBanter_satlantic.Rmd
Update manuscript and hawaii data
Extensive work on Manuscript and preparation of Hawaii data (in folder). Hawaii data *should* be good to knit. Manuscript has references, needs updated results, figures, inline data, and updated discussion.

--- 2023-02-07 natlantic/beakerBanter_natlantic.Rmd
@@
 ---
-title: 'BANTER for Beaked Whales: North Atlantic'
+title: 'BANTER for Beaked Whales: South Atlantic'
 author: "Shannon Rankin"
-date: "2023-01-26"
+date: "2023-02-03"
 output: html_document
@@
 
-1.  **Set up our PPS** (PAMPal Settings Object) for the NAtlantic
+1.  **Set up our PPS** (PAMPal Settings Object) for the satlantic
     dataset
@@
 ```{r, read acoustic study, include=FALSE, eval=!freshRun}
-data <- readRDS('natlantic_study.rds')
+data <- readRDS('satlantic_study.rds')
 # Double check warning messages
@@
 ```{r, process detections, eval=freshRun}
-data <- processPgDetections(pps, mode='db', id='natlantic_bw')
-saveRDS(data, 'natlantic_study.rds')
+data <- processPgDetections(pps, mode='db', id='satlantic_bw')
+saveRDS(data, 'satlantic_study.rds')
 # Double check warning messages
@@
 ```{r, assign species ID}
-HB1603_BW <- read_excel("HB1603_BWs.xlsx", sheet = "Alldata")#read in data
+GU1605_BW <- read_excel("GU1605_PG_OfflineEvents_20210809.xlsx", sheet = "AllData")#read in data
 #create formatted event/species dataframe 
-speciesID <- HB1603_BW[c("PGId", "Final Species Classification", "Database")] %>% 
-  mutate("event" = paste(Database, PGId, sep=".OE")) %>% 
-  rename("species" = "Final Species Classification") %>% 
+speciesID <- GU1605_BW[c("Id", "Species", "Database")] %>% 
+  mutate("event" = paste(Database, Id, sep=".OE")) %>% 
+  rename("species" = "Species") %>% 
   mutate("species" = str_replace_all(species, "'", ""),
-         "species" = str_replace(species, "Gervais/Trues", "BW")) %>% 
+         "species" = str_replace(species, "Sowerby", "Sowerbys"),
+         "species" = str_replace(species, "Blainville", "Blainvilles"),
+         "species" = str_replace(species, "Cuvier", "Cuviers")) %>% 
   select(all_of(c("event", "species")))
@@
 ```{r, filter unwanted data}
-goodSpecies<- c(c("Trues", "Cuviers", "Sowerbys", "Gervais"))
+goodSpecies<- c(c("Blainvilles", "Trues", "Cuviers", "Sowerbys", "Gervais"))
 data <- filter(data, species %in% goodSpecies)
@@
                          dropVars = c('All_ici'), training=TRUE)
-saveRDS(banter_data, file='natlantic_banter_data.rds')
+saveRDS(banter_data, file='satlantic_banter_data.rds')
 
 banter_data_ici <- export_banter(data, dropSpecies = 'unid', training=TRUE)
-saveRDS(banter_data_ici, file='natlantic_banter_data_ici.rds')
+saveRDS(banter_data_ici, file='satlantic_banter_data_ici.rds')
 ```
@@
 ```{r, read banter model, include=FALSE, eval=!freshRun}
-banter_model <- readRDS('natlantic_banter_model_t5e3s3_t5e3s4.rds')
+banter_model <- readRDS('satlantic_banter_model_t5e3s3_t5e3s4.rds')
 ```
@@
 banter_model <- initBanterModel(banter_data$events)
-banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=5e3, sampsize=3, importance = TRUE)
+banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=5e3, sampsize=4, importance = TRUE)
 ```
@@
 ```{r,save banter model, eval=freshRun}
-saveRDS(banter_model, 'natlantic_banter_model_t5e3s3_t5e3s4.rds')
+saveRDS(banter_model, 'satlantic_banter_model_t5e3s3_t5e3s4.rds')
 ```
@@
 ```{r, read banter ici model, include=FALSE, eval=!freshRun}
-banter_model_ici <- readRDS('natlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
+banter_model_ici <- readRDS('satlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
 ```
@@
 ```{r, save banter ici model, eval=freshRun}
-saveRDS(banter_model_ici, 'natlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
+saveRDS(banter_model_ici, 'satlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
 ```
@@
 ```{r, confusion matrix, include=TRUE}
-natlantic_confuseMatrix <- rfPermute::confusionMatrix(banter_model_RF)
-natlantic_confuseMatrix <- kable(natlantic_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+satlantic_confuseMatrix <- rfPermute::confusionMatrix(banter_model_RF)
+satlantic_confuseMatrix <- kable(satlantic_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
   kable_classic()%>%
@@
   row_spec(4, bold = TRUE)%>%
-  save_kable(paste(modelname, "natlantic_confuseMatrix.png", sep="_"), zoom = 2)
+  save_kable(paste(modelname, "satlantic_confuseMatrix.png", sep="_"), zoom = 2)
 
-natlantic_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
-natlantic_ici_confuseMatrix <- kable(natlantic_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+satlantic_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
+satlantic_ici_confuseMatrix <- kable(satlantic_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
   kable_classic()%>%
@@
   row_spec(4, bold = TRUE)%>%
-  save_kable(paste(modelname_ici, "natlantic_ici_confuseMatrix.png", sep="_"), zoom = 2)
+  save_kable(paste(modelname_ici, "satlantic_ici_confuseMatrix.png", sep="_"), zoom = 2)
 
@@
 
-![](banter_model_natlantic_confuseMatrix.png)
+![](banter_model_hawaii_confuseMatrix.png)
 
@@
 ```{r, proximity plot, include=TRUE}
-png(paste(modelname, "natlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
-natlantic_proximityPlot <- plotProximity(banter_model_RF)
+png(paste(modelname, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
+hawaii_proximityPlot <- plotProximity(banter_model_RF)
 dev.off()
-natlantic_proximityPlot <- plotProximity(banter_model_RF)
+hawaii_proximityPlot <- plotProximity(banter_model_RF)
 
-png(paste(modelname_ici, "natlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
-ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
+png(paste(modelname_ici, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
 dev.off()
-ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
 ```
@@
 ```{r, importance heat map, include=TRUE}
-png(paste(modelname, "natlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
+png(paste(modelname, "satlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
 plotImportance(banter_model_RF, plot.type="heatmap")
 dev.off()
-natlantic_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
+hawaii_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
 
-png(paste(modelname_ici, "natlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
+png(paste(modelname_ici, "satlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
 plotImportance(banter_model_ici_RF, plot.type="heatmap")
 dev.off()
-ici_natlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
 ```
@@
 ```{r, plot votes, include=TRUE}
-png(paste(modelname, "natlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
+png(paste(modelname, "satlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
 plotVotes(banter_model_RF)
 dev.off()
-natlantic_votes <- plotVotes(banter_model_RF)
+hawaii_votes <- plotVotes(banter_model_RF)
 
-png(paste(modelname_ici, "natlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
+png(paste(modelname_ici, "satlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
 plotVotes(banter_model_ici_RF)
 dev.off()
-ici_natlantic_votes <- plotVotes(banter_model_ici_RF)
+ici_hawaii_votes <- plotVotes(banter_model_ici_RF)
 ```
@@
 ```{r, class priors, include=TRUE}
-natlantic_priors <- classPriors(banter_model_RF, NULL)
+hawaii_priors <- classPriors(banter_model_RF, NULL)
 ```

### 2023-02-07 epacific/spot/SPOTgps2Pamguard.R

@@ -1,0 +1,58 @@
+##Spotgps2Pamguard.R
+#Jay gave me a spot track for each drift (2 spots for most drifts), but these are
+#not in the correct time. He then gave me a script that he used to fix the files, as
+#well as a 'fixed' file with UTC time (this is AllSpotTrackswUTC.csv )
+#Unfortunately, the column names are not suitable for addPgGps
+#so, first I must change those columns, then resave, then I can add the GPS
+#to the pamguard database
+
+#clean if needed
+rm(list=ls())
+
+#Run if your computer is set to GMT
+Sys.setenv(TZ='GMT')
+
+library(dplyr)
+library(plyr)
+library(purrr)
+library(tidyr)
+
+##Take all *.csv files and merge into one dataframe, and add the filename as an extra column
+setwd("E:/beaker.banter.2/pacwest/pascal/spot")
+temp <- purrr::map_dfr(list.files(pattern="*.csv", full.names = TRUE),
+                       ~read.csv(.x) %>% mutate(file = sub(".csv$", "", basename(.x))))
+
+#Separate the filename column to separate the station#
+pascal_AllSpot_UTC <-temp %>%
+  separate(file, c("info", "station"), sep="station")
+
+#Rename filenames
+colnames(pascal_AllSpot_UTC)
+pascal_AllSpot_UTC <- rename(pascal_AllSpot_UTC, c("dateTime" ="UTC", "spotID" = "Name", 
+                                               "long" ="Longitude",  "lat" = "Latitude"))
+colnames(pascal_AllSpot_UTC)
+#write as a csv
+write.csv(pascal_AllSpot_UTC, "pascal_AllSpot_UTC.csv")
+
+#view unique stations/SpotIDs to help select appropriate spotIDs
+pascalSpotStations <- read.csv("pascal_AllSpot_UTC.csv")
+spotSumm <- distinct(pascal_AllSpot_UTC[,c("Name", "station")])
+
+
+#Change working directory to databases
+setwd("E:/beaker.banter.2/pacwest/pascal/databases")
+library(PAMmisc)
+
+#Add GPS to Pamguard
+#Note: repeat for each database
+myDb <- "Station-28_Soundtrap-J_MASTER-BW.sqlite3"
+myGps <- "pascal_AllSpot_UTC.csv"
+addPgGps(db = myDb, gps = myGps, source = "csv", format = '%m/%d/%Y %H:%M:%S')
+
+
+
+
+
+
+
+

### 2023-02-07 epacific/spot/old/CleanupPASCAL_SpotData.r

@@ -1,0 +1,62 @@
+# Cleans up Spot location data from "CleanupPASCAL_SpotData.r"
+#    to convert to UTC, eliminate speed outliers and filter out locations before and after deployment
+#    also estimates deployment duration and distances traveled.
+
+library(geosphere)
+
+##############################################################################################
+# get deployment and retrieve times in Excel format and convert to R format
+DeployRetrTimes= read.csv("j:/Jay/ACOUSTIC/Buoy Recorder/PASCAL/DateTimeDataReconciliation.csv")
+DeployTime= as.POSIXct((DeployRetrTimes$UTC_DeployTime-2)*24*60*60,origin='1900-01-01',tz='gmt')
+RetrTime= as.POSIXct((DeployRetrTimes$UTC_RetrTime-2)*24*60*60,origin='1900-01-01',tz='gmt')
+
+
+######################################################################################################
+## read Spot geolocation info and save in a single file with UTC times added
+# NOTE downloaded Spot data were in local time (UTC - 7)
+setwd("j:\\Jay\\ACOUSTIC\\Buoy Recorder\\PASCAL\\SpotLocations by Station")
+sumKms= rep(NA,30); sumHrs= rep(NA,30)
+for (iStation in 1:30) {
+  SpotData= read.csv(file=paste("2017-08-04_spotTrack-station",as.character(iStation),".csv",sep=""))
+  
+  # convert date/time to UTC and store as number  
+  UTC=  strptime(SpotData$dateTime, "%m/%d/%Y %H:%M", tz="gmt") + (7*60*60)
+  # eliminate positions that are < deploy and > retrieve time
+  SpotData= SpotData[(UTC > DeployTime[iStation]) & (UTC < RetrTime[iStation]),]  
+  
+  #eliminate distance outliers
+  n= length(SpotData$lat)
+  dist= distGeo(cbind(SpotData$long[1:(n-1)],SpotData$lat[1:(n-1)]),cbind(SpotData$long[2:n],SpotData$lat[2:n])) / 1000   #distance in km
+  dist[n]= 0
+  #cat(iStation,max(dist),which.max(dist),"\n")
+  outliers= which(dist>2) + 1
[…]

### 2023-02-07 epacific/updateEventNames.R

@@ -1,0 +1,28 @@
+updateEventNames <- function(x, study=c('cces', 'pascal')) {
+    subFun <- switch(match.arg(study),
+                     'cces' = function(i) {
+                         cfile <- gsub('PamGuard64 2_00_16e |Drift-|_Final|_JST|_GoodAngles|\\.sqlite3$', '', i)
+                         cfile <- paste0('CCES_', cfile)
+                         cfile
+                     },
+                     'pascal' = function(i) {
+                         pfile <- gsub('_MASTER-BW| 15dB| Part|_ETG|\\.sqlite3$', '', i)
+                         pfile <- gsub('Station-([0-9\\-]{,6})_(Soundtrap|Card)-([A-z0-9]{,2})(.*)', 'PASCAL_\\1\\3\\4', pfile)
+                         pfile
+                     }
+    )
+    newNames <- sapply(names(events(x)), subFun)
+    names(newNames) <- NULL
+    names(events(x)) <- newNames
+    for(e in seq_along(events(x))) {
+        id(x[[e]]) <- newNames[e]
+        if(!is.null(ancillary(x[[e]])$environmental$event)) {
+            ancillary(x[[e]])$environmental$event <- newNames[e]
+        }
+    }
+    x
+}
+# updated 11/27/2020 to adjust for environmental event names
+# how to use - source this file first, then:
+# myData_pascal <- updateEventNames(myData_pascal, study='pascal')
+# myData_cces <- updateEventNames(myData_cces, study='cces')

### 2023-02-07 natlantic/beakerBanter_natlantic.Rmd
natlantic data
Add revised natlantic data w/ updated data from Annamaria

@@
 library(stringr)
+library(dplyr)
 library(here)
@@
 ```{r, assign species ID}
-HB1603_BW <- read_excel("HB1603_BWs.xlsx", sheet = "Alldata")#read in data
+HB1603 <- read_excel("HB1603_BWs.xlsx", sheet = "Alldata")#read in data
+HB1603 <- filter(HB1603, !eventType == "BRAN")
 #create formatted event/species dataframe 
-speciesID <- HB1603_BW[c("PGId", "Final Species Classification", "Database")] %>% 
+speciesID <- HB1603[c("PGId", "Final Species Classification", "Database")] %>% 
   mutate("event" = paste(Database, PGId, sep=".OE")) %>% 
@@
 data <- setSpecies(data, method='manual', value = speciesID)
-#Note: 130 detections are not given species IDs and will be filtered out at a later step.
+#Note: 156 detections are not given species IDs and will be filtered out at a later step.
 ```
@@
 png(paste(modelname, "natlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
-natlantic_proximityPlot <- plotProximity(banter_model_RF)
+hawaii_proximityPlot <- plotProximity(banter_model_RF)
 dev.off()
-natlantic_proximityPlot <- plotProximity(banter_model_RF)
+hawaii_proximityPlot <- plotProximity(banter_model_RF)
 
 png(paste(modelname_ici, "natlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
-ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
 dev.off()
-ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
 ```
@@
 dev.off()
-natlantic_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
+hawaii_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
 
@@
 dev.off()
-ici_natlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
 ```
@@
 dev.off()
-natlantic_votes <- plotVotes(banter_model_RF)
+hawaii_votes <- plotVotes(banter_model_RF)
 
@@
 dev.off()
-ici_natlantic_votes <- plotVotes(banter_model_ici_RF)
+ici_hawaii_votes <- plotVotes(banter_model_ici_RF)
 ```

### 2023-03-02 epacific/beakerBanter_epacific.Rmd
add epacific analysis
Added updated epacific analysis using CCES

---
title: "BANTER for Beaked Whales: EPacific (CCES survey)"
author: "Shannon Rankin"
date: "2023-03-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# When this rmarkdown file is knit:
# FALSE means it will used saved .rds files instead of re-processing
# TRUE will run everything from scratch - this will take a long time
freshRun <- FALSE
```

## PAMpal Data Processing

Start by loading the required packages

```{r, load packages, message=FALSE}
library(PAMpal)
library(banter)
library(rfPermute)
library(kableExtra)
library(here)
here()
```

1.  **Set up our PPS** (PAMPal Settings Object) for the EPacific dataset

```{r, pampal settings, eval=freshRun}
pps <- PAMpalSettings(db='Databases/', 
                      binaries = 'Binaries/',
                      sr_hz='auto', 
                      winLen_sec=.0025, 
                      filterfrom_khz=10, 
                      filterto_khz=NULL)
```

2.  **Process data** and save to file to eliminate repeated processing.

[If]{.underline} you have already run the processing code, ensure you
have set 'freshRun = FALSE' at top of this document to read in the
existing .rds file for downstream processing.

```{r, read acoustic study, eval=!freshRun}
data <- readRDS('epacific_study.rds')
# Double check warning messages
print(getWarnings(data)$message)
```

[If]{.underline} this is the [initial processing]{.underline}, ensure
you have set 'freshRun = TRUE' at top of this document to process and
save data. This will take some time to run.

```{r, process detections, eval=freshRun}
data <- processPgDetections(pps, mode='db', id='epacific_bw')
saveRDS(data, 'epacific_study.rds')
# Double check warning messages
print(getWarnings(data)$message)
```

3.  **Add GPS data**: Add GPS data from PAMGuard table

    ``` {r, add GPS}
    addGps(data)
    getWarnings(data)
    ```

4.  **Assign species identity** according to original PAMguard labels,
    then relabel for consistency across projects.

```{r, assign species ID}
data <- setSpecies(data, 'pamguard')
reSpecies <- readRDS('species.rds')
print(reSpecies)
data <- setSpecies(data, 'reassign', value=reSpecies)
```

5.  **Filter out unwanted data**: (1) subset and rename species for
    BANTER model and (2) filter to retain only Channel 1.

```{r, filter unwanted data}
goodSpecies<- c("ZC", "MS", "BB", "BW", "BW37V", "BW43", "possBW", "BWC")
data <- filter(data, species %in% goodSpecies)
data <- setSpecies(data, method='reassign',
                   value=data.frame(old=c('ZC', 'MS', 'BB', 'BWC'), new=c("Cuviers", "Stejnegers", "Bairds", "CrossSeamount")))
data_ch1only <- filter(data, Channel == '1')
```

6. **Calculate Inter-Click Interval** (ICI).

```{r, calculate ICI}
data_ch1only <- calculateICI(data_ch1only, time='peakTime')
```

7. **Export Event Wav Files** Wav files for events will be exported to a local drive, and then uploaded to Figshare.

``` {r, export events wav, eval=freshRun}
addRecordings (data_ch1only, log = TRUE )

wavDir <- "./ccesEventWav" #identify directory to write clips to
writeEventClips(data_ch1only, buffer = 5, mode = 'event', useSample = TRUE, folder = wavDir)
#Once this is finished, this data needs to be uploaded to a project in Figshare. Obtain the article ID and post below in annomate section

```

8. **Export to Annomate** Export annotation information to a csv for upload to Ann-O-Mate.
**NOTE** DO NOT SAVE THIS CODE CHUNK PUBLICLY--

```{r, export to annomate, include=TRUE, eval=freshRun}
#First, run:
unique(species(data)) #to see unique species and order in which they are named
#Then update line 119 to include full species names in order in which they occur
specMap <- data.frame(
    old = unique(species(data)),
    new = NA)
specMap$new <- c('Possible Beaked Whale', 'Ziphius cavirostris', 'Mesoplodon stejnegeri', 
                 'BW39V', 'Unidentified Beaked Whale', 'Berardius bairdii', 'BW43')
anno <- prepAnnotation(data, specMap = specMap, mode='event',
                       # following args are optional, fill in
                       # any that will be the same for all entries
                       source = 'figshare.com',
                       contact = 'shannon.rankin@noaa.gov')
#Export Annomate file, does this info save to acoustic study?

# to get figshare data you will need the article id and your personal token
# i recommend storing your token in an external file instead of typing it in here
# also DO NOT commit this file to any repository
figToken <- yaml::read_yaml('secrets.yaml')$fig_token
figId <- 13393360
figshareData <- getFigshareInfo(figToken, figId)

anno <- matchRecordingUrl(anno, figshareData)
# will give you warnings and messages about missing fields
checkAnnotation(anno)
# if you need to fix any manually, write to csv then read back in
annoFile <- 'AnnomateExport.csv'
write.csv(anno, file=annoFile, row.names = FALSE)
# go fix stuff
anno <- read.csv(annoFile, stringsAsFactors = FALSE)
# recheck to see if all seems fine
checkAnnotation(anno)
# add to acoustic study for storage
data <- addAnnotation(data, anno)
# this creates CSV ready for figshare, will also repeat messages from the check
export_annomate(data, file=annoFile)

```

9. Export data for BANTER (and drop species codes that will not be used
    for training). We will create two datasets: one with ICI and one
    without ICI, and save these for import into

```{r, banter export, eval=freshRun}
banter_data <- export_banter(data_ch1only, dropSpecies = c('BW', 'possBW'), 
                         dropVars = c('All_ici'), training=TRUE)
saveRDS(banter_data, file='epacific_banter_data.rds')

banter_data_ici <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), training=TRUE)
saveRDS(banter_data_ici, file='epacific_banter_data_ici.rds')

#save update of Acoustic Study
saveRDS(data_ch1only, 'epacific_study.rds')
saveRDS(data, 'epacificALL_study.rds')
```

## Build a BANTER Classification Model

### *Simple Model*

Load data, initialize a BANTER model, and create an initial BANTER
detector model (stage 1).

```{r, read banter model, include=FALSE, eval=!freshRun}
banter_model <- readRDS('epacific_banter_model_t1e4s3_t1e4s4.rds')
```

```{r, banter detector model, eval=freshRun}
banter_model <- initBanterModel(banter_data$events)
banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=1e4, sampsize=3, importance = TRUE)
```

Check for stability (are more trees needed?) before moving on

```{r, banter detector model stability, eval=freshRun}
plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 1:3))
plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 4:6))
summary(banter_model)
```

Run BANTER Event Model (stage 2)

```{r, banter event model, eval=freshRun}
banter_model <- runBanterModel(banter_model, ntree=1e4, sampsize=4)
summary(banter_model)
```

Check for stability (are more trees needed?). If acceptable, save model
with tree/sampsize info in the filename.

```{r,save banter model, eval=freshRun}
saveRDS(banter_model, 'epacific_banter_model_t1e4s3_t1e4s4.rds')
```

### *ICI Model*

Load data (if previously run)

```{r, read banter ici model, include=FALSE, eval=!freshRun}
banter_model_ici <- readRDS('epacific_banter_model_ici_t1e4s3_t1e4s4.rds')
```

Fresh Run: Initialize, Run & Evaluate Detector Model

```{r, banter ici detector model, eval=freshRun}
banter_model_ici <- initBanterModel(banter_data_ici$events)
banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=1e4, sampsize=3, importance = TRUE)
plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 1:3))
plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 4:6))
summary(banter_model_ici)
```

Run Event Model

```{r, banter ici event model, eval=freshRun}
banter_model_ici <- runBanterModel(banter_model_ici, ntree=1e4, sampsize=4)
summary(banter_model_ici)
```

Save BANTER model

```{r, save banter ici model, eval=freshRun}
saveRDS(banter_model_ici, 'epacific_banter_model_ici_t1e4s3_t1e4s4.rds')
```

## BANTER Analytics

There are a number of visualizations/data products that allow us to
visualize our BANTER classifier; most use the rfPermute package (see
[BANTER
Guidelines](https://taikisan21.github.io/PAMpal/banterGuide.html) for
more information. First, identify the model you would like to examine
(comment out the model you [do not]{.underline} want to examine).

```{r, identify model}
model <- banter_model
modelname <- "banter_model"

model_ici <- banter_model_ici
modelname_ici <- "banter_model_ici"
```

Extract the Random Forest model object from our BANTER model for
analysis.

```{r, extract RF model}
banter_model_RF <- getBanterModel(model)
banter_model_ici_RF <- getBanterModel(model_ici)
```

Confusion Matrix

```{r, confusion matrix, include=TRUE}
epacific_confuseMatrix <- rfPermute::confusionMatrix(banter_model_RF)
epacific_confuseMatrix <- kable(epacific_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
  kable_classic()%>%
  column_spec(6, border_right = TRUE)%>%
  row_spec(0, bold = TRUE)%>%
  row_spec(4,hline_after = TRUE)%>%
  row_spec(6, bold = TRUE)%>%
  save_kable(paste(modelname, "epacific_confuseMatrix.png", sep="_"), zoom = 2)

epacific_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
epacific_ici_confuseMatrix <- kable(epacific_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
  kable_classic()%>%
  column_spec(6, border_right = TRUE)%>%
  row_spec(0, bold = TRUE)%>%
  row_spec(4,hline_after = TRUE)%>%
  row_spec(6, bold = TRUE)%>%
  save_kable(paste(modelname_ici, "epacific_ici_confuseMatrix.png", sep="_"), zoom = 2)
```

Proximity Plot

```{r, proximity plot, include=TRUE}
png(paste(modelname, "epacific_proximity.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
epacific_proximityPlot <- plotProximity(banter_model_RF)
dev.off()
epacific_proximityPlot <- plotProximity(banter_model_RF)

png(paste(modelname_ici, "epacific_proximity.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
ici_epacific_proximityPlot <- plotProximity(banter_model_ici_RF)
dev.off()
ici_epacific_proximityPlot <- plotProximity(banter_model_ici_RF)

```

Importance Heatmap

```{r, importance heat map, include=TRUE}
png(paste(modelname, "epacific_importance.png", sep="_"), width = 30, height = 25, units = 'cm',  res = 300)
plotImportance(banter_model_RF, plot.type="heatmap")
dev.off()
epacific_importance <- plotImportance(banter_model_RF, plot.type="heatmap")

png(paste(modelname_ici, "epacific_importance.png", sep="_"), width = 30, height = 25, units = 'cm',  res = 300)
plotImportance(banter_model_ici_RF, plot.type="heatmap")
dev.off()
ici_epacific_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
```

PlotVotes

```{r, plot votes, include=TRUE}
png(paste(modelname, "epacific_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
plotVotes(banter_model_RF)
dev.off()
epacific_votes <- plotVotes(banter_model_RF)

png(paste(modelname_ici, "epacific_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
plotVotes(banter_model_ici_RF)
dev.off()
ici_epacific_votes <- plotVotes(banter_model_ici_RF)
```

Class Priors (Expected Error Rate)

```{r, class priors, include=TRUE}
epacific_priors <- classPriors(banter_model_RF, NULL)
```

Plot Confusion Matrix Heat Map

```{r, confusion matrix heat map, include=TRUE}
plotConfMat(banter_model_RF, title="ConfusionMatrix", plot=TRUE)
```

Plot Error Trace for Banter Model

```{r, model error trace, include=TRUE}
plot(banter_model_RF)
```

Plot Predicted Probabilities

```{r, predicted probabilities, include=TRUE}
plotPredictedProbs(banter_model_RF, bins=30, plot=TRUE)
```

### 2023-03-02 hawaii/beakerBanter_hawaii.Rmd
add epacific analysis
Added updated epacific analysis using CCES

@@
 author: "Shannon Rankin"
-date: "2022-02-06"
+date: "2023-02-06"
 output: html_document
@@
 saveRDS(banter_data_ici, file='hawaii_banter_data_ici.rds')
+
+#save update of Acoustic Study
+saveRDS(data_ch1only, 'hawaii_study.rds')
 ```

### 2023-03-02 manuscript/manuscript.qmd
add epacific analysis
Added updated epacific analysis using CCES

@@
 ---
+
 ```{r, hide=TRUE}
 library(tidyverse)
+library(PAMpal)
 library(knitr)
 library(here)
+here
 ```
+
 # Introduction
@@
 
-Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods ([@macleod2018]).
+Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod2018].
 However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering2013a].
@@
 
+```{r, hide = TRUE}
+setwd("..")
+natlanticData<-readRDS("natlantic/natlantic_study.rds")
+#Note: Species IDs not saved. Need to rerun? Not sure whats up.
+
+```
+
 Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n=70), Sowerby's beaked whales (n=5), True's beaked whales (n=65), and Gervais' beaked whales (n=3).
@@
 
+```{r, hide = TRUE}
+setwd("..")
+satlanticData<-readRDS("satlantic/satlantic_study.rds")
+```
+
 Species encountered in the SAtlantic study dataset include Blainsville's beaked whales (n=35), Cuvier's beaked whales (n=66), Gervais' beaked whales (n=45), Sowerby's beaked whales (n=2), and True's beaked whales (n-12).
@@
 ## Hawaii
+
 The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (Yano et al.

### 2023-03-02 manuscript/manuscript.tex
add epacific analysis
Added updated epacific analysis using CCES

@@
 K. McCullough \and Erin Oleson \and Melissa Soldevilla}
-\date{2/6/23}
+\date{2/7/23}
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 Passive acoustic monitoring (PAM) from has proven to be a valuable tool
-for studying populations of marine mammals ((Parijs et al. 2009));
+for studying populations of marine mammals (Parijs et al. 2009);
 however, the value of these studies depends on our ability to identify
@@
 attributed to a given species, based on their spectral or temporal
-characteristics ((Baumann-Pickering et al. 2013a); (Bittle and Duncan
-2013); (Rankin and Barlow 2005); (Soldevilla et al. 2008)). The dramatic
+characteristics (Baumann-Pickering et al. 2013a); (Bittle and Duncan
+2013); (Rankin and Barlow 2005); (Soldevilla et al. 2008). The dramatic
 increase in recordings make it impossible for experienced acousticians
@@
 difficult to study using typical shipboard visual observation methods
-((MacLeod 2018)). However, beaked whales make stereotyped echolocation
+(MacLeod 2018). However, beaked whales make stereotyped echolocation
 clicks that have temporal and spatial characteristics that can be used
-to differentiate species ((Baumann-Pickering et al. 2013b)). Detection
-of beaked whale signals in large datasets is greatly expanding our
+to differentiate species (Baumann-Pickering et al. 2013b). Detection of
+beaked whale signals in large datasets is greatly expanding our
 understanding of their population structure and the potential impact of
-human activities on these species ((Barlow et al. 2022);
-(Baumann-Pickering et al. 2014); (Simonis et al. 2020)). Unfortunately,
+human activities on these species (Barlow et al. 2022);
+(Baumann-Pickering et al. 2014); (Simonis et al. 2020). Unfortunately,
 manual classification of beaked whale echolocation clicks requires
@@
 originally developed for classifying dolphin schools from acoustic
-recordings ((Rankin et al. 2017a)). This hierarchical random forest
-event classifier relies on an initial call classifier (in the dolphin
-case study, this included one each for whistles, echolocation clicks,
-and burst pulses) followed by an event classifier. The event classifier
+recordings (Rankin et al. 2017a). This hierarchical random forest event
+classifier relies on an initial call classifier (in the dolphin case
+study, this included one each for whistles, echolocation clicks, and
+burst pulses) followed by an event classifier. The event classifier
 considers the probability of assignment to species for each call in the
@@
 call types, relatively minor changes to a call detector could yield
-different detector results ((Rankin et al. 2017b)). In this case,
-settings for a whistle \& moan detector were modified to improve
-performance at detecting burst pulses. This approach of applying
-multiple call detectors (of the same type but with different settings)
-suggests that perhaps BANTER could be used on species that only produce
-one call type (or where only one call type was analyzed), as long as the
-settings of the automated call detector were modified such that the
-overall number of detections was different. This approach was
-successfully applied to classified echolocation clicks for narwhal and
-belugas ((Zahn et al. 2021)). For species who primarily (or exclusively)
-produce echolocation clicks, BANTER may serve as an option for automated
-machine learning classification.
+different detector results (Rankin et al. 2017b). In this case, settings
+for a whistle \& moan detector were modified to improve performance at
+detecting burst pulses. This approach of applying multiple call
+detectors (of the same type but with different settings) suggests that
+perhaps BANTER could be used on species that only produce one call type
+(or where only one call type was analyzed), as long as the settings of
+the automated call detector were modified such that the overall number
+of detections was different. This approach was successfully applied to
+classified echolocation clicks for narwhal and belugas (Zahn et al.
+2021). For species who primarily (or exclusively) produce echolocation
+clicks, BANTER may serve as an option for automated machine learning
+classification.
 
@@
 2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels)
-survey ((J. Keating et al. 2018a)). Buoys were deployed at a number of
+survey (J. Keating et al. 2018a). Buoys were deployed at a number of
 stations in the California Current, including pre-determined locations
@@
 some sampling at 96 kHz for ocean noise measurements). Individual buoys
-were deployed for 11-23 days (see (J. Keating et al. 2018b) for more
-detailed information).
+were deployed for 11-23 days (see {[}J. Keating et al. (2018b){]} for
+more detailed information).
 
@@
 database and binaries for downstream processing using PAMpal package
-((Sakai 2021)) in R programming language ((R Core Team 2022)).
+(Sakai 2021) in R programming language (R Core Team 2022).
 
@@
 BANTER models was created for each trial using the BANTER package in R
-programming language ((Archer 2022a)). Standard click calculations were
+programming language (Archer 2022a). Standard click calculations were
 used for the BANTER call classifier; event level measures of ICI were
@@
 emphasis on lower sample size and higher number of trees for
-computational performance ((Rankin and Archer 2021)).
+computational performance (Rankin and Archer 2021).
 
@@
 visualizations provided by the rfpermute package in R programming
-language ((Archer 2022b)).
+language (Archer 2022b).
 
@@
 (\emph{Mesoplodon stejnegeri}), and unidentified beaked whale `BW43'
-((Baumann-Pickering et al. 2013a)). Analyses were conducted separately
-for the different study areas due to differences in the data collection
+(Baumann-Pickering et al. 2013a). Analyses were conducted separately for
+the different study areas due to differences in the data collection
 methods that precluded combining of datasets. Specific species
@@
 expertise to classify manually, using a variety of available visuals and
-metrics ((DeAngelis et al. 2018c)). Our results suggest that BANTER may
+metrics (DeAngelis et al. 2018c). Our results suggest that BANTER may
 serve as an efficient and effective means of classifying and

### 2023-03-02 natlantic/beakerBanter_natlantic.Rmd
add epacific analysis
Added updated epacific analysis using CCES

@@
 saveRDS(banter_data_ici, file='natlantic_banter_data_ici.rds')
+
+#save update of Acoustic Study
+saveRDS(data, 'natlantic_study.rds')
 ```

### 2023-03-02 satlantic/beakerBanter_satlantic.Rmd
add epacific analysis
Added updated epacific analysis using CCES

@@
 author: "Shannon Rankin"
-date: "2023-02-03"
+date: "2023-02-08"
 output: html_document
@@
 ```{r, assign species ID}
-GU1605_BW <- read_excel("GU1605_PG_OfflineEvents_20210809.xlsx", sheet = "AllData")#read in data
+GU1605 <- read_excel("GU1605_PG_OfflineEvents_20210809.xlsx", sheet = "BWonly")#read in data
+GU1605 <- filter(GU1605, !eventType == "BRAN")
 #create formatted event/species dataframe 
-speciesID <- GU1605_BW[c("Id", "Species", "Database")] %>% 
+speciesID <- GU1605[c("Id", "Species", "Database")] %>% 
   mutate("event" = paste(Database, Id, sep=".OE")) %>% 
@@
 data <- setSpecies(data, method='manual', value = speciesID)
-#Note: 130 detections are not given species IDs and will be filtered out at a later step.
+#Note: 238 detections are not given species IDs and will be filtered out at a later step.
 ```
@@
 saveRDS(banter_data_ici, file='satlantic_banter_data_ici.rds')
+
+#save update of Acoustic Study
+saveRDS(data, 'satlantic_study.rds')
 ```
@@
 ```{r, read banter model, include=FALSE, eval=!freshRun}
-banter_model <- readRDS('satlantic_banter_model_t5e3s3_t5e3s4.rds')
+banter_model <- readRDS('satlantic_banter_model_t1e4s2_t1e5s1.rds')
 ```
@@
 banter_model <- initBanterModel(banter_data$events)
-banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=5e3, sampsize=4, importance = TRUE)
+banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=1e4, sampsize=2, importance = TRUE)
 ```
@@
 ```{r, banter event model}
-banter_model <- runBanterModel(banter_model, ntree=5e3, sampsize=4)
+banter_model <- runBanterModel(banter_model, ntree=1e5, sampsize=1)
 summary(banter_model)
@@
 ```{r,save banter model, eval=freshRun}
-saveRDS(banter_model, 'satlantic_banter_model_t5e3s3_t5e3s4.rds')
+saveRDS(banter_model, 'satlantic_banter_model_t1e4s2_t1e5s1.rds')
 ```
@@
 ```{r, read banter ici model, include=FALSE, eval=!freshRun}
-banter_model_ici <- readRDS('satlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
+banter_model_ici <- readRDS('satlantic_banter_model_ici_t1e4s4_t1e5s1.rds')
 ```
@@
 banter_model_ici <- initBanterModel(banter_data_ici$events)
-banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=10e3, sampsize=2, importance = TRUE)
+banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=1e4, sampsize=4, importance = TRUE)
 ```
@@
 ```{r, banter ici event model}
-banter_model_ici <- runBanterModel(banter_model_ici, ntree=10e3, sampsize=4)
+banter_model_ici <- runBanterModel(banter_model_ici, ntree=1e5, sampsize=1)
 summary(banter_model_ici)
@@
 ```{r, save banter ici model, eval=freshRun}
-saveRDS(banter_model_ici, 'satlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
+saveRDS(banter_model_ici, 'satlantic_banter_model_ici_t1e4s4_t1e5s1.rds')
 ```
@@
 
-![](banter_model_hawaii_confuseMatrix.png)
+![](banter_model_satlantic_confuseMatrix.png)
 
@@
 png(paste(modelname, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
-hawaii_proximityPlot <- plotProximity(banter_model_RF)
+satlantic_proximityPlot <- plotProximity(banter_model_RF)
 dev.off()
-hawaii_proximityPlot <- plotProximity(banter_model_RF)
+satlantic_proximityPlot <- plotProximity(banter_model_RF)
 
 png(paste(modelname_ici, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
-ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_satlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
 dev.off()
-ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_satlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
 ```
@@
 dev.off()
-hawaii_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
+satlantic_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
 
@@
 dev.off()
-ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_satlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
 ```
@@
 dev.off()
-hawaii_votes <- plotVotes(banter_model_RF)
+satlantic_votes <- plotVotes(banter_model_RF)
 
@@
 dev.off()
-ici_hawaii_votes <- plotVotes(banter_model_ici_RF)
+ici_satlantic_votes <- plotVotes(banter_model_ici_RF)
 ```
@@
 ```{r, class priors, include=TRUE}
-hawaii_priors <- classPriors(banter_model_RF, NULL)
+satlantic_priors <- classPriors(banter_model_RF, NULL)
 ```

### 2023-03-02 epacific/spot/SPOTgps2Pamguard.R

@@ -1,58 +1,0 @@
-##Spotgps2Pamguard.R
-#Jay gave me a spot track for each drift (2 spots for most drifts), but these are
-#not in the correct time. He then gave me a script that he used to fix the files, as
-#well as a 'fixed' file with UTC time (this is AllSpotTrackswUTC.csv )
-#Unfortunately, the column names are not suitable for addPgGps
-#so, first I must change those columns, then resave, then I can add the GPS
-#to the pamguard database
-
-#clean if needed
-rm(list=ls())
-
-#Run if your computer is set to GMT
-Sys.setenv(TZ='GMT')
-
-library(dplyr)
-library(plyr)
-library(purrr)
-library(tidyr)
-
-##Take all *.csv files and merge into one dataframe, and add the filename as an extra column
-setwd("E:/beaker.banter.2/pacwest/pascal/spot")
-temp <- purrr::map_dfr(list.files(pattern="*.csv", full.names = TRUE),
-                       ~read.csv(.x) %>% mutate(file = sub(".csv$", "", basename(.x))))
-
-#Separate the filename column to separate the station#
-pascal_AllSpot_UTC <-temp %>%
-  separate(file, c("info", "station"), sep="station")
-
-#Rename filenames
-colnames(pascal_AllSpot_UTC)
-pascal_AllSpot_UTC <- rename(pascal_AllSpot_UTC, c("dateTime" ="UTC", "spotID" = "Name", 
-                                               "long" ="Longitude",  "lat" = "Latitude"))
-colnames(pascal_AllSpot_UTC)
-#write as a csv
-write.csv(pascal_AllSpot_UTC, "pascal_AllSpot_UTC.csv")
-
-#view unique stations/SpotIDs to help select appropriate spotIDs
-pascalSpotStations <- read.csv("pascal_AllSpot_UTC.csv")
-spotSumm <- distinct(pascal_AllSpot_UTC[,c("Name", "station")])
-
-
-#Change working directory to databases
-setwd("E:/beaker.banter.2/pacwest/pascal/databases")
-library(PAMmisc)
-
-#Add GPS to Pamguard
-#Note: repeat for each database
-myDb <- "Station-28_Soundtrap-J_MASTER-BW.sqlite3"
-myGps <- "pascal_AllSpot_UTC.csv"
-addPgGps(db = myDb, gps = myGps, source = "csv", format = '%m/%d/%Y %H:%M:%S')
-
-
-
-
-
-
-
-

### 2023-03-02 epacific/spot/old/CleanupPASCAL_SpotData.r

@@ -1,62 +1,0 @@
-# Cleans up Spot location data from "CleanupPASCAL_SpotData.r"
-#    to convert to UTC, eliminate speed outliers and filter out locations before and after deployment
-#    also estimates deployment duration and distances traveled.
-
-library(geosphere)
-
-##############################################################################################
-# get deployment and retrieve times in Excel format and convert to R format
-DeployRetrTimes= read.csv("j:/Jay/ACOUSTIC/Buoy Recorder/PASCAL/DateTimeDataReconciliation.csv")
-DeployTime= as.POSIXct((DeployRetrTimes$UTC_DeployTime-2)*24*60*60,origin='1900-01-01',tz='gmt')
-RetrTime= as.POSIXct((DeployRetrTimes$UTC_RetrTime-2)*24*60*60,origin='1900-01-01',tz='gmt')
-
-
-######################################################################################################
-## read Spot geolocation info and save in a single file with UTC times added
-# NOTE downloaded Spot data were in local time (UTC - 7)
-setwd("j:\\Jay\\ACOUSTIC\\Buoy Recorder\\PASCAL\\SpotLocations by Station")
-sumKms= rep(NA,30); sumHrs= rep(NA,30)
-for (iStation in 1:30) {
-  SpotData= read.csv(file=paste("2017-08-04_spotTrack-station",as.character(iStation),".csv",sep=""))
-  
-  # convert date/time to UTC and store as number  
-  UTC=  strptime(SpotData$dateTime, "%m/%d/%Y %H:%M", tz="gmt") + (7*60*60)
-  # eliminate positions that are < deploy and > retrieve time
-  SpotData= SpotData[(UTC > DeployTime[iStation]) & (UTC < RetrTime[iStation]),]  
-  
-  #eliminate distance outliers
-  n= length(SpotData$lat)
-  dist= distGeo(cbind(SpotData$long[1:(n-1)],SpotData$lat[1:(n-1)]),cbind(SpotData$long[2:n],SpotData$lat[2:n])) / 1000   #distance in km
-  dist[n]= 0
-  #cat(iStation,max(dist),which.max(dist),"\n")
-  outliers= which(dist>2) + 1
[…]

### 2023-03-02 epacific/updateEventNames.R

@@ -1,28 +1,0 @@
-updateEventNames <- function(x, study=c('cces', 'pascal')) {
-    subFun <- switch(match.arg(study),
-                     'cces' = function(i) {
-                         cfile <- gsub('PamGuard64 2_00_16e |Drift-|_Final|_JST|_GoodAngles|\\.sqlite3$', '', i)
-                         cfile <- paste0('CCES_', cfile)
-                         cfile
-                     },
-                     'pascal' = function(i) {
-                         pfile <- gsub('_MASTER-BW| 15dB| Part|_ETG|\\.sqlite3$', '', i)
-                         pfile <- gsub('Station-([0-9\\-]{,6})_(Soundtrap|Card)-([A-z0-9]{,2})(.*)', 'PASCAL_\\1\\3\\4', pfile)
-                         pfile
-                     }
-    )
-    newNames <- sapply(names(events(x)), subFun)
-    names(newNames) <- NULL
-    names(events(x)) <- newNames
-    for(e in seq_along(events(x))) {
-        id(x[[e]]) <- newNames[e]
-        if(!is.null(ancillary(x[[e]])$environmental$event)) {
-            ancillary(x[[e]])$environmental$event <- newNames[e]
-        }
-    }
-    x
-}
-# updated 11/27/2020 to adjust for environmental event names
-# how to use - source this file first, then:
-# myData_pascal <- updateEventNames(myData_pascal, study='pascal')
-# myData_cces <- updateEventNames(myData_cces, study='cces')

### 2023-06-08 epacific/beakerBanter_epacific.Rmd
Major revisions for draft
Revised all datasets and created draft manuscript

@@
 author: "Shannon Rankin"
-date: "2023-03-01"
+date: "2023-05-25"
 output: html_document
@@
 
+###Organizational Instructions Save this file in an R Project that
+includes the following foldering: /Project \|-Database/ \|-Binaries/
+\|-Figures/ \|-epacific \|\_ beakerBanter_epacific.Rmd \|\_
+epacific_banter_data.rds \|\_ epacific_banter_data_ici.rds \|\_
+epacific_banter_model_ec_t1e4s3_t1e4s4.rds \|\_
+epacific_banter_model_ici_t1e4s3_t1e4s4
+
 ## PAMpal Data Processing
@@
 ```{r, load packages, message=FALSE}
-library(PAMpal)
-library(banter)
-library(rfPermute)
-library(kableExtra)
-library(here)
+library("easypackages")
+libraries("PAMpal", "banter", "rfPermute", "kableExtra", "magick", "magrittr", "here")
 here()
@@
 
-[If]{.underline} you have already run the processing code, ensure you
-have set 'freshRun = FALSE' at top of this document to read in the
-existing .rds file for downstream processing.
-
-```{r, read acoustic study, eval=!freshRun}
-data <- readRDS('epacific_study.rds')
-# Double check warning messages
-print(getWarnings(data)$message)
-```
-
 [If]{.underline} this is the [initial processing]{.underline}, ensure
@@
 
-3.  **Add GPS data**: Add GPS data from PAMGuard table
+[If]{.underline} you have already run the processing code, ensure you
+have set 'freshRun = FALSE' at top of this document to read in the
+existing .rds file for downstream processing.
 
-    ``` {r, add GPS}
-    addGps(data)
-    getWarnings(data)
+```{r, read acoustic study, include=FALSE, eval=!freshRun}
+data_ch1only <- readRDS('epacific_study.rds')
+# Double check warning messages
+print(getWarnings(data_ch1only)$message)#Note: Detection >7200sec apart is a sperm whale detection. Anne was informed of the error.
 ```
 
-4.  **Assign species identity** according to original PAMguard labels,
+3.  **Assign species identity** according to original PAMguard labels,
     then relabel for consistency across projects.
 
-```{r, assign species ID}
+```{r, assign species ID, eval=freshRun}
 data <- setSpecies(data, 'pamguard')
@@
 
-5.  **Filter out unwanted data**: (1) subset and rename species for
+4.  **Filter out unwanted data**: (1) subset and rename species for
     BANTER model and (2) filter to retain only Channel 1.
 
-```{r, filter unwanted data}
-goodSpecies<- c("ZC", "MS", "BB", "BW", "BW37V", "BW43", "possBW", "BWC")
+```{r, filter unwanted data, eval=freshRun}
+data <- setSpecies(data, method='reassign',
+                   value=data.frame(old=c('ZC', 'MS', 'BB', 'BWC', 'BW37V'), new=c("Cuviers", "Stejnegers", "Bairds", "CrossSeamount", "Hubbs")))
+goodSpecies<- c("Cuviers", "Stejnegers", "Bairds", "CrossSeamount", "Hubbs", "BW", "BW43", "possBW")
 data <- filter(data, species %in% goodSpecies)
-data <- setSpecies(data, method='reassign',
-                   value=data.frame(old=c('ZC', 'MS', 'BB', 'BWC'), new=c("Cuviers", "Stejnegers", "Bairds", "CrossSeamount")))
 data_ch1only <- filter(data, Channel == '1')
@@
 
-6. **Calculate Inter-Click Interval** (ICI).
+5.  **Calculate Inter-Click Interval** (ICI).
 
-```{r, calculate ICI}
+```{r, calculate ICI, eval=freshRun}
 data_ch1only <- calculateICI(data_ch1only, time='peakTime')
@@
 
-7. **Export Event Wav Files** Wav files for events will be exported to a local drive, and then uploaded to Figshare.
+6.  **Add GPS data**: Add GPS data from PAMGuard table (2 hour
+    threshold), then filter out data without GPS
 
-``` {r, export events wav, eval=freshRun}
-addRecordings (data_ch1only, log = TRUE )
+```{r, add GPS, eval=freshRun}
+data_ch1only <- addGps(data_ch1only, thresh = 7200)#11 events over the threshold
+getWarnings(data_ch1only)
 
-wavDir <- "./ccesEventWav" #identify directory to write clips to
-writeEventClips(data_ch1only, buffer = 5, mode = 'event', useSample = TRUE, folder = wavDir)
-#Once this is finished, this data needs to be uploaded to a project in Figshare. Obtain the article ID and post below in annomate section
+data_ch1only <- filter(data_ch1only, !is.na(Latitude))#filter out events w/o GPS
+clicks <- getClickData(data_ch1only)
+sum(is.na(clicks$Latitude))
+unique(clicks[is.na(clicks$Latitude), 'eventId'])#confirm there are no remaining events w/o GPS
+```
+
+7. **Add Environmental Data**
+```{r, Add Environmental Data, eval=freshRun}
+#SST
+data_ch1only <- matchEnvData(data_ch1only, nc='jplMURSST41mday', var='sst')
+#Seafloor Depth
+data_ch1only <- matchEnvData(data_ch1only, nc='erdSrtm30plusSeafloorGradient', var='sea_floor_depth')
+#Seafloor Gradient
+data_ch1only <- matchEnvData(data_ch1only, nc='erdSrtm30plusSeafloorGradient', var='magnitude_gradient')
 
@@
 
-8. **Export to Annomate** Export annotation information to a csv for upload to Ann-O-Mate.
-**NOTE** DO NOT SAVE THIS CODE CHUNK PUBLICLY--
+7.  **Export Event Wav Files** Wav files for events will be exported to
+    a local drive, and then uploaded to Figshare.
 
-```{r, export to annomate, include=TRUE, eval=freshRun}
-#First, run:
-unique(species(data)) #to see unique species and order in which they are named
-#Then update line 119 to include full species names in order in which they occur
-specMap <- data.frame(
-    old = unique(species(data)),
-    new = NA)
-specMap$new <- c('Possible Beaked Whale', 'Ziphius cavirostris', 'Mesoplodon stejnegeri', 
-                 'BW39V', 'Unidentified Beaked Whale', 'Berardius bairdii', 'BW43')
-anno <- prepAnnotation(data, specMap = specMap, mode='event',
-                       # following args are optional, fill in
-                       # any that will be the same for all entries
-                       source = 'figshare.com',
-                       contact = 'shannon.rankin@noaa.gov')
-#Export Annomate file, does this info save to acoustic study?
+```{r, export events wav, eval=freshRun}
+#This should only be run one time, with access to original wav files. Will be commented out to ensure it is not accidentally run
+#data <- readRDS(here('epacific', 'epacific_study.rds'))
+#Check that all events are <120s
+#checkStudy(data, maxLength = 120)
 
-# to get figshare data you will need the article id and your personal token
-# i recommend storing your token in an external file instead of typing it in here
-# also DO NOT commit this file to any repository
-figToken <- yaml::read_yaml('secrets.yaml')$fig_token
-figId <- 13393360
-figshareData <- getFigshareInfo(figToken, figId)
+#Identify location of all recording files and log files
+#data<- addRecordings (data, log = TRUE )
+#saveRDS(data, 'epacific_study.rds')
 
-anno <- matchRecordingUrl(anno, figshareData)
-# will give you warnings and messages about missing fields
-checkAnnotation(anno)
-# if you need to fix any manually, write to csv then read back in
-annoFile <- 'AnnomateExport.csv'
-write.csv(anno, file=annoFile, row.names = FALSE)
-# go fix stuff
-anno <- read.csv(annoFile, stringsAsFactors = FALSE)
-# recheck to see if all seems fine
-checkAnnotation(anno)
-# add to acoustic study for storage
-data <- addAnnotation(data, anno)
-# this creates CSV ready for figshare, will also repeat messages from the check
-export_annomate(data, file=annoFile)
+#Identify directory to save wav files, and write event Wav Clips
+#wavDir <- here('epacific', 'ccesEventWav')#identify directory to write clips to
+#writeEventClips(data, buffer = 5, mode = 'event', useSample = FALSE, outDir = wavDir)
+#saveRDS(data, 'epacific_study.rds')
 
 ```
+
+8.  **Export to Annomate** Export annotation information to a csv for
+    upload to Ann-O-Mate. Run code in annomate.R, but do not share this code publicly.
 
@@
 banter_data <- export_banter(data_ch1only, dropSpecies = c('BW', 'possBW'), 
-                         dropVars = c('All_ici'), training=TRUE)
+                         dropVars = c('All_ici', 'Click_Detector_3_ici', 'sea_floor_depth_mean', 'sst_mean', 'productivity_mean', 'chlorophyll_mean',  'magnitude_gradient_mean'), training=TRUE)
 saveRDS(banter_data, file='epacific_banter_data.rds')
 
-banter_data_ici <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), training=TRUE)
+banter_data_ici <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), dropVars = c('sea_floor_depth_mean', 'sst_mean', 'productivity_mean', 'chlorophyll_mean',  'magnitude_gradient_mean'),training=TRUE)
 saveRDS(banter_data_ici, file='epacific_banter_data_ici.rds')
+
+banter_data_env <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), training=TRUE)
+saveRDS(banter_data_env, file='epacific_banter_data_env.rds')
 
@@
 
-### *Simple Model*
+### *EC (only) Model*
 
-Load data, initialize a BANTER model, and create an initial BANTER
-detector model (stage 1).
+Initialize, Run & Evaluate Detector Model (stage 1).
 
-```{r, read banter model, include=FALSE, eval=!freshRun}
-banter_model <- readRDS('epacific_banter_model_t1e4s3_t1e4s4.rds')
-```
+```{r, banter ec detector model, eval=freshRun}
+banter_model_ec <- initBanterModel(banter_data$events)
+banter_model_ec <- addBanterDetector(banter_model_ec, banter_data$detectors, ntree=1e4, sampsize=3, importance = TRUE)
 
-```{r, banter detector model, eval=freshRun}
-banter_model <- initBanterModel(banter_data$events)
-banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=1e4, sampsize=3, importance = TRUE)
-```
-
-Check for stability (are more trees needed?) before moving on
-
-```{r, banter detector model stability, eval=freshRun}
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 1:3))
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 4:6))
-summary(banter_model)
+plotDetectorTrace(banter_model_ec, detector = paste0('Click_Detector_', 1:3))
+plotDetectorTrace(banter_model_ec, detector = paste0('Click_Detector_', 4:6))
+summary(banter_model_ec)
 ```
@@
 
-```{r, banter event model, eval=freshRun}
-banter_model <- runBanterModel(banter_model, ntree=1e4, sampsize=4)
-summary(banter_model)
+```{r, banter ec event model, eval=freshRun}
+banter_model_ec <- runBanterModel(banter_model_ec, ntree=1e4, sampsize=4)
+summary(banter_model_ec)
 ```
 
-Check for stability (are more trees needed?). If acceptable, save model
-with tree/sampsize info in the filename.
+Once a stable model is identified, save model with tree/sampsize info in
+the filename.
 
-```{r,save banter model, eval=freshRun}
-saveRDS(banter_model, 'epacific_banter_model_t1e4s3_t1e4s4.rds')
+```{r,save banter ec model, eval=freshRun}
+saveRDS(banter_model_ec, 'epacific_banter_model_ec_t1e4s3_t1e4s4.rds')
 ```
@@
 
-Load data (if previously run)
-
-```{r, read banter ici model, include=FALSE, eval=!freshRun}
-banter_model_ici <- readRDS('epacific_banter_model_ici_t1e4s3_t1e4s4.rds')
-```
-
-Fresh Run: Initialize, Run & Evaluate Detector Model
+Initialize, Run & Evaluate Detector Model (stage 1).
 
@@
 banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=1e4, sampsize=3, importance = TRUE)
+
 plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 1:3))
@@
 
-Run Event Model
+Run Event Model (stage 2)
 
@@
 
-Save BANTER model
+Once a stable model is identified, save model with tree/sampsize info in
+the filename.
 
@@
 ```
+### *ENV Model*
 
+Initialize, Run & Evaluate Detector Model (stage 1).
+```{r, banter env detector model, eval=freshRun}
+banter_model_env <- initBanterModel(banter_data_env$events)
+banter_model_env <- addBanterDetector(banter_model_env, banter_data_env$detectors, ntree=1e4, sampsize=1, importance = TRUE)
+
+plotDetectorTrace(banter_model_env, detector = paste0('Click_Detector_', 1:3))
+plotDetectorTrace(banter_model_env, detector = paste0('Click_Detector_', 4:6))
+summary(banter_model_env)
+```
+
+Run Event Model (stage 2)
+```{r, banter env event model, eval=freshRun}
+banter_model_env <- runBanterModel(banter_model_env, ntree=1e4, sampsize=5)
+summary(banter_model_env)
+```
+
+Once a stable model is identified, save model with tree/sampsize info in
+the filename.
+
+```{r, save banter env model, eval=freshRun}
+saveRDS(banter_model_env, 'epacific_banter_model_env_t1e4s1_t1e4s5.rds')
+```
 ## BANTER Analytics
@@
 Guidelines](https://taikisan21.github.io/PAMpal/banterGuide.html) for
-more information. First, identify the model you would like to examine
-(comment out the model you [do not]{.underline} want to examine).
+more information.
 
-```{r, identify model}
-model <- banter_model
-modelname <- "banter_model"
+First, load the models (if not a fresh run)
+
+```{r, read banter model, include=FALSE, eval=!freshRun}
+banter_model_ec <- readRDS('epacific_banter_model_ec_t1e4s3_t1e4s4.rds')
+banter_model_ici <- readRDS('epacific_banter_model_ici_t1e4s3_t1e4s4.rds')
+banter_model_env <- readRDS('epacific_banter_model_env_t1e4s1_t1e4s5.rds')
+```
+
+Identify the model you would like to examine (comment out the model you
+[do not]{.underline} want to examine).
+
+```{r, identify model, eval=!freshRun}
+model_ec <- banter_model_ec
+modelname_ec <- "banter_model_ec"
 
@@
 modelname_ici <- "banter_model_ici"
+
+model_env <- banter_model_env
+modelname_env <- "banter_model_env"
 ```
@@
 
-```{r, extract RF model}
-banter_model_RF <- getBanterModel(model)
+```{r, extract RF model, eval=!freshRun}
+banter_model_ec_RF <- getBanterModel(model_ec)
 banter_model_ici_RF <- getBanterModel(model_ici)
+banter_model_env_RF <- getBanterModel(model_env)
+```
+
+Class Priors (Expected Error Rate)
+
+```{r, class priors, include=TRUE, eval=!freshRun}
+epacific_ec_priors <- classPriors(banter_model_ec_RF, NULL)[,1]
+epacific_ici_priors <- classPriors(banter_model_ici_RF, NULL)[,1]
+epacific_env_priors <- classPriors(banter_model_env_RF, NULL)[,1]
 ```
@@
 
-```{r, confusion matrix, include=TRUE}
-epacific_confuseMatrix <- rfPermute::confusionMatrix(banter_model_RF)
-epacific_confuseMatrix <- kable(epacific_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+```{r, confusion matrix, include=TRUE, eval=!freshRun}
+epacific_ec_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ec_RF)
+epacific_ec_confuseMatrix <- cbind(epacific_ec_confuseMatrix, priors = epacific_ec_priors)
+epacific_ec_confuseMatrix <- kable(epacific_ec_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
   kable_classic()%>%
-  column_spec(6, border_right = TRUE)%>%
+   column_spec(7, border_right = TRUE)%>%
    row_spec(0, bold = TRUE)%>%
    row_spec(4,hline_after = TRUE)%>%
-  row_spec(6, bold = TRUE)%>%
-  save_kable(paste(modelname, "epacific_confuseMatrix.png", sep="_"), zoom = 2)
+   row_spec(7, bold = TRUE)%>%
+   save_kable("../manuscript/manuscript_files/epacific_ec_confuseMatrix.png", zoom = 9)
 
 epacific_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
+epacific_ici_confuseMatrix <- cbind(epacific_ici_confuseMatrix, priors = epacific_ici_priors)
 epacific_ici_confuseMatrix <- kable(epacific_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
   kable_classic()%>%
-  column_spec(6, border_right = TRUE)%>%
+  column_spec(7, border_right = TRUE)%>%
   row_spec(0, bold = TRUE)%>%
   row_spec(4,hline_after = TRUE)%>%
-  row_spec(6, bold = TRUE)%>%
-  save_kable(paste(modelname_ici, "epacific_ici_confuseMatrix.png", sep="_"), zoom = 2)
+  row_spec(7, bold = TRUE)%>%
+  save_kable("../manuscript/manuscript_files/epacific_ici_confuseMatrix.png", zoom = 9)
+
+epacific_env_confuseMatrix <- rfPermute::confusionMatrix(banter_model_env_RF)
+epacific_env_confuseMatrix <- cbind(epacific_env_confuseMatrix, priors = epacific_env_priors)
+epacific_env_confuseMatrix <- kable(epacific_env_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(7, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(4,hline_after = TRUE)%>%
+  row_spec(7, bold = TRUE)%>%
+  save_kable("../manuscript/manuscript_files/epacific_env_confuseMatrix.png", zoom = 9)
 ```
+
+![BANTER Model E.Pacific EC Confusion
+Matrix](../manuscript/manuscript_files/epacific_ec_confuseMatrix.png)
+
+![BANTER Model E.Pacific ICI Confusion
+Matrix](../manuscript/manuscript_files/epacific_ici_confuseMatrix.png)
+
+![BANTER Model E.Pacific ENV Confusion
+Matrix](../manuscript/manuscript_files/epacific_env_confuseMatrix.png)
 
@@
 
-```{r, proximity plot, include=TRUE}
-png(paste(modelname, "epacific_proximity.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
-epacific_proximityPlot <- plotProximity(banter_model_RF)
+```{r, proximity plot, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/epacific_ec_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotProximity(banter_model_ec_RF)
 dev.off()
-epacific_proximityPlot <- plotProximity(banter_model_RF)
+ec_epacific_proximityPlot <- plotProximity(banter_model_ec_RF)
 
-png(paste(modelname_ici, "epacific_proximity.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
+png(('../manuscript/manuscript_files/epacific_ici_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
 ici_epacific_proximityPlot <- plotProximity(banter_model_ici_RF)
@@
 
+png(('../manuscript/manuscript_files/epacific_env_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
+env_epacific_proximityPlot <- plotProximity(banter_model_env_RF)
+dev.off()
+env_epacific_proximityPlot <- plotProximity(banter_model_env_RF)
 ```
@@
 
-```{r, importance heat map, include=TRUE}
-png(paste(modelname, "epacific_importance.png", sep="_"), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_RF, plot.type="heatmap")
+```{r, importance heat map, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/epacific_ec_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
+plotImportance(banter_model_ec_RF, plot.type="heatmap")
 dev.off()
-epacific_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
+ec_epacific_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap")
 
-png(paste(modelname_ici, "epacific_importance.png", sep="_"), width = 30, height = 25, units = 'cm',  res = 300)
+png(('../manuscript/manuscript_files/epacific_ici_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
 plotImportance(banter_model_ici_RF, plot.type="heatmap")
@@
 ici_epacific_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+
+png(('../manuscript/manuscript_files/epacific_env_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
+plotImportance(banter_model_env_RF, plot.type="heatmap")
+dev.off()
+env_epacific_importance <- plotImportance(banter_model_env_RF, plot.type="heatmap")
 ```
@@
 
-```{r, plot votes, include=TRUE}
-png(paste(modelname, "epacific_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_RF)
+```{r, plot votes, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/epacific_ec_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_ec_RF)
 dev.off()
-epacific_votes <- plotVotes(banter_model_RF)
+epacific_votes <- plotVotes(banter_model_ec_RF)
 
-png(paste(modelname_ici, "epacific_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
+png(('../manuscript/manuscript_files/epacific_ici_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
 plotVotes(banter_model_ici_RF)
@@
 ici_epacific_votes <- plotVotes(banter_model_ici_RF)
-```
 
-Class Priors (Expected Error Rate)
-
-```{r, class priors, include=TRUE}
-epacific_priors <- classPriors(banter_model_RF, NULL)
-```
-
-Plot Confusion Matrix Heat Map
-
-```{r, confusion matrix heat map, include=TRUE}
-plotConfMat(banter_model_RF, title="ConfusionMatrix", plot=TRUE)
-```
-
-Plot Error Trace for Banter Model
-
-```{r, model error trace, include=TRUE}
-plot(banter_model_RF)
+png(('../manuscript/manuscript_files/epacific_env_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_env_RF)
+dev.off()
+env_epacific_votes <- plotVotes(banter_model_env_RF)
 ```
@@
 ```{r, predicted probabilities, include=TRUE}
-plotPredictedProbs(banter_model_RF, bins=30, plot=TRUE)
+plotPredictedProbs(banter_model_ec_RF, bins=30, plot=TRUE)
+plotPredictedProbs(banter_model_ici_RF, bins=30, plot=TRUE)
+plotPredictedProbs(banter_model_env_RF, bins=30, plot=TRUE)
 ```
+
+Create Figure for Publication
+
+```{r, Publication Figures, include=TRUE}
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+epacific_ec_Figure <-image_append(c(prox, heat, vote))
+epacific_ec_Figure<- image_append(c(confuse, epacific_ec_Figure), stack=TRUE)
+image_write(epacific_ec_Figure, path = here('manuscript', 'manuscript_files','epacific_ec_Figure.png'), format ='png')
+print(epacific_ec_Figure, info=FALSE)
+
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+epacific_ici_Figure <-image_append(c(prox, heat, vote))
+epacific_ici_Figure<- image_append(c(confuse, epacific_ici_Figure), stack=TRUE)
+image_write(epacific_ici_Figure, path = here('manuscript', 'manuscript_files','epacific_ici_Figure.png'), format ='png')
+print(epacific_ici_Figure, info=FALSE)
+
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+epacific_env_Figure <-image_append(c(prox, heat, vote))
+epacific_env_Figure<- image_append(c(confuse, epacific_env_Figure), stack=TRUE)
+image_write(epacific_env_Figure, path = here('manuscript', 'manuscript_files','epacific_env_Figure.png'), format ='png')
+print(epacific_env_Figure, info=FALSE)
+```

### 2023-06-08 hawaii/beakerBanter_hawaii.Rmd
Major revisions for draft
Revised all datasets and created draft manuscript

--- 2022-12-14 hawaii/beakerBanter_hawaii.Rmd
@@
 author: "Shannon Rankin"
-date: "2022-12-07"
+date: "2023-05-24"
 output: html_document
@@
 ```
+/Project
+  |-Database/
+  |-Binaries/
+  |-Figures/
+  |-hawaii
+    |_ beakerBanter_hawaii.Rmd
+    |_ hawaii_banter_data.rds
+    |_ hawaii_banter_data_ici.rds
+    |_ hawaii_banter_model_t5e3s10_t1e4s4.rds
+    |_ hawaii_banter_model_ici_t5e3s5_t1e4s4.rds
 
@@
 ```{r, load packages, message=FALSE}
-library(PAMpal)
-library(banter)
-library(rfPermute)
-library(here)
+library("easypackages")
+libraries("PAMpal", "banter", "rfPermute", "kableExtra", "magick", "magrittr", "here")
 here()
@@
 
-[If]{.underline} you have already run the processing code, ensure you
-have set 'freshRun = FALSE' at top of this document to read in the
-existing .rds file for downstream processing.
-
-```{r, read acoustic study, include=FALSE, eval=!freshRun}
-data <- readRDS('hawaii_study.rds')
-# Double check warning messages
-print(getWarnings(data)$message)
-```
-
 [If]{.underline} this is the [initial processing]{.underline}, ensure
@@
 
+[If]{.underline} you have already run the processing code, ensure you
+have set 'freshRun = FALSE' at top of this document to read in the
+existing .rds file for downstream processing.
+
+```{r, read acoustic study, include=FALSE, eval=!freshRun}
+data_ch1only <- readRDS('hawaii_study.rds')
+# Double check warning messages
+print(getWarnings(data_ch1only)$message)
+```
+
 3.  Assign species identity according to original PAMguard labels, then
@@
 
-```{r, assign species ID}
+```{r, assign species ID, eval=freshRun}
 data <- setSpecies(data, 'pamguard')
@@
 
-```{r, filter unwanted data}
+```{r, filter unwanted data, eval=freshRun}
 goodSpecies<- c("ZC", "MD", "BW", "BWC", "IP", "BW", "possBW")
@@
 
-```{r, calculate ICI}
+```{r, calculate ICI, eval=freshRun}
 data_ch1only <- calculateICI(data_ch1only, time='peakTime')
@@
 
-### *Model without ICI*
+### *EC (only) Model*
 
-Load data, initialize a BANTER model, and create an initial BANTER
-detector model (stage 1).
+Initialize, Run & Evaluate Detector Model (stage 1).
+```{r, banter ec detector model, eval=freshRun}
+banter_model_ec <- initBanterModel(banter_data$events)
+banter_model_ec <- addBanterDetector(banter_model_ec, banter_data$detectors, ntree=5e3, sampsize=10, importance = TRUE)
 
-```{r, read banter model, include=FALSE, eval=!freshRun}
-banter_model <- readRDS('hawaii_banter_model_t5e3s10_t10e3s4.rds')
-```
-
-```{r, banter detector model, eval=freshRun}
-banter_model <- initBanterModel(banter_data$events)
-banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=5e3, sampsize=10, importance = TRUE)
-```
-
-Check for stability (are more trees needed?) before moving on
-
-```{r, banter detector model stability}
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 0:2))
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 3:5))
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 6))
-summary(banter_model)
+plotDetectorTrace(banter_model_ec, detector = paste0('Click_Detector_', 0:3))
+plotDetectorTrace(banter_model_ec, detector = paste0('Click_Detector_', 4:6))
+summary(banter_model_ec)
 ```
@@
 Run BANTER Event Model (stage 2)
-
-```{r, banter event model, eval=freshRun}
-banter_model <- runBanterModel(banter_model, ntree=10e3, sampsize=4)
-summary(banter_model)
+```{r, banter ec event model, eval=freshRun}
+banter_model_ec <- runBanterModel(banter_model_ec, ntree=1e4, sampsize=4)
+summary(banter_model_ec)
 ```
 
-Check for stability (are more trees needed?). If acceptable, save model
-with tree/sampsize info in the filename.
-
-```{r,save banter model, eval=freshRun}
-saveRDS(banter_model, 'hawaii_banter_model_t5e3s10_t10e3s4.rds')
+Once a stable model is identified, save model with tree/sampsize info in the filename.
+```{r,save banter ec model, eval=freshRun}
+saveRDS(banter_model_ec, 'hawaii_banter_model_ec_t5e3s10_t1e4s4.rds')
 ```
@@
 
-Load data, initialize a BANTER model, and create an initial BANTER
-detector model (stage 1).
-
-```{r, read banter ici model, include=FALSE, eval=!freshRun}
-banter_model_ici <- readRDS('hawaii_banter_model_ici_t5e3s5_t10e3s4.rds')
-```
-
+Initialize, Run & Evaluate Detector Model (stage 1)
 ```{r, banter ici detector model, eval=freshRun}
@@
 banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=5e3, sampsize=5, importance = TRUE)
-```
 
-Check for stability to see if we need more trees
-
-```{r, banter ici detector model stability}
-plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 0:2))
-plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 3:5))
-plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 6))
+plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 0:3))
+plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 4:6))
 summary(banter_model_ici)
@@
 Run BANTER Event Model (stage 2)
-
 ```{r, banter ici event model, eval=freshRun}
-banter_model_ici <- runBanterModel(banter_model_ici, ntree=10e3, sampsize=4)
+banter_model_ici <- runBanterModel(banter_model_ici, ntree=1e4, sampsize=4)
 summary(banter_model_ici)
@@
 
-Check for stability (are more trees needed?). If acceptable, save model
-with tree/sampsize info in the filename.
-
+Once a stable model is identified, save model with tree/sampsize info in the filename.
 ```{r, save banter ici model, eval=freshRun}
-saveRDS(banter_model_ici, 'hawaii_banter_model_ici_t5e3s5_t10e3s4.rds')
+saveRDS(banter_model_ici, 'hawaii_banter_model_ici_t5e3s5_t1e4s4.rds')
 ```
@@
 
+```{r, read banter model, include=FALSE, eval=!freshRun}
+banter_model_ec <- readRDS('hawaii_banter_model_ec_t5e3s10_t1e4s4.rds')
+banter_model_ici <- readRDS('hawaii_banter_model_ici_t5e3s5_t1e4s4.rds')
+```
+
 First, identify the model you would like to examine (comment out the
@@
 ```{r, identify model}
-#model <- banter_model_ici
-model <- banter_model
+model_ec <- banter_model_ec
+modelname_ec <- "banter_model_ec"
+
+model_ici <- banter_model_ici
+modelname_ici <- "banter_model_ici"
 ```
@@
 ```{r, extract RF model}
-banter_model_RF <- getBanterModel(model)
+banter_model_ec_RF <- getBanterModel(model_ec)
+banter_model_ici_RF <- getBanterModel(model_ici)
+```
+
+Class Priors (Expected Error Rate)
+
+```{r, class priors, include=TRUE}
+hawaii_ec_priors <- classPriors(banter_model_ec_RF, NULL)[,1]
+hawaii_ici_priors <- classPriors(banter_model_ici_RF, NULL)[,1]
 ```
@@
 ```{r, confusion matrix, include=TRUE}
-hawaii_confuseMatrix <- confusionMatrix(banter_model_RF)
+hawaii_ec_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ec_RF)
+hawaii_ec_confuseMatrix <- cbind(hawaii_ec_confuseMatrix, priors = hawaii_ec_priors)
+hawaii_ec_confuseMatrix <- kable(hawaii_ec_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(5, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(5,hline_after = TRUE)%>%
+  row_spec(5, bold = TRUE)%>%
+  save_kable('../manuscript/manuscript_files/hawaii_ec_confuseMatrix.png', zoom = 9)
+
+hawaii_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
+hawaii_ici_confuseMatrix <- cbind(hawaii_ici_confuseMatrix, priors = hawaii_ici_priors)
+hawaii_ici_confuseMatrix <- kable(hawaii_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(5, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(5,hline_after = TRUE)%>%
+  row_spec(5, bold = TRUE)%>%
+  save_kable('../manuscript/manuscript_files/hawaii_ici_confuseMatrix.png', zoom = 9)
+
 ```
 
-Class Priors (Expected Error Rate)
+![BANTER Model Hawaii EC Confusion
+Matrix](../manuscript/manuscript_files/hawaii_ec_confuseMatrix.png)
 
-```{r, class priors, include=TRUE}
-hawaii_priors <- classPriors(banter_model_RF, NULL)
+![BANTER Model Hawaii ICI Confusion
+Matrix](../manuscript/manuscript_files/hawaii_ici_confuseMatrix.png)
+
+Proximity Plot
+
+```{r, proximity plot, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/hawaii_ec_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotProximity(banter_model_ec_RF)
+dev.off()
+ec_hawaii_proximityPlot <- plotProximity(banter_model_ec_RF)
+
+png(('../manuscript/manuscript_files/hawaii_ici_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
+dev.off()
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
 ```
 
-Plot Confusion Matrix Heat Map
+Importance Heatmap
 
-```{r, confusion matrix heat map, include=TRUE}
-plotConfMat(banter_model_RF, title="ConfusionMatrix", plot=TRUE)
-```
+```{r, importance heat map, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/hawaii_ec_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
+plotImportance(banter_model_ec_RF, plot.type="heatmap")
+dev.off()
+ec_hawaii_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap")
 
-Plot Error Trace for Banter Model
-
-```{r, model error trace, include=TRUE}
-plot(banter_model_RF)
+png(('../manuscript/manuscript_files/hawaii_ici_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
+plotImportance(banter_model_ici_RF, plot.type="heatmap")
+dev.off()
+ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
 ```
@@
 
-```{r, plot votes, include=TRUE}
-png("hawaii_votes.png", width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_RF)
+```{r, plot votes, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/hawaii_ec_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_ec_RF)
 dev.off()
-plotVotes(banter_model_RF)
-```
+hawaii_votes <- plotVotes(banter_model_ec_RF)
 
-Importance Heatmap
-
-```{r, importance heat map, include=TRUE}
-png("hawaii_importance.png", width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_RF, plot.type="heatmap")
+png(('../manuscript/manuscript_files/hawaii_ici_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_ici_RF)
 dev.off()
-plotImportance(banter_model_RF, plot.type="heatmap")
-```
-
-Proximity Plot
-
-```{r, proximity plot, include=TRUE}
-png("hawaii_proximity.png", width = 20, height = 20, units = 'cm',  res = 300)
-hawaii_proximityPlot <- plotProximity(banter_model_RF)
-dev.off()
-hawaii_proximityPlot <- plotProximity(banter_model_RF)
+ici_hawaii_votes <- plotVotes(banter_model_ici_RF)
 ```
@@
 
-```{r, predicted probabilities, include=TRUE}
-plotPredictedProbs(banter_model_RF, bins=30, plot=TRUE)
+```{r, predicted probabilities, include=TRUE, results=FALSE}
+plotPredictedProbs(banter_model_ec_RF, bins=30, plot=TRUE)
+plotPredictedProbs(banter_model_ici_RF, bins=30, plot=TRUE)
 ```
+
+Create Figure for Publication
+```{r, Publication Figure, include=TRUE}
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+hawaii_ec_Figure <-image_append(c(prox, heat, vote))
+hawaii_ec_Figure<- image_append(c(confuse, hawaii_ec_Figure), stack=TRUE)
+image_write(hawaii_ec_Figure, path = here('manuscript', 'manuscript_files','hawaii_ec_Figure.png'), format ='png')
+print(hawaii_ec_Figure, info=FALSE)
+
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+hawaii_ici_Figure <-image_append(c(prox, heat, vote))
+hawaii_ici_Figure<- image_append(c(confuse, hawaii_ici_Figure), stack=TRUE)
+image_write(hawaii_ici_Figure, path = here('manuscript', 'manuscript_files','hawaii_ici_Figure.png'), format ='png')
+print(hawaii_ici_Figure, info=FALSE)
+```

### 2023-06-08 manuscript/Supplement.qmd
Major revisions for draft
Revised all datasets and created draft manuscript

---
title: Supplement for Open-Source machine learning BANTER acoustic classification of a cryptic echolocating species
author: Shannon Rankin
date: last-modified
output: pdf
---

```{r, load packages, include=FALSE}
library("easypackages")
libraries("tidyverse", "knitr", "kableExtra", "magick", "magrittr", "here")
here()
```

# NAtlantic
## EC Model

Detector Model:  ntree=1e4, sampsize=3; 
Event Model: ntree=1e4, sampsize=3
```{r,  NAtlantic EC Model Figure, echo=FALSE}
confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_confuseMatrix.png'))%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_votes.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_proximity.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_importance.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("d) Importance Heat Map", size=100, color = "black")

natlantic_ec_Figure <-image_append(c(prox, heat, vote))
natlantic_ec_Figure<- image_append(c(confuse, natlantic_ec_Figure), stack=TRUE)
print(natlantic_ec_Figure, info=FALSE)
```

## EC_ICI_ALT Model
Detector Model:  ntree=1e4, sampsize=5; 
Event Model: ntree=1e4, sampsize=5
```{r, NAtlantic EC_ICI_ALT Model Figure,  echo=FALSE}
confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_confuseMatrix.png'))%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_votes.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_proximity.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_importance.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("d) Importance Heat Map", size=100, color = "black")

natlantic_alt_Figure <-image_append(c(prox, heat, vote))
natlantic_alt_Figure<- image_append(c(confuse, natlantic_alt_Figure), stack=TRUE)
print(natlantic_alt_Figure, info=FALSE)
```

# SAtlantic
## EC Model
Detector Model:  ntree=1e4, sampsize=2; 
Event Model: ntree=1e5, sampsize=1
```{r, SAtlantic EC Model Figure,  echo=FALSE}
confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_confuseMatrix.png'))%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_votes.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_proximity.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_importance.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("d) Importance Heat Map", size=100, color = "black")

satlantic_ec_Figure <-image_append(c(prox, heat, vote))
satlantic_ec_Figure<- image_append(c(confuse, satlantic_ec_Figure), stack=TRUE)
print(satlantic_ec_Figure, info=FALSE)
```

## EC_ICI_ALT Model
Detector Model:  ntree=1e4, sampsize=4; 
Event Model: ntree=1e5, sampsize=2
```{r, SAtlantic EC_ICI_ALT Model Figure,  echo=FALSE}
confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_confuseMatrix.png'))%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_votes.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_proximity.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_importance.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("d) Importance Heat Map", size=100, color = "black")

satlantic_alt_Figure <-image_append(c(prox, heat, vote))
satlantic_alt_Figure<- image_append(c(confuse, satlantic_alt_Figure), stack=TRUE)
print(satlantic_alt_Figure, info=FALSE)
```

# Hawaii
## EC_ICI Model
Detector Model:  ntree=5e3, sampsize=5; 
Event Model: ntree=1e4, sampsize=4
```{r, Hawaii EC_ICI Model Figure,  echo=FALSE}
confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_confuseMatrix.png'))%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_votes.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_proximity.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_importance.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("d) Importance Heat Map", size=100, color = "black")

hawaii_ici_Figure <-image_append(c(prox, heat, vote))
hawaii_ici_Figure<- image_append(c(confuse, hawaii_ici_Figure), stack=TRUE)
print(hawaii_ici_Figure, info=FALSE)
```

# EPacific
## EC Model
Detector Model:  ntree=1e4, sampsize=3; 
Event Model: ntree=1e4, sampsize=4

```{r, EPacific EC Model Figure,  echo=FALSE}
confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_confuseMatrix.png'))%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_votes.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_proximity.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_importance.png'))%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("d) Importance Heat Map", size=100, color = "black")

epacific_ec_Figure <-image_append(c(prox, heat, vote))
epacific_ec_Figure<- image_append(c(confuse, epacific_ec_Figure), stack=TRUE)
print(epacific_ec_Figure, info=FALSE)
```

### 2023-06-08 manuscript/manuscript.qmd
Major revisions for draft
Revised all datasets and created draft manuscript

--- 2023-02-07 manuscript/manuscript.qmd
@@
           postal-code: 92037
-    note: |
-      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
   - name: Frederick Archer
@@
           postal-code: 92037
-    note: Yet another author footnote.
   - name: Jay Barlow
@@
           postal-code: 96818
+  - name: Anne Simonis
+    email: anne.simonis@noaa.gov
+    affiliations:
+        - id: SWFSC
+          name: Southwest Fisheries Science Center
+          department: Marine Mammal & Turtle Division
+          address: 8901 La Jolla Shores Dr.
+          city: La Jolla
+          state: CA
+          postal-code: 92037
   - name: Melissa Soldevilla
@@
           postal-code: 33143
+  - name: Jenny Trickey
+    email: jenny.trickey@noaa.gov
+    affilations:
+        - id: PIFSC
+          name: Pacific Islands Fisheries Science Center
+          department: 
+          address: 1845 Wasp Blvd
+          city: Honolulu
+          state: HI
+          postal-code: 96818
 abstract: |
@@
 Data from four large scale surveys of beaked whales (North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and BANTER (hierarchical random forest classifier).
-Overall correct classification results ranged from 86% for the South Atlantic data to 99% for the North Atlantic.
-Results for many species could likely be improved with increased sample sizes and consideration of alternative automated detectors.
+Overall correct classification results ranged from 87% for the South Atlantic data to 98% for the North Atlantic.
+Results for many species could likely be improved with increased sample sizes, consideration of alternative automated detectors, and addition of relevant environmental features.
 These methods provide a highly automated approach to acoustic detection and classification using open source methods that can be readily adopted for species and geographic regions.
@@
 date: last-modified
-bibliography: bibliography.bib
+bibliography: ../manuscript/bibliography.bib
+link-citations: TRUE
 format:
@@
 ---
-```{r, hide=TRUE}
-library(tidyverse)
-library(knitr)
-library(here)
+
+```{r, load packages, include=FALSE}
+library("easypackages")
+libraries("tidyverse", "knitr", "kableExtra", "magick", "magrittr", "here")
+here()
 ```
+
 # Introduction
 
-Passive acoustic monitoring (PAM) from has proven to be a valuable tool for studying populations of marine mammals [@parijs2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring.
-Historically, experienced acousticians manually identify stereotyped sounds that could be reliably attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering2013]; [@bittle2013]; [@rankin2005]; [@soldevilla2008].
+Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring.
+Historically, experienced acousticians manually identify stereotyped sounds that could be reliably attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008].
 The dramatic increase in recordings make it impossible for experienced acousticians to manually annotate all data, and there has been an increasing need for the development of automated classification routines that can provide accurate determinations of the species responsible for the sounds.
 
-Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods ([@macleod2018]).
-However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering2013a].
-Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow2022]; [@baumann-pickering2014]; [@simonis2020].
+Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod_beaked_2018].
+However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_spatio-temporal_2014].
+Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022]; [@baumann-pickering_spatio-temporal_2014]; [@simonis_passive_2020].
 Unfortunately, manual classification of beaked whale echolocation clicks requires analysis by trained observers using a number of visual aids to examine the call characteristics.
@@
 
-BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings [@rankin2017].
-This hierarchical random forest event classifier relies on an initial call classifier (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier.
+BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings [@rankin_acoustic_2017].
+This hierarchical random forest event classifier relies on an initial call classifier for each call type (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier.
 The event classifier considers the probability of assignment to species for each call in the call classifier, along with any event-level characteristics (such as call rate).
 BANTER is very flexible and can accommodate any number of measures from any number of detectors.
-While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results [@rankin2017a].
+While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results [@rankin_acoustic_2017].
 In this case, settings for a whistle & moan detector were modified to improve performance at detecting burst pulses.
 This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the overall number of detections was different.
-This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn2021].
+This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn_acoustic_2021].
 For species who primarily (or exclusively) produce echolocation clicks, BANTER may serve as an option for automated machine learning classification.
 
-Here we applied BANTER acoustic classification to beaked whale detections from four large acoustic datasets off the US East Coast, Hawaii, and the US West Coast.
+Here we applied BANTER acoustic classification to beaked whale detections from four large acoustic datasets including the US East Coast (north and south), Hawaii, and the US West Coast.
 Acoustic data were analyzed by experienced acousticians to determine species identity, and these 'ground truth' classifications served as training data for the supervised BANTER trials.
@@
 AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey area.
-The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (HTI-96-min, High Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see [@deangelis2018b] for more information).
+The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (HTI-96-min, High Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see [@deangelis_description_2018] for more information).
 The SAtlantic area ranged from Delaware south to central Florida (GU1605, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (Reson \[MODEL\], Teledyne Marine, Slangerup, Denmark) recorded 500 kHz sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass filter (see \[CITE\] for more information).
+Variation in hydrophone sensitivities precluded combining the NATlantic and SAtlantic datasets.
 
-Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, [@yano2018]).  Drifting recording buoys were deployed at randomly selected locations within the main Hawaiian Islands.
+Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, [@yano_cetacean_2018]).  Drifting acoustic recorders were deployed at randomly selected locations within the main Hawaiian Islands.
 Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA).
@@
 SM3M data were sampled at 256 kHz with continuous recordings.
-Individual buoys were deployed for 10 to 50 days (see [@yano2018]for more detailed information).
+Individual buoys were deployed for 10 to 50 days (see [@yano_cetacean_2018] for more detailed information).
 
 Recordings from the Eastern North Pacific Ocean off the U.S.
-west coast (EPacific) were collected using drifting acoustic recorders during the 2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels) survey [@keating2018].
-Buoys were deployed at a number of stations in the California Current, including pre-determined locations (for abundance estimation studies) and in close proximity to seamounts (Ad Hoc Seamount Experiment).
+west coast (EPacific) were collected using drifting acoustic recorders during the 2018 CCES (California Current Ecosystem Survey) survey [@simonis_passive_2020].
+Buoys were deployed at 23 stations distributed throughout the California Current region; data from 15 of these included high quality acoustic data used in this analysis.
 Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA).
-Recordings from pre-determined locations were duty cycled and recorded 2 minutes out of a variable 'off' time; recordings from the Seamount Experiment were continuous.
-Soundtrap recorders were sampled at 288 kHz and the SM3M recorders were recorded at 256 sample rate (with some sampling at 96 kHz for ocean noise measurements).
-Individual buoys were deployed for 11-23 days (see [@keating2018a] for more detailed information).
+Recordings were duty cycled and recorded 2 minutes out of a variable 'off' time; recorders sampled at a minimum 256 sample rate.
+Individual buoys were deployed for 5-79 days (see [@simonis_passive_2020] for more detailed information).
 
@@
 
-All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c) with a suite of generic click classifiers within the Click Detector module (see [@keating2013]).  Classifier sets were saved such that any click may be classified as more than one click type.
-The general click classifiers were effectively click detectors, as each implemented a simple spectral band settings (e.g., 2- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz).
-An additional detector/classifier within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see [@keating2013a]).
+All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c, 2.00.16e) with a suite of generic click classifiers within the Click Detector module (see [@keating_summary_2013]).  Classifier sets were saved such that any click may be classified as more than one click type.
+The general click classifiers were effectively click detectors, as each implemented a simple spectral band settings (e.g., 2 -- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz).
+An additional detector within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see [@keating_summary_2013]).
 
@@
 Acoustic researchers with experience in classifying beaked whale echolocation clicks examined the bearing-time display for the click detector module in PAMGuard to assess the species identity of the acoustic event.
-For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see [@deangelis2018]).
-Data were saved within the PAMGuard database and binaries for downstream processing using PAMpal package [@sakai2021] in R programming language [@rcoreteam2022].
+For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see [@deangelis_description_2018]).
+Data were saved within the PAMGuard database and binaries for downstream processing using PAMpal package [@sakai_pampal_2021] in R programming language [@r_core_team_r_2022].
 
 Standard click calculations were measured from echolocation clicks using default values in PAMpal.
-For EPacific data, any events \> 2 minutes were subdivided into 2-minute events to ensure that all events were of equal value (because most of our data was duty cycled into 2-minute blocks).
 Inter-click interval (ICI) was calculated using the 'calculateICI' function in PAMpal; the mode of the ICI was identified for each beaked whale acoustic event.
+In addition, for EPacific data, we used the ‘matchEnvData’ function in PAMpal to include ERDAPP environmental data (@simons_erddap_2022), including seafloor depth and seafloor gradient (both from erdSrtm30plusSeafloorGradient), as well as sea surface temperature (from jplMURSST41).
 
@@
 Data subsets for each trial were exported for BANTER classification using the export_banter function in PAMpal.
-At a minimum, each trial included standard click calculations for all clicks in a beaked whale acoustic event (EC); and some trials included the mode of the ICI for each event (EC_ICI).
+Trials included standard click calculations for all clicks in a beaked whale acoustic event (EC); the addition of  ICI for each event (EC_ICI), and for the EPacific dataset we also included environmental variables (EC_ICI_ENV).
 
@@
 
-BANTER models was created for each trial using the BANTER package in R programming language [@archer2022].
-Standard click calculations were used for the BANTER call classifier; event level measures of ICI were applied to the BANTER event classifier.
-A minimum of two events per species and complete suite of variables (absence of NAs) were required for training and testing of the classification model.
+BANTER models were created for each trial using the BANTER package in R programming language [@archer_banter_2022].
+Standard click calculations were used for the BANTER call classifier; event level measures of ICI and environmental variables were applied to the BANTER event classifier.
+A minimum of two events per species and the complete suite of variables (absence of NAs) were required for training and testing of the classification model.
 The BANTER random forest models contain two parameters, ntree and sampsize, which must be selected to ensure stability of the model.
 The sample size (sampsize) is the number of events considered for each trial of the model, and the ntree is the number of trees (iterations of the model) run in a given model.
-These values were modified to obtain the best results, with an emphasis on lower sample size and higher number of trees for computational performance [@rankin2021].
+These values were modified to obtain the best results, with an emphasis on lower sample size and higher number of trees for computational performance [@rankin_banter_2021].
 
-The model output include the confusion matrix, and a number of visualizations provided by the rfpermute package in R programming language [@archer2022a].
+The model output include the confusion matrix, and a number of visualizations provided by the rfpermute package in R programming language [@archer_rfpermute_2022].
 
@@
 
-Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study; species included: Baird's beaked whales (*Berardius bairdii*), Blainsville's beaked whales (*Mesoplodon densirostris*), Cuvier's beaked whales (*Ziphius cavirostris*), Cross Seamount beaked whale (McDonald et al., 2009), Gervais' beaked whales (*Mesoplodon europaeus*), Longman's beaked whale (*Indopacetus pacificus*), Sowerby's beaked whales (*Mesoplodon bidens*), True's beaked whales (*Mesoplodon mirus*), Stejneger's beaked whale (*Mesoplodon stejnegeri*), and unidentified beaked whale 'BW43' [@baumann-pickering2013].
+Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study; species included: Baird's beaked whales (*Berardius bairdii*), Blainsville's beaked whales (*Mesoplodon densirostris*), Cross Seamount beaked whale (@mcdonald_acoustic_2009), Cuvier's beaked whales (*Ziphius cavirostris*),  Gervais' beaked whales (*Mesoplodon europaeus*), Hubb's beaked whale (*Mesoplodon carlhubbsi*), Longman's beaked whale (*Indopacetus pacificus*), Sowerby's beaked whales (*Mesoplodon bidens*), Stejneger's beaked whale (*Mesoplodon stejnegeri*), True's beaked whales (*Mesoplodon mirus*), and unidentified beaked whale 'BW43' [@baumann-pickering_species-specific_2013].
 Analyses were conducted separately for the different study areas due to differences in the data collection methods that precluded combining of datasets.
@@
 
-Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n=70), Sowerby's beaked whales (n=5), True's beaked whales (n=65), and Gervais' beaked whales (n=3).
+Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n = 120), Gervais' beaked whales (n = 4), Sowerby's beaked whales (n = 6), and True's beaked whales (n = 76).
 BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).
 
-For the NAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (EC_ICI, Detector Model: sampsize = 4, ntrees = 30,000; BANTER Model: sampsize = 4, trees=300,000) and provided an overall correct classification rate of 100% for all four species (Fig.
-1).
-There were insufficient samples of Gervais' beaked whales to include in the NAtlantic model (n=2).
-All classification results were greater than the expected error rate (Prior in Fig.
-1a).
-Results from the EC model are presented here; results from the EC_ICI model can be found in supplementary materials (Supplement Fig.
+For the NAtlantic dataset, the model with ICI had a modest increase in accuracy over the EC (only) model (EC_ICI model, Detector Model: sampsize = 3, ntrees = 10,000; Event Model: sampsize = 3, trees = 100,000).
+This model provided an overall correct classification rate of 98% for all four species (pct.correct in Confusion Matrix, @fig-natlantic a).
+Classification scores ranged from 75% for Gervais' beaked whales to 100% for Cuvier's and Sowerby's beaked whales.
+All classification results were greater than the expected error rate (Priors in @fig-natlantic a).
+Results from the EC_ICI model are presented here; results from the EC model were similar and can be found in supplementary materials (Supplement Fig.
 1).
 
-\[Insert Figure 1\]\[Figure 1.
-BANTER classification results from the NAtlantic dataset (EC).
-Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
-Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+The proximity plot (@fig-natlantic b) provides a view of the distribution of events within the feature space of the two 'most important' dimensions, or predictors.
+For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result.
+The importance heat map (@fig-natlantic c) shows the most important features for predicting each species.
+The strength of the classifications by species can be seen in the plot votes (@fig-natlantic d), which show the distribution of the votes for each of the 100,000 trials for each event.
+For each plot, the events are represented as vertical slices (with plot subdivided along the x-axes according to the event sample size), and the percentage of votes for each species is represented by their color along that vertical slice.
+In a perfect scenario, all 100,000 trees would have voted for the correct species and therefore the classification would be correct and the strength of these classifications would be maximized.
 
-The proximity plot (Fig.1b) provides a view of the distribution of events within the feature space of the two 'most important' dimensions, or predictors.
-For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result.
-The importance heat map (Fig.
-1c) shows that the most important features for predicting each species.
-
-The strength of the classifications by species can be seen in the plot votes (Fig.
-1d), which show the istribution of the votes for each of the 300,000 trials.
-For each plot, the events are represented as vertical slices (with plot subdivided along the x-axes according to the event sample size), and the percentage of votes for each species is represented by their color along that vertical slice.
-In a perfect scenario, all 300,000 trees would have voted for the correct species and therefore the classification would be correct AND the strength of these classifications would be maximized.
-The plot vote for the NAtlantic dataset show that all species
+The proximity plot (@fig-natlantic b) shows overlap between Gervais', Sowerby's and True's beaked whales in this two dimension feature space and the species classifications are resolved using additional features, as shown in the importance heat map (@fig-natlantic c).
+The plot vote figure (@fig-natlantic d) shows strong classification strength for Cuvier's and Sowerby's beaked whales, with increased variability in voting distributions for Gervais' and True's beaked whales.
 
@@
 
-Species encountered in the SAtlantic study dataset include Blainsville's beaked whales (n=35), Cuvier's beaked whales (n=66), Gervais' beaked whales (n=45), Sowerby's beaked whales (n=2), and True's beaked whales (n-12).
+Species encountered in the SAtlantic study dataset include Blainsville's beaked whales (n = 46), Cuvier's beaked whales (n = 111), Gervais' beaked whales (n = 77), Sowerby's beaked whales (n = 2), and True's beaked whales (n = 13).
 BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).
 
-For the SAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (Detector Model: sampsize = 2, ntrees = 30,000; BANTER Model: sampsize = 1, ntrees = 300,000) and provided an overall correct classification rate of 86.9% for all four species.
+For the SAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (EC_ICI Model, Detector Model: sampsize = 4, ntrees = 10,000; Event Model: sampsize = 1, ntrees = 100,000) and provided an overall correct classification rate of 87.1% for all five species.
+Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-satlantic); all classification results were greater than the expected error rate (Priors in @fig-natlantic a).
 The sample sizes for the BANTER model was kept low (sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2.
-Analysis with a larger sample size resulted in improved classification results for Cuvier's beaked whale (see Supplement Fig.
+Analysis with a larger sample size resulted in improved classification results for Cuvier's and Gervais' beaked whale (see Supplement Fig.
 3 for EC_ICI_ALT), but resulted in the loss of Sowerby's in the final model.
-Classification scores ranged from a low of 74.2% (Cuvier's beaked whale) to 100% (Sowerby's and True's beaked whale, Fig.
-2); all classification results were greater than the expected error rate (Prior in Fig.
-2a).
-Results from the EC_ICI model are presented here; results from the EC model can be found in supplementary materials (Supplement Fig.
+Results from the EC_ICI model are presented here; results from the EC and EC_ICI_ALT model can be found in supplementary materials (Supplement Fig.
 2).
 
-Figure 2.
-BANTER classification results from the SAtlantic dataset.
-Confusion matrix provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).
-Proximity plot (a) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (b) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (c) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
-
-For SAtlantic, the proximity plot (Fig.
-2b) shows that True's beaked whales are differentiated from the other species based solely on the two most important features.
-The importance heat map (Fig.
-2c) shows that Cuvier's, Gervais' and Sowerby's beaked whales rely on a number of predictors that are lower in importance and there is likely greater distinction between these species than is represented by the proximity plot (Fig.
-2b).
-The plot votes (Fig.
-2d) shows a moderate number of trees in the forest were misclassified and these misclassifications included all species in the model.
+For SAtlantic, the two most important features are insufficient to differentiate species, as seen by the overlap in species distribution with these feature spaces in the proximity plot (@fig-satlantic b).
+Importance variables vary by species (importance heat map @fig-satlantic c), with ICI measurements having relatively low importance for Blainsville's and Sowerby's beaked whales (@fig-satlantic c).
+The plot votes graph (@fig-satlantic d) shows relatively weak classification results; most events consisted of \<50% of the trees voting for the correct species.
 
 ## Hawaii
-The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (Yano et al.
-2018); species included Longman's beaked whale (n=121), Blainsville's beaked whale (n=518), Cross Seamount beaked whale (n=84) and Cuvier's beaked whale (n=201).
+
+The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (@yano_cetacean_2018); species included Blainsville's beaked whale (n = 521), Cross Seamount beaked whale (n = 76), Cuvier's beaked whale (n = 201), and Longman's beaked whale (n = 122).
 BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).
 
-The most accurate model considered the echolocation clicks only (Detector Model: sampsize = 10, ntrees = 5,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 91.8% for all four species.
-Classification scores ranged from a low of 84.5% (Cross Seamount Beaked Whale) to 97.5% (Longman's beaked whale, Fig.
-3).
- All classification results were greater than the expected error rate (Prior in Fig.
-3a).
+The most accurate model considered the echolocation clicks only (EC Model, Detector Model: sampsize = 10, ntrees = 5,000; Event model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.3% for all four species.
+Classification scores ranged from a low of 86% (Cuvier's Beaked Whale) to 97.5% (Longman's beaked whale, @fig-hawaii).
+All classification results were greater than the expected error rate (Prior in @fig-hawaii a).
 Results from the EC model are presented here; results from EC_ICI model can be found in supplementary materials (Supplement Fig.
@@
 
-::: {layout="[[100], [30,30,30]]"}
-a)  ![Fig.1](../hawaii/banter_model_hawaii_confuseMatrix.png){#fig-a}
-
-b)  ![Fig.1](../hawaii/banter_model_hawaii_proximity.png){#fig-b} :::
-
-c)  ![Fig.1](../hawaii/banter_model_hawaii_importance.png)
-
-d)  ![Fig.1](../hawaii/banter_model_hawaii_votes.png)
-:::
-
-\[Insert Figure 3\]\[Figure 3.
-BANTER classification results from the Hawaii dataset.
-Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).
-Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
-Species codes include Cross Seamount beaked whale (BWC), Longman's beaked whale (IP), Blainsville's beaked whale (MD) and Cuvier's beaked whale (ZC).
-
-There are clear distinctions in the primary features used to classify three of the four species as shown in the Proximity Plot (Fig.
-3b) which shows the classification results based on the two most 'important' predictors.
-Distinctive clusters based on the two most important features can be found for Longman's (IP), Blainsville's (MD), and Cuvier's beaked whales (ZC), and there is overlap in the feature space of the clusters for Cuvier's and Cross Seamount beaked whales (BWC).
-This overlap is resolved with different features for ZC (3^rd^, 7^th^ in Fig.
-3c heat map) and BWC (5^th^, 8^th^ feature in heat map), as evidence in the low misclassification rates for these species (Fig.
-3a confusion matrix).
-The plot votes graph (Fig.
-3d) show complicated results with variation in the proportion of trees that correctly classified the species for each event.
+Distinctive clusters based on the two most important dimensions can be found for Longman's and Blainsville's beaked whales, and there is overlap in this feature space of the clusters for Cuvier's and Cross Seamount beaked whales as shown in the Proximity Plot (@fig-hawaii b).
+This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the Importance Heat Map (@fig-hawaii c), resulting in the low misclassification rates for these species (@fig-hawaii a confusion matrix).
+The plot votes graph (@fig-hawaii d) show strong classification results for most Blainsville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.
 
@@
 
-The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States ([@keating2018]); species included Baird's beaked whale (n=91), Unidentified Beaked Whale 'BW43' (n=11), Stejneger's beaked whale (n=41) and Cuvier's beaked whale (n=1,332).
-BANTER classification trials included (1) echolocation clicks (EC) and (2) echolocation clicks and ICI (EC_ICI).
+The EPacific dataset consisted of 15 drifting recording buoys deployed off the west coast of the United States [@simonis_passive_2020]; species included Baird's beaked whale (n = 29), Unidentified Beaked Whale 'BW43' (n = 125), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 926), Hubbs beaked whale (n = 66) and Stejneger's beaked whale (n = 42).
+BANTER classification trials included (1) echolocation clicks (EC), (2) echolocation clicks and ICI (EC_ICI), and (3) echolocation clicks, ICI, and environmental features (EC_ICI_ENV).
 
-The best model considered the echolocation clicks and the ICI values (Detector Model: sampsize = 5, ntrees = 30,000; BANTER model: sampsize = 4, ntree = 200,000) and provided an overall correct classification rate of 93.7% for all four species.
-Classification scores ranged from a low of 93.2% (Cuvier's beaked whale) to 100% (BW43, Fig.
-4a).
-All classification results were greater than the expected error rate (Prior in Fig.
-4a).
-Results from the EC_ICI model are presented here; complete results can be found in supplementary materials (Supplement Figs.
-5-7).
+The best model considered the echolocation clicks, ICI, and environmental features (EC_ICI_ENV Model, Detector Model: sampsize = 3, ntrees = 10,000; Event model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.2% for all six species.
+Classification scores ranged from a low of 91.2% (BW43) to 100% for Baird's beaked whales(@fig-epacific a).
+All classification results were greater than the expected error rate (Prior in @fig-epacific a).
+Results from the EC_ICI_ENV model are presented here; results from the EC and EC_ICI model can be found in supplementary materials (Supplement Figs.
+5).
 
-\[Insert Figure 4\]\[Figure 4.
-BANTER classification results from the EPacific dataset.
-Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).  Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
-Species codes include Baird's beaked whale (BB), unidentified beaked whale (BW43), Stejneger's beaked whale (MS) and Cuvier's beaked whale (ZC).\]
-
-There are clear distinctions in the primary features used to classify the species as shown in the Proximity Plot (Fig.
-4b) which shows the classification results based on the two most 'important' predictors.
-The species each form distinctive clusters based on these two features; however, many events fall outside these distinct bounds (detections shown in area between species classification clusters).
-Both BB and BW43 relied on a wide range of variables for classification (Fig.
-4c heat map).
-The plot votes (Fig.
-4d) show strong classification results for most events of most species.
+There is considerable overlap in the primary feature space for all but BW43, as shown in the Proximity Plot (@fig-epacific b).
+ICI was among the most important classification variable for all species (@fig-epacific c heat map).
+The plot votes (@fig-epacific d) show relatively strong classification results Baird's, Hubbs, and Stejneger's beaked whales.
 
@@
 
-The results for the four study areas varied from a low of 86.9% correct classification in the SAtlantic dataset to a high of 100% correct classification in the NAtlantic dataset.
-All species in all datasets had classifications rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), BWC (Hawaii), and Blainsville's beaked whales (Hawaii) had classification rates below 90%.
-There were insufficient sample sizes of Gervais to include in the NAtlantic dataset.
+The results for the four study areas varied from a low of 87.1% correct classification in the SAtlantic dataset to a high of 98% correct classification in the NAtlantic dataset.
+All species in all datasets had classifications rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais beaked whales (NAtlantic) had classification rates below 90%.
 
 Several of our datasets included very low sample sizes for some species; however, low sample sizes did not always result in low classification scores.
-In the Atlantic dataset, Sowerby's beaked whale was consistently correctly classified in both models (NAtlantic and SAtlantic); however, while the plot votes showed very strong classification strength in the NAtlantic, the classification strength (represented by the number of trees that voted for the correct classification) was poor in the SAtlantic study area.This is likely due to both a small sample size of Sowerby's detected in the S.
-study area (n=2) combined with the small sampsize used in the BANTER model.
+In the NAtlantic dataset, Sowerby's beaked whale (n=6) was correctly classified in 100% of the samples, and classification results were very strong.
+The Sowerby's beaked whale events in the SAtlantic study and the Gervais beaked whale events in the NAtlantic study consisted of weak classifications.
+This is likely due to both a small number of events (n = 2 for Sowerby's in SAtlantic and n = 4 for Gervais in NAtlantic) combined with the small sampsize used in the BANTER model.
 In general, larger sample sizes should lead to improved overall classification results, and improved strength of these classification results (as identified by plot vote graph).
@@
  For species with high variability in the predictor variables and an overlap in the range of these variables with other species, a large sample size is required to describe the true variability of these call measurements.
-Even with reasonably large sample sizes, the classification may suffer (e.g, SAtlantic Cuvier's beaked whale = 74%).
-We found by increasing the sampsize in BANTER, we could increase the classification rate for this species in this area to (97%, see Supplement Fig.3); however, this resulted in the loss of Sowerby's beaked whale in the final model.
+Even with reasonably large sample sizes, the classification may suffer (e.g, SAtlantic Cuvier's beaked whale = 79.2% correct classification).
+We found by increasing the sampsize in BANTER, we could increase the classification rate for this species in this area to (91.8%, see Supplement Fig.3); however, this resulted in the loss of Sowerby's beaked whale in the final model.
 So, an increased sample size in Sowerby's would likely lead to increased classification results for other species in the model.
 
-Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (long-beaked and short-beaked common dolphins, [@rankin2017b]).
-This suggests that large increases in sample size may improve classification results for Cuvier's beaked whales.
+Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (long-beaked and short-beaked common dolphins, [@rankin_acoustic_2017]).
+This suggests that large increases in sample sizes may improve classification results for Cuvier's beaked whales.
 
@@
 Future work should examine if these Cuvier's beaked whale events can be clustered in an objective and stable manner; if so, then these can be input into BANTER models as separate types of Cuvier's beaked whales.
-We recommend caution in this approach- this should only be considered if there is reasonable evidence that these different groups represent authentic cluster.
+We recommend caution in this approach- this should only be considered if there is reasonable evidence that these different groups represent authentic acoustic groups.
 We do not recommend segregation of misclassified subgroups simply to improve classification scores.
 
-Gervais' and True's beaked whales have similar click characteristics that make them difficult to differentiate, and require considerable expertise to classify manually, using a variety of available visuals and metrics [@deangelis2018a].
+Gervais' and True's beaked whales have similar click characteristics that make them difficult to differentiate and they require considerable expertise to classify manually [@deangelis_description_2018].
 Our results suggest that BANTER may serve as an efficient and effective means of classifying and differentiating these two species.
-Despite modest sample sizes, Gervais' and True's beaked whales showed high classification scores (95.6% and 100%, respectively).
+Despite modest sample sizes, Gervais' and True's beaked whales showed high classification scores in the SAtlantic (93.5% and 100%, respectively), and True's performed well in the NAtlantic.
 Unfortunately, we were unable to combine the N and S Atlantic datasets (which would increase sample sizes of True's beaked whales) due to differences in hydrophone characteristics that result in differences in call measures.
@@
 
-For the SAtlantic models, ICI was found to be an important predictor for many species \[*talk about the 1 change in sowerby's when the ICI was added- do you think it was because of an error in ICI?*\] .
-For Hawaii and the EPacific, however, we found that adding ICI into the BANTER classifier did not improve final classification results.
-The differences here are likely due to different analysis methods for identifying events.
-Beaked whale events were defined by encounter time for both the Hawaii and EPacific data, which were collected using drifting acoustic recorders where animals could not be localized.
-The Atlantic data were collected using a towed hydrophone array with localization capabilities.
-Each event was defined as an individual click train, so ICI for these events more likely represented true ICI, and was found to be more informative.
-[@baumann-pickering2013] showed an overlap in the ICI between Cuvier's and Blainsville's beaked whales, but a strong difference in ICI between these species and BWC.
-Likewise, [@baumann-pickering2013] found a difference in ICI between Baird's beaked whales and Cuvier's beaked whales.
-Subdividing click trains in PAMGuard is time consuming, but alternative options include consideration of the PAMGuard click train detector module in BANTER, or by developing a similar function in R to apply to events.
-Integration of ICI from individual click trains may serve to improve classification results for some species in Hawaii and EPacific.
-
-Gervais were eliminated from the NAtlantic BANTER model due to low sample size (n=2); results were extremely poor when they were included in the model.
-Analysis notes indicate uncertainty in manual species classification on one of the two events in this dataset.
+In addition to poor classification of Gervais in the NAtlantic study area, the plot votes show that for one of the four events, only a few of the 100,000 trees 'voted' for Gervais, and the majority of the votes were for True's beaked whales.
+Analysis notes indicate uncertainty in manual species classification of one of the events in this dataset.
 While small sample size can at times provide good classification scores, as we found in Sowerby's beaked whale in this SAtlantic dataset, they can be devastating if there are inaccuracies in the training data.
@@
 In the case of beaked whales, we recommend that training data include a reasonable level of confidence for inclusion.
-When sample sizes are low, such as in these cases, it may be wise to consider the strength of the classifications when including them for model training (as we have done here).
+This is especially critical for species with small sample sizes.
 An alternative is to require agreement from multiple analysts, as provided by the EPacific datasets.
 This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified.
-For example, (cite jays new paper) recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (including the EPacific data presented here).
+For example, [@barlow_unique_2022] recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (including the EPacific data presented here).
+
+Interclick Interval (ICI) was found to be the most important variable for NAtlantic and EPacific, and ICI were the third and fourth ranked variable for SAtlantic (see Importance Heat Maps for each survey area).
+While the addition of ICI did not improve results Hawaii, results were similar and ICI ranked #9 in importance for the Hawaii study area.
+There was a difference in how ICI was calculated for these surveys, and it is possible these differences may affect the value of the ICI variable in species classification for some species.
+Beaked whale events were defined by encounter time for both the Hawaii and EPacific data, which were collected using drifting acoustic recorders where animals could not be localized.
+The Atlantic data were collected using a towed hydrophone array with localization capabilities.
+Each event was defined as an individual click train, so ICI for these events more likely represented true ICI.
+[@baumann-pickering_species-specific_2013] showed an overlap in the ICI between Cuvier's and Blainsville's beaked whales, but a strong difference in ICI between these species and Cross Seamount beaked whales.
+It is possible that classification of these species could be improved by refining an event to consider ICI from individual click trains.
+Subdividing click trains in PAMGuard is time consuming (and complicated in the absence of localization), but alternative options include consideration of the PAMGuard click train detector module in BANTER, or by developing a similar function in R to apply to events.
+
+For this study, only a few simple environmental features were considered for one of the study areas (seafloor depth, seafloor gradient, and sea surface temperature).
+In the EPacific study area, inclusion of these variables increased the overall classification rate from 89.2% to 92.2%.
+These variables had the most impact on classification results for the Cross Seamount beaked whale (n = 6), raising the classification scores to from 83.3% to 100%.
+Alternative environmental features may improve results for other species.
 
@@
 Training data should consist of labels with strong confidence in species identify (ideally determined by agreement from more than one analyst) and sample sizes should be large enough to explain the natural variability in the data.
-In fact, consideration of the strength of classification scores may help identify cases where the training data were of poor quality, or where there may be evidence to identify different acoustic populations or behavior states.
 
 The workflow presented here provides a highly automated approach to detection of acoustic events (PAMGuard), integration of environmental data (PAMpal) and acoustic event classification (BANTER).
-These methods significantly reduce manual analysis, provide more consistent classification results with fewer biases, and provide an estimate of classification error.  The greatest improvement to classification results for beaked whales would likely result from improved sample sizes and examination of individual click trains in measuring ICI.
+These methods significantly reduce manual analysis, provide more consistent classification results with fewer biases, and provide an estimate of classification error.  The greatest improvement to classification results for beaked whales would likely result from improved sample sizes, and examination of individual click trains in measuring ICI.
 Consideration of additional detectors (e.g.
 matched filter detector, click train detector) or additional environmental variables may further improve classification results.
+Improved alignment of detection methods across studies (i.e., definition of events, hydrophone calibration) may allow for combination of data within geographic regions to improve assessments within regions for all species, and across regions for species with global distributions, such as Cuvier's beaked whales.
 These highly automated methods may allow for analysis of data across large spatial and temporal scales to address large ecological and population level questions.
@@
 The authors would like to acknowledge the large number of scientists and shipboard crew who were responsible for data collection during multiple large scale studies.
+Sea surface temperature data were provided by JPL under support by NASA MEaSUREs program.
 Funding for data collection and analysis were provided by U.S.
 Navy, Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration.
-The manuscript was improved thanks to reviews from \[XXXX\].
-
-Here are two sample references: @Feynman1963118 @Dirac1953888.
-
-By default, natbib will be used with the `authoryear` style, set in `classoption` variable in YAML. You can sets extra options with `natbiboptions` variable in YAML header. Example
-
-    natbiboptions: longnamesfirst,angle,semicolon
-
-There are various more specific bibliography styles available at <https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files>.
-To use one of these, add it in the header using, for example, `biblio-style: model1-num-names`.
-
-## Using CSL
-
-If `cite-method` is set to `citeproc` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`.
-In this case, the `csl` option is used to format the references.
-By default, this template will provide an appropriate style, but alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>.
-These can be downloaded and stored locally, or the url can be used as in the example header.
+The manuscript was improved thanks to reviews from XX and XX.
 
@@
 
-@fig-meaningless is generated using an R chunk.
+```{r, natlantic}
+#| label: fig-natlantic
+#| fig-cap: BANTER classification results from the NAtlantic dataset including echolocation clicks and interclick interval (EC_ICI).
+Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
+confuse <- magick::image_read('manuscript_files/natlantic_ici_confuseMatrix.png')%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read('manuscript_files/natlantic_ici_votes.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read('manuscript_files/natlantic_ici_proximity.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read('manuscript_files/natlantic_ici_importance.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("c) Importance Heat Map", size=100, color = "black")
 
-```{r}
-#| label: fig-meaningless
-#| fig-cap: A meaningless scatterplot
-#| fig-width: 5
-#| fig-height: 5
-#| fig-align: center
-#| out-width: 50%
-#| echo: false
-plot(runif(25), runif(25))
+natlantic_ici_Figure <-image_append(c(prox, heat, vote))
+natlantic_ici_Figure<- image_append(c(confuse, natlantic_ici_Figure), stack=TRUE)
+print(natlantic_ici_Figure, info=FALSE)
 ```
 
-# Tables coming from R
+```{r, satlantic}
+#| label: fig-satlantic
+#| fig-cap: BANTER classification results from the SAtlantic dataset (EC_ICI).
+Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
+confuse <- magick::image_read('manuscript_files/satlantic_ici_confuseMatrix.png')%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read('manuscript_files/satlantic_ici_votes.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read('manuscript_files/satlantic_ici_proximity.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read('manuscript_files/satlantic_ici_importance.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("c) Importance Heat Map", size=100, color = "black")
 
-Tables can also be generated using R chunks, as shown in @tbl-simple example.
-
-```{r}
-#| label: tbl-simple
-#| tbl-cap: Caption centered above table
-#| echo: true
-knitr::kable(head(mtcars)[,1:4])
+satlantic_ici_Figure <-image_append(c(prox, heat, vote))
+satlantic_ici_Figure<- image_append(c(confuse, satlantic_ici_Figure), stack=TRUE)
+print(satlantic_ici_Figure, info=FALSE)
 ```
 
+```{r, hawaii}
+#| label: fig-hawaii
+#| fig-cap: BANTER classification results from the Hawaii dataset (EC).
+Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
+confuse <- magick::image_read('manuscript_files/hawaii_ec_confuseMatrix.png')%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read('manuscript_files/hawaii_ec_votes.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read('manuscript_files/hawaii_ec_proximity.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read('manuscript_files/hawaii_ec_importance.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("c) Importance Heat Map", size=100, color = "black")
+
+hawaii_ec_Figure <-image_append(c(prox, heat, vote))
+hawaii_ec_Figure<- image_append(c(confuse, hawaii_ec_Figure), stack=TRUE)
+print(hawaii_ec_Figure, info=FALSE)
+```
+
+```{r, epacific-env}
+#| label: fig-epacific
+#| fig-cap: BANTER classification results from the EPacific dataset with environmental data (EC_ICI_ENV).
+Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
+confuse <- magick::image_read('manuscript_files/epacific_env_confuseMatrix.png')%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read('manuscript_files/epacific_env_votes.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read('manuscript_files/epacific_env_proximity.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read('manuscript_files/epacific_env_importance.png')%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("c) Importance Heat Map", size=100, color = "black")
+
+epacific_env_Figure <-image_append(c(prox, heat, vote))
+epacific_env_Figure<- image_append(c(confuse, epacific_env_Figure), stack=TRUE)
+print(epacific_env_Figure, info=FALSE)
+```
 # References {.unnumbered}

### 2023-06-08 manuscript/manuscript_18May2023.qmd
Major revisions for draft
Revised all datasets and created draft manuscript

--- 2023-02-07 manuscript/manuscript.qmd
@@
 ---
-```{r, hide=TRUE}
+
+```{r, include=FALSE}
 library(tidyverse)
+library(PAMpal)
 library(knitr)
+library(magick)
 library(here)
+here()
 ```
+
 # Introduction
 
-Passive acoustic monitoring (PAM) from has proven to be a valuable tool for studying populations of marine mammals [@parijs2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring.
-Historically, experienced acousticians manually identify stereotyped sounds that could be reliably attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering2013]; [@bittle2013]; [@rankin2005]; [@soldevilla2008].
+Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]\]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring.
+Historically, experienced acousticians manually identify stereotyped sounds that could be reliably attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle2013]; [@rankin_source_2005]; [@soldevilla_classification_2008].
 The dramatic increase in recordings make it impossible for experienced acousticians to manually annotate all data, and there has been an increasing need for the development of automated classification routines that can provide accurate determinations of the species responsible for the sounds.
 
-Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods ([@macleod2018]).
-However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering2013a].
-Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow2022]; [@baumann-pickering2014]; [@simonis2020].
+Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod_beaked_2018].
+However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_spatio-temporal_2014].
+Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022]; [@baumann-pickering_spatio-temporal_2014]; [@simonis_passive_2020].
 Unfortunately, manual classification of beaked whale echolocation clicks requires analysis by trained observers using a number of visual aids to examine the call characteristics.
@@
 
-BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings [@rankin2017].
-This hierarchical random forest event classifier relies on an initial call classifier (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier.
+BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings [@rankin_acoustic_2017].
+This hierarchical random forest event classifier relies on an initial call classifier for each call type (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier.
 The event classifier considers the probability of assignment to species for each call in the call classifier, along with any event-level characteristics (such as call rate).
 BANTER is very flexible and can accommodate any number of measures from any number of detectors.
-While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results [@rankin2017a].
+While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results [@rankin_acoustic_2017].
 In this case, settings for a whistle & moan detector were modified to improve performance at detecting burst pulses.
@@
 
-Here we applied BANTER acoustic classification to beaked whale detections from four large acoustic datasets off the US East Coast, Hawaii, and the US West Coast.
+Here we applied BANTER acoustic classification to beaked whale detections from four large acoustic datasets including the US East Coast (north and south), Hawaii, the US West Coast.
 Acoustic data were analyzed by experienced acousticians to determine species identity, and these 'ground truth' classifications served as training data for the supervised BANTER trials.
@@
 AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey area.
-The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (HTI-96-min, High Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see [@deangelis2018b] for more information).
+The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (HTI-96-min, High Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see [@deangelis_description_2018] for more information).
 The SAtlantic area ranged from Delaware south to central Florida (GU1605, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (Reson \[MODEL\], Teledyne Marine, Slangerup, Denmark) recorded 500 kHz sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass filter (see \[CITE\] for more information).
 
-Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, [@yano2018]).  Drifting recording buoys were deployed at randomly selected locations within the main Hawaiian Islands.
+Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, [@yano_cetacean_2018]).  Drifting recording buoys were deployed at randomly selected locations within the main Hawaiian Islands.
 Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA).
@@
 SM3M data were sampled at 256 kHz with continuous recordings.
-Individual buoys were deployed for 10 to 50 days (see [@yano2018]for more detailed information).
+Individual buoys were deployed for 10 to 50 days (see [@yano_cetacean_2018]for more detailed information).
 
 Recordings from the Eastern North Pacific Ocean off the U.S.
-west coast (EPacific) were collected using drifting acoustic recorders during the 2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels) survey [@keating2018].
+west coast (EPacific) were collected using drifting acoustic recorders during the 2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels) survey [@keating_passive_2018].
 Buoys were deployed at a number of stations in the California Current, including pre-determined locations (for abundance estimation studies) and in close proximity to seamounts (Ad Hoc Seamount Experiment).
@@
 Soundtrap recorders were sampled at 288 kHz and the SM3M recorders were recorded at 256 sample rate (with some sampling at 96 kHz for ocean noise measurements).
-Individual buoys were deployed for 11-23 days (see [@keating2018a] for more detailed information).
+Individual buoys were deployed for 11-23 days (see [@keating_passive_2018] for more detailed information).
 
@@
 
-All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c) with a suite of generic click classifiers within the Click Detector module (see [@keating2013]).  Classifier sets were saved such that any click may be classified as more than one click type.
+All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c) with a suite of generic click classifiers within the Click Detector module (see [@keating_summary_2013]).  Classifier sets were saved such that any click may be classified as more than one click type.
 The general click classifiers were effectively click detectors, as each implemented a simple spectral band settings (e.g., 2- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz).
-An additional detector/classifier within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see [@keating2013a]).
+An additional detector/classifier within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see [@keating_summary_2013]).
 
@@
 Acoustic researchers with experience in classifying beaked whale echolocation clicks examined the bearing-time display for the click detector module in PAMGuard to assess the species identity of the acoustic event.
-For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see [@deangelis2018]).
-Data were saved within the PAMGuard database and binaries for downstream processing using PAMpal package [@sakai2021] in R programming language [@rcoreteam2022].
+For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see [@deangelis_description_2018]).
+Data were saved within the PAMGuard database and binaries for downstream processing using PAMpal package [@sakai_pampal_2021] in R programming language [@rcoreteam2022].
 
@@
 
-BANTER models was created for each trial using the BANTER package in R programming language [@archer2022].
+BANTER models was created for each trial using the BANTER package in R programming language [@archer_banter_2022].
 Standard click calculations were used for the BANTER call classifier; event level measures of ICI were applied to the BANTER event classifier.
@@
 The sample size (sampsize) is the number of events considered for each trial of the model, and the ntree is the number of trees (iterations of the model) run in a given model.
-These values were modified to obtain the best results, with an emphasis on lower sample size and higher number of trees for computational performance [@rankin2021].
+These values were modified to obtain the best results, with an emphasis on lower sample size and higher number of trees for computational performance [@rankin_banter_2021].
 
-The model output include the confusion matrix, and a number of visualizations provided by the rfpermute package in R programming language [@archer2022a].
+The model output include the confusion matrix, and a number of visualizations provided by the rfpermute package in R programming language [@archer_rfpermute_2022].
 
@@
 
-Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study; species included: Baird's beaked whales (*Berardius bairdii*), Blainsville's beaked whales (*Mesoplodon densirostris*), Cuvier's beaked whales (*Ziphius cavirostris*), Cross Seamount beaked whale (McDonald et al., 2009), Gervais' beaked whales (*Mesoplodon europaeus*), Longman's beaked whale (*Indopacetus pacificus*), Sowerby's beaked whales (*Mesoplodon bidens*), True's beaked whales (*Mesoplodon mirus*), Stejneger's beaked whale (*Mesoplodon stejnegeri*), and unidentified beaked whale 'BW43' [@baumann-pickering2013].
+Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study; species included: Baird's beaked whales (*Berardius bairdii*), Blainsville's beaked whales (*Mesoplodon densirostris*), Cuvier's beaked whales (*Ziphius cavirostris*), Cross Seamount beaked whale (McDonald et al., 2009), Gervais' beaked whales (*Mesoplodon europaeus*), Longman's beaked whale (*Indopacetus pacificus*), Sowerby's beaked whales (*Mesoplodon bidens*), True's beaked whales (*Mesoplodon mirus*), Stejneger's beaked whale (*Mesoplodon stejnegeri*), and unidentified beaked whale 'BW43' [@baumann-pickering_species-specific_2013].
 Analyses were conducted separately for the different study areas due to differences in the data collection methods that precluded combining of datasets.
@@
 
-\[Insert Figure 1\]\[Figure 1.
-BANTER classification results from the NAtlantic dataset (EC).
+![BANTER classification results from the NAtlantic dataset (EC).
 Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
@@
 Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/natlantic_Figure.png){fig-align="center"}
 
@@
 
-Species encountered in the SAtlantic study dataset include Blainsville's beaked whales (n=35), Cuvier's beaked whales (n=66), Gervais' beaked whales (n=45), Sowerby's beaked whales (n=2), and True's beaked whales (n-12).
+Species encountered in the SAtlantic study dataset include Blainsville's beaked whales (n = 35), Cuvier's beaked whales (n = 66), Gervais' beaked whales (n = 45), Sowerby's beaked whales (n = 2), and True's beaked whales (n = 12).
 BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).
@@
 
-Figure 2.
-BANTER classification results from the SAtlantic dataset.
+![BANTER classification results from the SAtlantic dataset.
 Confusion matrix provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).
@@
 Heat map (b) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (c) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+Vote Plot (c) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/satlantic_ici_Figure.png){fig-align="center"}
 
@@
 ## Hawaii
+
 The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (Yano et al.
-2018); species included Longman's beaked whale (n=121), Blainsville's beaked whale (n=518), Cross Seamount beaked whale (n=84) and Cuvier's beaked whale (n=201).
+2018); species included Blainsville's beaked whale (n = 521), Cross Seamount beaked whale (n = 76), Cuvier's beaked whale (n = 201), and Longman's beaked whale (n = 122).
 BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).
 
-The most accurate model considered the echolocation clicks only (Detector Model: sampsize = 10, ntrees = 5,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 91.8% for all four species.
-Classification scores ranged from a low of 84.5% (Cross Seamount Beaked Whale) to 97.5% (Longman's beaked whale, Fig.
+The most accurate model considered the echolocation clicks only (Detector Model: sampsize = 10, ntrees = 5,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.1% for all four species.
+Classification scores ranged from a low of 86% (Cuvier's Beaked Whale) to 97.5% (Longman's beaked whale, Fig.
 3).
@@
 
-::: {layout="[[100], [30,30,30]]"}
-a)  ![Fig.1](../hawaii/banter_model_hawaii_confuseMatrix.png){#fig-a}
-
-b)  ![Fig.1](../hawaii/banter_model_hawaii_proximity.png){#fig-b} :::
-
-c)  ![Fig.1](../hawaii/banter_model_hawaii_importance.png)
-
-d)  ![Fig.1](../hawaii/banter_model_hawaii_votes.png)
-:::
-
-\[Insert Figure 3\]\[Figure 3.
-BANTER classification results from the Hawaii dataset.
+![BANTER classification results from the Hawaii dataset.
 Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).
@@
 Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
-Species codes include Cross Seamount beaked whale (BWC), Longman's beaked whale (IP), Blainsville's beaked whale (MD) and Cuvier's beaked whale (ZC).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/hawaii_ici_Figure.png)
 
@@
 3b) which shows the classification results based on the two most 'important' predictors.
-Distinctive clusters based on the two most important features can be found for Longman's (IP), Blainsville's (MD), and Cuvier's beaked whales (ZC), and there is overlap in the feature space of the clusters for Cuvier's and Cross Seamount beaked whales (BWC).
-This overlap is resolved with different features for ZC (3^rd^, 7^th^ in Fig.
-3c heat map) and BWC (5^th^, 8^th^ feature in heat map), as evidence in the low misclassification rates for these species (Fig.
+Distinctive clusters based on the two most important features can be found for Longman's, Blainsville's, and Cuvier's beaked whales, and there is overlap in the feature space of the clusters for Cuvier's and Cross Seamount beaked whales.
+This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the Importance Heat Map (Fig.
+3c), resulting in the low misclassification rates for these species (Fig.
 3a confusion matrix).
 The plot votes graph (Fig.
-3d) show complicated results with variation in the proportion of trees that correctly classified the species for each event.
+3d) show strong classification results for Blainsville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.
 
@@
 
-The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States ([@keating2018]); species included Baird's beaked whale (n=91), Unidentified Beaked Whale 'BW43' (n=11), Stejneger's beaked whale (n=41) and Cuvier's beaked whale (n=1,332).
+The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States ([@simonis_passive_2020]); species included Baird's beaked whale (n = 31), Unidentified Beaked Whale 'BW43' (n = 135), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 935), Hubbs beaked whale (n = 66) and Stejneger's beaked whale (n = 42).
 BANTER classification trials included (1) echolocation clicks (EC) and (2) echolocation clicks and ICI (EC_ICI).
 
-The best model considered the echolocation clicks and the ICI values (Detector Model: sampsize = 5, ntrees = 30,000; BANTER model: sampsize = 4, ntree = 200,000) and provided an overall correct classification rate of 93.7% for all four species.
-Classification scores ranged from a low of 93.2% (Cuvier's beaked whale) to 100% (BW43, Fig.
+The best model considered the echolocation clicks and the ICI values (Detector Model: sampsize = 3, ntrees = 10,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 91.4% for all six species.
+Classification scores ranged from a low of 83.3% (Cross Seamount beaked whale) to 96.9% (Hubbs beaked whale, Fig.
 4a).
@@
 
-\[Insert Figure 4\]\[Figure 4.
-BANTER classification results from the EPacific dataset.
-Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).  Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+![BANTER classification results from the EPacific dataset.
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate), and .
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
 Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
-Species codes include Baird's beaked whale (BB), unidentified beaked whale (BW43), Stejneger's beaked whale (MS) and Cuvier's beaked whale (ZC).\]
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/epacific_ici_Figure.png){fig-align="center"}
 
-There are clear distinctions in the primary features used to classify the species as shown in the Proximity Plot (Fig.
-4b) which shows the classification results based on the two most 'important' predictors.
-The species each form distinctive clusters based on these two features; however, many events fall outside these distinct bounds (detections shown in area between species classification clusters).
-Both BB and BW43 relied on a wide range of variables for classification (Fig.
+There is considerable overlap in the primary features used to classify the Cuvier's, Cross Seamount, and BW43 beaked whales, as shown in the Proximity Plot (Fig.
+4b), which shows the classification results based on the two most 'important' predictors.
+ICI was among the most important classification variable for all species (Fig.
 4c heat map).
 The plot votes (Fig.
-4d) show strong classification results for most events of most species.
+4d) show strong classification results for most events of all species except Cross Seamount beaked whale.
 
@@
 
-Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (long-beaked and short-beaked common dolphins, [@rankin2017b]).
+Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (long-beaked and short-beaked common dolphins, [@rankin_acoustic_2017]).
 This suggests that large increases in sample size may improve classification results for Cuvier's beaked whales.
@@
 
-Gervais' and True's beaked whales have similar click characteristics that make them difficult to differentiate, and require considerable expertise to classify manually, using a variety of available visuals and metrics [@deangelis2018a].
+Gervais' and True's beaked whales have similar click characteristics that make them difficult to differentiate, and require considerable expertise to classify manually, using a variety of available visuals and metrics [@deangelis_description_2018].
 Our results suggest that BANTER may serve as an efficient and effective means of classifying and differentiating these two species.
@@
 Each event was defined as an individual click train, so ICI for these events more likely represented true ICI, and was found to be more informative.
-[@baumann-pickering2013] showed an overlap in the ICI between Cuvier's and Blainsville's beaked whales, but a strong difference in ICI between these species and BWC.
-Likewise, [@baumann-pickering2013] found a difference in ICI between Baird's beaked whales and Cuvier's beaked whales.
+[@baumann-pickering_species-specific_2013] showed an overlap in the ICI between Cuvier's and Blainsville's beaked whales, but a strong difference in ICI between these species and BWC.
+Likewise, [@baumann-pickering_species-specific_2013] found a difference in ICI between Baird's beaked whales and Cuvier's beaked whales.
 Subdividing click trains in PAMGuard is time consuming, but alternative options include consideration of the PAMGuard click train detector module in BANTER, or by developing a similar function in R to apply to events.
@@
 Navy, Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration.
-The manuscript was improved thanks to reviews from \[XXXX\].
-
-Here are two sample references: @Feynman1963118 @Dirac1953888.
-
-By default, natbib will be used with the `authoryear` style, set in `classoption` variable in YAML. You can sets extra options with `natbiboptions` variable in YAML header. Example
-
-    natbiboptions: longnamesfirst,angle,semicolon
-
-There are various more specific bibliography styles available at <https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files>.
-To use one of these, add it in the header using, for example, `biblio-style: model1-num-names`.
-
-## Using CSL
-
-If `cite-method` is set to `citeproc` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`.
-In this case, the `csl` option is used to format the references.
-By default, this template will provide an appropriate style, but alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>.
-These can be downloaded and stored locally, or the url can be used as in the example header.
+The manuscript was improved thanks to reviews from XX and XX.
 
@@
 
-```{r}
+```{r, test}
 #| label: fig-meaningless
@@
 #| echo: false
-plot(runif(25), runif(25))
+#plot(runif(25), runif(25))
 ```

### 2023-06-08 manuscript/manuscript_19May2023.qmd
Major revisions for draft
Revised all datasets and created draft manuscript

--- 2023-06-08 manuscript/manuscript_18May2023.qmd
@@
 Data from four large scale surveys of beaked whales (North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and BANTER (hierarchical random forest classifier).
-Overall correct classification results ranged from 86% for the South Atlantic data to 99% for the North Atlantic.
+Overall correct classification results ranged from 87% for the South Atlantic data to 98% for the North Atlantic.
 Results for many species could likely be improved with increased sample sizes and consideration of alternative automated detectors.
@@
 This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the overall number of detections was different.
-This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn2021].
+This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn_acoustic_2021].
 For species who primarily (or exclusively) produce echolocation clicks, BANTER may serve as an option for automated machine learning classification.
@@
 
-Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n=70), Sowerby's beaked whales (n=5), True's beaked whales (n=65), and Gervais' beaked whales (n=3).
+Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n=120), Sowerby's beaked whales (n=6), True's beaked whales (n=76), and Gervais' beaked whales (n=3).
 BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).
 
-For the NAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (EC_ICI, Detector Model: sampsize = 4, ntrees = 30,000; BANTER Model: sampsize = 4, trees=300,000) and provided an overall correct classification rate of 100% for all four species (Fig.
-1).
+For the NAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (EC_ICI, Detector Model: sampsize = 4, ntrees = 30,000; BANTER Model: sampsize = 4, trees=300,000) and provided an overall correct classification rate of 100% for all four species (@fig-natlantic).
 There were insufficient samples of Gervais' beaked whales to include in the NAtlantic model (n=2).
-All classification results were greater than the expected error rate (Prior in Fig.
-1a).
+All classification results were greater than the expected error rate (Prior in @fig-natlantic a).
 Results from the EC model are presented here; results from the EC_ICI model can be found in supplementary materials (Supplement Fig.
@@
 
-![BANTER classification results from the NAtlantic dataset (EC).
-Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
-Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/natlantic_Figure.png){fig-align="center"}
+The proximity plot (@fig-natlantic b) provides a view of the distribution of events within the feature space of the two 'most important' dimensions, or predictors.
+For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result.
+The importance heat map (@fig-natlantic c) shows that the most important features for predicting each species.
 
-The proximity plot (Fig.1b) provides a view of the distribution of events within the feature space of the two 'most important' dimensions, or predictors.
-For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result.
-The importance heat map (Fig.
-1c) shows that the most important features for predicting each species.
-
-The strength of the classifications by species can be seen in the plot votes (Fig.
-1d), which show the istribution of the votes for each of the 300,000 trials.
+The strength of the classifications by species can be seen in the plot votes (@fig-natlantic d), which show the istribution of the votes for each of the 300,000 trials.
 For each plot, the events are represented as vertical slices (with plot subdivided along the x-axes according to the event sample size), and the percentage of votes for each species is represented by their color along that vertical slice.
@@
 3 for EC_ICI_ALT), but resulted in the loss of Sowerby's in the final model.
-Classification scores ranged from a low of 74.2% (Cuvier's beaked whale) to 100% (Sowerby's and True's beaked whale, Fig.
-2); all classification results were greater than the expected error rate (Prior in Fig.
-2a).
+Classification scores ranged from a low of 74.2% (Cuvier's beaked whale) to 100% (Sowerby's and True's beaked whale, @fig-satlantic); all classification results were greater than the expected error rate (Prior in @fig-natlantic a).
 Results from the EC_ICI model are presented here; results from the EC model can be found in supplementary materials (Supplement Fig.
@@
 
-![BANTER classification results from the SAtlantic dataset.
-Confusion matrix provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).
-Proximity plot (a) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (b) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (c) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/satlantic_ici_Figure.png){fig-align="center"}
-
-For SAtlantic, the proximity plot (Fig.
-2b) shows that True's beaked whales are differentiated from the other species based solely on the two most important features.
-The importance heat map (Fig.
-2c) shows that Cuvier's, Gervais' and Sowerby's beaked whales rely on a number of predictors that are lower in importance and there is likely greater distinction between these species than is represented by the proximity plot (Fig.
-2b).
-The plot votes (Fig.
-2d) shows a moderate number of trees in the forest were misclassified and these misclassifications included all species in the model.
+For SAtlantic, the proximity plot (@fig-natlantic b) shows that True's beaked whales are differentiated from the other species based solely on the two most important features.
+The importance heat map (@fig-natlantic c) shows that Cuvier's, Gervais' and Sowerby's beaked whales rely on a number of predictors that are lower in importance and there is likely greater distinction between these species than is represented by the proximity plot (@fig-natlantic b).
+The plot votes (@fig-natlantic d) shows a moderate number of trees in the forest were misclassified and these misclassifications included all species in the model.
 
@@
 The most accurate model considered the echolocation clicks only (Detector Model: sampsize = 10, ntrees = 5,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.1% for all four species.
-Classification scores ranged from a low of 86% (Cuvier's Beaked Whale) to 97.5% (Longman's beaked whale, Fig.
-3).
+Classification scores ranged from a low of 86% (Cuvier's Beaked Whale) to 97.5% (Longman's beaked whale, @fig-hawaii).
 All classification results were greater than the expected error rate (Prior in Fig.
@@
 
-![BANTER classification results from the Hawaii dataset.
-Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate).
-Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/hawaii_ici_Figure.png)
-
-There are clear distinctions in the primary features used to classify three of the four species as shown in the Proximity Plot (Fig.
-3b) which shows the classification results based on the two most 'important' predictors.
+There are clear distinctions in the primary features used to classify three of the four species as shown in the Proximity Plot (@fig-hawaii b) which shows the classification results based on the two most 'important' predictors.
 Distinctive clusters based on the two most important features can be found for Longman's, Blainsville's, and Cuvier's beaked whales, and there is overlap in the feature space of the clusters for Cuvier's and Cross Seamount beaked whales.
-This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the Importance Heat Map (Fig.
-3c), resulting in the low misclassification rates for these species (Fig.
-3a confusion matrix).
-The plot votes graph (Fig.
-3d) show strong classification results for Blainsville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.
+This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the Importance Heat Map (@fig-hawaii c), resulting in the low misclassification rates for these species (@fig-hawaii a confusion matrix).
+The plot votes graph (@fig-hawaii d) show strong classification results for Blainsville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.
 
@@
 
-The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States ([@simonis_passive_2020]); species included Baird's beaked whale (n = 31), Unidentified Beaked Whale 'BW43' (n = 135), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 935), Hubbs beaked whale (n = 66) and Stejneger's beaked whale (n = 42).
+The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States [@simonis_passive_2020]; species included Baird's beaked whale (n = 31), Unidentified Beaked Whale 'BW43' (n = 135), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 935), Hubbs beaked whale (n = 66) and Stejneger's beaked whale (n = 42).
 BANTER classification trials included (1) echolocation clicks (EC) and (2) echolocation clicks and ICI (EC_ICI).
@@
 The best model considered the echolocation clicks and the ICI values (Detector Model: sampsize = 3, ntrees = 10,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 91.4% for all six species.
-Classification scores ranged from a low of 83.3% (Cross Seamount beaked whale) to 96.9% (Hubbs beaked whale, Fig.
-4a).
-All classification results were greater than the expected error rate (Prior in Fig.
-4a).
+Classification scores ranged from a low of 83.3% (Cross Seamount beaked whale) to 96.9% (Hubbs beaked whale, @fig-epacific a).
+All classification results were greater than the expected error rate (Prior in @fig-epacific a).
 Results from the EC_ICI model are presented here; complete results can be found in supplementary materials (Supplement Figs.
@@
 
-![BANTER classification results from the EPacific dataset.
-Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate), and .
-Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
-Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/epacific_ici_Figure.png){fig-align="center"}
-
-There is considerable overlap in the primary features used to classify the Cuvier's, Cross Seamount, and BW43 beaked whales, as shown in the Proximity Plot (Fig.
-4b), which shows the classification results based on the two most 'important' predictors.
-ICI was among the most important classification variable for all species (Fig.
-4c heat map).
-The plot votes (Fig.
-4d) show strong classification results for most events of all species except Cross Seamount beaked whale.
+There is considerable overlap in the primary features used to classify the Cuvier's, Cross Seamount, and BW43 beaked whales, as shown in the Proximity Plot (@fig-epacific b), which shows the classification results based on the two most 'important' predictors.
+ICI was among the most important classification variable for all species (@fig-epacific c heat map).
+The plot votes (@fig-epacific d) show strong classification results for most events of all species except Cross Seamount beaked whale.
 
@@
 
-@fig-meaningless is generated using an R chunk.
+[![BANTER classification results from the NAtlantic dataset (EC).
+Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](./images/natlantic_Figure.png){#fig-natlantic="" fig-align="center"}](#fig-natlantic)
 
-```{r, test}
-#| label: fig-meaningless
-#| fig-cap: A meaningless scatterplot
-#| fig-width: 5
-#| fig-height: 5
-#| fig-align: center
-#| out-width: 50%
-#| echo: false
-#plot(runif(25), runif(25))
-```
+[![BANTER classification results from the SAtlantic dataset.
+Confusion matrix provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (a) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (b) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (c) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](./images/satlantic_ici_Figure.png){#fig-satlantic="" fig-align="center"}](#fig-satlantic)
 
-# Tables coming from R
+[![BANTER classification results from the Hawaii dataset.
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](./images/hawaii_ici_Figure.png){#fig-hawaii="" fig-align="center"}](#fig-hawaii)
 
-Tables can also be generated using R chunks, as shown in @tbl-simple example.
-
-```{r}
-#| label: tbl-simple
-#| tbl-cap: Caption centered above table
-#| echo: true
-knitr::kable(head(mtcars)[,1:4])
-```
+[![BANTER classification results from the EPacific dataset.
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](./images/epacific_ici_Figure.png){#fig-epacific="" fig-align="center"}](#fig-epacific)
 

### 2023-06-08 natlantic/beakerBanter_natlantic.Rmd
Major revisions for draft
Revised all datasets and created draft manuscript

@@
 author: "Shannon Rankin"
-date: "2023-01-26"
+date: "2023-05-25"
 output: html_document
@@
 ```
+/Project
+  |-Database/
+  |-Binaries/
+  |-Figures/
+  |-natlantic
+    |_ beakerBanter_natlantic.Rmd
+    |_ natlantic_banter_data.rds
+    |_ natlantic_banter_data_ici.rds
+    |_ natlantic_banter_ec_model_t1e4s3_t1e4s3.rds
+    |_ natlantic_banter_model_ici_t1e4s3_t1e5s3.rds
+    |_ natlantic_banter_model_alt_t1e4s5_t1e4s5.rds
 
@@
 ```{r, load packages, message=FALSE}
-library (PAMpal)
-library(banter)
-library(rfPermute)
-library(kableExtra)
-library(readxl)
-library(stringr)
-library(dplyr)
-library(here)
+library("easypackages")
+libraries("PAMpal", "banter", "rfPermute", "readxl", "stringr", "dplyr", "kableExtra", "magick", "magrittr", "here")
 here()
@@
 
-[If]{.underline} you have already run the processing code, ensure you
-have set 'freshRun = FALSE' at top of this document to read in the
-existing .rds file for downstream processing.
-
-```{r, read acoustic study, include=FALSE, eval=!freshRun}
-data <- readRDS('natlantic_study.rds')
-# Double check warning messages
-print(getWarnings(data)$message)
-```
-
 [If]{.underline} this is the [initial processing]{.underline}, ensure
@@
 saveRDS(data, 'natlantic_study.rds')
+# Double check warning messages
+print(getWarnings(data)$message)
+```
+[If]{.underline} you have already run the processing code, ensure you
+have set 'freshRun = FALSE' at top of this document to read in the
+existing .rds file for downstream processing.
+
+```{r, read acoustic study, include=FALSE, eval=!freshRun}
+data <- readRDS('natlantic_study.rds')
 # Double check warning messages
@@
 
-```{r, assign species ID}
+```{r, assign species ID, eval=freshRun}
 HB1603 <- read_excel("HB1603_BWs.xlsx", sheet = "Alldata")#read in data
@@
 
-```{r, filter unwanted data}
+```{r, filter unwanted data, eval=freshRun}
 goodSpecies<- c(c("Trues", "Cuviers", "Sowerbys", "Gervais"))
@@
 
-```{r, calculate ICI}
+```{r, calculate ICI, eval=freshRun}
 data <- calculateICI(data, time='peakTime')
@@
 
-### *Model without ICI*
+### *EC (only) Model*
 
-1.  Load data, initialize a BANTER model, and create an initial BANTER
-    detector model (stage 1).
+Initialize, Run & Evaluate Detector Model (stage 1).
+```{r, banter ec detector model, eval=freshRun}
+banter_model_ec <- initBanterModel(banter_data$events)
+banter_model_ec <- addBanterDetector(banter_model_ec, banter_data$detectors, ntree=1e4, sampsize=3, importance = TRUE)
 
-```{r, read banter model, include=FALSE, eval=!freshRun}
-banter_model <- readRDS('natlantic_banter_model_t5e3s3_t5e3s4.rds')
+plotDetectorTrace(banter_model_ec, detector = paste0('Click_Detector_', 3:4))
+plotDetectorTrace(banter_model_ec, detector = paste0('Click_Detector_', 5:6))
+summary(banter_model_ec)
 ```
 
-```{r, banter detector model, eval=freshRun}
-banter_model <- initBanterModel(banter_data$events)
-banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=5e3, sampsize=3, importance = TRUE)
+Run BANTER Event Model (stage 2)
+```{r, banter ec event model, eval=freshRun}
+banter_model_ec <- runBanterModel(banter_model_ec, ntree=1e4, sampsize=3)
+summary(banter_model_ec)
 ```
 
-Check for stability (are more trees needed?) before moving on
-
-```{r, banter detector model stability}
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 3:4))
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 5:6))
-summary(banter_model)
-```
-
-2.  Run BANTER Event Model (stage 2)
-
-```{r, banter event model}
-banter_model <- runBanterModel(banter_model, ntree=5e3, sampsize=4)
-summary(banter_model)
-```
-
-Check for stability (are more trees needed?). If acceptable, save model
-with tree/sampsize info in the filename.
-
-```{r,save banter model, eval=freshRun}
-saveRDS(banter_model, 'natlantic_banter_model_t5e3s3_t5e3s4.rds')
+Once a stable model is identified, save model with tree/sampsize info in the filename.
+```{r,save banter ec model, eval=freshRun}
+saveRDS(banter_model_ec, 'natlantic_banter_ec_model_t1e4s3_t1e4s3.rds')
 ```
@@
 
-1.  Load data, initialize a BANTER model, and create an initial BANTER
-    detector model (stage 1).
-
-```{r, read banter ici model, include=FALSE, eval=!freshRun}
-banter_model_ici <- readRDS('natlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
-```
-
+Initialize, Run & Evaluate Detector Model (stage 1).
 ```{r, banter ici detector model, eval=freshRun}
 banter_model_ici <- initBanterModel(banter_data_ici$events)
-banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=10e3, sampsize=2, importance = TRUE)
-```
+banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=1e4, sampsize=3, importance = TRUE)
 
-Check for stability to see if we need more trees
-
-```{r, banter ici detector model stability}
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 3:4))
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 5:6))
+plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 3:4))
+plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 5:6))
 summary(banter_model_ici)
@@
 
-2.  Run BANTER Event Model (stage 2)
-
-```{r, banter ici event model}
-banter_model_ici <- runBanterModel(banter_model_ici, ntree=10e3, sampsize=4)
+Run BANTER Event Model (stage 2)
+```{r, banter ici event model, eval=freshRun}
+banter_model_ici <- runBanterModel(banter_model_ici, ntree=1e5, sampsize=3)
 summary(banter_model_ici)
@@
 
-Check for stability (are more trees needed?). If acceptable, save model
-with tree/sampsize info in the filename.
+Once a stable model is identified, save model with tree/sampsize info in the filename.
+```{r, save banter ici model, eval=freshRun}
+saveRDS(banter_model_ici, 'natlantic_banter_model_ici_t1e4s3_t1e5s3.rds')
+```
+### *Alternative Model (EC_ICI_ALT) with larger samp size*
 
-```{r, save banter ici model, eval=freshRun}
-saveRDS(banter_model_ici, 'natlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
+Initialize, Run & Evaluate Detector Model (stage 1).
+```{r, banter alt detector model, eval=freshRun}
+banter_model_alt <- initBanterModel(banter_data_ici$events)
+banter_model_alt <- addBanterDetector(banter_model_alt, banter_data_ici$detectors, ntree=1e4, sampsize=5, importance = TRUE)
+
+plotDetectorTrace(banter_model_alt, detector = paste0('Click_Detector_', 3:4))
+plotDetectorTrace(banter_model_alt, detector = paste0('Click_Detector_', 5:6))
+summary(banter_model_alt)
 ```
 
+Run BANTER Event Model (stage 2)
+```{r, banter alt event model, eval=freshRun}
+banter_model_alt <- runBanterModel(banter_model_alt, ntree=1e4, sampsize=5)
+summary(banter_model_alt)
+```
+
+Once a stable model is identified, save model with tree/sampsize info in the filename.
+```{r, save banter alt model, eval=freshRun}
+saveRDS(banter_model_alt, 'natlantic_banter_model_alt_t1e4s5_t1e4s5.rds')
+```
 ## BANTER Analytics
@@
 
+```{r, read banter model, include=FALSE, eval=!freshRun}
+banter_model_ec <- readRDS('natlantic_banter_ec_model_t1e4s3_t1e4s3.rds')
+banter_model_ici <- readRDS('natlantic_banter_model_ici_t1e4s3_t1e5s3.rds')
+banter_model_alt <- readRDS('natlantic_banter_model_alt_t1e4s5_t1e4s5.rds')
+```
+
 First, identify the model you would like to examine (comment out the
@@
 ```{r, identify model}
-model <- banter_model
-modelname <- "banter_model"
+model_ec <- banter_model_ec
+modelname_ec <- "banter_model_ec"
 
@@
 modelname_ici <- "banter_model_ici"
+
+model_alt <- banter_model_alt
+modelname_alt <- "banter_model_alt"
 ```
@@
 ```{r, extract RF model, include=TRUE}
-banter_model_RF <- getBanterModel(model)
+banter_model_ec_RF <- getBanterModel(model_ec)
 banter_model_ici_RF <- getBanterModel(model_ici)
+banter_model_alt_RF <- getBanterModel(model_alt)
+```
+
+Class Priors (Expected Error Rate)
+
+```{r, class priors, include=TRUE}
+natlantic_ec_priors <- classPriors(banter_model_ec_RF, NULL)[,1]
+natlantic_ici_priors <- classPriors(banter_model_ici_RF, NULL)[,1]
+natlantic_alt_priors <- classPriors(banter_model_alt_RF, NULL)[,1]
 ```
@@
 ```{r, confusion matrix, include=TRUE}
-natlantic_confuseMatrix <- rfPermute::confusionMatrix(banter_model_RF)
-natlantic_confuseMatrix <- kable(natlantic_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+natlantic_ec_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ec_RF)
+natlantic_ec_confuseMatrix <- cbind(natlantic_ec_confuseMatrix, priors = natlantic_ec_priors)
+
+natlantic_ec_confuseMatrix <- kable(natlantic_ec_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(5, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(3,hline_after = TRUE)%>%
+  row_spec(5, bold = TRUE)%>%
+  save_kable('../manuscript/manuscript_files/natlantic_ec_confuseMatrix.png', zoom = 9)
+
+natlantic_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
+natlantic_ici_confuseMatrix <- cbind(natlantic_ici_confuseMatrix, priors = natlantic_ici_priors)
+
+natlantic_ici_confuseMatrix <- kable(natlantic_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(5, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(3,hline_after = TRUE)%>%
+  row_spec(5, bold = TRUE)%>%
+  save_kable('../manuscript/manuscript_files/natlantic_ici_confuseMatrix.png', zoom = 9)
+
+natlantic_alt_confuseMatrix <- rfPermute::confusionMatrix(banter_model_alt_RF)
+natlantic_alt_confuseMatrix <- cbind(natlantic_alt_confuseMatrix, priors = natlantic_alt_priors)
+
+natlantic_alt_confuseMatrix <- kable(natlantic_alt_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
   kable_classic()%>%
@@
   row_spec(4, bold = TRUE)%>%
-  save_kable(paste(modelname, "natlantic_confuseMatrix.png", sep="_"), zoom = 2)
-
-natlantic_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
-natlantic_ici_confuseMatrix <- kable(natlantic_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
-  kable_classic()%>%
-  column_spec(4, border_right = TRUE)%>%
-  row_spec(0, bold = TRUE)%>%
-  row_spec(3,hline_after = TRUE)%>%
-  row_spec(4, bold = TRUE)%>%
-  save_kable(paste(modelname_ici, "natlantic_ici_confuseMatrix.png", sep="_"), zoom = 2)
+  save_kable('../manuscript/manuscript_files/natlantic_alt_confuseMatrix.png', zoom = 9)
 
@@
 
-![](banter_model_hawaii_confuseMatrix.png)
+![BANTER Model N. Atlantic EC Confusion
+Matrix](../manuscript/manuscript_files/natlantic_ec_confuseMatrix.png)
+
+![BANTER Model N. Atlantic ICI Confusion
+Matrix](../manuscript/manuscript_files/natlantic_ici_confuseMatrix.png)
+
+![BANTER Model N. Atlantic Alt Confusion
+Matrix](../manuscript/manuscript_files/natlantic_alt_confuseMatrix.png)
 
@@
 
-```{r, proximity plot, include=TRUE}
-png(paste(modelname, "natlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
-hawaii_proximityPlot <- plotProximity(banter_model_RF)
+```{r, proximity plot, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/natlantic_ec_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
+natlantic_ec_proximityPlot <- plotProximity(banter_model_ec_RF)
 dev.off()
-hawaii_proximityPlot <- plotProximity(banter_model_RF)
+natlantic_ec_proximityPlot <- plotProximity(banter_model_ec_RF)
 
-png(paste(modelname_ici, "natlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
-ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
+png(('../manuscript/manuscript_files/natlantic_ici_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
+ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
 dev.off()
-ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
+
+png(('../manuscript/manuscript_files/natlantic_alt_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
+alt_natlantic_proximityPlot <- plotProximity(banter_model_alt_RF)
+dev.off()
+alt_natlantic_proximityPlot <- plotProximity(banter_model_alt_RF)
 ```
@@
 
-```{r, importance heat map, include=TRUE}
-png(paste(modelname, "natlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_RF, plot.type="heatmap")
+```{r, importance heat map, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/natlantic_ec_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
+plotImportance(banter_model_ec_RF, plot.type="heatmap")
 dev.off()
-hawaii_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
+natlantic_ec_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap")
 
-png(paste(modelname_ici, "natlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
+png(('../manuscript/manuscript_files/natlantic_ici_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
 plotImportance(banter_model_ici_RF, plot.type="heatmap")
 dev.off()
-ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_natlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+
+png(('../manuscript/manuscript_files/natlantic_alt_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
+plotImportance(banter_model_alt_RF, plot.type="heatmap")
+dev.off()
+alt_natlantic_importance <- plotImportance(banter_model_alt_RF, plot.type="heatmap")
 ```
@@
 
-```{r, plot votes, include=TRUE}
-png(paste(modelname, "natlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_RF)
+```{r, plot votes, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/natlantic_ec_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_ec_RF)
 dev.off()
-hawaii_votes <- plotVotes(banter_model_RF)
+natlantic_ec_votes <- plotVotes(banter_model_ec_RF)
 
-png(paste(modelname_ici, "natlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
+png(('../manuscript/manuscript_files/natlantic_ici_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
 plotVotes(banter_model_ici_RF)
 dev.off()
-ici_hawaii_votes <- plotVotes(banter_model_ici_RF)
-```
+ici_natlantic_votes <- plotVotes(banter_model_ici_RF)
 
-Class Priors (Expected Error Rate)
-
-```{r, class priors, include=TRUE}
-hawaii_priors <- classPriors(banter_model_RF, NULL)
-```
-
-Plot Confusion Matrix Heat Map
-
-```{r, confusion matrix heat map, include=TRUE}
-plotConfMat(banter_model_RF, title="ConfusionMatrix", plot=TRUE)
-```
-
-Plot Error Trace for Banter Model
-
-```{r, model error trace, include=TRUE}
-plot(banter_model_RF)
+png(('../manuscript/manuscript_files/natlantic_alt_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_alt_RF)
+dev.off()
+alt_natlantic_votes <- plotVotes(banter_model_alt_RF)
 ```
@@
 
-```{r, predicted probabilities, include=TRUE}
-plotPredictedProbs(banter_model_RF, bins=30, plot=TRUE)
+```{r, predicted probabilities, include=TRUE, results=FALSE}
+plotPredictedProbs(banter_model_ec_RF, bins=30, plot=TRUE)
+plotPredictedProbs(banter_model_ici_RF, bins=30, plot=TRUE)
+plotPredictedProbs(banter_model_alt_RF, bins=30, plot=TRUE)
 ```
+
+Create Figure for Publication
+```{r, Publication Figure, include=TRUE}
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+natlantic_ec_Figure <-image_append(c(prox, heat, vote))
+natlantic_ec_Figure<- image_append(c(confuse, natlantic_ec_Figure), stack=TRUE)
+image_write(natlantic_ec_Figure, path = here('manuscript', 'manuscript_files','natlantic_ec_Figure.png'), format ='png')
+print(natlantic_ec_Figure, info=FALSE)
+
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+natlantic_ici_Figure <-image_append(c(prox, heat, vote))
+natlantic_ici_Figure<- image_append(c(confuse, natlantic_ici_Figure), stack=TRUE)
+image_write(natlantic_ici_Figure, path = here('manuscript', 'manuscript_files','natlantic_ici_Figure.png'), format ='png')
+print(natlantic_ici_Figure, info=FALSE)
+
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+natlantic_alt_Figure <-image_append(c(prox, heat, vote))
+natlantic_alt_Figure<- image_append(c(confuse, natlantic_ici_Figure), stack=TRUE)
+image_write(natlantic_alt_Figure, path = here('manuscript', 'manuscript_files','natlantic_alt_Figure.png'), format ='png')
+print(natlantic_alt_Figure, info=FALSE)
+```

### 2023-06-08 satlantic/beakerBanter_satlantic.Rmd
Major revisions for draft
Revised all datasets and created draft manuscript

@@
 author: "Shannon Rankin"
-date: "2023-02-08"
+date: "2023-05-17"
 output: html_document
@@
 ```
+###Organizational Instructions
+Save this file in an R Project that includes the following foldering:
+/Project
+  |-Database/
+  |-Binaries/
+  |-Figures/
+  |-satlantic
+    |_ beakerBanter_satlantic.Rmd
+    |_ GU1605_PG_OfflineEvents_20210809.xlsx
+    |_ satlantic_banter_data.rds
+    |_ satlantic_banter_data_ici.rds
+    |_ satlantic_banter_model_ec_t1e4s2_t1e5s1.rds
+    |_ satlantic_banter_model_ici_t1e4s4_t1e5s1.rds
+    |_ satlantic_banter_model_alt_t1e4s4_t1e5s2.rds
 
@@
 ```{r, load packages, message=FALSE}
-library (PAMpal)
-library(banter)
-library(rfPermute)
-library(kableExtra)
-library(readxl)
-library(stringr)
-library(here)
+library("easypackages")
+libraries("PAMpal", "banter", "rfPermute", "kableExtra", "readxl", "stringr", "magick", "magrittr", "here")
 here()
@@
 
-[If]{.underline} you have already run the processing code, ensure you
-have set 'freshRun = FALSE' at top of this document to read in the
-existing .rds file for downstream processing.
-
-```{r, read acoustic study, include=FALSE, eval=!freshRun}
-data <- readRDS('satlantic_study.rds')
-# Double check warning messages
-print(getWarnings(data)$message)
-```
-
 [If]{.underline} this is the [initial processing]{.underline}, ensure
@@
 saveRDS(data, 'satlantic_study.rds')
+# Double check warning messages
+print(getWarnings(data)$message)
+```
+
+[If]{.underline} you have already run the processing code, ensure you
+have set 'freshRun = FALSE' at top of this document to read in the
+existing .rds file for downstream processing.
+
+```{r, read acoustic study, include=FALSE, eval=!freshRun}
+data <- readRDS(here('satlantic', 'satlantic_study.rds'))
 # Double check warning messages
@@
 
-```{r, assign species ID}
+```{r, assign species ID, eval=freshRun}
 GU1605 <- read_excel("GU1605_PG_OfflineEvents_20210809.xlsx", sheet = "BWonly")#read in data
@@
 
-```{r, filter unwanted data}
+```{r, filter unwanted data, eval=freshRun}
 goodSpecies<- c(c("Blainvilles", "Trues", "Cuviers", "Sowerbys", "Gervais"))
@@
 
-### *Model without ICI*
+### *EC (only) Model*
 
-1.  Load data, initialize a BANTER model, and create an initial BANTER
-    detector model (stage 1).
+Initialize, Run & Evaluate Detector Model (stage 1).
+```{r, banter ec detector model, eval=freshRun}
+banter_model_ec <- initBanterModel(banter_data$events)
+banter_model_ec <- addBanterDetector(banter_model_ec, banter_data$detectors, ntree=1e4, sampsize=2, importance = TRUE)
 
-```{r, read banter model, include=FALSE, eval=!freshRun}
-banter_model <- readRDS('satlantic_banter_model_t1e4s2_t1e5s1.rds')
+plotDetectorTrace(banter_model_ec, detector = paste0('Click_Detector_', 3:4))
+plotDetectorTrace(banter_model_ec, detector = paste0('Click_Detector_', 5:6))
+summary(banter_model_ec)
 ```
 
-```{r, banter detector model, eval=freshRun}
-banter_model <- initBanterModel(banter_data$events)
-banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=1e4, sampsize=2, importance = TRUE)
+Run BANTER Event Model (stage 2)
+```{r, banter ec event model, eval=freshRun}
+banter_model_ec <- runBanterModel(banter_model_ec, ntree=1e5, sampsize=1)
+summary(banter_model_ec)
 ```
 
-Check for stability (are more trees needed?) before moving on
-
-```{r, banter detector model stability}
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 3:4))
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 5:6))
-summary(banter_model)
-```
-
-2.  Run BANTER Event Model (stage 2)
-
-```{r, banter event model}
-banter_model <- runBanterModel(banter_model, ntree=1e5, sampsize=1)
-summary(banter_model)
-```
-
-Check for stability (are more trees needed?). If acceptable, save model
-with tree/sampsize info in the filename.
-
-```{r,save banter model, eval=freshRun}
-saveRDS(banter_model, 'satlantic_banter_model_t1e4s2_t1e5s1.rds')
+Once a stable model is identified, save model with tree/sampsize info in the filename.
+```{r,save banter ec model, eval=freshRun}
+saveRDS(banter_model_ec, 'satlantic_banter_model_ec_t1e4s2_t1e5s1.rds')
 ```
@@
 
-1.  Load data, initialize a BANTER model, and create an initial BANTER
-    detector model (stage 1).
-
-```{r, read banter ici model, include=FALSE, eval=!freshRun}
-banter_model_ici <- readRDS('satlantic_banter_model_ici_t1e4s4_t1e5s1.rds')
-```
-
+Initialize, Run & Evaluate Detector Model (stage 1)
 ```{r, banter ici detector model, eval=freshRun}
@@
 banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=1e4, sampsize=4, importance = TRUE)
-```
 
-Check for stability to see if we need more trees
-
-```{r, banter ici detector model stability}
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 3:4))
-plotDetectorTrace(banter_model, detector = paste0('Click_Detector_', 5:6))
+plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 3:4))
+plotDetectorTrace(banter_model_ici, detector = paste0('Click_Detector_', 5:6))
 summary(banter_model_ici)
@@
 
-2.  Run BANTER Event Model (stage 2)
-
-```{r, banter ici event model}
+Run BANTER Event Model (stage 2)
+```{r, banter ici event model, eval=freshRun}
 banter_model_ici <- runBanterModel(banter_model_ici, ntree=1e5, sampsize=1)
@@
 
-Check for stability (are more trees needed?). If acceptable, save model
-with tree/sampsize info in the filename.
+Once a stable model is identified, save model with tree/sampsize info in the filename.
+```{r, save banter ici model, eval=freshRun, eval=freshRun}
+saveRDS(banter_model_ici, 'satlantic_banter_model_ici_t1e4s4_t1e5s1.rds')
+```
 
-```{r, save banter ici model, eval=freshRun}
-saveRDS(banter_model_ici, 'satlantic_banter_model_ici_t1e4s4_t1e5s1.rds')
+### Alternative Model (EC_ICI_ALT) with larger sampsize
+
+Initialize, Run & Evaluate Detector Model (stage 1)
+```{r, banter alt detector model, eval=freshRun}
+banter_model_alt <- initBanterModel(banter_data_ici$events)
+banter_model_alt <- addBanterDetector(banter_model_alt, banter_data_ici$detectors, ntree=1e4, sampsize=4, importance = TRUE)
+
+plotDetectorTrace(banter_model_alt, detector = paste0('Click_Detector_', 3:4))
+plotDetectorTrace(banter_model_alt, detector = paste0('Click_Detector_', 5:6))
+summary(banter_model_alt)
+```
+
+Run BANTER Event Model (stage 2)
+```{r, banter alt event model, eval=freshRun}
+banter_model_alt <- runBanterModel(banter_model_alt, ntree=1e5, sampsize=2)
+summary(banter_model_alt)
+```
+
+Once a stable model is identified, save model with tree/sampsize info in the filename.
+```{r, save banter alt model, eval=freshRun, eval=freshRun}
+saveRDS(banter_model_alt, 'satlantic_banter_model_alt_t1e4s4_t1e5s2.rds')
 ```
@@
 
-First, identify the model you would like to examine (comment out the
+First, load the models (if not a fresh run)
+```{r, read banter models, include=FALSE, eval=!freshRun}
+banter_model_ec <- readRDS(here('satlantic', 'satlantic_banter_model_ec_t1e4s2_t1e5s1.rds'))
+banter_model_ici <- readRDS(here('satlantic', 'satlantic_banter_model_ici_t1e4s4_t1e5s1.rds'))
+banter_model_alt <- readRDS(here('satlantic', 'satlantic_banter_model_alt_t1e4s4_t1e5s2.rds'))
+```
+
+Identify the model you would like to examine (comment out the
 model you [do not]{.underline} want to examine).
@@
 ```{r, extract RF model, include=TRUE}
-banter_model_RF <- getBanterModel(model)
+banter_model_ec_RF <- getBanterModel(model_ec)
 banter_model_ici_RF <- getBanterModel(model_ici)
+banter_model_alt_RF <- getBanterModel(model_alt)
+```
+
+Class Priors (Expected Error Rate)
+
+```{r, class priors, include=TRUE}
+satlantic_ec_priors <- classPriors(banter_model_ec_RF, NULL)[,1]
+satlantic_ici_priors <- classPriors(banter_model_ici_RF, NULL)[,1]
+satlantic_alt_priors <- classPriors(banter_model_alt_RF, NULL)[,1]
 ```
@@
 ```{r, confusion matrix, include=TRUE}
-satlantic_confuseMatrix <- rfPermute::confusionMatrix(banter_model_RF)
-satlantic_confuseMatrix <- kable(satlantic_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+satlantic_ec_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ec_RF)
+satlantic_ec_confuseMatrix <- cbind(satlantic_ec_confuseMatrix, priors = satlantic_ec_priors)
+satlantic_ec_confuseMatrix <- kable(satlantic_ec_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
   kable_classic()%>%
-  column_spec(4, border_right = TRUE)%>%
+  column_spec(6, border_right = TRUE)%>%
   row_spec(0, bold = TRUE)%>%
   row_spec(3,hline_after = TRUE)%>%
-  row_spec(4, bold = TRUE)%>%
-  save_kable(paste(modelname, "satlantic_confuseMatrix.png", sep="_"), zoom = 2)
+  row_spec(6, bold = TRUE)%>%
+  save_kable('../manuscript/manuscript_files/satlantic_ec_confuseMatrix.png', zoom = 9)
 
 satlantic_ici_confuseMatrix <- rfPermute::confusionMatrix(banter_model_ici_RF)
+satlantic_ici_confuseMatrix <- cbind(satlantic_ici_confuseMatrix, priors = satlantic_ici_priors)
 satlantic_ici_confuseMatrix <- kable(satlantic_ici_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
   kable_classic()%>%
-  column_spec(4, border_right = TRUE)%>%
+  column_spec(6, border_right = TRUE)%>%
   row_spec(0, bold = TRUE)%>%
   row_spec(3,hline_after = TRUE)%>%
-  row_spec(4, bold = TRUE)%>%
-  save_kable(paste(modelname_ici, "satlantic_ici_confuseMatrix.png", sep="_"), zoom = 2)
+  row_spec(6, bold = TRUE)%>%
+  save_kable('../manuscript/manuscript_files/satlantic_ici_confuseMatrix.png', zoom = 9)
+
+satlantic_alt_confuseMatrix <- rfPermute::confusionMatrix(banter_model_alt_RF)
+satlantic_alt_confuseMatrix <- cbind(satlantic_alt_confuseMatrix, priors = satlantic_alt_priors)
+satlantic_alt_confuseMatrix <- kable(satlantic_alt_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(5, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(3,hline_after = TRUE)%>%
+  row_spec(5, bold = TRUE)%>%
+  save_kable('../manuscript/manuscript_files/satlantic_alt_confuseMatrix.png', zoom = 9)
 
@@
 
-![](banter_model_satlantic_confuseMatrix.png)
+![BANTER Model S.Atlantic EC Confusion
+Matrix](../manuscript/manuscript_files/satlantic_ec_confuseMatrix.png)
+
+![BANTER Model S.Atlantic ICI Confusion
+Matrix](../manuscript/manuscript_files/satlantic_ici_confuseMatrix.png)
+
+![BANTER Model S.Atlantic Alternative Model Confusion
+Matrix](../manuscript/manuscript_files/satlantic_alt_confuseMatrix.png)
 
@@
 
-```{r, proximity plot, include=TRUE}
-png(paste(modelname, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
-satlantic_proximityPlot <- plotProximity(banter_model_RF)
+```{r, proximity plot, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/satlantic_ec_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
+satlantic_ec_proximityPlot <- plotProximity(banter_model_ec_RF)
 dev.off()
-satlantic_proximityPlot <- plotProximity(banter_model_RF)
+satlantic_ec_proximityPlot <- plotProximity(banter_model_ec_RF)
 
-png(paste(modelname_ici, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
+png(('../manuscript/manuscript_files/satlantic_ici_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
 ici_satlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
@@
 ici_satlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
+
+png(('../manuscript/manuscript_files/satlantic_alt_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
+alt_satlantic_proximityPlot <- plotProximity(banter_model_alt_RF)
+dev.off()
+alt_satlantic_proximityPlot <- plotProximity(banter_model_alt_RF)
 ```
@@
 
-```{r, importance heat map, include=TRUE}
-png(paste(modelname, "satlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_RF, plot.type="heatmap")
+```{r, importance heat map, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/satlantic_ec_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
+plotImportance(banter_model_ec_RF, plot.type="heatmap")
 dev.off()
-satlantic_importance <- plotImportance(banter_model_RF, plot.type="heatmap")
+satlantic_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap")
 
-png(paste(modelname_ici, "satlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
+png(('../manuscript/manuscript_files/satlantic_ici_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
 plotImportance(banter_model_ici_RF, plot.type="heatmap")
@@
 ici_satlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+
+png(('../manuscript/manuscript_files/satlantic_alt_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
+plotImportance(banter_model_alt_RF, plot.type="heatmap")
+dev.off()
+alt_satlantic_importance <- plotImportance(banter_model_alt_RF, plot.type="heatmap")
 ```
@@
 
-```{r, plot votes, include=TRUE}
-png(paste(modelname, "satlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_RF)
+```{r, plot votes, include=TRUE, results=FALSE}
+png(('../manuscript/manuscript_files/satlantic_ec_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_ec_RF)
 dev.off()
-satlantic_votes <- plotVotes(banter_model_RF)
+satlantic_votes <- plotVotes(banter_model_ec_RF)
 
-png(paste(modelname_ici, "satlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
+png(('../manuscript/manuscript_files/satlantic_ici_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
 plotVotes(banter_model_ici_RF)
@@
 ici_satlantic_votes <- plotVotes(banter_model_ici_RF)
-```
 
-Class Priors (Expected Error Rate)
-
-```{r, class priors, include=TRUE}
-satlantic_priors <- classPriors(banter_model_RF, NULL)
-```
-
-Plot Confusion Matrix Heat Map
-
-```{r, confusion matrix heat map, include=TRUE}
-plotConfMat(banter_model_RF, title="ConfusionMatrix", plot=TRUE)
-```
-
-Plot Error Trace for Banter Model
-
-```{r, model error trace, include=TRUE}
-plot(banter_model_RF)
+png(('../manuscript/manuscript_files/satlantic_alt_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_alt_RF)
+dev.off()
+alt_satlantic_votes <- plotVotes(banter_model_alt_RF)
 ```
@@
 
-```{r, predicted probabilities, include=TRUE}
-plotPredictedProbs(banter_model_RF, bins=30, plot=TRUE)
+```{r, predicted probabilities, include=TRUE, results=FALSE}
+plotPredictedProbs(banter_model_ec_RF, bins=30, plot=TRUE)
+plotPredictedProbs(banter_model_ici_RF, bins=30, plot=TRUE)
+plotPredictedProbs(banter_model_alt_RF, bins=30, plot=TRUE)
 ```
+
+Create Figure for Publication
+```{r, Publication Figure, include=TRUE}
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+satlantic_ec_Figure <-image_append(c(prox, heat, vote))
+satlantic_ec_Figure<- image_append(c(confuse, satlantic_ec_Figure), stack=TRUE)
+image_write(satlantic_ec_Figure, path = here('manuscript', 'manuscript_files','satlantic_ec_Figure.png'), format ='png')
+print(satlantic_ec_Figure, info=FALSE)
+
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+satlantic_ici_Figure <-image_append(c(prox, heat, vote))
+satlantic_ici_Figure<- image_append(c(confuse, satlantic_ici_Figure), stack=TRUE)
+image_write(satlantic_ici_Figure, path = here('manuscript', 'manuscript_files','satlantic_ici_Figure.png'), format ='png')
+print(satlantic_ici_Figure, info=FALSE)
+
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+satlantic_alt_Figure <-image_append(c(prox, heat, vote))
+satlantic_alt_Figure<- image_append(c(confuse, satlantic_alt_Figure), stack=TRUE)
+image_write(satlantic_alt_Figure, path = here('manuscript', 'manuscript_files','satlantic_alt_Figure.png'), format ='png')
+print(satlantic_alt_Figure, info=FALSE)
+```

### 2023-06-08 epacific/AnnoMatePaperCode.R

@@ -1,0 +1,35 @@
+# load appropriate programs
+library (PAMpal)
+library (PAMmisc)
+library (rfigshare)
+
+#Authenticate Figshare access, if needed
+rfigshare::fs_auth() #authenticate Figshare access, if needed
+
+#get information on repository files for https://figshare.com/account/projects/94511/articles/23319938
+data.details <- rfigshare::fs_details(23319938) 
+#download acoustics study from Figshare
+data <- download.file( url = file.path("https://ndownloader.figshare.com/files", data.details$files[[3]]$id), destfile = file.path(data.dir, data.details$files[[3]]$name))
+
+# reassign species names for Ann-O-Mate
+specMap <- data.frame(
+  old = unique(species(data)),
+  new = NA)
+specMap$new <- c('Ziphius cavirostris', 'possible Ziphiidae', 'Ziphiidae', 
+                 'Mesoplodon carlhubbsi', 'BW43', 'Berardius bairdii', 
+                 'Cross Seamount beaked whale', 'Mesoplodon stejnegeri')
+
+# save beaked whale data as myData
+anno <- prepAnnotation(data, specMap = specMap, mode='event',
+                       source = 'figshare.com',
+                       source_id = 22959786, 
+                       annotator = 'Anne Simonis, https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036',
+                       annotation_id = anno$event,
+                       annotation_date = 2019,
+                       annotation_info_url = 'https://figshare.com/articles/dataset/BWEventClips_CCES2018/22959786',
+                       recording_info_url = 'https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036',
+                       type = 'echolocation click event',
+                       contact = 'shannon.rankin@noaa.gov')
+
+
+

### 2023-06-08 epacific/annomate.R

@@ -1,0 +1,53 @@
+#### Annomate Example ####
+# Updated 2023-02-06
+# requires PAMpal 0.18.0 and PAMmisc 1.11.0
+
+# data must have GPS and species assigned
+library(PAMpal)
+library(here)
+here()
+
+data <- readRDS('epacific/epacific_study.rds')
+specMap <- data.frame(
+  old = unique(species(data)),
+  new = NA)
+specMap$new <- c('Ziphius cavirostris', 'possible Ziphiidae', 'Ziphiidae', 
+                 'Mesoplodon carlhubbsi', 'BW43', 'Berardius bairdii', 
+                 'Cross Seamount beaked whale', 'Mesoplodon stejnegeri')
+anno <- prepAnnotation(data, specMap = specMap, mode='event',
+                       # following args are optional, fill in
+                       # any that will be the same for all entries
+                       source = 'figshare.com',
+                       contact = 'shannon.rankin@noaa.gov')
+# or you can add them later
+anno$source = 'figshare.com'
+anno$source_id = 22959786
+anno$annotator = 'Anne Simonis, https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036'
+anno$annotation_id = anno$event
+anno$annotation_date = 2019
+anno$annotation_info_url = 'https://figshare.com/articles/dataset/BWEventClips_CCES2018/22959786'
+anno$recording_info_url = 'https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036'
+anno$type = 'echolocation click event'
+
+library(PAMmisc)
+# to get figshare data you will need the article id and your personal token
+# i recommend storing your token in an external file instead of typing it in here
+# also DO NOT commit this file to any repository
+figToken <- readRDS(file = "epacific/figtoken.rds")
+figId <- 22959786
+figshareData <- getFigshareInfo(figToken, figId)
+
+anno <- matchRecordingUrl(anno, figshareData)
+# will give you warnings and messages about missing fields
+checkAnnotation(anno)
+# if you need to fix any manually, write to csv then read back in
+annoFile <- 'AnnomateExport.csv'
+write.csv(anno, file=annoFile, row.names = FALSE)
+# go fix stuff
+anno <- read.csv(annoFile, stringsAsFactors = FALSE)
+# recheck to see if all seems fine
+checkAnnotation(anno)
+# add to acoustic study for storage
+data <- addAnnotation(data, anno)
[…]

### 2023-06-12 manuscript/manuscript.qmd
Edits
Modify according to edits by Taiki Sakai

@@
 In this case, settings for a whistle & moan detector were modified to improve performance at detecting burst pulses.
-This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the overall number of detections was different.
+This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the results were different.
 This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn_acoustic_2021].
@@
 All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c, 2.00.16e) with a suite of generic click classifiers within the Click Detector module (see [@keating_summary_2013]).  Classifier sets were saved such that any click may be classified as more than one click type.
-The general click classifiers were effectively click detectors, as each implemented a simple spectral band settings (e.g., 2 -- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz).
+The general click classifiers were treated as spectral band click detectors (e.g., 2 -- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz).
 An additional detector within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see [@keating_summary_2013]).
@@
 Acoustic detection of beaked whale events were then classified to species by experienced acoustic researchers using multiple lines of evidence.
-If beaked whales were sighted in close proximity to the recording device, the visual confirmation of species identify was used to determine species identity.
-Otherwise, experienced acousticians followed strict protocol to define species identify based on acoustic characteristics.
+If beaked whales were sighted in close proximity to the recording device, the visual confirmation of species identity was used to determine species identity.
+Otherwise, experienced acousticians followed strict protocol to define species identity based on acoustic characteristics.
 When the available characteristics were inconclusive, the species was considered an unidentified beaked whale; only acoustic events classified to species were considered for this study.
@@
 For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see [@deangelis_description_2018]).
-Data were saved within the PAMGuard database and binaries for downstream processing using PAMpal package [@sakai_pampal_2021] in R programming language [@r_core_team_r_2022].
+Data were saved within the PAMGuard database and binaries for downstream processing using the PAMpal package [@sakai_pampal_2021] in the R programming language [@r_core_team_r_2022].
 
@@
 
-BANTER models were created for each trial using the BANTER package in R programming language [@archer_banter_2022].
+BANTER models were created for each trial using the BANTER package in the R programming language [@archer_banter_2022].
 Standard click calculations were used for the BANTER call classifier; event level measures of ICI and environmental variables were applied to the BANTER event classifier.
@@
 
-The model output include the confusion matrix, and a number of visualizations provided by the rfpermute package in R programming language [@archer_rfpermute_2022].
+The model output includes the confusion matrix, and a number of visualizations provided by the rfpermute package in the R programming language [@archer_rfpermute_2022].
 
@@
 The results for the four study areas varied from a low of 87.1% correct classification in the SAtlantic dataset to a high of 98% correct classification in the NAtlantic dataset.
-All species in all datasets had classifications rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais beaked whales (NAtlantic) had classification rates below 90%.
+All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais beaked whales (NAtlantic) had classification rates below 90%.
 
@@
 While small sample size can at times provide good classification scores, as we found in Sowerby's beaked whale in this SAtlantic dataset, they can be devastating if there are inaccuracies in the training data.
-These two examples of small sample sizes highlights a conflict in preferred protocol.
+These two examples of small sample sizes highlight a conflict in preferred protocol.
 Ideally, (1) all [species]{.underline} would be included in a classification model, so as to better represent the local species diversity, (2) all [events]{.underline} would ideally be included in the classification model, so as to better represent the variability found in the area, and (3) only confident 'ground truth' classifications would be considered in the training model.
@@
 Interclick Interval (ICI) was found to be the most important variable for NAtlantic and EPacific, and ICI were the third and fourth ranked variable for SAtlantic (see Importance Heat Maps for each survey area).
-While the addition of ICI did not improve results Hawaii, results were similar and ICI ranked #9 in importance for the Hawaii study area.
+While the addition of ICI did not improve results for Hawaii, results were similar and ICI ranked #9 in importance for the Hawaii study area.
 There was a difference in how ICI was calculated for these surveys, and it is possible these differences may affect the value of the ICI variable in species classification for some species.
@@
 BANTER is a supervised machine learning tool and requires reliable training data for success.
-Training data should consist of labels with strong confidence in species identify (ideally determined by agreement from more than one analyst) and sample sizes should be large enough to explain the natural variability in the data.
+Training data should consist of labels with strong confidence in species identity (ideally determined by agreement from more than one analyst) and sample sizes should be large enough to explain the natural variability in the data.
 
@@
 Sea surface temperature data were provided by JPL under support by NASA MEaSUREs program.
-Funding for data collection and analysis were provided by U.S.
+Funding for data collection and analysis were provided by the U.S.
 Navy, Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration.
@@
 #| fig-cap: BANTER classification results from the NAtlantic dataset including echolocation clicks and interclick interval (EC_ICI).
-Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
 Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
@@
 #| fig-cap: BANTER classification results from the SAtlantic dataset (EC_ICI).
-Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
 Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
@@
 #| fig-cap: BANTER classification results from the Hawaii dataset (EC).
-Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
 Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
@@
 #| fig-cap: BANTER classification results from the EPacific dataset with environmental data (EC_ICI_ENV).
-Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
 Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).

### 2023-06-12 epacific/annomate.R

@@ -7,6 +7,7 @@ library(PAMpal)
 library(here)
 here()
 
+
 data <- readRDS('epacific/epacific_study.rds')
 specMap <- data.frame(
   old = unique(species(data)),
@@ -14,20 +15,18 @@ specMap <- data.frame(
 specMap$new <- c('Ziphius cavirostris', 'possible Ziphiidae', 'Ziphiidae', 
                  'Mesoplodon carlhubbsi', 'BW43', 'Berardius bairdii', 
                  'Cross Seamount beaked whale', 'Mesoplodon stejnegeri')
+
 anno <- prepAnnotation(data, specMap = specMap, mode='event',
-                       # following args are optional, fill in
-                       # any that will be the same for all entries
                        source = 'figshare.com',
+                       source_id = 22959786, 
+                       annotator = 'Anne Simonis, Jenny Trickey',
+                       annotation_date = 2019,
+                       annotation_info_url = 'https://figshare.com/articles/dataset/BWEventClips_CCES2018/22959786',
+                       recording_info_url = 'https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036',
+                       type = 'echolocation click event',
                        contact = 'shannon.rankin@noaa.gov')
-# or you can add them later
-anno$source = 'figshare.com'
-anno$source_id = 22959786
-anno$annotator = 'Anne Simonis, https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036'
+
 anno$annotation_id = anno$event
-anno$annotation_date = 2019
-anno$annotation_info_url = 'https://figshare.com/articles/dataset/BWEventClips_CCES2018/22959786'
-anno$recording_info_url = 'https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036'
-anno$type = 'echolocation click event'
 
 library(PAMmisc)
 # to get figshare data you will need the article id and your personal token
@@ -41,7 +40,7 @@ anno <- matchRecordingUrl(anno, figshareData)
 # will give you warnings and messages about missing fields
 checkAnnotation(anno)
 # if you need to fix any manually, write to csv then read back in
-annoFile <- 'AnnomateExport.csv'
+annoFile <- 'AnnomateExport_epacific_rankin2023.csv'
 write.csv(anno, file=annoFile, row.names = FALSE)
 # go fix stuff
 anno <- read.csv(annoFile, stringsAsFactors = FALSE)

### 2023-08-21 epacific/beakerBanter_epacific.Rmd
manuscript update
Updated Manuscript

@@
 banter_data <- export_banter(data_ch1only, dropSpecies = c('BW', 'possBW'), 
-                         dropVars = c('All_ici', 'Click_Detector_3_ici', 'sea_floor_depth_mean', 'sst_mean', 'productivity_mean', 'chlorophyll_mean',  'magnitude_gradient_mean'), training=TRUE)
+                         dropVars = c('All_ici', 'Click_Detector_3_ici', 'sea_floor_depth_mean', 'sst_mean', 'productivity_mean', 'chlorophyll_mean',  'magnitude_gradient_mean', 'Latitude', 'Longitude', 'gpsUncertainty'), training=TRUE)
 saveRDS(banter_data, file='epacific_banter_data.rds')
 
-banter_data_ici <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), dropVars = c('sea_floor_depth_mean', 'sst_mean', 'productivity_mean', 'chlorophyll_mean',  'magnitude_gradient_mean'),training=TRUE)
+banter_data_ici <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), dropVars = c('sea_floor_depth_mean', 'sst_mean', 'productivity_mean', 'chlorophyll_mean',  'magnitude_gradient_mean','Latitude', 'Longitude', 'gpsUncertainty'),training=TRUE)
 saveRDS(banter_data_ici, file='epacific_banter_data_ici.rds')
 
-banter_data_env <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), training=TRUE)
+banter_data_env <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), dropVars = c('Latitude', 'Longitude', 'gpsUncertainty'), training=TRUE)
 saveRDS(banter_data_env, file='epacific_banter_data_env.rds')
@@
 banter_model_ec <- initBanterModel(banter_data$events)
-banter_model_ec <- addBanterDetector(banter_model_ec, banter_data$detectors, ntree=1e4, sampsize=3, importance = TRUE)
+banter_model_ec <- addBanterDetector(banter_model_ec, banter_data$detectors, ntree=1e4, sampsize=4, importance = TRUE)
 
@@
 ```{r,save banter ec model, eval=freshRun}
-saveRDS(banter_model_ec, 'epacific_banter_model_ec_t1e4s3_t1e4s4.rds')
+saveRDS(banter_model_ec, 'epacific_banter_model_ec_t1e4s4_t1e4s4.rds')
 ```
@@
 ```{r, read banter model, include=FALSE, eval=!freshRun}
-banter_model_ec <- readRDS('epacific_banter_model_ec_t1e4s3_t1e4s4.rds')
+banter_model_ec <- readRDS('epacific_banter_model_ec_t1e4s4_t1e4s4.rds')
 banter_model_ici <- readRDS('epacific_banter_model_ici_t1e4s3_t1e4s4.rds')
@@
 png(('../manuscript/manuscript_files/epacific_ec_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_ec_RF, plot.type="heatmap")
+plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
 dev.off()
-ec_epacific_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap")
+ec_epacific_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
 
 png(('../manuscript/manuscript_files/epacific_ici_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_ici_RF, plot.type="heatmap")
+plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
 dev.off()
-ici_epacific_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_epacific_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
 
 png(('../manuscript/manuscript_files/epacific_env_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_env_RF, plot.type="heatmap")
+plotImportance(banter_model_env_RF, plot.type="heatmap", n=10)
 dev.off()
-env_epacific_importance <- plotImportance(banter_model_env_RF, plot.type="heatmap")
+env_epacific_importance <- plotImportance(banter_model_env_RF, plot.type="heatmap", n=10)
 ```
@@
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_importance.png'))%>%
@@
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=300, color = "black")
 
@@
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_importance.png'))%>%
@@
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_importance.png'))%>%

### 2023-08-21 hawaii/beakerBanter_hawaii.Rmd
manuscript update
Updated Manuscript

@@
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_importance.png'))%>%
@@
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_importance.png'))%>%

### 2023-08-21 manuscript/Supplement.qmd
manuscript update
Updated Manuscript

@@
 Detector Model:  ntree=1e4, sampsize=5; 
-Event Model: ntree=1e4, sampsize=5
+Event Model: ntree=1e5, sampsize=5
 ```{r, NAtlantic EC_ICI_ALT Model Figure,  echo=FALSE}
@@
 ## EC Model
-Detector Model:  ntree=1e4, sampsize=2; 
+Detector Model:  ntree=1e4, sampsize=8; 
 Event Model: ntree=1e5, sampsize=1
@@
 Detector Model:  ntree=1e4, sampsize=4; 
-Event Model: ntree=1e5, sampsize=2
+Event Model: ntree=1e5, sampsize=3
 ```{r, SAtlantic EC_ICI_ALT Model Figure,  echo=FALSE}
@@
 ```
+
+## EC_ICI Model
+Detector Model:  ntree=1e4, sampsize=3; 
+Event Model: ntree=1e4, sampsize=4
+
+```{r, EPacific EC ICI Model Figure,  echo=FALSE}
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+epacific_ici_Figure <-image_append(c(prox, heat, vote))
+epacific_ici_Figure<- image_append(c(confuse, epacific_ici_Figure), stack=TRUE)
+print(epacific_ici_Figure, info=FALSE)
+```

### 2023-08-21 manuscript/manuscript.qmd
manuscript update
Updated Manuscript

@@
 ---
-title: Open-Source machine learning BANTER acoustic classification of a cryptic echolocating species
+title: Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses
 author:
@@
           postal-code: 02543
-  - name: Annamaria DeAngelis
+  - name: Annamaria I. DeAngelis
     email: annamaria.deangelis@noaa.gov
@@
           postal-code: 33143
-  - name: Jenny Trickey
+  - name: Jennifer S. Trickey
     email: jenny.trickey@noaa.gov
@@
   Passive acoustic monitoring is increasingly used for assessing populations of marine mammals; however, analysis of large datasets is limited by our ability to easily classify sounds detected.
-Classification of beaked whale acoustic events, in particular, require evaluation of multiple lines of evidence by expert analysts.
+Classification of beaked whale acoustic events, in particular, requires evaluation of multiple lines of evidence by expert analysts.
 Here we present a highly automated approach to acoustic detection and classification using supervised machine learning and open source software methods.
-Data from four large scale surveys of beaked whales (North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and BANTER (hierarchical random forest classifier).
-Overall correct classification results ranged from 87% for the South Atlantic data to 98% for the North Atlantic.
+Data from four large scale surveys of beaked whales (northwestern North Atlantic, southwestern North Atlantic, Hawaii, and eastern North Pacific) were analyzed using PAMGuard (acoustic detection), PAMpal (acoustic analysis) and BANTER (hierarchical random forest classifier).
+Overall classification accuracy ranged from 88% for the southwestern North Atlantic data to 97% for the northwestern North Atlantic.
 Results for many species could likely be improved with increased sample sizes, consideration of alternative automated detectors, and addition of relevant environmental features.
-These methods provide a highly automated approach to acoustic detection and classification using open source methods that can be readily adopted for species and geographic regions.
+These methods provide a highly automated approach to acoustic detection and classification using open source methods that can be readily adopted for other species and geographic regions.
 keywords: 
@@
 link-citations: TRUE
-format:
-  pdf: default
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Ecological Informatics
-      formatting: preprint
-      model: 1p
-      cite-style: authoryear
+format: docx
+
+#  pdf: default
+#  elsevier-pdf:
+#    keep-tex: true
+#    journal:
+#      name: Ecological Informatics
+#      formatting: preprint
+#      model: 1p
+#      cite-style: authoryear
 ---
@@
 Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring.
-Historically, experienced acousticians manually identify stereotyped sounds that could be reliably attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008].
-The dramatic increase in recordings make it impossible for experienced acousticians to manually annotate all data, and there has been an increasing need for the development of automated classification routines that can provide accurate determinations of the species responsible for the sounds.
+Historically, experienced acousticians have manually identified stereotyped sounds that could reliably be attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008].
+However, the dramatic increase in the quantity of recordings being made make it impossible for experienced acousticians to manually annotate it all.
+This has led to an increased need for the development of automated classification routines that can provide accurate species determinations from acoustic recordings.
 
 Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod_beaked_2018].
-However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_spatio-temporal_2014].
+However, beaked whales make stereotyped echolocation signals in the form of frequency-modulated (FM) pulses that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_species-specific_2013].
 Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022]; [@baumann-pickering_spatio-temporal_2014]; [@simonis_passive_2020].
-Unfortunately, manual classification of beaked whale echolocation clicks requires analysis by trained observers using a number of visual aids to examine the call characteristics.
-Development of automated classification routines, if accurate, serve to improve the efficiency and decrease the cost of analyzing large datasets.
+Unfortunately, manual classification of beaked whale echolocation pulses requires analysis by trained analysts using a number of visual aids to examine the call characteristics.
+Development of automated classification routines, if accurate, serves to improve the efficiency, reduce the subjectivity, and decrease the cost of analyzing large datasets.
 
-BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings [@rankin_acoustic_2017].
-This hierarchical random forest event classifier relies on an initial call classifier for each call type (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier.
-The event classifier considers the probability of assignment to species for each call in the call classifier, along with any event-level characteristics (such as call rate).
+BANTER (**B**io-**A**coustic eve**NT** classifi**ER**) is a supervised machine learning method originally developed for species identification of dolphin signals from acoustic recordings [@rankin_acoustic_2017].
+This hierarchical Random Forest [@breiman_random_2001] event classifier combines information from independent classification models for each call type (e.g., whistles, echolocation pulses, and burst pulses for the dolphin-based model) into a model classifying a full acoustic event, which is defined as a discrete collection of calls during an encounter.
+The event classifier uses the distribution of species assignment probabilities for each call type, along with any event-level metrics (such as call rate).
 BANTER is very flexible and can accommodate any number of measures from any number of detectors.
-While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results [@rankin_acoustic_2017].
-In this case, settings for a whistle & moan detector were modified to improve performance at detecting burst pulses.
-This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the results were different.
-This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn_acoustic_2021].
-For species who primarily (or exclusively) produce echolocation clicks, BANTER may serve as an option for automated machine learning classification.
 
-Here we applied BANTER acoustic classification to beaked whale detections from four large acoustic datasets including the US East Coast (north and south), Hawaii, and the US West Coast.
-Acoustic data were analyzed by experienced acousticians to determine species identity, and these 'ground truth' classifications served as training data for the supervised BANTER trials.
-Our goals are to identify an efficient and accurate automated approach to acoustic classification of beaked whales, and to provide a framework for analysis that may serve for other PAM studies.
+While BANTER was designed to classify events by integrating information from multiple types of call detectors, relatively minor changes to a detector for a specific call type could yield different detector results.
+In the dolphin-based model [@rankin_acoustic_2017], settings for a whistle and moan detector were modified to improve performance at detecting burst pulses.
+This approach of applying output from multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type is to be analyzed), as long as the settings of the automated call detector were modified such that the results were different.
+This approach was successfully applied to classified echolocation pulses for narwhals and belugas [@zahn_acoustic_2021].
+For species that primarily (or exclusively) produce echolocation pulses, BANTER may serve as an option for automated machine learning classification.
+
+Here we applied BANTER acoustic classification to beaked whale detections from large acoustic datasets from the U.S.
+East Coast, Hawaii, and the U.S.
+West Coast.
+Acoustic data were analyzed by experienced acousticians to determine species identity, and these 'ground truth' classifications served as training data for the supervised BANTER models.
+Our goals were to identify an efficient and accurate automated approach to acoustic species identification of beaked whales, and to provide a framework for analysis that may serve for other PAM studies.
 
@@
 
-Passive acoustic data was collected during three surveys conducted by National Oceanographic and Atmospheric Administrations' research operations off the US East Coast, the Hawaiian Islands (Hawaii) and the US West Coast.
+Passive acoustic data were collected during surveys conducted by the National Oceanographic and Atmospheric Administrations' (NOAA) research operations off the U.S.
+East Coast, the Hawaiian Islands, and the U.S.
+West Coast.
 
-Recordings from the western North Atlantic Ocean (U.S.
-east coast) were collected using towed hydrophone arrays during the 2016 AMAPPS II (Atlantic Marine Assessment Program for Protected Species) survey (cite).
-AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey area.
-The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (HTI-96-min, High Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see [@deangelis_description_2018] for more information).
-The SAtlantic area ranged from Delaware south to central Florida (GU1605, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (Reson \[MODEL\], Teledyne Marine, Slangerup, Denmark) recorded 500 kHz sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass filter (see \[CITE\] for more information).
-Variation in hydrophone sensitivities precluded combining the NATlantic and SAtlantic datasets.
+Recordings from the U.S.
+East Coast (western North Atlantic Ocean) were collected using towed hydrophone arrays during the 2016 AMAPPS II (Atlantic Marine Assessment Program for Protected Species) survey by Northeast Fisheries Science Center and Southeast Fisheries Science Center [@united_states_national_marine_fisheries_service_2016_2016].
+AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey areas.
+The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, Appendix A, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (HTI-96-min, High Tech Inc., Long Beach, MS) which recorded at a 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see @deangelis_description_2018 for more information).
+The SAtlantic area ranged from Delaware south to central Florida (GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (Reson TC4013, Teledyne Marine, Slangerup, Denmark) which recorded at a 500 kHz sample rate (custom 12 channel SailDAQ soundcard) and were decimated to 192 kHz with a 1 kHz high pass filter (see GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016] for more information).
+The NAtlantic and SAtlantic recordings were treated as two separate datasets due to differences in hydrophone frequency responses and sensitivities.
 
-Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, [@yano_cetacean_2018]).  Drifting acoustic recorders were deployed at randomly selected locations within the main Hawaiian Islands.
-Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA).
-Soundtrap recorders were sampled at 288 kHz with a duty cycle of 2 min on for every 10 minutes.
-SM3M data were sampled at 256 kHz with continuous recordings.
-Individual buoys were deployed for 10 to 50 days (see [@yano_cetacean_2018] for more detailed information).
+Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, @mccullough_acoustic_2021, @yano_cetacean_2018).
+Drifting acoustic recorders were deployed at randomly selected locations within the Main Hawaiian Islands.
+Data were collected using SoundTrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) sampled at 288 kHz with a duty cycle of 2 min on for every 10 minutes.
+Individual drifting recorders were deployed for 11 to 23 days (see @mccullough_acoustic_2021, @yano_cetacean_2018 for more detailed information).
 
-Recordings from the Eastern North Pacific Ocean off the U.S.
-west coast (EPacific) were collected using drifting acoustic recorders during the 2018 CCES (California Current Ecosystem Survey) survey [@simonis_passive_2020].
-Buoys were deployed at 23 stations distributed throughout the California Current region; data from 15 of these included high quality acoustic data used in this analysis.
-Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA).
-Recordings were duty cycled and recorded 2 minutes out of a variable 'off' time; recorders sampled at a minimum 256 sample rate.
-Individual buoys were deployed for 5-79 days (see [@simonis_passive_2020] for more detailed information).
+Recordings from the eastern North Pacific Ocean off the U.S.
+West Coast (EPacific) were collected using drifting acoustic recorders during the 2018 CCES (California Current Ecosystem Survey) [@simonis_passive_2020].
+Drifting recorders were deployed at 23 stations distributed throughout the California Current region; data from 15 of these included high quality acoustic data used in this analysis.
+Two types of drifting recorders were used: (1) SoundTrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA).
+Recordings were duty cycled and recorded 2 minutes out of a variable 'off' time; recorders sampled at a minimum 256 kHz sample rate.
+Individual drifting recorders were deployed for 5-79 days (see @simonis_passive_2020 for more detailed information).
 
@@
 
-All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c, 2.00.16e) with a suite of generic click classifiers within the Click Detector module (see [@keating_summary_2013]).  Classifier sets were saved such that any click may be classified as more than one click type.
-The general click classifiers were treated as spectral band click detectors (e.g., 2 -- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz).
-An additional detector within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see [@keating_summary_2013]).
+All acoustic recordings were analyzed using PAMGuard software (versions 2.00.15c, 2.00.16e) with a suite of generic click classifiers within the Click Detector module (see @keating_summary_2013).
+Classifier sets were saved such that any click may be classified as more than one click type (using 'save classifier set' in Pamguard click classification window).
+Five general click classifiers were treated as spectral band click detectors (i.e., 2 -- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz).
+An additional sixth detector within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see @keating_summary_2013).
 
-Acoustic detection of beaked whale events were then classified to species by experienced acoustic researchers using multiple lines of evidence.
-If beaked whales were sighted in close proximity to the recording device, the visual confirmation of species identity was used to determine species identity.
-Otherwise, experienced acousticians followed strict protocol to define species identity based on acoustic characteristics.
-When the available characteristics were inconclusive, the species was considered an unidentified beaked whale; only acoustic events classified to species were considered for this study.
-Acoustic researchers with experience in classifying beaked whale echolocation clicks examined the bearing-time display for the click detector module in PAMGuard to assess the species identity of the acoustic event.
-For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see [@deangelis_description_2018]).
-Data were saved within the PAMGuard database and binaries for downstream processing using the PAMpal package [@sakai_pampal_2021] in the R programming language [@r_core_team_r_2022].
+Echolocation signals that were identified as frequency-modulated pulses from beaked whales were then linked as PAMGuard "events" by experienced acoustic researchers.
+Events included all the pulses that were close to each other in time, had similar spectral and waveform characteristics, and were received at consistent bearing angles.
+For NAtlantic and SAtlantic datasets, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into putative individuals by identification of consecutive pulses along the same bearing angle in the bearing time plot (see @deangelis_using_2017).
 
-Standard click calculations were measured from echolocation clicks using default values in PAMpal.
-Inter-click interval (ICI) was calculated using the 'calculateICI' function in PAMpal; the mode of the ICI was identified for each beaked whale acoustic event.
-In addition, for EPacific data, we used the ‘matchEnvData’ function in PAMpal to include ERDAPP environmental data (@simons_erddap_2022), including seafloor depth and seafloor gradient (both from erdSrtm30plusSeafloorGradient), as well as sea surface temperature (from jplMURSST41).
+Experienced acoustic researchers used multiple lines of evidence (PAMGuard click detector bearing time display, click waveform display, click spectrum, and wigner plots) to assess the species identity of the acoustic event.
+When the available characteristics were inconclusive, the species was considered an unidentified beaked whale.
+Data were saved within the PAMGuard database and binaries for downstream processing using the *PAMpal package* [@sakai_pampal_2021] in R v2022.07.02 [@r_core_team_r_2022].
 
-Multiple trials were conducted to better understand the value (and potential limitations) of the different suite of measurements.
-Data subsets for each trial were exported for BANTER classification using the export_banter function in PAMpal.
-Trials included standard click calculations for all clicks in a beaked whale acoustic event (EC); the addition of  ICI for each event (EC_ICI), and for the EPacific dataset we also included environmental variables (EC_ICI_ENV).
+Standard click metrics describing the spectral and temporal characteristics were measured from echolocation pulses using default values in *PAMpal* [@sakai_pampal_2023].
+Inter-pulse interval (IPI) was calculated using the 'calculateICI' function in *PAMpal* and the mode of the IPI was identified for each beaked whale acoustic event.
+In addition, for the EPacific data, we used the 'matchEnvData' function in *PAMpal* to include ERDAPP environmental data [@simons_erddap_2022], including seafloor depth and seafloor gradient (both from erdSrtm30plusSeafloorGradient), as well as sea surface temperature (from jplMURSST41).
+A full consideration of environmental data for all regions was beyond the scope of this study.
+
+Multiple models were created to better understand the value (and potential limitations) of the different suite of measurements.
+Models included standard click metrics (only) for all pulses in a beaked whale acoustic event (EC); the addition of IPI for each event (EC_IPI), and for the EPacific dataset we also included environmental variables (EC_IPI_ENV).
 
@@
 
-BANTER models were created for each trial using the BANTER package in the R programming language [@archer_banter_2022].
-Standard click calculations were used for the BANTER call classifier; event level measures of ICI and environmental variables were applied to the BANTER event classifier.
-A minimum of two events per species and the complete suite of variables (absence of NAs) were required for training and testing of the classification model.
-The BANTER random forest models contain two parameters, ntree and sampsize, which must be selected to ensure stability of the model.
-The sample size (sampsize) is the number of events considered for each trial of the model, and the ntree is the number of trees (iterations of the model) run in a given model.
-These values were modified to obtain the best results, with an emphasis on lower sample size and higher number of trees for computational performance [@rankin_banter_2021].
-
-The model output includes the confusion matrix, and a number of visualizations provided by the rfpermute package in the R programming language [@archer_rfpermute_2022].
+BANTER models were created using the *banter* package [@archer_banter_2022] in R v2022.07.02 [@r_core_team_r_2022].
+Output from PAMGuard click detector modules were used to create the BANTER detector models, for a total of 6 detector models (one for each click detector used).
+For each detector model, BANTER creates a mean of the species classification probabilities for each call type, which is applied to the BANTER event model (in conjunction with any event-level variables).
+A minimum of two events per species and the complete suite of variables (no missing data) were required for training and testing of the classification model.
+Event level variables included IPI and environmental variables (for the EPacific dataset).
+The Random Forest models that BANTER is based on are largely driven by two parameters, the number of trees in the forest (*ntree*) and the number of samples randomly drawn to build each tree (*sampsize*).
+Small values of *sampsize* were selected to improve computational performance, but large values of *ntree* were used to obtain models with stable classification results [@rankin_banter_2021].
+Model results were summarized using the *rfpermute* package [@archer_rfpermute_2022].
 
@@
 
-Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study; species included: Baird's beaked whales (*Berardius bairdii*), Blainsville's beaked whales (*Mesoplodon densirostris*), Cross Seamount beaked whale (@mcdonald_acoustic_2009), Cuvier's beaked whales (*Ziphius cavirostris*),  Gervais' beaked whales (*Mesoplodon europaeus*), Hubb's beaked whale (*Mesoplodon carlhubbsi*), Longman's beaked whale (*Indopacetus pacificus*), Sowerby's beaked whales (*Mesoplodon bidens*), Stejneger's beaked whale (*Mesoplodon stejnegeri*), True's beaked whales (*Mesoplodon mirus*), and unidentified beaked whale 'BW43' [@baumann-pickering_species-specific_2013].
+Four datasets from four different regions (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study, and species included: Baird's beaked whales (*Berardius bairdii*), Blainville's beaked whales (*Mesoplodon densirostris*), Cross Seamount beaked whales [@mcdonald_acoustic_2009], Cuvier's beaked whales (*Ziphius cavirostris*), Gervais' beaked whales (*M.
+europaeus*), Hubbs' beaked whales (*M.
+carlhubbsi*), Longman's beaked whales (*Indopacetus pacificus*), Sowerby's beaked whales (*M.
+bidens*), Stejneger's beaked whales (*M.
+stejnegeri*), True's beaked whales (*M.
+mirus*), and unidentified beaked whale 'BW43' [@baumann-pickering_species-specific_2013].
 Analyses were conducted separately for the different study areas due to differences in the data collection methods that precluded combining of datasets.
-Specific species combinations and sample sizes varied by study area.
+Species and sample sizes varied by study area.
 
@@
 Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n = 120), Gervais' beaked whales (n = 4), Sowerby's beaked whales (n = 6), and True's beaked whales (n = 76).
-BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).
+BANTER classification models included (1) echolocation pulses (EC), and (2) echolocation pulses and IPI (EC_IPI).
 
-For the NAtlantic dataset, the model with ICI had a modest increase in accuracy over the EC (only) model (EC_ICI model, Detector Model: sampsize = 3, ntrees = 10,000; Event Model: sampsize = 3, trees = 100,000).
-This model provided an overall correct classification rate of 98% for all four species (pct.correct in Confusion Matrix, @fig-natlantic a).
+For the NAtlantic dataset, the most accurate model included echolocation pulse metrics with IPI; this model had a modest increase in accuracy over the EC (only) model (EC_IPI model, Detector Model: *sampsize* = 3, *ntree* = 10,000; Event Model: *sampsize* = 3, *ntree* = 100,000).
+This model provided an overall correct classification rate of 97.5% for all four species (pct.correct in Confusion Matrix, @fig-natlantic a).
 Classification scores ranged from 75% for Gervais' beaked whales to 100% for Cuvier's and Sowerby's beaked whales.
-All classification results were greater than the expected error rate (Priors in @fig-natlantic a).
-Results from the EC_ICI model are presented here; results from the EC model were similar and can be found in supplementary materials (Supplement Fig.
+All classification results were greater than expected (Priors in @fig-natlantic a).
+Results from the EC_IPI model are presented here; results from the EC model were similar and can be found in supplementary materials (Supplement Fig.
 1).
 
-The proximity plot (@fig-natlantic b) provides a view of the distribution of events within the feature space of the two 'most important' dimensions, or predictors.
+The proximity plot (@fig-natlantic b) provides a view of the distribution of events within the classification model space.
 For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result.
-The importance heat map (@fig-natlantic c) shows the most important features for predicting each species.
-The strength of the classifications by species can be seen in the plot votes (@fig-natlantic d), which show the distribution of the votes for each of the 100,000 trials for each event.
-For each plot, the events are represented as vertical slices (with plot subdivided along the x-axes according to the event sample size), and the percentage of votes for each species is represented by their color along that vertical slice.
-In a perfect scenario, all 100,000 trees would have voted for the correct species and therefore the classification would be correct and the strength of these classifications would be maximized.
-
-The proximity plot (@fig-natlantic b) shows overlap between Gervais', Sowerby's and True's beaked whales in this two dimension feature space and the species classifications are resolved using additional features, as shown in the importance heat map (@fig-natlantic c).
-The plot vote figure (@fig-natlantic d) shows strong classification strength for Cuvier's and Sowerby's beaked whales, with increased variability in voting distributions for Gervais' and True's beaked whales.
+This plot shows the degree of overlap between Gervais', Sowerby's, and True's beaked whales.
+Species classifications of events are resolved using additional features (mean assignment probabilities for each of the detectors in the detector model, as well as any event level variables), as seen in the ten most important features for predicting each species (importance heat map, @fig-natlantic c).
+The strength of classifications for each event can be seen in the distribution of votes for each species across the forest (@fig-natlantic d).
+In each frame of this figure, the events are represented as vertical slices along the x-axis, and the percentage of votes for each species is represented by their color along that vertical slice in the y-axis.
+This distribution shows strong assignment probabilities for most Cuvier's and Sowerby's beaked whale events, with greater variability in assignment for Gervais' and True's beaked whale events.
 
@@
 
-Species encountered in the SAtlantic study dataset include Blainsville's beaked whales (n = 46), Cuvier's beaked whales (n = 111), Gervais' beaked whales (n = 77), Sowerby's beaked whales (n = 2), and True's beaked whales (n = 13).
-BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).
+Species encountered in the SAtlantic study dataset include Blainville's beaked whales (n = 46), Cuvier's beaked whales (n = 111), Gervais' beaked whales (n = 77), Sowerby's beaked whales (n = 2), and True's beaked whales (n = 13).
+BANTER classification models included (1) echolocation pulses (EC), and (2) echolocation pulses and IPI (EC_IPI).
 
-For the SAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (EC_ICI Model, Detector Model: sampsize = 4, ntrees = 10,000; Event Model: sampsize = 1, ntrees = 100,000) and provided an overall correct classification rate of 87.1% for all five species.
-Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-satlantic); all classification results were greater than the expected error rate (Priors in @fig-natlantic a).
-The sample sizes for the BANTER model was kept low (sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2.
+For the SAtlantic dataset, the most accurate model considered the echolocation pulse metrics and the IPI for an event (EC_IPI Model, Detector Model: *sampsize* = 4, *ntree* = 10,000; Event Model: *sampsize* = 1, *ntree* = 100,000) and provided an overall correct classification rate of 88.7% for all five species.
+Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-satlantic); all classification results were greater than expected (Priors in @fig-natlantic a).
+The sample sizes for the BANTER model were kept low (event sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2.
 Analysis with a larger sample size resulted in improved classification results for Cuvier's and Gervais' beaked whale (see Supplement Fig.
-3 for EC_ICI_ALT), but resulted in the loss of Sowerby's in the final model.
-Results from the EC_ICI model are presented here; results from the EC and EC_ICI_ALT model can be found in supplementary materials (Supplement Fig.
+3 for EC_IPI_ALT), but resulted in the loss of Sowerby's in the final model.
+Results from the EC_IPI model are presented here; results from the EC and EC_IPI_ALT model can be found in supplementary materials (Supplement Fig.
 2).
 
-For SAtlantic, the two most important features are insufficient to differentiate species, as seen by the overlap in species distribution with these feature spaces in the proximity plot (@fig-satlantic b).
-Importance variables vary by species (importance heat map @fig-satlantic c), with ICI measurements having relatively low importance for Blainsville's and Sowerby's beaked whales (@fig-satlantic c).
-The plot votes graph (@fig-satlantic d) shows relatively weak classification results; most events consisted of \<50% of the trees voting for the correct species.
+For this dataset, the two most important features are insufficient to differentiate species, as seen by the overlap in species distribution with these feature spaces in the proximity plot (@fig-satlantic b).
+Importance variables varied by species (importance heat map @fig-satlantic c), with IPI measurements having relatively low importance (@fig-satlantic c).
+The distribution of votes (@fig-satlantic d) shows relatively weak classification results; most events consisted of \<50% of the trees voting for the correct species.
 
@@
 
-The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (@yano_cetacean_2018); species included Blainsville's beaked whale (n = 521), Cross Seamount beaked whale (n = 76), Cuvier's beaked whale (n = 201), and Longman's beaked whale (n = 122).
-BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).
+The Hawaii dataset consisted of 13 drifting recorders deployed within the Main Hawaiian Islands; species included Blainville's beaked whale (n = 521), Cross Seamount beaked whale (n = 76), Cuvier's beaked whale (n = 201), and Longman's beaked whale (n = 122).
+BANTER classification models included (1) echolocation pulses (EC), and (2) echolocation pulses and IPI (EC_IPI).
 
-The most accurate model considered the echolocation clicks only (EC Model, Detector Model: sampsize = 10, ntrees = 5,000; Event model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.3% for all four species.
-Classification scores ranged from a low of 86% (Cuvier's Beaked Whale) to 97.5% (Longman's beaked whale, @fig-hawaii).
-All classification results were greater than the expected error rate (Prior in @fig-hawaii a).
-Results from the EC model are presented here; results from EC_ICI model can be found in supplementary materials (Supplement Fig.
+The most accurate model considered the echolocation pulses only (EC Model, Detector Model: *sampsize* = 10, *ntree* = 5,000; Event model: *sampsize* = 4, *ntree* = 10,000) and provided an overall correct classification rate of 92.3% for all four species.
+Classification scores ranged from a low of 86% (Cuvier's beaked whale) to 97.5% (Longman's beaked whale, @fig-hawaii).
+All classification results were greater than expected (Prior in @fig-hawaii a).
+Results from the EC model are presented here; results from EC_IPI model can be found in supplementary materials (Supplement Fig.
 4).
 
-Distinctive clusters based on the two most important dimensions can be found for Longman's and Blainsville's beaked whales, and there is overlap in this feature space of the clusters for Cuvier's and Cross Seamount beaked whales as shown in the Proximity Plot (@fig-hawaii b).
-This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the Importance Heat Map (@fig-hawaii c), resulting in the low misclassification rates for these species (@fig-hawaii a confusion matrix).
-The plot votes graph (@fig-hawaii d) show strong classification results for most Blainsville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.
+Distinctive clusters can be seen for Longman's and Blainville's beaked whales, although there is overlap on these first two dimensions of the clusters for Cuvier's and Cross Seamount beaked whales (@fig-hawaii b).
+This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the importance heat map (@fig-hawaii c), resulting in the low misclassification rates for these species (@fig-hawaii a confusion matrix).
+The distribution of votes (@fig-hawaii d) show strong classification results for most Blainville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.
 
-## Eastern Pacific
+## EPacific
 
-The EPacific dataset consisted of 15 drifting recording buoys deployed off the west coast of the United States [@simonis_passive_2020]; species included Baird's beaked whale (n = 29), Unidentified Beaked Whale 'BW43' (n = 125), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 926), Hubbs beaked whale (n = 66) and Stejneger's beaked whale (n = 42).
-BANTER classification trials included (1) echolocation clicks (EC), (2) echolocation clicks and ICI (EC_ICI), and (3) echolocation clicks, ICI, and environmental features (EC_ICI_ENV).
+The EPacific dataset consisted of 15 drifting recorders deployed off the U.S.
+West Coast, and species included Baird's beaked whale (n = 29), unidentified beaked whale 'BW43' (n = 125), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 926), Hubbs' beaked whale (n = 66), and Stejneger's beaked whale (n = 42).
+BANTER classification models included (1) echolocation pulses (EC), (2) echolocation pulses and IPI (EC_IPI), and (3) echolocation pulses, IPI, and environmental features (EC_IPI_ENV).
 
-The best model considered the echolocation clicks, ICI, and environmental features (EC_ICI_ENV Model, Detector Model: sampsize = 3, ntrees = 10,000; Event model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.2% for all six species.
-Classification scores ranged from a low of 91.2% (BW43) to 100% for Baird's beaked whales(@fig-epacific a).
-All classification results were greater than the expected error rate (Prior in @fig-epacific a).
-Results from the EC_ICI_ENV model are presented here; results from the EC and EC_ICI model can be found in supplementary materials (Supplement Figs.
+The best model considered the echolocation pulses, IPI, and environmental features (EC_IPI_ENV Model, Detector Model: *sampsize* = 1, *ntree* = 10,000; Event model: *sampsize* = 5, *ntree* = 10,000) and provided an overall correct classification rate of 91.9% for all six species.
+Classification scores ranged from a low of 91.2% (BW43) to 100% for Cross Seamount beaked whale (@fig-epacific a).
+All classification results were greater than expected (Prior in @fig-epacific a).
+Results from the EC_IPI_ENV model are presented here; results from the EC and EC_IPI model can be found in supplementary materials (Supplement Figs.
 5).
 
-There is considerable overlap in the primary feature space for all but BW43, as shown in the Proximity Plot (@fig-epacific b).
-ICI was among the most important classification variable for all species (@fig-epacific c heat map).
-The plot votes (@fig-epacific d) show relatively strong classification results Baird's, Hubbs, and Stejneger's beaked whales.
+There is considerable overlap in the primary feature space for all but BW43, as shown in the proximity plot (@fig-epacific b).
+IPI was among the most important classification variable for all species (@fig-epacific c importance heat map).
+The distribution of votes (@fig-epacific d) show relatively strong classification results for Baird's, Hubbs', and Stejneger's beaked whales.
 
@@
 
-The results for the four study areas varied from a low of 87.1% correct classification in the SAtlantic dataset to a high of 98% correct classification in the NAtlantic dataset.
-All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais beaked whales (NAtlantic) had classification rates below 90%.
+The results for the four study areas varied from a low of 88.7% correct classification in the SAtlantic dataset to a high of 97.5% correct classification in the NAtlantic dataset.
+All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%.
 
 Several of our datasets included very low sample sizes for some species; however, low sample sizes did not always result in low classification scores.
-In the NAtlantic dataset, Sowerby's beaked whale (n=6) was correctly classified in 100% of the samples, and classification results were very strong.
-The Sowerby's beaked whale events in the SAtlantic study and the Gervais beaked whale events in the NAtlantic study consisted of weak classifications.
-This is likely due to both a small number of events (n = 2 for Sowerby's in SAtlantic and n = 4 for Gervais in NAtlantic) combined with the small sampsize used in the BANTER model.
-In general, larger sample sizes should lead to improved overall classification results, and improved strength of these classification results (as identified by plot vote graph).
+In the NAtlantic dataset, Sowerby's beaked whale (n=6) was correctly classified in 100% of the events, with strong assignment probabilities.
+The Sowerby's beaked whale events in the SAtlantic study and the Gervais' beaked whale events in the NAtlantic study consisted of weak classifications.
+This is likely due to both a small number of events (n = 2 for Sowerby's in SAtlantic and n = 4 for Gervais' in NAtlantic) combined with the small *sampsize* used in the BANTER model.
+In general, larger sample sizes should lead to improved overall classification results, and improved strength of these classification results (as indicated in the vote distributions).
 
-While some species have distinctive differences in their click characteristics that lead to strong classification despite small sample sizes (e.g.
-Sowerby's beaked whales in NAtlantic); other species, such as Cuvier's beaked whales, have significant variation in their click measurements.
+While some species have distinctive differences in their pulse characteristics that can lead to strong classification despite small sample sizes (e.g.
+Sowerby's beaked whales in NAtlantic), other species, such as Cuvier's beaked whales, have significant variation in their pulse measurements.
  For species with high variability in the predictor variables and an overlap in the range of these variables with other species, a large sample size is required to describe the true variability of these call measurements.
-Even with reasonably large sample sizes, the classification may suffer (e.g, SAtlantic Cuvier's beaked whale = 79.2% correct classification).
-We found by increasing the sampsize in BANTER, we could increase the classification rate for this species in this area to (91.8%, see Supplement Fig.3); however, this resulted in the loss of Sowerby's beaked whale in the final model.
+Even with reasonably large sample sizes, the classification may suffer (e.g., SAtlantic Cuvier's beaked whale = 82.8% correct classification).
+We found that by increasing the *sampsize* in BANTER, we could increase the classification rate for this species in this area to 91.8% (see Supplement Fig.3).
+However, this resulted in the inability to include Sowerby's beaked whale in the final model.
 So, an increased sample size in Sowerby's would likely lead to increased classification results for other species in the model.
 
-Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (long-beaked and short-beaked common dolphins, [@rankin_acoustic_2017]).
+Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (e.g., long-beaked and short-beaked common dolphins, @rankin_acoustic_2017).
 This suggests that large increases in sample sizes may improve classification results for Cuvier's beaked whales.
 
-Researchers have noted several 'types' of click characteristics for Cuvier's beaked whales (ADA, pers.
-comm) which may be related to variability in the propagation environment, different animal behaviors, gender or age classes, or acoustic populations.
-Future work should examine if these Cuvier's beaked whale events can be clustered in an objective and stable manner; if so, then these can be input into BANTER models as separate types of Cuvier's beaked whales.
-We recommend caution in this approach- this should only be considered if there is reasonable evidence that these different groups represent authentic acoustic groups.
-We do not recommend segregation of misclassified subgroups simply to improve classification scores.
-
-Gervais' and True's beaked whales have similar click characteristics that make them difficult to differentiate and they require considerable expertise to classify manually [@deangelis_description_2018].
+Gervais' and True's beaked whales have similar pulse characteristics that make them difficult to differentiate and they require considerable expertise to classify manually [@deangelis_description_2018].
 Our results suggest that BANTER may serve as an efficient and effective means of classifying and differentiating these two species.
 Despite modest sample sizes, Gervais' and True's beaked whales showed high classification scores in the SAtlantic (93.5% and 100%, respectively), and True's performed well in the NAtlantic.
-Unfortunately, we were unable to combine the N and S Atlantic datasets (which would increase sample sizes of True's beaked whales) due to differences in hydrophone characteristics that result in differences in call measures.
+Unfortunately, we were unable to combine the Atlantic datasets (which would increase sample sizes of True's beaked whales) due to differences in hydrophone characteristics that resulted in differences in call metrics.
 Calibration of signals (to make them comparable) may allow for combining datasets to improve sample sizes.
 
-In addition to poor classification of Gervais in the NAtlantic study area, the plot votes show that for one of the four events, only a few of the 100,000 trees 'voted' for Gervais, and the majority of the votes were for True's beaked whales.
-Analysis notes indicate uncertainty in manual species classification of one of the events in this dataset.
-While small sample size can at times provide good classification scores, as we found in Sowerby's beaked whale in this SAtlantic dataset, they can be devastating if there are inaccuracies in the training data.
+In addition to poor classification of Gervais' in the NAtlantic study area, the plot votes show that for one of the four events, only a few of the 100,000 trees 'voted' for Gervais', and the majority of the votes were for True's beaked whales.
+Analysts' notes indicated uncertainty in manual species classification of this event in this dataset.
+While small sample size can at times provide good classification scores, as we found for Sowerby's beaked whale in this SAtlantic dataset, they can be heavily influential if there are inaccuracies in the training data.
+
 These two examples of small sample sizes highlight a conflict in preferred protocol.
@@
 Unfortunately, in species where identity must be determined based on call characteristics (rather than visual confirmation of species identity in the field), it can sometimes be difficult to confidently determine species identity when calls are highly variable and do not include at least one call that provides high confidence of species identity.
-In the case of beaked whales, we recommend that training data include a reasonable level of confidence for inclusion.
+In the case of beaked whales, we recommend that training data include a high level of confidence for inclusion.
 This is especially critical for species with small sample sizes.
-An alternative is to require agreement from multiple analysts, as provided by the EPacific datasets.
+An alternative is to require agreement from multiple analysts.
 This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified.
-For example, [@barlow_unique_2022] recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (including the EPacific data presented here).
+For example, [@barlow_unique_2022] recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (expert analysis did not identify these in the EPacific data presented here).
 
-Interclick Interval (ICI) was found to be the most important variable for NAtlantic and EPacific, and ICI were the third and fourth ranked variable for SAtlantic (see Importance Heat Maps for each survey area).
-While the addition of ICI did not improve results for Hawaii, results were similar and ICI ranked #9 in importance for the Hawaii study area.
-There was a difference in how ICI was calculated for these surveys, and it is possible these differences may affect the value of the ICI variable in species classification for some species.
-Beaked whale events were defined by encounter time for both the Hawaii and EPacific data, which were collected using drifting acoustic recorders where animals could not be localized.
-The Atlantic data were collected using a towed hydrophone array with localization capabilities.
-Each event was defined as an individual click train, so ICI for these events more likely represented true ICI.
-[@baumann-pickering_species-specific_2013] showed an overlap in the ICI between Cuvier's and Blainsville's beaked whales, but a strong difference in ICI between these species and Cross Seamount beaked whales.
-It is possible that classification of these species could be improved by refining an event to consider ICI from individual click trains.
-Subdividing click trains in PAMGuard is time consuming (and complicated in the absence of localization), but alternative options include consideration of the PAMGuard click train detector module in BANTER, or by developing a similar function in R to apply to events.
+Inter-pulse interval (IPI) was found to be the most important variable for NAtlantic and EPacific, and IPI was the fourth and fifth ranked variables for SAtlantic (see importance heat maps for each survey area).
+While the addition of IPI did not improve results for data from Hawaii, results were similar and IPI ranked #9 in importance for the Hawaii study area.
+@baumann-pickering_species-specific_2013 showed an overlap in the IPI between Cuvier's and Blainville's beaked whales, but a strong difference in IPI between these species and Cross Seamount beaked whales.
+It is possible that classification of these species could be improved by refining an event to annotation at the individual level, as was done with the NAtlantic and SAtlantic datasets.
+Subdividing pulse trains into individuals in PAMGuard is time consuming, but alternative options include consideration of the PAMGuard click train detector module as an additional first stage detector in BANTER, or by developing a similar function in R to apply to events.
 
 For this study, only a few simple environmental features were considered for one of the study areas (seafloor depth, seafloor gradient, and sea surface temperature).
-In the EPacific study area, inclusion of these variables increased the overall classification rate from 89.2% to 92.2%.
-These variables had the most impact on classification results for the Cross Seamount beaked whale (n = 6), raising the classification scores to from 83.3% to 100%.
+In the EPacific study area, inclusion of these variables increased the overall classification rate from 91.1% to 91.9%.
+These variables had the most impact on classification results for the Cross Seamount beaked whale (n = 6), raising the classification scores from 83.3% to 100%.
 Alternative environmental features may improve results for other species.
@@
 
-The workflow presented here provides a highly automated approach to detection of acoustic events (PAMGuard), integration of environmental data (PAMpal) and acoustic event classification (BANTER).
-These methods significantly reduce manual analysis, provide more consistent classification results with fewer biases, and provide an estimate of classification error.  The greatest improvement to classification results for beaked whales would likely result from improved sample sizes, and examination of individual click trains in measuring ICI.
-Consideration of additional detectors (e.g.
-matched filter detector, click train detector) or additional environmental variables may further improve classification results.
-Improved alignment of detection methods across studies (i.e., definition of events, hydrophone calibration) may allow for combination of data within geographic regions to improve assessments within regions for all species, and across regions for species with global distributions, such as Cuvier's beaked whales.
+The workflow presented here provides a highly automated approach to detection of acoustic events (PAMGuard), integration of environmental data (PAMpal), and acoustic event classification (BANTER).
+These methods significantly reduce manual analysis, provide more consistent classification results with fewer biases, and provide an estimate of classification error.
+The greatest improvement to classification results for beaked whales would likely result from improved sample sizes, and examination of individuals to accurately measure IPI.
+Consideration of additional detectors (e.g., matched filter detector, click train detector) or additional environmental variables may further improve classification results.
+Improved alignment of detection methods across studies (i.e., definition of events, hydrophone calibration) may allow for combining data within geographic regions to improve assessments within regions for all species, and across regions for species with global distributions, such as Cuvier's beaked whales.
 These highly automated methods may allow for analysis of data across large spatial and temporal scales to address large ecological and population level questions.
@@
 The authors would like to acknowledge the large number of scientists and shipboard crew who were responsible for data collection during multiple large scale studies.
-Sea surface temperature data were provided by JPL under support by NASA MEaSUREs program.
+Thanks to Liam Mueller-Brennan for their help with data analysis.
+Sea surface temperature data were provided by JPL under support by the NASA MEaSUREs program.
 Funding for data collection and analysis were provided by the U.S.
-Navy, Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration.
+Navy, the Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration.
+The AMAPPS study (NAtlantic/SAtantic) was funded in part by the U.S.
+Department of the Interior, Bureau of Ocean Energy Management, Environmental Studies Program, Washington, DC, through Inter-Agency Agreement Number M10PG00075 with the National Marine Fisheries Service as the Atlantic Marine Assessment Program for Protected Species.
+The HICEAS 2017 Survey (Hawaii) was funded by Bureau of Ocean Energy Management under (IAA M17PG00024) and NMFS Pacific Islands Fisheries Science Center.
+The CCES Survey (EPacific) was funded by Bureau of Ocean Energy Management (No.
+M17PG00025), U.S.
+Navy Pacific Fleet Environmental Readiness Division (IAA No.
+N00070-18-MP-4C560) and Chief of Naval Operations N45, and Southwest Fisheries Science Center.
 The manuscript was improved thanks to reviews from XX and XX.
@@
 #| label: fig-natlantic
-#| fig-cap: BANTER classification results from the NAtlantic dataset including echolocation clicks and interclick interval (EC_ICI).
+#| fig-cap: BANTER classification results from the NAtlantic dataset including echolocation pulses and inter-pulse interval (EC_IPI).
 Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
-Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
 Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
@@
 ```
+\newpage
 
@@
 #| label: fig-satlantic
-#| fig-cap: BANTER classification results from the SAtlantic dataset (EC_ICI).
+#| fig-cap: BANTER classification results from the SAtlantic dataset (EC_IPI).
 Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
-Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
 Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
@@
 ```
-
+\newpage
 ```{r, hawaii}
@@
 Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
-Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
 Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
@@
 ```
-
+\newpage
 ```{r, epacific-env}
 #| label: fig-epacific
-#| fig-cap: BANTER classification results from the EPacific dataset with environmental data (EC_ICI_ENV).
+#| fig-cap: BANTER classification results from the EPacific dataset with environmental data (EC_IPI_ENV).
 Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
-Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification).
-Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
 Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
@@
 ```
+\newpage
 # References {.unnumbered}

### 2023-08-21 natlantic/beakerBanter_natlantic.Rmd
manuscript update
Updated Manuscript

@@
 
-6.  **Export data for BANTER** (and drop species codes that will not be
+6.  **Add GPS data**: Add GPS data from PAMGuard table (2 hour
+    threshold), then filter out data without GPS
+
+```{r, add GPS, eval=freshRun}
+data <- addGps(data, thresh = 7200)#11 events over the threshold
+getWarnings(data)
+```
+
+7. **Add Environmental Data**
+```{r, Add Environmental Data, eval=freshRun}
+#SST
+data <- matchEnvData(data, nc='jplMURSST41mday', var='sst')
+#Seafloor Depth
+data <- matchEnvData(data, nc='usgsCeSS111', var='topo')
+```
+
+8.  **Export data for BANTER** (and drop species codes that will not be
     used for training). We will create two datasets: one with ICI and
@@
 banter_data <- export_banter(data, dropSpecies = 'unid', 
-                         dropVars = c('All_ici'), training=TRUE)
+                         dropVars = c('All_ici', 'topo', 'sst', 'topo_mean', 'sst_mean', 
+                                      'Latitude', 'Longitude', 'gpsUncertainty'), training=TRUE)
 saveRDS(banter_data, file='natlantic_banter_data.rds')
 
-banter_data_ici <- export_banter(data, dropSpecies = 'unid', training=TRUE)
+banter_data_ici <- export_banter(data, dropSpecies = 'unid', 
+                                 dropVars = c('topo', 'sst', 'topo_mean', 'sst_mean', 
+                                              'Latitude', 'Longitude', 'gpsUncertainty'),training=TRUE)
 saveRDS(banter_data_ici, file='natlantic_banter_data_ici.rds')
+
+banter_data_env <- export_banter(data, dropSpecies = c('unid'), 
+                                 dropVars = c('Latitude', 'Longitude', 'gpsUncertainty'), training=TRUE)
+saveRDS(banter_data_env, file='natlantic_banter_data_env.rds')
 
@@
 ```
+
+### *ENV Model*
+
+Initialize, Run & Evaluate Detector Model (stage 1).
+```{r, banter env detector model, eval=freshRun}
+banter_model_env <- initBanterModel(banter_data_env$events)
+banter_model_env <- addBanterDetector(banter_model_env, banter_data_env$detectors, ntree=1e5, sampsize=3, importance = TRUE)
+
+plotDetectorTrace(banter_model_env, detector = paste0('Click_Detector_', 3:4))
+plotDetectorTrace(banter_model_env, detector = paste0('Click_Detector_', 5:6))
+summary(banter_model_env)
+```
+
+Run Event Model (stage 2)
+```{r, banter env event model, eval=freshRun}
+banter_model_env <- runBanterModel(banter_model_env, ntree=1e5, sampsize=3)
+summary(banter_model_env)
+```
+
+Once a stable model is identified, save model with tree/sampsize info in
+the filename.
+
+```{r, save banter env model, eval=freshRun}
+saveRDS(banter_model_env, 'natlantic_banter_model_env_t1e5s3_t1e5s3.rds')
+```
+
 ### *Alternative Model (EC_ICI_ALT) with larger samp size*
@@
 ```{r, banter alt event model, eval=freshRun}
-banter_model_alt <- runBanterModel(banter_model_alt, ntree=1e4, sampsize=5)
+banter_model_alt <- runBanterModel(banter_model_alt, ntree=1e5, sampsize=5)
 summary(banter_model_alt)
@@
 ```{r, save banter alt model, eval=freshRun}
-saveRDS(banter_model_alt, 'natlantic_banter_model_alt_t1e4s5_t1e4s5.rds')
+saveRDS(banter_model_alt, 'natlantic_banter_model_alt_t1e4s5_t1e5s5.rds')
 ```
@@
 banter_model_ici <- readRDS('natlantic_banter_model_ici_t1e4s3_t1e5s3.rds')
-banter_model_alt <- readRDS('natlantic_banter_model_alt_t1e4s5_t1e4s5.rds')
+banter_model_env <- readRDS('natlantic_banter_model_env_t1e5s3_t1e5s3.rds')
+banter_model_alt <- readRDS('natlantic_banter_model_alt_t1e4s5_t1e5s5.rds')
 ```
@@
 
+model_env <- banter_model_env
+modelname_env <- "banter_model_env"
+
 model_alt <- banter_model_alt
@@
 banter_model_ici_RF <- getBanterModel(model_ici)
+banter_model_env_RF <- getBanterModel(model_env)
 banter_model_alt_RF <- getBanterModel(model_alt)
@@
 natlantic_ici_priors <- classPriors(banter_model_ici_RF, NULL)[,1]
+natlantic_env_priors <- classPriors(banter_model_env_RF, NULL)[,1]
 natlantic_alt_priors <- classPriors(banter_model_alt_RF, NULL)[,1]
@@
 
+natlantic_env_confuseMatrix <- rfPermute::confusionMatrix(banter_model_env_RF)
+natlantic_env_confuseMatrix <- cbind(natlantic_env_confuseMatrix, priors = natlantic_env_priors)
+
+natlantic_env_confuseMatrix <- kable(natlantic_env_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(5, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(3,hline_after = TRUE)%>%
+  row_spec(5, bold = TRUE)%>%
+  save_kable('../manuscript/manuscript_files/natlantic_env_confuseMatrix.png', zoom = 9)
+
 natlantic_alt_confuseMatrix <- rfPermute::confusionMatrix(banter_model_alt_RF)
@@
 
+![BANTER Model N. Atlantic ENV Confusion
+Matrix](../manuscript/manuscript_files/natlantic_env_confuseMatrix.png)
+
 ![BANTER Model N. Atlantic Alt Confusion
@@
 
+png(('../manuscript/manuscript_files/natlantic_env_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
+env_natlantic_proximityPlot <- plotProximity(banter_model_env_RF)
+dev.off()
+env_natlantic_proximityPlot <- plotProximity(banter_model_env_RF)
+
 png(('../manuscript/manuscript_files/natlantic_alt_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
@@
 png(('../manuscript/manuscript_files/natlantic_ec_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_ec_RF, plot.type="heatmap")
+plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
 dev.off()
-natlantic_ec_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap")
+natlantic_ec_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
 
 png(('../manuscript/manuscript_files/natlantic_ici_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_ici_RF, plot.type="heatmap")
+plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
 dev.off()
-ici_natlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_natlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
+
+png(('../manuscript/manuscript_files/natlantic_env_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
+plotImportance(banter_model_env_RF, plot.type="heatmap", n=10)
+dev.off()
+env_natlantic_importance <- plotImportance(banter_model_env_RF, plot.type="heatmap", n=10)
 
 png(('../manuscript/manuscript_files/natlantic_alt_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_alt_RF, plot.type="heatmap")
+plotImportance(banter_model_alt_RF, plot.type="heatmap", n=10)
 dev.off()
-alt_natlantic_importance <- plotImportance(banter_model_alt_RF, plot.type="heatmap")
+alt_natlantic_importance <- plotImportance(banter_model_alt_RF, plot.type="heatmap", n=10)
 ```
@@
 
+png(('../manuscript/manuscript_files/natlantic_env_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_env_RF)
+dev.off()
+env_natlantic_votes <- plotVotes(banter_model_env_RF)
+
 png(('../manuscript/manuscript_files/natlantic_alt_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
@@
 plotPredictedProbs(banter_model_ici_RF, bins=30, plot=TRUE)
+plotPredictedProbs(banter_model_env_RF, bins=30, plot=TRUE)
 plotPredictedProbs(banter_model_alt_RF, bins=30, plot=TRUE)
@@
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_importance.png'))%>%
@@
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_importance.png'))%>%
@@
 
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_env_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_env_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=300, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_env_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=300, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_env_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=300, color = "black")
+
+natlantic_env_Figure <-image_append(c(prox, heat, vote))
+natlantic_env_Figure<- image_append(c(confuse, natlantic_env_Figure), stack=TRUE)
+image_write(natlantic_env_Figure, path = here('manuscript', 'manuscript_files','natlantic_env_Figure.png'), format ='png')
+print(natlantic_env_Figure, info=FALSE)
+
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_confuseMatrix.png'))%>%
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_importance.png'))%>%
@@
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=300, color = "black")
 
 natlantic_alt_Figure <-image_append(c(prox, heat, vote))
-natlantic_alt_Figure<- image_append(c(confuse, natlantic_ici_Figure), stack=TRUE)
+natlantic_alt_Figure<- image_append(c(confuse, natlantic_alt_Figure), stack=TRUE)
 image_write(natlantic_alt_Figure, path = here('manuscript', 'manuscript_files','natlantic_alt_Figure.png'), format ='png')

### 2023-08-21 satlantic/beakerBanter_satlantic.Rmd
manuscript update
Updated Manuscript

@@
 banter_data <- export_banter(data, dropSpecies = 'unid', 
-                         dropVars = c('All_ici'), training=TRUE)
+                         dropVars = c('All_ici', 'topo', 'topo_mean', 'sst_mean', 'sst', 'Click_Detector_3_ici',
+                                      'Latitude', 'Longitude', 'gpsUncertainty'), training=TRUE)
 saveRDS(banter_data, file='satlantic_banter_data.rds')
 
-banter_data_ici <- export_banter(data, dropSpecies = 'unid', training=TRUE)
+banter_data_ici <- export_banter(data, dropSpecies = 'unid', 
+                                 dropVars = c('topo', 'topo_mean', 'sst_mean','sst', 
+                                              'Latitude', 'Longitude', 'gpsUncertainty'), training=TRUE)
 saveRDS(banter_data_ici, file='satlantic_banter_data_ici.rds')
+
+banter_data_env <- export_banter(data, dropSpecies = c('unid'), 
+                                 dropVars = c('Latitude', 'Longitude', 'gpsUncertainty'), training=TRUE)
+saveRDS(banter_data_env, file='satlantic_banter_data_env.rds')
 
@@
 banter_model_ec <- initBanterModel(banter_data$events)
-banter_model_ec <- addBanterDetector(banter_model_ec, banter_data$detectors, ntree=1e4, sampsize=2, importance = TRUE)
+banter_model_ec <- addBanterDetector(banter_model_ec, banter_data$detectors, ntree=1e4, sampsize=8, importance = TRUE)
 
@@
 ```{r,save banter ec model, eval=freshRun}
-saveRDS(banter_model_ec, 'satlantic_banter_model_ec_t1e4s2_t1e5s1.rds')
+saveRDS(banter_model_ec, 'satlantic_banter_model_ec_t1e4s8_t1e5s1.rds')
 ```
@@
 ```
+### *ENV Model*
 
-### Alternative Model (EC_ICI_ALT) with larger sampsize
+Initialize, Run & Evaluate Detector Model (stage 1).
+```{r, banter env detector model, eval=freshRun}
+banter_model_env <- initBanterModel(banter_data_env$events)
+banter_model_env <- addBanterDetector(banter_model_env, banter_data_env$detectors, ntree=1e5, sampsize=2, importance = TRUE)
+
+plotDetectorTrace(banter_model_env, detector = paste0('Click_Detector_', 3:4))
+plotDetectorTrace(banter_model_env, detector = paste0('Click_Detector_', 5:6))
+summary(banter_model_env)
+```
+
+Run Event Model (stage 2)
+```{r, banter env event model, eval=freshRun}
+banter_model_env <- runBanterModel(banter_model_env, ntree=1e5, sampsize=1)
+summary(banter_model_env)
+```
+
+Once a stable model is identified, save model with tree/sampsize info in
+the filename.
+
+```{r, save banter env model, eval=freshRun}
+saveRDS(banter_model_env, 'satlantic_banter_model_env_t1e5s2_t1e5s1.rds')
+```
+
+### *Alternative Model (EC_ICI_ALT)* with larger sampsize
 
@@
 ```{r, banter alt event model, eval=freshRun}
-banter_model_alt <- runBanterModel(banter_model_alt, ntree=1e5, sampsize=2)
+banter_model_alt <- runBanterModel(banter_model_alt, ntree=1e5, sampsize=3)
 summary(banter_model_alt)
@@
 ```{r, save banter alt model, eval=freshRun, eval=freshRun}
-saveRDS(banter_model_alt, 'satlantic_banter_model_alt_t1e4s4_t1e5s2.rds')
+saveRDS(banter_model_alt, 'satlantic_banter_model_alt_t1e4s4_t1e5s3.rds')
 ```
@@
 ```{r, read banter models, include=FALSE, eval=!freshRun}
-banter_model_ec <- readRDS(here('satlantic', 'satlantic_banter_model_ec_t1e4s2_t1e5s1.rds'))
+banter_model_ec <- readRDS(here('satlantic', 'satlantic_banter_model_ec_t1e4s8_t1e5s1.rds'))
 banter_model_ici <- readRDS(here('satlantic', 'satlantic_banter_model_ici_t1e4s4_t1e5s1.rds'))
-banter_model_alt <- readRDS(here('satlantic', 'satlantic_banter_model_alt_t1e4s4_t1e5s2.rds'))
+banter_model_env <- readRDS(here('satlantic', 'satlantic_banter_model_env_t1e5s2_t1e5s1.rds'))
+banter_model_alt <- readRDS(here('satlantic', 'satlantic_banter_model_alt_t1e4s4_t1e5s3.rds'))
 ```
@@
 satlantic_ici_priors <- classPriors(banter_model_ici_RF, NULL)[,1]
+satlantic_env_priors <- classPriors(banter_model_env_RF, NULL)[,1]
 satlantic_alt_priors <- classPriors(banter_model_alt_RF, NULL)[,1]
@@
 
+satlantic_env_confuseMatrix <- rfPermute::confusionMatrix(banter_model_env_RF)
+satlantic_env_confuseMatrix <- cbind(satlantic_env_confuseMatrix, priors = satlantic_env_priors)
+satlantic_env_confuseMatrix <- kable(satlantic_env_confuseMatrix, align = "c", digits = c(0,0,0,0,2,2,2))%>%
+  kable_classic()%>%
+  column_spec(6, border_right = TRUE)%>%
+  row_spec(0, bold = TRUE)%>%
+  row_spec(3,hline_after = TRUE)%>%
+  row_spec(6, bold = TRUE)%>%
+  save_kable("../manuscript/manuscript_files/satlantic_env_confuseMatrix.png", zoom = 9)
+
 satlantic_alt_confuseMatrix <- rfPermute::confusionMatrix(banter_model_alt_RF)
@@
 
+![BANTER Model S.Atlantic ENV Confusion
+Matrix](../manuscript/manuscript_files/satlantic_env_confuseMatrix.png)
+
 ![BANTER Model S.Atlantic Alternative Model Confusion
@@
 
+png(('../manuscript/manuscript_files/satlantic_env_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
+env_satlantic_proximityPlot <- plotProximity(banter_model_env_RF)
+dev.off()
+env_satlantic_proximityPlot <- plotProximity(banter_model_env_RF)
+
 png(('../manuscript/manuscript_files/satlantic_alt_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
@@
 png(('../manuscript/manuscript_files/satlantic_ec_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_ec_RF, plot.type="heatmap")
+plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
 dev.off()
-satlantic_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap")
+satlantic_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
 
 png(('../manuscript/manuscript_files/satlantic_ici_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_ici_RF, plot.type="heatmap")
+plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
 dev.off()
-ici_satlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_satlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
+
+png(('../manuscript/manuscript_files/satlantic_env_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
+plotImportance(banter_model_env_RF, plot.type="heatmap", n=10)
+dev.off()
+env_satlantic_importance <- plotImportance(banter_model_env_RF, plot.type="heatmap", n=10)
 
 png(('../manuscript/manuscript_files/satlantic_alt_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_alt_RF, plot.type="heatmap")
+plotImportance(banter_model_alt_RF, plot.type="heatmap", n=10)
 dev.off()
-alt_satlantic_importance <- plotImportance(banter_model_alt_RF, plot.type="heatmap")
+alt_satlantic_importance <- plotImportance(banter_model_alt_RF, plot.type="heatmap", n=10)
 ```
@@
 
+png(('../manuscript/manuscript_files/satlantic_env_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
+plotVotes(banter_model_env_RF)
+dev.off()
+env_satlantic_votes <- plotVotes(banter_model_env_RF)
+
 png(('../manuscript/manuscript_files/satlantic_alt_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
@@
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_importance.png'))%>%
@@
   image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_annotate("a) Confusion Matrix", size=300, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=300, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=300, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_importance.png'))%>%
@@
 print(satlantic_ici_Figure, info=FALSE)
+
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_env_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_env_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_env_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_env_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+satlantic_env_Figure <-image_append(c(prox, heat, vote))
+satlantic_env_Figure<- image_append(c(confuse, satlantic_env_Figure), stack=TRUE)
+image_write(satlantic_env_Figure, path = here('manuscript', 'manuscript_files','satlantic_env_Figure.png'), format ='png')
+print(satlantic_env_Figure, info=FALSE)
 

### 2023-08-28 hawaii/beakerBanter_hawaii.Rmd
Revise Hawaii and Manuscript
Revise Hawaii to reduce HeatPlot to show top 10; revise manuscript based on Cory's comments, update supplement.

@@
 png(('../manuscript/manuscript_files/hawaii_ec_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_ec_RF, plot.type="heatmap")
+plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
 dev.off()
-ec_hawaii_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap")
+ec_hawaii_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
 
 png(('../manuscript/manuscript_files/hawaii_ici_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_ici_RF, plot.type="heatmap")
+plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
 dev.off()
-ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
 ```

### 2023-08-28 manuscript/Supplement.qmd
Revise Hawaii and Manuscript
Revise Hawaii to reduce HeatPlot to show top 10; revise manuscript based on Cory's comments, update supplement.

--- 2023-06-08 manuscript/Supplement.qmd
@@
 ---
-title: Supplement for Open-Source machine learning BANTER acoustic classification of a cryptic echolocating species
+title: Supplement for Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses
 author: Shannon Rankin
@@
 # NAtlantic
-## EC Model
+Results for the NAtlantic EC (only) model are provided in @fig-natlantic_EC (Detector Model: *sampsize* = 3, *ntree* = 10,000; Event Model: *sampsize* = 3, *ntree* = 10,000).
 
-Detector Model:  ntree=1e4, sampsize=3; 
-Event Model: ntree=1e4, sampsize=3
 ```{r,  NAtlantic EC Model Figure, echo=FALSE}
+#| label: fig-natlantic_EC
+#| fig-cap: BANTER classification results from the NAtlantic dataset including echolocation pulses (EC).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| #| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_confuseMatrix.png'))%>%
@@
 
-## EC_ICI_ALT Model
-Detector Model:  ntree=1e4, sampsize=5; 
-Event Model: ntree=1e4, sampsize=5
-```{r, NAtlantic EC_ICI_ALT Model Figure,  echo=FALSE}
-confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
-vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_votes.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
-prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_proximity.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
-heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+# SAtlantic
 
-natlantic_alt_Figure <-image_append(c(prox, heat, vote))
-natlantic_alt_Figure<- image_append(c(confuse, natlantic_alt_Figure), stack=TRUE)
-print(natlantic_alt_Figure, info=FALSE)
-```
+Results for the SAtlantic EC (only) model are provided in @fig-satlantic_EC (Detector Model: *sampsize* = 8, *ntree* = 10,000; Event Model: *sampsize* = 1, *ntree* = 100,000).
 
-# SAtlantic
-## EC Model
-Detector Model:  ntree=1e4, sampsize=2; 
-Event Model: ntree=1e5, sampsize=1
 ```{r, SAtlantic EC Model Figure,  echo=FALSE}
+#| label: fig-satlantic_EC
+#| fig-cap: BANTER classification results from the SAtlantic dataset including echolocation pulses (EC).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| #| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_confuseMatrix.png'))%>%
@@
 
-## EC_ICI_ALT Model
-Detector Model:  ntree=1e4, sampsize=4; 
-Event Model: ntree=1e5, sampsize=2
-```{r, SAtlantic EC_ICI_ALT Model Figure,  echo=FALSE}
+Results for the SAtlantic EC_IPI_ALT model are provided in @fig-satlantic_EC_IPI_ALT (Detector Model: *sampsize* = 4, *ntree* = 10,000; Event Model: *sampsize* = 3, *ntree* = 100,000).
+
+```{r, SAtlantic EC_IPI_ALT Model Figure,  echo=FALSE}
+#| label: fig-satlantic_EC_IPI_ALT
+#| fig-cap: BANTER classification results from the alternative SAtlantic dataset including echolocation pulses and inter-pulse interval with a large sampsize (EC_IPI_ALT).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| #| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_confuseMatrix.png'))%>%
@@
 # Hawaii
-## EC_ICI Model
-Detector Model:  ntree=5e3, sampsize=5; 
-Event Model: ntree=1e4, sampsize=4
-```{r, Hawaii EC_ICI Model Figure,  echo=FALSE}
+
+Results for the Hawaii EC_IPI model are provided in @fig-hawaii_EC_IPI (Detector Model: *sampsize* = 5, *ntree* = 100,000; Event Model: *sampsize* = 4, *ntree* = 10,000).
+
+```{r, Hawaii EC_IPI Model Figure,  echo=FALSE}
+#| label: fig-hawaii_EC_IPI
+#| fig-cap: BANTER classification results from the Hawaii dataset including echolocation pulses and inter-pulse interval (EC_IPI).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| #| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_confuseMatrix.png'))%>%
@@
 # EPacific
-## EC Model
-Detector Model:  ntree=1e4, sampsize=3; 
-Event Model: ntree=1e4, sampsize=4
+
+Results for the EPacific EC (only) model are provided in @fig-epacific_EC (Detector Model: *sampsize* = 4, *ntree* = 10,000; Event Model: *sampsize* = 4, *ntree* = 10,000).
 
 ```{r, EPacific EC Model Figure,  echo=FALSE}
+#| label: fig-epacific_EC
+#| fig-cap: BANTER classification results from the EPacific dataset including echolocation pulses (EC).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| #| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_confuseMatrix.png'))%>%
@@
 ```
+
+Results for the EPacific EC (only) model are provided in @fig-epacific_EC_IPI (Detector Model: *sampsize* = 3, *ntree* = 10,000; Event Model: *sampsize* = 4, *ntree* = 10,000).
+
+```{r, EPacific EC IPI Model Figure,  echo=FALSE}
+#| label: fig-epacific_EC_IPI
+#| fig-cap: BANTER classification results from the EPacific dataset including echolocation pulses and inter-pulse interval (EC_IPI).
+Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate).
+Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification).
+Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue).
+Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
+#| #| fig-width: 10
+#| fig-height: 10
+#| fig-align: center
+#| out-width: 100%
+#| echo: false
+confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_confuseMatrix.png'))%>%
+  image_border(color="#ffffff", geometry = "50x130")%>%
+  image_annotate("a) Confusion Matrix", size=100, color = "black")
+vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_votes.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("d) Vote Plot", size=100, color = "black")
+prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_proximity.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_annotate("b) Proximity Plot", size=100, color = "black")
+heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_importance.png'))%>%
+  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_scale("3300")%>%
+  image_annotate("d) Importance Heat Map", size=100, color = "black")
+
+epacific_ici_Figure <-image_append(c(prox, heat, vote))
+epacific_ici_Figure<- image_append(c(confuse, epacific_ici_Figure), stack=TRUE)
+print(epacific_ici_Figure, info=FALSE)
+```

### 2023-08-28 manuscript/manuscript.qmd
Revise Hawaii and Manuscript
Revise Hawaii to reduce HeatPlot to show top 10; revise manuscript based on Cory's comments, update supplement.

@@
 Historically, experienced acousticians have manually identified stereotyped sounds that could reliably be attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008].
-However, the dramatic increase in the quantity of recordings being made make it impossible for experienced acousticians to manually annotate it all.
+However, the substantial increase in the quantity of recordings being made make it impossible for experienced acousticians to manually annotate it all.
 This has led to an increased need for the development of automated classification routines that can provide accurate species determinations from acoustic recordings.
@@
 AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey areas.
-The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, Appendix A, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (HTI-96-min, High Tech Inc., Long Beach, MS) which recorded at a 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see @deangelis_description_2018 for more information).
-The SAtlantic area ranged from Delaware south to central Florida (GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (Reson TC4013, Teledyne Marine, Slangerup, Denmark) which recorded at a 500 kHz sample rate (custom 12 channel SailDAQ soundcard) and were decimated to 192 kHz with a 1 kHz high pass filter (see GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016] for more information).
+The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, Appendix A, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (HTI-96-min, High Tech Inc., Long Beach, MS), which recorded at a 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see @deangelis_description_2018 for more information).
+The SAtlantic area ranged from Delaware south to central Florida (GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (Reson TC4013, Teledyne Marine, Slangerup, Denmark), which recorded at a 500 kHz sample rate (custom 12 channel SailDAQ soundcard) and were decimated to 192 kHz with a 1 kHz high pass filter (see GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016] for more information).
 The NAtlantic and SAtlantic recordings were treated as two separate datasets due to differences in hydrophone frequency responses and sensitivities.
@@
 
-Echolocation signals that were identified as frequency-modulated pulses from beaked whales were then linked as PAMGuard "events" by experienced acoustic researchers.
+Echolocation signals that were identified as frequency-modulated pulses from beaked whales were then linked to PAMGuard "events" by experienced acoustic analysts.
 Events included all the pulses that were close to each other in time, had similar spectral and waveform characteristics, and were received at consistent bearing angles.
@@
 
-Experienced acoustic researchers used multiple lines of evidence (PAMGuard click detector bearing time display, click waveform display, click spectrum, and wigner plots) to assess the species identity of the acoustic event.
+Experienced acoustic analysts used multiple lines of evidence (PAMGuard click detector bearing time display, click waveform display, click spectrum, and wigner plots) to assess the species identity of the acoustic event.
 When the available characteristics were inconclusive, the species was considered an unidentified beaked whale.
@@
 Results from the EC_IPI model are presented here; results from the EC and EC_IPI_ALT model can be found in supplementary materials (Supplement Fig.
-2).
+2, 3).
 
@@
 Results from the EC_IPI_ENV model are presented here; results from the EC and EC_IPI model can be found in supplementary materials (Supplement Figs.
-5).
+5, 6).
 
@@
 
-Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (e.g., long-beaked and short-beaked common dolphins, @rankin_acoustic_2017).
+Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced analysts are unable to differentiate species (e.g., long-beaked and short-beaked common dolphins, @rankin_acoustic_2017).
 This suggests that large increases in sample sizes may improve classification results for Cuvier's beaked whales.
@@
 This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified.
-For example, [@barlow_unique_2022] recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (expert analysis did not identify these in the EPacific data presented here).
+For example, [@barlow_unique_2022] recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (expert analysts did not identify these in the EPacific data presented here).
 
@@
 N00070-18-MP-4C560) and Chief of Naval Operations N45, and Southwest Fisheries Science Center.
-The manuscript was improved thanks to reviews from XX and XX.
+The manuscript was improved thanks to reviews from Cory Hom-Weaver.
 

### 2023-08-31 manuscript/Supplement.qmd
update draft manuscript for submission to RPTS
update draft manuscript and RPTS submission form. Modify Hawaii figures to show top 10 heat map.

@@
 output: pdf
+#format: docx
 ---

### 2023-08-31 manuscript/manuscript.qmd
update draft manuscript for submission to RPTS
update draft manuscript and RPTS submission form. Modify Hawaii figures to show top 10 heat map.

@@
 For the SAtlantic dataset, the most accurate model considered the echolocation pulse metrics and the IPI for an event (EC_IPI Model, Detector Model: *sampsize* = 4, *ntree* = 10,000; Event Model: *sampsize* = 1, *ntree* = 100,000) and provided an overall correct classification rate of 88.7% for all five species.
-Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-satlantic); all classification results were greater than expected (Priors in @fig-natlantic a).
+Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-natlantic); all classification results were greater than expected (Priors in @fig-natlantic a).
 The sample sizes for the BANTER model were kept low (event sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2.

### 2023-09-14 manuscript/manuscript.qmd
Manuscript Preparation
Draft manuscript and declaration statement for submission.

@@
 Results from the EC_IPI model are presented here; results from the EC model were similar and can be found in supplementary materials (Supplement Fig.
-1).
+S1).
 
@@
 Analysis with a larger sample size resulted in improved classification results for Cuvier's and Gervais' beaked whale (see Supplement Fig.
-3 for EC_IPI_ALT), but resulted in the loss of Sowerby's in the final model.
+S3 for EC_IPI_ALT), but resulted in the loss of Sowerby's in the final model.
 Results from the EC_IPI model are presented here; results from the EC and EC_IPI_ALT model can be found in supplementary materials (Supplement Fig.
-2, 3).
+S2, S3).
 
@@
 Results from the EC model are presented here; results from EC_IPI model can be found in supplementary materials (Supplement Fig.
-4).
+S4).
 
@@
 Results from the EC_IPI_ENV model are presented here; results from the EC and EC_IPI model can be found in supplementary materials (Supplement Figs.
-5, 6).
+S5, S6).
 
@@
 Even with reasonably large sample sizes, the classification may suffer (e.g., SAtlantic Cuvier's beaked whale = 82.8% correct classification).
-We found that by increasing the *sampsize* in BANTER, we could increase the classification rate for this species in this area to 91.8% (see Supplement Fig.3).
+We found that by increasing the *sampsize* in BANTER, we could increase the classification rate for this species in this area to 91.8% (see Supplement Fig.S3).
 However, this resulted in the inability to include Sowerby's beaked whale in the final model.

### 2024-01-09 manuscript/manuscript_files/ReviewResponse_DetailedRebuttal.Rmd
draft rebuttal
draft rebuttal

---
title: "DetailedManuscriptRevisions"
author: "Shannon Rankin"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

When resubmitting your manuscript, please carefully consider all issues mentioned in the reviewers' comments, outline every change made point by point, and provide suitable rebuttals for any comments not addressed.

To submit your revised manuscript go to <https://www.editorialmanager.com/ecoinf/> and log in as an Author where you will see a menu item called 'Submission Needing Revision'.

Please resubmit your manuscript by Feb 01, 2024.

```{r cars}
summary(cars)
```

## Requested Revisions

-   Research Elements (optional)

This journal encourages you to share research objects - including your raw data, methods, protocols, software, hardware and more -- which support your original research article in a Research Elements journal. Research Elements are open access, multidisciplinary, peer-reviewed journals which make the objects associated with your research more discoverable, trustworthy and promote replicability and reproducibility. As open access journals, there may be an Article Publishing Charge if your paper is accepted for publication. Find out more about the Research Elements journals at <https://www.elsevier.com/authors/tools-and-resources/research-elements-journals?dgcid=ec_em_research_elements_email>.

-   the Introduction does not cover the pertinent literature and fails to highlight the novelty of your study. You are missing important work that has been published on the topic, including the journal Ecological Informatics. This is not a journal that solely focuses on the technical aspects of the published scholarship, we also want to see context, broader implications of your work, critical discussion of the uncertainties etc.

    -   Add a paragraph about different approaches to automated classification, including CNNs, and what the strengths/limitations are. Discuss how even though some of these are really advancing and are becoming increasingly accessible to non-programmers, there is still a high barrier of entry. Banter is stupid easy & can be used by people with limited programming experience. but how does it was developed for working with multiple call types-- how does it work for just clicks?

-   The quality of your Figures is highly problematic and needs to improve significantly.

-   The Discussion is limited and fails to draw parallels with relevant studies from the peer-reviewed literature or put the presented results into the right context. You need to augment significantly the Discussion section in your paper and evaluate your work with reference to similar studies from the field of ecological informatics; several examples are also available in the journal Ecological Informatics.

-   You are missing the Research Highlights, which should not exceed the maximum of 85 characters!!

-   The justification for employing BANTER models is clear. **One could argue that more recent, deep learning-based methodologies could have been incorporated.** However, given that the detection performance in the study appears quite robust, the decision to utilize random forest detectors seems justified, as it doesn't seem to have adversely affected the results. [Discuss value of these methods given advances in CNNs, etc]

-   Improve quality of ALL images

    ## Detailed Responses and Rebuttals

Research Highlights (maybe 3-5 highlights):

-   [Banter is easy to use open source classifier for bioacoustic data to be used on multiple call types]

-   [Banter can be applied to single call types when multiple detectors are used to identify those individual call types]

-   [how well did banter do at classification?]

### 2024-01-09 manuscript/manuscript.qmd
draft rebuttal
draft rebuttal

@@
 Development of automated classification routines, if accurate, serves to improve the efficiency, reduce the subjectivity, and decrease the cost of analyzing large datasets.
+
+Advances in computing have led to a dramatic increase in the opportunities to use machine learning for bioacoustic classification tasks.
+In particular, use of deep neural networks show great promise for large datasets on a wide variety of taxa, including birds (cite), mammals (cite), insects (cite), reptiles (cite), fish (cite) and marine mammals (Franier et al.
+, Roman Ruiz et al 2023).
+Developing a neural network is non-trivial and has typically required extensive programming experience, but there are more and more tools that allow people to more readily use them (Kahl et al.
+2021)\].
+Neural networks typically require significant sample sizes, although there are increasingly new tools to enhance data or combined disparate datasets (Padovese et al.
+2023, Nanni et al.
+2020, Shorten and Khoshgoftaar 2019, Dufourq et al.
+2022).
+\[There is a significant learning curve to understand how to properly apply and test neural network classification methods on a novel dataset\].
+For many bioacoustic classification tasks, modification of existing machine learning methods may be sufficient and have a lower barrier to entry.
+Random Forest based acoustic classifiers have been found to be relatively easy, efficient, and effective approach to bioacoustic classification (Ross and Allen 2014, Lopez-Baucells et al.
+2019).
 
@@
 ```
+
 \newpage
@@
 ```
+
 \newpage
+
 ```{r, hawaii}
@@
 ```
+
 \newpage
+
 ```{r, epacific-env}
@@
 ```
+
 \newpage
+
 # References {.unnumbered}

### 2024-01-09 epacific/annomate.R

@@ -3,6 +3,7 @@
 # requires PAMpal 0.18.0 and PAMmisc 1.11.0
 
 # data must have GPS and species assigned
+#devtools::install_github('TaikiSan21/PAMpal')
 library(PAMpal)
 library(here)
 here()
@@ -32,7 +33,7 @@ library(PAMmisc)
 # to get figshare data you will need the article id and your personal token
 # i recommend storing your token in an external file instead of typing it in here
 # also DO NOT commit this file to any repository
-figToken <- readRDS(file = "epacific/figtoken.rds")
+figToken <- readRDS(file = "../shush/figtoken.rds")
 figId <- 22959786
 figshareData <- getFigshareInfo(figToken, figId)
 
@@ -49,4 +50,7 @@ checkAnnotation(anno)
 # add to acoustic study for storage
 data <- addAnnotation(data, anno)
 # this creates CSV ready for figshare, will also repeat messages from the check
-export_annomate(data, file=annoFile)
+#export_annomate(data, file=annoFile)#Needs a bug fix on PAMpal first
+#Run this temporarily until the bug is fixed:
+annoExport <- export_annomate(data)
+write.csv(annoExport, file=annoFile, row.names=FALSE)

### 2024-01-12 manuscript/manuscript_files/ReviewResponse_DetailedRebuttal.Rmd
revisions
revisions to manuscript based on reviewers comments

@@
 -   [how well did banter do at classification?]
+
+I would like to thank the reviewers and editors, especially the Associate Editor, for their review of our manuscript.
+
+-   ***Request to include expanded review of research on the research topic (Introduction).*** The associate editor requested that we expand our Introduction to include work that has been published on the topic.
+Our intention was to focus on a new application of an existing tool, and I assume that the associate editor would like increased consideration of deep learning methods for acoustic classification.
+To that end, I have included an additional paragraph in the introduction that includes consideration of these methods and a justification of why despite the incredibly advances in deep learning in recent years, consideration of machine learning methods (such as models based on Random Forest) may be preferred for some classification tasks.
+
+-   **Request to evaluate our results and draw parallels with relevant studies in the peer reviewed literature to provide important context (Discussion).** I included a brief comparison of our results to other machine learning methods to acoustically classify beaked whales, and highlighted how the number of classes and the sample size per class differentiates these studies.
+***I also included a brief discussion of how the ease of use of BANTER allows biologists to readily adopt these methods, whereas the barrier to entry for most other machine learning models is steep.*** I hesitate to dive deeper into this discussion as these methods are evolving extremely rapidly, and while deep learning methods are rapidly advancing and are arguably the future of our science, at this point they have requirements that are beyond the scope of many scientists.
+Our focus is on how to use existing tools to provide reasonable results for imperfect data.
+I have added an additional statement encouraging the preparation of training sets for the eventual consideration in developing deep learning routines.
+
+-   Provide Context, broader implications of our work, and critical discussion of the uncertainties
+
+-   Improve resolution of Figures
+
+-   Include Research Highlights

### 2024-01-12 manuscript/manuscript.qmd
revisions
revisions to manuscript based on reviewers comments

--- 2023-09-14 manuscript/manuscript.qmd
@@
 
-Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod_beaked_2018].
-However, beaked whales make stereotyped echolocation signals in the form of frequency-modulated (FM) pulses that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_species-specific_2013].
-Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022]; [@baumann-pickering_spatio-temporal_2014]; [@simonis_passive_2020].
-Unfortunately, manual classification of beaked whale echolocation pulses requires analysis by trained analysts using a number of visual aids to examine the call characteristics.
-Development of automated classification routines, if accurate, serves to improve the efficiency, reduce the subjectivity, and decrease the cost of analyzing large datasets.
+Advances in computing have led to a dramatic increase in the opportunities to use machine learning for bioacoustic classification [@bianco_machine_2019, @mutanu_review_2022, @stowell_computational_2022].
+In particular, use of deep neural networks show great promise for large datasets on a wide variety of taxa, including marine mammals [@frainer_automatic_2023, @frasier_automated_2017, @li_model-based_2020, @roman_ruiz_fin_2023, @zhong_beluga_2020].
+Neural networks typically require significant sample sizes for all classes, although there are increasingly new tools to enhance data or combined disparate datasets [@dufourq_passive_2022, @nanni_data_2020, @nolasco_learning_2023, @padovese_adapting_2023, @shorten_survey_2019].
+There is a significant learning curve to understand how to properly apply and test neural network classification methods on a novel dataset, although development of more user friendly tools are in development [@kahl_birdnet_2021, @kirsebom_meridian_2022].
+Despite the great potential for deep learning methods, modification of existing machine learning methods with a lower barrier to entry may be sufficient for many bioacoustic classification tasks.
 
-BANTER (**B**io-**A**coustic eve**NT** classifi**ER**) is a supervised machine learning method originally developed for species identification of dolphin signals from acoustic recordings [@rankin_acoustic_2017].
-This hierarchical Random Forest [@breiman_random_2001] event classifier combines information from independent classification models for each call type (e.g., whistles, echolocation pulses, and burst pulses for the dolphin-based model) into a model classifying a full acoustic event, which is defined as a discrete collection of calls during an encounter.
+Random Forest based acoustic classifiers have been found to be relatively easy, efficient, and effective approach to bioacoustic classification [@lopez-baucells_stronger_2019, @ross_random_2014].
+BANTER (**B**io-**A**coustic eve**NT** classifi**ER**) is a hierarchical Random Forest [@breiman_random_2001] event classifier originally developed for species identification of dolphin signals from acoustic recordings [@rankin_acoustic_2017].
+BANTER combines information from independent classification models for each call type (e.g., whistles, echolocation pulses, and burst pulses for the dolphin-based model) into a model classifying a full acoustic event, which is defined as a discrete collection of calls during an encounter.
 The event classifier uses the distribution of species assignment probabilities for each call type, along with any event-level metrics (such as call rate).
-BANTER is very flexible and can accommodate any number of measures from any number of detectors.
+BANTER is very flexible and can accommodate any number of measures from any number of detectors used to identify calls from any species.
+BANTER is easy to use and is readily available as a stand-alone R package [@archer_banter_2022].
 
@@
 For species that primarily (or exclusively) produce echolocation pulses, BANTER may serve as an option for automated machine learning classification.
+
+Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod_beaked_2018].
+However, beaked whales make stereotyped echolocation signals in the form of frequency-modulated (FM) pulses that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_species-specific_2013].
+Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022]; [@baumann-pickering_spatio-temporal_2014]; [@simonis_passive_2020].
+Unfortunately, manual classification of beaked whale echolocation pulses requires analysis by trained analysts using a number of visual aids to examine the call characteristics.
+Development of automated classification routines, if accurate, serves to improve the efficiency, reduce the subjectivity, and decrease the cost of analyzing large datasets.
 
@@
 All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%.
+Other studies on machine learning acoustic classification of beaked whales have yielded stronger classification results, although these studies included fewer beaked whale species classes and much larger sample sizes for each species class \[@li_model-based_2020-1, @lebien_species-level_2018, @cohen_identification_2022\].
+While there is great room for improvement in our models, and we suspect that modest increases in sample sizes for some species will result in significant improvements to the model results, it is likely that the low sample sizes presented in these datasets would be insufficient to develop and test deep learning models.
 

### 2024-01-30 manuscript/manuscript_files/ReviewResponse_DetailedRebuttal.Rmd
Revisions and Rebuttal
Revisions to manuscript + Rebuttal draft

--- 2024-01-09 manuscript/manuscript_files/ReviewResponse_DetailedRebuttal.Rmd
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
+To the Editors of Ecological Informatics:
 
-## Instructions
+Thank you for your consideration and review of my research article submission entitled "Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses".
+I appreciate your time and the feedback to improve this manuscript for publication.
+We agree with all comments and suggestions provided by the reviewers/editors, and we provide an explanation of how we addressed each comment.
 
-When resubmitting your manuscript, please carefully consider all issues mentioned in the reviewers' comments, outline every change made point by point, and provide suitable rebuttals for any comments not addressed.
+-   ***Request to include 3-5 Research Highlights (optional).*** We suggest the following Research Highlights:
 
-To submit your revised manuscript go to <https://www.editorialmanager.com/ecoinf/> and log in as an Author where you will see a menu item called 'Submission Needing Revision'.
+    -   BANTER is an open source, accessible machine learning classifier for data with multiple call types
 
-Please resubmit your manuscript by Feb 01, 2024.
+    -   BANTER can use multiple detectors on single call types for acoustic event classification
 
-```{r cars}
-summary(cars)
-```
+    -   BANTER can classify beaked whale species with relatively low sample sizes
 
-## Requested Revisions
+    -   BANTER is easy to use and can be easily adopted to address current classification needs
 
--   Research Elements (optional)
+-   ***Request to include expanded review of research on the research topic (Introduction).*** Our intention was to focus on a new application of an existing tool, and I assume the request was in relation to relatively new deep learning methods for acoustic classification.
+To that end, I have included an additional Introduction paragraph that includes consideration of these methods and a justification of why despite the incredible advances in deep learning in recent years, consideration of machine learning methods (such as models based on Random Forest) may be preferred for some classification tasks.
 
-This journal encourages you to share research objects - including your raw data, methods, protocols, software, hardware and more -- which support your original research article in a Research Elements journal.
-Research Elements are open access, multidisciplinary, peer-reviewed journals which make the objects associated with your research more discoverable, trustworthy and promote replicability and reproducibility.
-As open access journals, there may be an Article Publishing Charge if your paper is accepted for publication.
-Find out more about the Research Elements journals at <https://www.elsevier.com/authors/tools-and-resources/research-elements-journals?dgcid=ec_em_research_elements_email>.
+-   ***Request to evaluate our results and draw parallels with relevant studies in the peer reviewed literature to provide important context (Discussion)*****.** In order to provide additional context in the discussion without losing sight of the more narrow goals of this study, I focused on discussion points related specifically to acoustic classification of beaked whales.
+To that end, I first added a paragraph outlining some specific needs as it relates to acoustic classification of beaked whales.
+I then provided a comparison of our general results to those provided by other automated methods for acoustic classification of beaked whales providing context regarding the specific needs outlined in the previous paragraph (most importantly, considering all relevant species in a region, including species with low sample sizes).
+Finally, I modified the final paragraph to acknowledge that the limitations in deep learning methods are most likely temporary, and that the methods provided here can be readily adopted now to meet existing acoustic classification needs.
 
--   the Introduction does not cover the pertinent literature and fails to highlight the novelty of your study.
-You are missing important work that has been published on the topic, including the journal Ecological Informatics.
-This is not a journal that solely focuses on the technical aspects of the published scholarship, we also want to see context, broader implications of your work, critical discussion of the uncertainties etc.
+-   ***Improve Resolution of Figures.*** I have increased the font size on most of the graphs, which required introducing an angle for the x-axis for the heatmaps (c) for all Figures.
+The page layout for these figures was shifted to Landscape to allow for the wide image format.
+If additional modifications (minor or major) are needed, these can be accommodated.
 
-    -   Add a paragraph about different approaches to automated classification, including CNNs, and what the strengths/limitations are.
-Discuss how even though some of these are really advancing and are becoming increasingly accessible to non-programmers, there is still a high barrier of entry.
-Banter is stupid easy & can be used by people with limited programming experience.
-but how does it was developed for working with multiple call types-- how does it work for just clicks?
+-   ***Research Elements (optional).*** This project is a collaboration, and as such not all researchers are fully prepared to make their data publicly available.
+The data provided my my research lab (epacific in this study) will be publicly available on Figshare (I am only waiting for final acceptance of this paper).
+This manuscript, and the associated data products will be made publicly available on github (if not in whole, at least the epacific component).
+We strongly believe in Open Science and strive towards that goal.
 
--   The quality of your Figures is highly problematic and needs to improve significantly.
+I hope that these revisions are acceptable, and that you accept this draft manuscript for publication.
+If you require additional revisions, I will be happy to consider them.
+I greatly appreciate you and your team for your time and energy considering this manuscript for publication.
 
--   The Discussion is limited and fails to draw parallels with relevant studies from the peer-reviewed literature or put the presented results into the right context.
-You need to augment significantly the Discussion section in your paper and evaluate your work with reference to similar studies from the field of ecological informatics; several examples are also available in the journal Ecological Informatics.
+Sincerely,
 
--   You are missing the Research Highlights, which should not exceed the maximum of 85 characters!!
+Shannon Rankin
 
--   The justification for employing BANTER models is clear.
-**One could argue that more recent, deep learning-based methodologies could have been incorporated.** However, given that the detection performance in the study appears quite robust, the decision to utilize random forest detectors seems justified, as it doesn't seem to have adversely affected the results.
-[Discuss value of these methods given advances in CNNs, etc]
+Wildlife Research Biologist
 
--   Improve quality of ALL images
+Southwest Fisheries Science Center
 
-    ## Detailed Responses and Rebuttals
-
-Research Highlights (maybe 3-5 highlights):
-
--   [Banter is easy to use open source classifier for bioacoustic data to be used on multiple call types]
-
--   [Banter can be applied to single call types when multiple detectors are used to identify those individual call types]
-
--   [how well did banter do at classification?]
+NOAA Fisheries

### 2024-01-30 epacific/beakerBanter_epacific.Rmd
Revisions and Rebuttal
Revisions to manuscript + Rebuttal draft

@@
 library("easypackages")
-libraries("PAMpal", "banter", "rfPermute", "kableExtra", "magick", "magrittr", "here")
+libraries("PAMpal", "banter", "rfPermute", "kableExtra", "magick", "magrittr", "ggplot2", "here")
 here()
@@
 png(('../manuscript/manuscript_files/epacific_ec_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotProximity(banter_model_ec_RF)
+ec_epacific_proximityPlot <- plotProximity(banter_model_ec_RF, plot = FALSE)
+ec_epacific_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
-ec_epacific_proximityPlot <- plotProximity(banter_model_ec_RF)
+ec_epacific_proximityPlot <- plotProximity(banter_model_ec_RF, label.size=16)
 
 png(('../manuscript/manuscript_files/epacific_ici_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
-ici_epacific_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_epacific_proximityPlot <- plotProximity(banter_model_ici_RF, plot = FALSE)
+ici_epacific_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
-ici_epacific_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_epacific_proximityPlot <- plotProximity(banter_model_ici_RF, label.size=16)
 
 png(('../manuscript/manuscript_files/epacific_env_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
-env_epacific_proximityPlot <- plotProximity(banter_model_env_RF)
+env_epacific_proximityPlot <- plotProximity(banter_model_env_RF, plot = FALSE)
+env_epacific_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
-env_epacific_proximityPlot <- plotProximity(banter_model_env_RF)
+env_epacific_proximityPlot <- plotProximity(banter_model_env_RF, label.size=16)
 ```
@@
 png(('../manuscript/manuscript_files/epacific_ec_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
+ec_epacific_plotImportance <- plotImportance(banter_model_ec_RF, plot.type="heatmap",n=10, plot = FALSE)
+ec_epacific_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
@@
 png(('../manuscript/manuscript_files/epacific_ici_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
+ici_epacific_plotImportance <- plotImportance(banter_model_ici_RF, plot.type = "heatmap", n=10, plot = FALSE)
+ici_epacific_plotImportance + theme(axis.text = element_text(size = 16),
+                                     axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/epacific_env_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_env_RF, plot.type="heatmap", n=10)
+env_epacific_plotImportance <- plotImportance(banter_model_env_RF, plot.type = "heatmap", n=10, plot = FALSE)
+env_epacific_plotImportance + theme(axis.text = element_text(size = 16),
+                                     axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/epacific_ec_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_ec_RF)
+ec_epacific_plotVotes <- plotVotes(banter_model_ec_RF, plot = FALSE)
+ec_epacific_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/epacific_ici_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_ici_RF)
+ici_epacific_plotVotes <- plotVotes(banter_model_ici_RF, plot = FALSE)
+ici_epacific_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/epacific_env_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_env_RF)
+env_epacific_plotVotes <- plotVotes(banter_model_env_RF, plot = FALSE)
+env_epacific_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_votes.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "290x130")%>%
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_env_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 

### 2024-01-30 hawaii/beakerBanter_hawaii.Rmd
Revisions and Rebuttal
Revisions to manuscript + Rebuttal draft

--- 2023-08-21 hawaii/beakerBanter_hawaii.Rmd
@@
 library("easypackages")
-libraries("PAMpal", "banter", "rfPermute", "kableExtra", "magick", "magrittr", "here")
+libraries("PAMpal", "banter", "rfPermute", "kableExtra", "magick", "magrittr", "ggplot1", "here")
 here()
@@
 png(('../manuscript/manuscript_files/hawaii_ec_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotProximity(banter_model_ec_RF)
+ec_hawaii_proximityPlot <- plotProximity(banter_model_ec_RF, plot = FALSE)
+ec_hawaii_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
-ec_hawaii_proximityPlot <- plotProximity(banter_model_ec_RF)
+ec_hawaii_proximityPlot <- plotProximity(banter_model_ec_RF, label.size=16)
 
 png(('../manuscript/manuscript_files/hawaii_ici_proximity.png'), width = 20, height = 20, units = 'cm',  res = 300)
-ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF, plot = FALSE)
+ici_hawaii_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
-ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF, label.size=16)
 ```
@@
 png(('../manuscript/manuscript_files/hawaii_ec_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_ec_RF, plot.type="heatmap")
+ec_hawaii_plotImportance <- plotImportance(banter_model_ec_RF, plot.type="heatmap",n=10, plot = FALSE)
+ec_hawaii_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
-ec_hawaii_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap")
+ec_hawaii_importance <- plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
 
 png(('../manuscript/manuscript_files/hawaii_ici_importance.png'), width = 30, height = 25, units = 'cm',  res = 300)
-plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_hawaii_plotImportance <- plotImportance(banter_model_ici_RF, plot.type="heatmap",n=10, plot = FALSE)
+ici_hawaii_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
-ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
+ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
 ```
@@
 png(('../manuscript/manuscript_files/hawaii_ec_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_ec_RF)
+ec_hawaii_plotVotes <- plotVotes(banter_model_ec_RF, plot = FALSE)
+ec_hawaii_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/hawaii_ici_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_ici_RF)
+ici_hawaii_plotVotes <- plotVotes(banter_model_ici_RF, plot = FALSE)
+ici_hawaii_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ec_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 

### 2024-01-30 manuscript/CopyOfmanuscript.qmd
Revisions and Rebuttal
Revisions to manuscript + Rebuttal draft

--- 2024-01-12 manuscript/manuscript.qmd
@@
 
+Development of automated acoustic classifiers is critical for efficient analysis of large passive acoustic datasets; these classification routines require representative training data created by experienced acoustic analysts.
+While beaked whales species can be identified based on their stereotyped frequency-modulated echolocation clicks @baumann-pickering_species-specific_2013, not all acoustic events include sounds that can be readily attributed to a specific species.
+Moreover, there are a number of beaked whales where little or nothing is known about their acoustic repertoire and new species have been detected in recent years [@deangelis_description_2018, @barlow_unique_2022].
+Therefore, development of acoustic classifiers for beaked whales should accommodate small and/or unbalanced sample sizes and have sufficient flexibility to allow for variable call characteristics.
+
 The results for the four study areas varied from a low of 88.7% correct classification in the SAtlantic dataset to a high of 97.5% correct classification in the NAtlantic dataset.
 All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%.
-Other studies on machine learning acoustic classification of beaked whales have yielded stronger classification results, although these studies included fewer beaked whale species classes and much larger sample sizes for each species class \[@li_model-based_2020-1, @lebien_species-level_2018, @cohen_identification_2022\].
-While there is great room for improvement in our models, and we suspect that modest increases in sample sizes for some species will result in significant improvements to the model results, it is likely that the low sample sizes presented in these datasets would be insufficient to develop and test deep learning models.
+Recent beaked whale classification studies using deep learning methods exhibit more robust results [@li_model-based_2020-1, @lebien_species-level_2018, @cohen_identification_2022, @ziegenhorn_discriminating_2022]; however, these models may not be appropriate for our dataset.
+Several studies included incomplete beaked whale species for the study area (@li_model-based_2020-1, @ziegenhorn_discriminating_2022 ) and all studies has sample large sample sizes (\>200) for each species class.
+The deep learning model developed by @cohen_identification_2022 included strong classification results with all beaked whale species found in our Atlantic models; however, it is unclear how this dataset trained on seafloor recorders would perform with towed array datasets.
+The BANTER models presented here only require a minimum of two samples per class and can therefore include all species within each region, making them useful to assess the presence of all beaked whales in the region.
+While deep learning methods may provide improved results once there are sufficient sample sizes for all relevant species within a study area, our results suggest that reasonable classification results can be obtained for all species using BANTER classification methods with as few as two samples per class.
 
@@
 Analysts' notes indicated uncertainty in manual species classification of this event in this dataset.
-While small sample size can at times provide good classification scores, as we found for Sowerby's beaked whale in this SAtlantic dataset, they can be heavily influential if there are inaccuracies in the training data.
+While small sample size can at times provide good classification scores, as we found for Sowerby's beaked whale in this SAtlantic dataset, they can be heavily influenced by inaccuracies in the training data.
 
@@
 
+Until deep learning methods become more accessible and practical for applications involving rare species and low sample sizes, or their applicability across recording platforms can be guaranteed, machine learning approaches such as BANTER can serve as an accessible means of developing acoustic classifiers for complicated datasets.
 The workflow presented here provides a highly automated approach to detection of acoustic events (PAMGuard), integration of environmental data (PAMpal), and acoustic event classification (BANTER).
@@
 
+{{< pagebreak >}}
 # Figures and tables
@@
 #| fig-align: center
-#| out-width: 100%
+#| out-width: 200%
 #| echo: false
 confuse <- magick::image_read('manuscript_files/natlantic_ici_confuseMatrix.png')%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read('manuscript_files/natlantic_ici_votes.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read('manuscript_files/natlantic_ici_proximity.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read('manuscript_files/natlantic_ici_importance.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("c) Importance Heat Map", size=100, color = "black")
+  image_annotate("c) Importance Heat Map", size=120, color = "black")
 
@@
 
-\newpage
+{{< pagebreak >}}
 
@@
 #| fig-align: center
-#| out-width: 100%
+#| out-width: 200%
 #| echo: false
 confuse <- magick::image_read('manuscript_files/satlantic_ici_confuseMatrix.png')%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read('manuscript_files/satlantic_ici_votes.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read('manuscript_files/satlantic_ici_proximity.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read('manuscript_files/satlantic_ici_importance.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  #image_border(color="#ffffff", geometry = "540x52")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("c) Importance Heat Map", size=100, color = "black")
+  image_annotate("c) Importance Heat Map", size=120, color = "black")
 
@@
 #| fig-align: center
-#| out-width: 100%
+#| out-width: 200%
 #| echo: false
 confuse <- magick::image_read('manuscript_files/hawaii_ec_confuseMatrix.png')%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read('manuscript_files/hawaii_ec_votes.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read('manuscript_files/hawaii_ec_proximity.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read('manuscript_files/hawaii_ec_importance.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("c) Importance Heat Map", size=100, color = "black")
+  image_annotate("c) Importance Heat Map", size=120, color = "black")
 
@@
 #| fig-align: center
-#| out-width: 100%
+#| out-width: 200%
 #| echo: false
 confuse <- magick::image_read('manuscript_files/epacific_env_confuseMatrix.png')%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read('manuscript_files/epacific_env_votes.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read('manuscript_files/epacific_env_proximity.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read('manuscript_files/epacific_env_importance.png')%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("c) Importance Heat Map", size=100, color = "black")
+  image_annotate("c) Importance Heat Map", size=120, color = "black")
 

### 2024-01-30 manuscript/manuscript.qmd
Revisions and Rebuttal
Revisions to manuscript + Rebuttal draft

@@
 Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring.
-Historically, experienced acousticians have manually identified stereotyped sounds that could reliably be attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008].
+Historically, experienced acousticians have manually identified stereotyped sounds that could reliably be attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013; @bittle_review_2013; @rankin_source_2005; @soldevilla_classification_2008].
 However, the substantial increase in the quantity of recordings being made make it impossible for experienced acousticians to manually annotate it all.
@@
 
-Advances in computing have led to a dramatic increase in the opportunities to use machine learning for bioacoustic classification [@bianco_machine_2019, @mutanu_review_2022, @stowell_computational_2022].
-In particular, use of deep neural networks show great promise for large datasets on a wide variety of taxa, including marine mammals [@frainer_automatic_2023, @frasier_automated_2017, @li_model-based_2020, @roman_ruiz_fin_2023, @zhong_beluga_2020].
-Neural networks typically require significant sample sizes for all classes, although there are increasingly new tools to enhance data or combined disparate datasets [@dufourq_passive_2022, @nanni_data_2020, @nolasco_learning_2023, @padovese_adapting_2023, @shorten_survey_2019].
-There is a significant learning curve to understand how to properly apply and test neural network classification methods on a novel dataset, although development of more user friendly tools are in development [@kahl_birdnet_2021, @kirsebom_meridian_2022].
+Advances in computing have led to a dramatic increase in the opportunities to use machine learning for bioacoustic classification [@bianco_machine_2019; @mutanu_review_2022; @stowell_computational_2022].
+In particular, use of deep neural networks show great promise for large datasets on a wide variety of taxa, including marine mammals [@frainer_automatic_2023; @frasier_automated_2017; @li_model-based_2020; @roman_ruiz_fin_2023; @zhong_beluga_2020].
+Neural networks typically require significant sample sizes for all classes, although there are increasingly new tools to enhance data or combined disparate datasets [@dufourq_passive_2022; @nanni_data_2020; @nolasco_learning_2023; @padovese_adapting_2023; @shorten_survey_2019].
+There is a significant learning curve to understand how to properly apply and test neural network classification methods on a novel dataset, although development of more user friendly tools are in development [@kahl_birdnet_2021; @kirsebom_meridian_2022].
 Despite the great potential for deep learning methods, modification of existing machine learning methods with a lower barrier to entry may be sufficient for many bioacoustic classification tasks.
 
-Random Forest based acoustic classifiers have been found to be relatively easy, efficient, and effective approach to bioacoustic classification [@lopez-baucells_stronger_2019, @ross_random_2014].
+Random Forest based acoustic classifiers have been found to be relatively easy, efficient, and effective approach to bioacoustic classification [@lopez-baucells_stronger_2019; @ross_random_2014].
 BANTER (**B**io-**A**coustic eve**NT** classifi**ER**) is a hierarchical Random Forest [@breiman_random_2001] event classifier originally developed for species identification of dolphin signals from acoustic recordings [@rankin_acoustic_2017].
@@
 However, beaked whales make stereotyped echolocation signals in the form of frequency-modulated (FM) pulses that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_species-specific_2013].
-Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022]; [@baumann-pickering_spatio-temporal_2014]; [@simonis_passive_2020].
+Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022; @baumann-pickering_spatio-temporal_2014; @simonis_passive_2020].
 Unfortunately, manual classification of beaked whale echolocation pulses requires analysis by trained analysts using a number of visual aids to examine the call characteristics.
@@
 When the available characteristics were inconclusive, the species was considered an unidentified beaked whale.
-Data were saved within the PAMGuard database and binaries for downstream processing using the *PAMpal package* [@sakai_pampal_2021] in R v2022.07.02 [@r_core_team_r_2022].
+Data were saved within the PAMGuard database and binaries for downstream processing using the *PAMpal package* [@sakai_pampal_2021] in R v2022.07.02 [@team_r_2013].
 
-Standard click metrics describing the spectral and temporal characteristics were measured from echolocation pulses using default values in *PAMpal* [@sakai_pampal_2023].
+Standard click metrics describing the spectral and temporal characteristics were measured from echolocation pulses using default values in *PAMpal* [@sakai_pampal_2021].
 Inter-pulse interval (IPI) was calculated using the 'calculateICI' function in *PAMpal* and the mode of the IPI was identified for each beaked whale acoustic event.
@@
 
-BANTER models were created using the *banter* package [@archer_banter_2022] in R v2022.07.02 [@r_core_team_r_2022].
+BANTER models were created using the *banter* package [@archer_banter_2022] in R v2022.07.02 [@team_r_2013].
 Output from PAMGuard click detector modules were used to create the BANTER detector models, for a total of 6 detector models (one for each click detector used).
@@
 
+Development of automated acoustic classifiers is critical for efficient analysis of large passive acoustic datasets; these classification routines require representative training data created by experienced acoustic analysts.
+While beaked whales species can be identified based on their stereotyped frequency-modulated echolocation clicks [@baumann-pickering_species-specific_2013], not all acoustic events include sounds that can be readily attributed to a specific species.
+Moreover, there are a number of beaked whales where little or nothing is known about their acoustic repertoire and new species have been detected in recent years [@deangelis_description_2018; @barlow_unique_2022].
+Therefore, development of acoustic classifiers for beaked whales should accommodate small and/or unbalanced sample sizes and have sufficient flexibility to allow for variable call characteristics.
+
 The results for the four study areas varied from a low of 88.7% correct classification in the SAtlantic dataset to a high of 97.5% correct classification in the NAtlantic dataset.
 All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%.
-Other studies on machine learning acoustic classification of beaked whales have yielded stronger classification results, although these studies included fewer beaked whale species classes and much larger sample sizes for each species class \[@li_model-based_2020-1, @lebien_species-level_2018, @cohen_identification_2022\].
-While there is great room for improvement in our models, and we suspect that modest increases in sample sizes for some species will result in significant improvements to the model results, it is likely that the low sample sizes presented in these datasets would be insufficient to develop and test deep learning models.
+Recent beaked whale classification studies using deep learning methods exhibit more robust results [@li_model-based_2020-1; @lebien_species-level_2018; @cohen_identification_2022; @ziegenhorn_discriminating_2022]; however, these models may not be appropriate for our dataset.
+Several studies included incomplete beaked whale species for the study area (@li_model-based_2020-1, @ziegenhorn_discriminating_2022) and all studies has sample large sample sizes (\>200) for each species class.
+The deep learning model developed by @cohen_identification_2022 included strong classification results with all beaked whale species found in our Atlantic models; however, it is unclear how this dataset trained on seafloor recorders would perform with towed array datasets.
+The BANTER models presented here only require a minimum of two samples per class and can therefore include all species within each region, making them useful to assess the presence of all beaked whales in the region.
+While deep learning methods may provide improved results once there are sufficient sample sizes for all relevant species within a study area, our results suggest that reasonable classification results can be obtained for all species using BANTER classification methods with as few as two samples per class.
 
@@
 This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified.
-For example, [@barlow_unique_2022] recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (expert analysts did not identify these in the EPacific data presented here).
+For example, @barlow_unique_2022 recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (expert analysts did not identify these in the EPacific data presented here).
 
@@
 N00070-18-MP-4C560) and Chief of Naval Operations N45, and Southwest Fisheries Science Center.
-The manuscript was improved thanks to reviews from Cory Hom-Weaver.
+The manuscript was improved thanks to reviews from Cory Hom-Weaver and an anonymous reviewer.
 
+{{< pagebreak >}}
 # Figures and tables

### 2024-01-30 natlantic/beakerBanter_natlantic.Rmd
Revisions and Rebuttal
Revisions to manuscript + Rebuttal draft

@@
 library("easypackages")
-libraries("PAMpal", "banter", "rfPermute", "readxl", "stringr", "dplyr", "kableExtra", "magick", "magrittr", "here")
+libraries("PAMpal", "banter", "rfPermute", "readxl", "stringr", "dplyr", "kableExtra", "magick", "magrittr", "ggplot2", "here")
 here()
@@
 png(('../manuscript/manuscript_files/natlantic_ec_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
-natlantic_ec_proximityPlot <- plotProximity(banter_model_ec_RF)
+ec_natlantic_proximityPlot <- plotProximity(banter_model_ec_RF, plot = FALSE)
+ec_natlantic_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
-natlantic_ec_proximityPlot <- plotProximity(banter_model_ec_RF)
+natlantic_ec_proximityPlot <- plotProximity(banter_model_ec_RF, label.size=16)
 
 png(('../manuscript/manuscript_files/natlantic_ici_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
-ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_natlantic_proximityPlot <- plotProximity(banter_model_ici_RF, plot = FALSE)
+ici_natlantic_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/natlantic_env_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
-env_natlantic_proximityPlot <- plotProximity(banter_model_env_RF)
+env_natlantic_proximityPlot <- plotProximity(banter_model_env_RF, plot = FALSE)
+env_natlantic_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
-env_natlantic_proximityPlot <- plotProximity(banter_model_env_RF)
+env_natlantic_proximityPlot <- plotProximity(banter_model_env_RF, label.size=16)
 
 png(('../manuscript/manuscript_files/natlantic_alt_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
-alt_natlantic_proximityPlot <- plotProximity(banter_model_alt_RF)
+alt_natlantic_proximityPlot <- plotProximity(banter_model_alt_RF, plot = FALSE)
+alt_natlantic_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
-alt_natlantic_proximityPlot <- plotProximity(banter_model_alt_RF)
+alt_natlantic_proximityPlot <- plotProximity(banter_model_alt_RF, label.size=16)
 ```
@@
 png(('../manuscript/manuscript_files/natlantic_ec_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
+ec_natlantic_plotImportance <- plotImportance(banter_model_ec_RF, plot.type="heatmap",n=10, plot = FALSE)
+ec_natlantic_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
@@
 png(('../manuscript/manuscript_files/natlantic_ici_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
+ici_natlantic_plotImportance <- plotImportance(banter_model_ici_RF, plot.type="heatmap",n=10, plot = FALSE)
+ici_natlantic_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
@@
 png(('../manuscript/manuscript_files/natlantic_env_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_env_RF, plot.type="heatmap", n=10)
+env_natlantic_plotImportance <- plotImportance(banter_model_env_RF, plot.type="heatmap",n=10, plot = FALSE)
+env_natlantic_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
@@
 png(('../manuscript/manuscript_files/natlantic_alt_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_alt_RF, plot.type="heatmap", n=10)
+alt_natlantic_plotImportance <- plotImportance(banter_model_alt_RF, plot.type="heatmap",n=10, plot = FALSE)
+alt_natlantic_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
@@
 png(('../manuscript/manuscript_files/natlantic_ec_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_ec_RF)
+ec_natlantic_plotVotes <- plotVotes(banter_model_ec_RF, plot = FALSE)
+ec_natlantic_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/natlantic_ici_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_ici_RF)
+ici_natlantic_plotVotes <- plotVotes(banter_model_ici_RF, plot = FALSE)
+ici_natlantic_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/natlantic_env_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_env_RF)
+env_natlantic_plotVotes <- plotVotes(banter_model_env_RF, plot = FALSE)
+env_natlantic_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/natlantic_alt_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_alt_RF)
+alt_natlantic_plotVotes <- plotVotes(banter_model_alt_RF, plot = FALSE)
+alt_natlantic_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ici_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_env_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_env_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_env_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_env_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_alt_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 

### 2024-01-30 satlantic/beakerBanter_satlantic.Rmd
Revisions and Rebuttal
Revisions to manuscript + Rebuttal draft

@@
 library("easypackages")
-libraries("PAMpal", "banter", "rfPermute", "kableExtra", "readxl", "stringr", "magick", "magrittr", "here")
+libraries("PAMpal", "banter", "rfPermute", "kableExtra", "readxl", "stringr", "magick", "magrittr", "ggplot2", "here")
 here()
@@
 png(('../manuscript/manuscript_files/satlantic_ec_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
-satlantic_ec_proximityPlot <- plotProximity(banter_model_ec_RF)
+ec_satlantic_proximityPlot <- plotProximity(banter_model_ec_RF, plot = FALSE)
+ec_satlantic_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_ici_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
-ici_satlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
+ici_satlantic_proximityPlot <- plotProximity(banter_model_ici_RF, plot = FALSE)
+ici_satlantic_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_env_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
-env_satlantic_proximityPlot <- plotProximity(banter_model_env_RF)
+env_satlantic_proximityPlot <- plotProximity(banter_model_env_RF, plot = FALSE)
+env_satlantic_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_alt_proximity.png'), width = 20, height = 20, units = 'cm', res = 300)
-alt_satlantic_proximityPlot <- plotProximity(banter_model_alt_RF)
+alt_satlantic_proximityPlot <- plotProximity(banter_model_alt_RF, plot = FALSE)
+alt_satlantic_proximityPlot$g + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_ec_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_ec_RF, plot.type="heatmap", n=10)
+ec_satlantic_plotImportance <- plotImportance(banter_model_ec_RF, plot.type="heatmap",n=10, plot = FALSE)
+ec_satlantic_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_ici_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_ici_RF, plot.type="heatmap", n=10)
+ici_satlantic_plotImportance <- plotImportance(banter_model_ici_RF, plot.type="heatmap",n=10, plot = FALSE)
+ici_satlantic_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_env_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_env_RF, plot.type="heatmap", n=10)
+env_satlantic_plotImportance <- plotImportance(banter_model_env_RF, plot.type="heatmap",n=10, plot = FALSE)
+env_satlantic_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_alt_importance.png'), width = 30, height = 25, units = 'cm', res = 300)
-plotImportance(banter_model_alt_RF, plot.type="heatmap", n=10)
+alt_satlantic_plotImportance <- plotImportance(banter_model_alt_RF, plot.type="heatmap",n=10, plot = FALSE)
+alt_satlantic_plotImportance + theme(axis.text = element_text(size = 16),
+                                      axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0),
+                                      legend.text = element_text(size = 16)
+                                      )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_ec_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_ec_RF)
+ec_satlantic_plotVotes <- plotVotes(banter_model_ec_RF, plot = FALSE)
+ec_satlantic_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_ici_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_ici_RF)
+ici_satlantic_plotVotes <- plotVotes(banter_model_ici_RF, plot = FALSE)
+ici_satlantic_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_env_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_env_RF)
+env_satlantic_plotVotes <- plotVotes(banter_model_env_RF, plot = FALSE)
+env_satlantic_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 png(('../manuscript/manuscript_files/satlantic_alt_votes.png'), width = 20, height = 20, units = 'cm',  res = 300)
-plotVotes(banter_model_alt_RF)
+alt_satlantic_plotVotes <- plotVotes(banter_model_alt_RF, plot = FALSE)
+alt_satlantic_plotVotes + theme(axis.text = element_text(size = 16),
+                                     axis.title = element_text(size = 16),
+                                     legend.text = element_text(size = 16)
+                                     )
 dev.off()
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=300, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=300, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=300, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ici_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=300, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_env_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_env_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_env_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_env_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "50x200")%>%
+  image_annotate("a) Confusion Matrix", size=150, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_votes.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_annotate("d) Vote Plot", size=150, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_proximity.png'))%>%
   image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_annotate("b) Proximity Plot", size=150, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "270x200")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=150, color = "black")
 

### 2024-02-02 manuscript/Supplement.qmd
Final Revision for Manuscript
Revise mansucript including supplement, figures, etc. This is as submitted.

--- 2023-08-28 manuscript/Supplement.qmd
@@
 date: last-modified
-output: pdf
+#output: pdf
+format: docx
 ---
@@
 library("easypackages")
-libraries("tidyverse", "knitr", "kableExtra", "magick", "magrittr", "here")
+libraries("tidyverse", "knitr", "kableExtra", "magick", "magrittr", "ggplot2", "here")
 here()
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_votes.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_proximity.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'natlantic_ec_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=120, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_votes.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_proximity.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_ec_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=120, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_votes.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_proximity.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'satlantic_alt_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=120, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_votes.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_proximity.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'hawaii_ici_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=120, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_votes.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_proximity.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ec_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=120, color = "black")
 
@@
 confuse <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_confuseMatrix.png'))%>%
-  image_border(color="#ffffff", geometry = "50x130")%>%
-  image_annotate("a) Confusion Matrix", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "80x208")%>%
+  image_annotate("a) Confusion Matrix", size=120, color = "black")
 vote <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_votes.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("d) Vote Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("d) Vote Plot", size=120, color = "black")
 prox <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_proximity.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
-  image_annotate("b) Proximity Plot", size=100, color = "black")
+  image_border(color="#ffffff", geometry = "260x125")%>%
+  image_annotate("b) Proximity Plot", size=120, color = "black")
 heat <- magick::image_read(here('manuscript', 'manuscript_files', 'epacific_ici_importance.png'))%>%
-  image_border(color="#ffffff", geometry = "270x130")%>%
+  image_border(color="#ffffff", geometry = "110x260")%>%
   image_scale("3300")%>%
-  image_annotate("d) Importance Heat Map", size=100, color = "black")
+  image_annotate("d) Importance Heat Map", size=120, color = "black")
 

### 2024-02-02 manuscript/manuscript.qmd
Final Revision for Manuscript
Revise mansucript including supplement, figures, etc. This is as submitted.

@@
 format: docx
+journal: 
+  name: Ecological Informatics
+  formatting: preprint
+  model: 1p
+  cite-style: authoryear
 
-#  pdf: default
-#  elsevier-pdf:
-#    keep-tex: true
-#    journal:
-#      name: Ecological Informatics
-#      formatting: preprint
-#      model: 1p
-#      cite-style: authoryear
 ---
@@
 heat <- magick::image_read('manuscript_files/satlantic_ici_importance.png')%>%
-  #image_border(color="#ffffff", geometry = "540x52")%>%
   image_border(color="#ffffff", geometry = "110x260")%>%

### 2024-02-02 epacific/AnnoMatePaperCode.R

@@ -1,35 +1,0 @@
-# load appropriate programs
-library (PAMpal)
-library (PAMmisc)
-library (rfigshare)
-
-#Authenticate Figshare access, if needed
-rfigshare::fs_auth() #authenticate Figshare access, if needed
-
-#get information on repository files for https://figshare.com/account/projects/94511/articles/23319938
-data.details <- rfigshare::fs_details(23319938) 
-#download acoustics study from Figshare
-data <- download.file( url = file.path("https://ndownloader.figshare.com/files", data.details$files[[3]]$id), destfile = file.path(data.dir, data.details$files[[3]]$name))
-
-# reassign species names for Ann-O-Mate
-specMap <- data.frame(
-  old = unique(species(data)),
-  new = NA)
-specMap$new <- c('Ziphius cavirostris', 'possible Ziphiidae', 'Ziphiidae', 
-                 'Mesoplodon carlhubbsi', 'BW43', 'Berardius bairdii', 
-                 'Cross Seamount beaked whale', 'Mesoplodon stejnegeri')
-
-# save beaked whale data as myData
-anno <- prepAnnotation(data, specMap = specMap, mode='event',
-                       source = 'figshare.com',
-                       source_id = 22959786, 
-                       annotator = 'Anne Simonis, https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036',
-                       annotation_id = anno$event,
-                       annotation_date = 2019,
-                       annotation_info_url = 'https://figshare.com/articles/dataset/BWEventClips_CCES2018/22959786',
-                       recording_info_url = 'https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036',
-                       type = 'echolocation click event',
-                       contact = 'shannon.rankin@noaa.gov')
-
-
-

### 2024-02-02 epacific/annomate.R

@@ -1,56 +1,0 @@
-#### Annomate Example ####
-# Updated 2023-02-06
-# requires PAMpal 0.18.0 and PAMmisc 1.11.0
-
-# data must have GPS and species assigned
-#devtools::install_github('TaikiSan21/PAMpal')
-library(PAMpal)
-library(here)
-here()
-
-
-data <- readRDS('epacific/epacific_study.rds')
-specMap <- data.frame(
-  old = unique(species(data)),
-  new = NA)
-specMap$new <- c('Ziphius cavirostris', 'possible Ziphiidae', 'Ziphiidae', 
-                 'Mesoplodon carlhubbsi', 'BW43', 'Berardius bairdii', 
-                 'Cross Seamount beaked whale', 'Mesoplodon stejnegeri')
-
-anno <- prepAnnotation(data, specMap = specMap, mode='event',
-                       source = 'figshare.com',
-                       source_id = 22959786, 
-                       annotator = 'Anne Simonis, Jenny Trickey',
-                       annotation_date = 2019,
-                       annotation_info_url = 'https://figshare.com/articles/dataset/BWEventClips_CCES2018/22959786',
-                       recording_info_url = 'https://figshare.com/articles/online_resource/CCES_2018_Final_Acoustic_Report/19358036',
-                       type = 'echolocation click event',
-                       contact = 'shannon.rankin@noaa.gov')
-
-anno$annotation_id = anno$event
-
-library(PAMmisc)
-# to get figshare data you will need the article id and your personal token
-# i recommend storing your token in an external file instead of typing it in here
-# also DO NOT commit this file to any repository
-figToken <- readRDS(file = "../shush/figtoken.rds")
-figId <- 22959786
-figshareData <- getFigshareInfo(figToken, figId)
-
-anno <- matchRecordingUrl(anno, figshareData)
-# will give you warnings and messages about missing fields
-checkAnnotation(anno)
-# if you need to fix any manually, write to csv then read back in
-annoFile <- 'AnnomateExport_epacific_rankin2023.csv'
-write.csv(anno, file=annoFile, row.names = FALSE)
-# go fix stuff
-anno <- read.csv(annoFile, stringsAsFactors = FALSE)
-# recheck to see if all seems fine
-checkAnnotation(anno)
-# add to acoustic study for storage
-data <- addAnnotation(data, anno)
-# this creates CSV ready for figshare, will also repeat messages from the check
[…]
